%TEXT FROM:
%http://www.newadvent.org/bible/1ma001.htm

\documentclass[10pt]{book} % use larger type; default would be 10pt

\usepackage[latin1]{luainputenc}

\usepackage[oldstyle]{libertine}
\usepackage{lettrine}
\usepackage{xcolor}

% dual language parallel text thing
\usepackage{eledmac}
\usepackage{eledpar}

% BLUES
\definecolor{benblue1}{HTML}{2B22C7}

% REDS
\definecolor{benred8}{HTML}{E82C00} %seems good, maybe the best?

% THREE COMMANDS TO CLEAN UP ALL THE ELEDPAR/ELEDMAC STUFF
\newcommand{\StartOfLatin}{
	\begin{pairs}
	\begin{Leftside}
	\firstlinenum{10000}\linenumincrement{10000}\beginnumbering\pstart
}

\newcommand{\StartOfEnglish}{
	\pend\endnumbering
	\end{Leftside}
	\begin{Rightside}

	\firstlinenum{10000}\linenumincrement{10000}\beginnumbering\pstart
}

\newcommand{\EndOfEnglish}{
	\pend\endnumbering
	\end{Rightside}
	\end{pairs}
	\Columns
}

\usepackage{fancyhdr}
\pagestyle{fancy}
\cfoot{\textcolor{benblue1}{\thepage}}
\renewcommand{\headrulewidth}{0pt}

%redefine 'plain' page styel to have blue page numbers too
\fancypagestyle{plain}{
	\chead{}
	\cfoot{\textcolor{benblue1}{\thepage}}

}

\begin{document}

\chapter*{1 Maccabees}

\begin{large}\begin{center}\textsc{Chapter i}\end{center}\end{large}
\lettrine[lines=3]{N}{ow} turn we to Alexander son of Philip, the Macedonian, that was the first to reign over all Greece. This Alexander marched out from his own land of Cethim, and overcame Darius, king of the Medes and Persians. \textcolor{benred8}{2}~Battles he waged a many; nor any fortress might hold out against him, nor any king escape with his life; \textcolor{benred8}{3}~and so he journeyed on to the world\textquotesingle s end, spoiling the nations everywhere; at his coming, silence fell on the earth. \textcolor{benred8}{4}~So great the power of him, so valiant his armies, what wonder if his heart grew proud? \textcolor{benred8}{5}~All those lands conquered, all those kings his tributaries! \textcolor{benred8}{6}~Then, all at once, he took to his bed, and the knowledge came to him he must die. \textcolor{benred8}{7}~Whereupon he summoned the noblest of his courtiers, men that had shared his own upbringing, and to these, while he had life in him yet, divided up his kingdom. \textcolor{benred8}{8}~So reigned Alexander for twelve years, and so died.
\textcolor{benred8}{9}~And what of these courtiers turned princes, each with a province of his own? \textcolor{benred8}{10}~Be sure they put on royal crowns, they and their sons after them, and so the world went from bad to worse. \textcolor{benred8}{11}~Burgeoned then from the stock of Antiochus a poisoned growth, another Antiochus, he that was called the Illustrious. He had been formerly a hostage at Rome, but now, in the hundred and thirty-seventh year of the Grecian empire, he came into his kingdom. \textcolor{benred8}{12}~In his day there were godless talkers abroad in Israel, that did not want for a hearing; Come, said they, let us make terms with the heathen that dwell about us! Ever since we forswore their company, nought but trouble has come our way. \textcolor{benred8}{13}~What would you? Such talk gained credit, \textcolor{benred8}{14}~and some were at pains to ask for the royal warrant; whereupon leave was given them, Gentile usages they should follow if they would. \textcolor{benred8}{15}~With that, they must have a game-place at Jerusalem, after the Gentile fashion, \textcolor{benred8}{16}~ay, and go uncircumcised; forgotten, their loyalty to the holy covenant, they must throw in their lot with the heathen, and become the slaves of impiety.
\textcolor{benred8}{17}~And now that he was firmly established on his throne, Antiochus would be lord of Egypt, and wear two crowns at once. \textcolor{benred8}{18}~So, with overwhelming force, with chariots and elephants and horsemen and a great array of ships, he marched on Egypt, \textcolor{benred8}{19}~and levied war against king Ptolemy, that could not hold his ground, but fled away, leaving many fallen. \textcolor{benred8}{20}~So Antiochus made himself master of all the strongholds in Egypt, and ransacked it for spoil; \textcolor{benred8}{21}~then, in the hundred and forty-third year, he turned his victorious march against Israel. \textcolor{benred8}{22}~With all that great army of his he came to Jerusalem \textcolor{benred8}{23}~and entered the sanctuary in royal state; the golden altar, the lamp-stand with its appurtenances, the table where bread was set out, beaker and goblet and golden bowl, curtain and capital and golden facings of the temple, all alike were stripped. \textcolor{benred8}{24}~Silver nor gold was spared, nor any ornament of price, nor hoarded treasures could he but find them; and thus laden he went back to his own country, \textcolor{benred8}{25}~first shedding a deal of blood, and speaking very blasphemously.
\textcolor{benred8}{26}~Loud mourning there was in Israel, mourning in all the country-side; \textcolor{benred8}{27}~wept ruler and elder, pined man and maid, and colour fled from woman\textquotesingle s cheeks; \textcolor{benred8}{28}~bridegroom took up the dirge, bride sat in her bower disconsolate; \textcolor{benred8}{29}~here was a land that trembled for its inhabitants, a whole race covered with confusion.
\textcolor{benred8}{30}~Two years passed, and then the king sent his chief collector of revenue to visit the cities of Juda. To Jerusalem he came, with a great rabble at his heels, \textcolor{benred8}{31}~and won credence with idle professions of friendship. \textcolor{benred8}{32}~Then he fell suddenly on the town and grievously mishandled it, slaying Israelites a many, \textcolor{benred8}{33}~plundering the city and setting fire to it. Houses and encircling walls of it were thrown down in ruins, \textcolor{benred8}{34}~women and children carried off into slavery, cattle driven away. \textcolor{benred8}{35}~And as for David\textquotesingle s Keep, they enclosed it with high, strong walls, and strong towers besides, to serve them for a fortress; \textcolor{benred8}{36}~garrisoned it with a godless crew of sinners like themselves, and made it fast, storing it with arms and provisions, besides the plunder they had amassed in Jerusalem, \textcolor{benred8}{37}~which they bestowed there for safety. Alas, what peril of treachery was here, \textcolor{benred8}{38}~what an ambush laid about the holy place, what devil\textquotesingle s work against Israel! \textcolor{benred8}{39}~What a tide of guiltless blood must flow about the sanctuary, till it was a sanctuary no more! \textcolor{benred8}{40}~Little wonder if the inhabitants of Jerusalem took to flight, leaving their city to strangers; mother so unnatural her own children must forsake. \textcolor{benred8}{41}~Her sanctuary a desert solitude, her feasts all lament, her sabbaths derided, her greatness brought low! \textcolor{benred8}{42}~Her pride was the measure of that abasement, her glory of that shame.
\textcolor{benred8}{43}~And now came a letter from king Antiochus to all the subjects of his realm, bidding them leave ancestral custom of this race or that, and become one nation instead. \textcolor{benred8}{44}~As for the heathen, they fell in readily enough with the royal will; \textcolor{benred8}{45}~and in Israel itself there were many that chose slavery, offering sacrifice to false gods and leaving the sabbath unobserved. \textcolor{benred8}{46}~Both in Jerusalem and in all the cities of Juda the king\textquotesingle s envoys published this edict; men must live by the law of the heathen round about, \textcolor{benred8}{47}~burnt-sacrifice, offering and atonement in God\textquotesingle s temple should be none, \textcolor{benred8}{48}~nor sabbath kept, nor feast-day. \textcolor{benred8}{49}~And, for the more profanation of the sanctuary, and of Israel\textquotesingle s holy people, \textcolor{benred8}{50}~altar and shrine and idol must be set up, swine\textquotesingle s flesh offered, and all manner of unhallowed meat; \textcolor{benred8}{51}~children be left uncircumcised, and their innocent lives contaminated with rites unclean, abominable; till the law should be forgotten, and the divine precepts fashioned anew. \textcolor{benred8}{52}~Durst any neglect the royal bidding, he must die.
\textcolor{benred8}{53}~Through the whole of his dominions the king\textquotesingle s writ ran, and commissioners were appointed besides to enforce it; \textcolor{benred8}{54}~no city of Juda but was ordered to do sacrifice. \textcolor{benred8}{55}~Many there were, traitors to the divine law, that took their part, and much mischief they did, \textcolor{benred8}{56}~driving the men of Israel to seek refuge in hiding, where refuge was to be had. \textcolor{benred8}{57}~It was on the fifteenth of Casleu, in the hundred and forty-fifth year, that king Antiochus set up an idol to desecrate God\textquotesingle s altar;\footnote[1]{Cf. Dan. 11.31, Mt. 24.15.} shrines there were in every township of Juda, \textcolor{benred8}{58}~offering of incense and of victims before house doors and in the open street; \textcolor{benred8}{59}~never a copy of the divine law but was torn up and burned; \textcolor{benred8}{60}~if any were found that kept the sacred record, or obeyed the Lord\textquotesingle s will, his life was forfeit to the king\textquotesingle s edict. \textcolor{benred8}{61}~Month by month such deeds of violence were done, in all townships where men of Israel dwelt, \textcolor{benred8}{62}~and on the twenty-fifth of the month sacrifice was made at the shrine that overshadowed the altar. \textcolor{benred8}{63}~Death it was for woman to have her child circumcised in defiance of the king; \textcolor{benred8}{64}~there in her own house she must be hung up, with the child about her neck, and the circumciser, too, must pay for it with his life. \textcolor{benred8}{65}~Many a son of Israel refused the unclean food, preferring death to defilement; \textcolor{benred8}{66}~and die they must, because they would not break God\textquotesingle s holy law. \textcolor{benred8}{67}~Grievous, most grievous was the doom that hung then over his people.
\begin{large}\begin{center}\textsc{Chapter ii}\end{center}\end{large}
\lettrine[lines=2]{I}{n} those days it was that Mattathias came forward, son of John, son of Simeon, a priest of Joarib\textquotesingle s family; he was for Jerusalem no more, but would take up his dwelling on the hill-side at Modin. \textcolor{benred8}{2}~Five sons he had, John, that was also called Gaddis, \textcolor{benred8}{3}~Simon (or Thasi), \textcolor{benred8}{4}~Judas (or Machabaeus), \textcolor{benred8}{5}~Eleazar (or Abaron), and Jonathan (or Apphus) \textcolor{benred8}{6}~and these saw well what foul things were a-doing in Juda\textquotesingle s country and the city of Jerusalem. \textcolor{benred8}{7}~Alas, what needed it, cried Mattathias, I should have been born into such an age as this? To see my people and the holy city alike brought to ruin, to sit by while the enemy overcame her, \textcolor{benred8}{8}~and in her very sanctuary the alien had his will? Temple of hers like a churl\textquotesingle s lot disregarded, \textcolor{benred8}{9}~rare treasure of hers into exile carried away; young and old, in the open streets of her, put to the sword! \textcolor{benred8}{10}~Never a race in heathendom but may parcel out her domains, grow rich with the spoil of her! \textcolor{benred8}{11}~Gone, all her fair adornment; the mistress is turned maid; \textcolor{benred8}{12}~laid waste, yonder sanctuary, that was our prize and pride, by Gentile feet dishonoured! \textcolor{benred8}{13}~And would we live yet?
\textcolor{benred8}{14}~With that, they tore their garments about them, Mattathias and his sons, and went clad in sackcloth, mourning right bitterly. \textcolor{benred8}{15}~And now the pursuivants of king Antiochus came to Modin; take refuge there who might, he must do sacrifice none the less, and burn incense, and leave the following of God\textquotesingle s law. \textcolor{benred8}{16}~Out went the folk of Israel to meet them, some complaisantly enough, but Mattathias and his sons firm in their resolve. \textcolor{benred8}{17}~And they singled out Mattathias from the rest; A man of mark, said they, and a great chieftain thou; brethren and sons thou hast a many. \textcolor{benred8}{18}~Wilt thou not be the first to come forward and do the king\textquotesingle s bidding, with the whole world, and the men of Juda everywhere, and what is left of Jerusalem? To be the king\textquotesingle s friend, thou and thy sons with thee, gold and silver and much else for thy reward! \textcolor{benred8}{19}~Loud rang the answer of Mattathias: What though king Antiochus have the whole world for his vassals? Obey the edict who will, forsaking the custom his fathers lived by, \textcolor{benred8}{20}~both I and son of mine, both I and clansman of mine, will obey the law handed down to us. \textcolor{benred8}{21}~Mercy of God! What needs it we should leave his will undone, his claims unhonoured? \textcolor{benred8}{22}~To deaf ears king Antiochus proclaims the sacrifice; we swerve not from the law\textquotesingle s path, right or left.
\textcolor{benred8}{23}~Before he had done speaking, a Jew came to offer the false gods sacrifice, there in full view of all, before the altar at Modin, as the king bade. \textcolor{benred8}{24}~Mattathias took fire at the sight of it; one heave of anger his heart gave, and his zeal for the law could contain itself no longer; there on the altar the sacrificer was slain. \textcolor{benred8}{25}~Nor spared he the pursuivant of king Antiochus that enjoined it; the altar, too, he pulled down. \textcolor{benred8}{26}~Not Phinees himself struck a better blow for the law, when he slew Zamri, the son of Salom!\footnote[1]{Cf. Num. 25.7.} \textcolor{benred8}{27}~And now Mattathias raised a cry in the city, Who loves the law? Who keeps the covenant unbroken? Out with you, and follow me! \textcolor{benred8}{28}~So fled he with his sons into the hill-country, leaving his possessions behind, there in the city.
\textcolor{benred8}{29}~Many there were that went out into the desert at this time, for love of truth and right; \textcolor{benred8}{30}~took children and women-folk and cattle with them, and settled down there, castaways in a flood of misfortune. \textcolor{benred8}{31}~But news of it reached Jerusalem, and the king\textquotesingle s men that were in David\textquotesingle s Keep; here were rebels lurking in the waste country, and drawing many over to their side. \textcolor{benred8}{32}~So they went out in pursuit, and offered battle; on a sabbath day, as it chanced.\footnote[2]{The words \textasciigrave as it chanced\textquotesingle  are not in the original, but it does not seem to be implied that the king\textquotesingle s men purposely chose the sabbath day for making their assault.} \textcolor{benred8}{33}~What, still stubborn? cried they. Come out, and yield yourselves to the king\textquotesingle s pleasure; your lives shall be spared. \textcolor{benred8}{34}~But the Jews\textquotesingle  answer was, come out and yield to the king\textquotesingle s pleasure they might not; law of the sabbath rest forbade it. \textcolor{benred8}{35}~So the attack began in good earnest; \textcolor{benred8}{36}~but the Jews made no resistance, never a stone flew, never a hiding-place of theirs was put in a state of defence; \textcolor{benred8}{37}~Die we all, they said, innocent men, and let heaven and earth bear witness, it was for no fault of ours we died. \textcolor{benred8}{38}~Thus, because it was a sabbath day when the attack was made, these men perished, and their wives and children and cattle with them; a thousand human lives lost.
\textcolor{benred8}{39}~Great grief it was to Mattathias and his company when they heard what had befallen them; \textcolor{benred8}{40}~and now there was high debate raised: Do we as our brethren did, forbear we to give battle for our lives and loyalties, and they will soon make an end of us! \textcolor{benred8}{41}~Then and there it was resolved, if any should attack them on the sabbath day, to engage him, else they should be put to death all of them, like those brethren of theirs in the covert of the hills. \textcolor{benred8}{42}~Now it was that the Assidaeans rallied to their side, a party that was of great consequence in Israel, lovers of the law one and all; \textcolor{benred8}{43}~and all who would escape from the evils of the time made common cause with them, and came to their assistance. \textcolor{benred8}{44}~So, mustering their forces, they wrought indignant vengeance upon sinners that were false to the law, till they were fain to take refuge among the heathen; \textcolor{benred8}{45}~wherever they went, Mattathias and his company, they threw the altars down, \textcolor{benred8}{46}~and whatever children they found uncircumcised, from one end of Israel to the other, they circumcised by right of conquest.\footnote[3]{It is not clear whether we are meant to understand that Gentile as well as Jewish children were circumcised.} \textcolor{benred8}{47}~Ere long, they drove the tyrant\textquotesingle s minions before them, and to such good purpose \textcolor{benred8}{48}~that Gentile was none, king though he were, could restrain the law\textquotesingle s observance; against their onslaught the powers of evil could not make head.
\textcolor{benred8}{49}~Meanwhile, the life of Mattathias was drawing to an end. And this charge he gave to his sons: Here be days when tyrant and blasphemer have their will, when all is calamity and bitter retribution. \textcolor{benred8}{50}~The more reason, my sons, why you should be jealous lovers of the law, ready to give your lives for that covenant your fathers knew. \textcolor{benred8}{51}~Your fathers, what deeds they did in their time! Great glory would you win, and a deathless name, let these be your models. \textcolor{benred8}{52}~See how Abraham was tested, and how trustfulness of his was counted virtue in him; \textcolor{benred8}{53}~see how Joseph in ill fortune was true to the commandment still, and came to be ruler of all Egypt. \textcolor{benred8}{54}~Here was Phinees, our own father, that grew hot in God\textquotesingle s cause, and earned the right of priesthood inalienable; \textcolor{benred8}{55}~and Josue, that for his loyalty was given command of Israel; \textcolor{benred8}{56}~and Caleb, that spoke out in the assembly, what broad acres were his! \textcolor{benred8}{57}~David, for the tender heart of him, left a dynasty that fails not; \textcolor{benred8}{58}~for Elias heaven opened, that was champion of the law; \textcolor{benred8}{59}~by faith Ananias, Azarias and Misael overcame the furnace, \textcolor{benred8}{60}~nor Daniel\textquotesingle s innocence might ravening lions devour. \textcolor{benred8}{61}~No generation but proves it; want they never for strength that trust in God. \textcolor{benred8}{62}~What, would you fear the tyrant\textquotesingle s threats? In dung and worms his glory shall end; \textcolor{benred8}{63}~all royal state to-day, and to-morrow there shall be no news of him; gone back to the dust he came from, and all his designs brought to nothing!
\textcolor{benred8}{64}~Nay, my sons, take courage; in the law\textquotesingle s cause rally you, in the law\textquotesingle s annals you shall win renown. \textcolor{benred8}{65}~Here is your brother Simon, trust me, a man of prudence; to him ever give heed, he is your father now. \textcolor{benred8}{66}~And here is Judas Machabaeus, from boyhood\textquotesingle s days a warrior; let him be your leader, and fight Israel\textquotesingle s battles. \textcolor{benred8}{67}~All lovers of the law make free of your fellowship; bring your country redress, \textcolor{benred8}{68}~and pay the Gentiles what they have earned; yet heeding ever what the law enjoins. \textcolor{benred8}{69}~With that, he gave them his blessing, and became part of his race. \textcolor{benred8}{70}~He was a hundred and forty-six years old when he died; his sons buried him where his fathers were buried, at Modin, and great lament all Israel made for the loss of him.
\begin{large}\begin{center}\textsc{Chapter iii}\end{center}\end{large}
\lettrine[lines=2]{A}{nd} now his son Judas, that was called Machabaeus, came forward to succeed him; \textcolor{benred8}{2}~nor any of Judas\textquotesingle  clan, nor any that had taken his father\textquotesingle s part, but lent him their aid still; right merrily they fought Israel\textquotesingle s battle. \textcolor{benred8}{3}~Here was one that brought his race renown; as great a warrior as ever donned breastplate, or armed himself for the fight, or drew sword to save his camp from peril; \textcolor{benred8}{4}~lion-hearted his deeds, not lion itself more relentless in pursuit. \textcolor{benred8}{5}~Traitors he ever sought out and hunted down, ever with fire-brand the oppressors of his people dislodged, \textcolor{benred8}{6}~till enemy was none but was daunted by the fear of him, traitor was none but fled in confusion, so well sped he the work of deliverance. \textcolor{benred8}{7}~Great deeds, that kings rued bitterly, Jacob with exultation heard, posterity holds blessed evermore! \textcolor{benred8}{8}~From city to city he went, ridding Juda of its law-breakers, averting the vengeance guilt of theirs had deserved; \textcolor{benred8}{9}~no corner of earth but he was renowned there, for one that had been able to rally a doomed people.
\textcolor{benred8}{10}~How sped Apollonius, that mustered a great force, of Gentiles and Samaritans both, to fight against Israel? \textcolor{benred8}{11}~No sooner Judas heard of it, than he met and routed and slew him; fell many and fled more, \textcolor{benred8}{12}~leaving their spoils behind them. The sword of Apollonius Judas himself carried away; and this it was he evermore used in battle.
\textcolor{benred8}{13}~And next it was Seron, captain of the armies in Syria, heard what a great retinue and faithful following Judas had; \textcolor{benred8}{14}~and nothing would serve, but he must win renown and high favour at court by crushing Judas, and all other his companions that defied the king\textquotesingle s edict. \textcolor{benred8}{15}~So he made all ready, and marched in with a strong muster of the ungodly at his heels, to be even with the men of Israel. \textcolor{benred8}{16}~As far as Bethoron pass they reached, and there Judas met them with his company, no better than a handful. \textcolor{benred8}{17}~These, when they saw such a host facing them, were for counsels of prudence; What, they said to Judas, should we offer battle to foes so many and so strong, faint as we be from a day of hungry marching? \textcolor{benred8}{18}~Nay, said Judas, nothing forbids great numbers should be at the mercy of small; what matter makes it to the God of heaven, few be his soldiers or many when he grants deliverance? \textcolor{benred8}{19}~Armed might avails not to win the day; victory is from above. \textcolor{benred8}{20}~What though they come to meet us in the proud confidence of superior strength, and think it an easy matter to slay us, slay our wives and children, plunder our goods? \textcolor{benred8}{21}~Life and loyalty at stake, we will offer battle none the less; \textcolor{benred8}{22}~and he, the Lord, will crush them to earth at our coming; never be afraid. \textcolor{benred8}{23}~And with that, all unawares, he fell upon Seron and his army, that were crushed, sure enough, by his onslaught; \textcolor{benred8}{24}~all down the pass of Bethoron he gave them chase, down into the plain, and eight hundred of them had fallen before ever they took refuge in the country of the Philistines.
\textcolor{benred8}{25}~By this, the neighbouring peoples had begun to take alarm, so formidable did Judas and his brethren appear to them, \textcolor{benred8}{26}~and the renown of him reached the king\textquotesingle s court; all the world was talking of Judas and his victories. \textcolor{benred8}{27}~An angry man was king Antiochus when the news came to him; he sent word round, and had all his army summoned together, a brave array, be sure of it. \textcolor{benred8}{28}~The treasury must be opened, to provide the troops with a whole year\textquotesingle s pay, and keep them in readiness for every need. \textcolor{benred8}{29}~Why, what was this? So heavily had Juda suffered, so great the discord he had aroused by the abolishing of its ancient usages, that scant revenue had come in from it, and the treasury was in default! \textcolor{benred8}{30}~Whence, now, to defray the cost of that largesse he had made so often, and with so lavish a hand? Never was king before him could rival his munificence.
\textcolor{benred8}{31}~Here was the king in great confusion of mind; and his thought was, to march into Persia and take toll of those countries; great store of money he might there amass. \textcolor{benred8}{32}~He left Lysias behind, that was a man of high rank and royal blood; he was to administer all the business of the kingdom, from Euphrates down to the Brook of Egypt, \textcolor{benred8}{33}~and have charge of the young prince Antiochus, until the king\textquotesingle s return. \textcolor{benred8}{34}~Half his army he entrusted to Lysias, and the elephants besides; and he signified all that he would have done, concerning Juda and Jerusalem particularly. \textcolor{benred8}{35}~A force must be sent to overpower all that fought for Israel, or were yet left in Jerusalem, and make a clean riddance of them; no trace of these must be left; \textcolor{benred8}{36}~all through the country settlers must be brought in from abroad, and the lands allotted between them. \textcolor{benred8}{37}~With that, the king left his capital of Antioch, taking the remainder of his army with him; it was the hundred and forty-seventh year of the empire. Soon he was across Euphrates river, and on the march through the high countries.
\textcolor{benred8}{38}~Three generals Lysias appointed for the task; Ptolemy son of Dorymenes, Nicanor and Gorgias, nobles all that were high in the royal favour; \textcolor{benred8}{39}~with forty thousand foot and seven thousand horse they were to march on Juda and make an end of it, as the king had ordered. \textcolor{benred8}{40}~So out they went, with all this army at their back, marched in, and pitched their tents near Emmaus, down in the valley. \textcolor{benred8}{41}~Be sure the traders all about were apprised of their coming, and made their way into the camp with great sums of silver and gold, and a retinue of servants besides, thinking to buy Israelite slaves; levies, too, from Syria and Philistia made common cause with the invader.
\textcolor{benred8}{42}~Judas, then, and his brethren found that matters had gone from bad to worse; here were the enemy encamped within their frontiers; they heard, besides, what orders the king had given for the destruction and taking away of their people. \textcolor{benred8}{43}~And the word went round among them, Now to restore the lost fortunes of our race; now to do battle for people of ours, sanctuary of ours! \textcolor{benred8}{44}~So a general assembly was called; they must make ready for the fight, and pray besides, to win mercy and pardon. \textcolor{benred8}{45}~Not at Jerusalem; Jerusalem lay there, no city but a desert waste, nor any of her sons came and went; her sanctuary defiled, her citadel garrisoned by the alien, she was but a haunt of the Gentiles. Sad days for the men of Jacob; pipe nor harp sounded there now. \textcolor{benred8}{46}~At Maspha, then, they gathered, looking across towards Jerusalem; time was when Maspha, too, had its place of prayer. \textcolor{benred8}{47}~All that day they fasted, and wore sackcloth, and covered their heads with ashes, and tore their garments about them.
\textcolor{benred8}{48}~What sights were these? Here, lying open, was a copy of the law, such as the heathen were ever making search for,~\ldots\  the counterpart of their own images.\footnote[1]{As the sentence stands, it can only mean that heathen pursuivants had confiscated the scriptures, as being objects of veneration to the Jews no less than idols were to themselves. But it seems possible that the text is deficient.} \textcolor{benred8}{49}~Here they had brought priestly vestments, and offering of first-fruits and tithes; here Nazirites were gathered, ripe and ready for the payment of their vows. \textcolor{benred8}{50}~And a loud cry rose to heaven, What shall we do for these, thy votaries? Whither escort them now? \textcolor{benred8}{51}~Sanctuary of thine is all profanation and defilement, priesthood of thine all misery and despair. \textcolor{benred8}{52}~And now, see where the heathen muster their armies to destroy us! Needs not we should tell thee, how murderous their intent. \textcolor{benred8}{53}~Lord, but for thy aid, how shall we resist their onslaught? \textcolor{benred8}{54}~Loudly their voices, and loud the trumpets rang.
\textcolor{benred8}{55}~Thereupon Judas chose out who should be their leaders, one with a thousand, one with a hundred, one with fifty, one with ten men to follow him; \textcolor{benred8}{56}~he sent home, too, all such as the law holds exempt; all that had but just built house, or married wife, or planted vineyard, and whoever had no stomach to the fight.\footnote[2]{See Deut. 20.5-8.} \textcolor{benred8}{57}~Then they moved camp, and pitched their tents southward of Emmaus. \textcolor{benred8}{58}~Now for girded loins, cried Judas, and brave hearts! By to-morrow\textquotesingle s light, you must engage yonder heathen, sworn enemies to us, and to the ground we hold sacred. \textcolor{benred8}{59}~Better die in battle, than live to see our race and our sanctuary overpowered. \textcolor{benred8}{60}~Be it what it may, heaven\textquotesingle s will be done!
\begin{large}\begin{center}\textsc{Chapter iv}\end{center}\end{large}
\lettrine[lines=2]{T}{hat} night, a detachment of five thousand foot and a thousand picked horsemen left their lines, under the command of Gorgias, \textcolor{benred8}{2}~thinking to reach the Jewish camp and strike a sudden blow at it; for guides, they had men of the Jerusalem garrison. \textcolor{benred8}{3}~But Judas had word of it; out he went, and all his valiant company with him, to attack the main body of the king\textquotesingle s army at Emmaus, \textcolor{benred8}{4}~while the defences of the camp were yet scattered. \textcolor{benred8}{5}~So Gorgias, making his night attack on the camp of Judas, and finding it empty, made no doubt they had given him the slip, and fell to scouring the hill-country for them; \textcolor{benred8}{6}~meanwhile, came day-break, and there was Judas down in the valley. True, there were but three thousand at his back, for defence and attack very ill arrayed; \textcolor{benred8}{7}~and here was this army of heathen folk, both strong and well protected, with cavalry circling about them, men bred to war! \textcolor{benred8}{8}~But Judas cried to his fellows, What, would you be daunted by the numbers of them? Would you give ground before their attack? \textcolor{benred8}{9}~Bethink you, what a host it was Pharao sent in pursuit of our fathers, there by the Red Sea, and they escaped none the less. \textcolor{benred8}{10}~Now, as then, besiege we heaven with our cries; will not the Lord have mercy? Will he not remember the covenant he had with our fathers, and rout, this day, yonder army at our coming? \textcolor{benred8}{11}~No doubt shall the world have thenceforward, but there is one claims Israel for his own, and grants her deliverance.
\textcolor{benred8}{12}~And now the heathen folk caught sight of them as they advanced to the attack, \textcolor{benred8}{13}~and left their lines to give battle. Thereupon Judas\textquotesingle  men sounded with the trumpet, \textcolor{benred8}{14}~and the two armies met. Routed the Gentiles were, sure enough, and took to their heels across the open country, \textcolor{benred8}{15}~sword of the pursuer ever catching the hindmost. All the way to Gezeron they were chased, and on into the plains by Idumaea,\footnote[1]{For \textasciigrave Idumaea\textquotesingle  some Greek manuscripts read \textasciigrave Judaea\textquotesingle . Neither reading gives a good sense, and it seems possible that the name of Accaron, or some other Philistine town, has been accidentally miswritten.} Azotus and Jamnia, with a loss of three thousand men. \textcolor{benred8}{16}~When Judas and his army came back from the pursuit, \textcolor{benred8}{17}~Not yours, he told them, to run greedily after the spoils of the camp; there is battle still awaiting us over yonder. \textcolor{benred8}{18}~Not far away, in the hill-country, lie Gorgias and his army; first meet you and beat you the enemy, and then you shall fall to your pillaging unafraid. \textcolor{benred8}{19}~Even as he spoke, they were ware of a company that watched them from the hill-side. \textcolor{benred8}{20}~But by now the camp was on fire, and it needed no more than the smoke of it to warn Gorgias of his defeat; \textcolor{benred8}{21}~that sight took the heart out of Syria, the more so when it proved that Judas and his army were in the valley, all appointed for battle, \textcolor{benred8}{22}~and they fled for their lives, down into the plain of Philistia. \textcolor{benred8}{23}~So to the pillaging of the camp Judas returned; what gold and silver they found there, what garments of blue and sea-purple, what rich treasures! \textcolor{benred8}{24}~Be sure there was singing of songs on their homeward journey, as they praised God in heaven, God who is gracious, whose mercy endures for ever. \textcolor{benred8}{25}~Here was a day of signal deliverance for Israel.
\textcolor{benred8}{26}~And what of Lysias? News reached him, through the survivors, of what had befallen, \textcolor{benred8}{27}~and he was both sick and sorry at the hearing; his own will crossed, and his master\textquotesingle s command ill carried out! \textcolor{benred8}{28}~So, in the following year, he made a muster of sixty thousand picked men, with five thousand horse, to crush the rebellion; \textcolor{benred8}{29}~into Judaea they marched, and encamped at Bethoron, where Judas met them with ten thousand.\footnote[2]{For \textasciigrave Judaea\textquotesingle  the Greek manuscripts have \textasciigrave Idumaea\textquotesingle , and for \textasciigrave Bethoron\textquotesingle , \textasciigrave Bethsura\textquotesingle . Cf. verse 61 below.} \textcolor{benred8}{30}~At the sight of their great numbers, this was Judas\textquotesingle  prayer: Blessed art thou, Saviour of Israel, who didst make use of thy servant David, a giant\textquotesingle s onset to overthrow! Victory thou didst give, over an invading army, to Saul\textquotesingle s son Jonathan and the squire that bore him company! \textcolor{benred8}{31}~So may yonder host, left at Israel\textquotesingle s mercy, unlearn its confidence in strength and in speed; \textcolor{benred8}{32}~strike terror into them, let their manhood melt away, as they tremble at the approach of doom; \textcolor{benred8}{33}~sword of thy true lovers be their undoing, triumph-song of thy worshippers their dirge! \textcolor{benred8}{34}~With that, battle was joined, and of Lysias\textquotesingle  men, five thousand were left dead on the field. \textcolor{benred8}{35}~What should he do? Here were his troops fled in disorder, here was Judas in command of brave men, that would as soon have an honourable death as life itself. Back he went to Antioch, and there levied soldiers for a greater expedition yet against Judaea.
\textcolor{benred8}{36}~And now Judas and his brethren had but one thought; the enemy vanquished, they would betake themselves to Jerusalem, to cleanse and restore the sanctuary. \textcolor{benred8}{37}~So the whole army fell into rank, and they climbed the hill of Sion together. \textcolor{benred8}{38}~What saw they? The holy place desolate, the altar profaned, charred gates, courts overgrown with brushwood, like forest clearing or mountain glen, the priests\textquotesingle  lodging in ruins. \textcolor{benred8}{39}~Upon this, there was rending of garments, and loud lament; dust they cast on their heads, \textcolor{benred8}{40}~and fell face to earth; then, as the trumpet\textquotesingle s note gave the summons, raised their cries to heaven.
\textcolor{benred8}{41}~And what did Judas? First, he sent a force to engage the citadel\textquotesingle s garrison, while the holy place was a-cleansing; \textcolor{benred8}{42}~then he chose priests, without blot or blemish, and true lovers of the law besides, \textcolor{benred8}{43}~who thereupon cleansed the sanctuary, nor any stone that was polluted with idolatry but they had it away into a place unclean. \textcolor{benred8}{44}~And next, he must concern himself with the altar of burnt-sacrifice, that was now all defiled. \textcolor{benred8}{45}~And it was good counsel they took; the altar must be destroyed, else the day when the Gentiles polluted it should be remembered to their shame. So destroy it they did, \textcolor{benred8}{46}~and laid up the stones in a place apt for their purpose, there on the temple hill. Here they must remain, until the coming of a prophet that should give sentence, what was to be done with them.
\textcolor{benred8}{47}~Then they raised a new altar in place of the old, using stones that had never felt the pick, as the law bade;\footnote[3]{See Ex. 20.25.} \textcolor{benred8}{48}~repaired shrine and inner walls, and rid both temple and temple courts of their defilement. \textcolor{benred8}{49}~New appurtenances, too, the temple must have, lamp-stand, incense-altar and table be restored to it; \textcolor{benred8}{50}~incense be put on the altar, lamps kindled to light the holy place, \textcolor{benred8}{51}~loaves set out on the table, and veils hung up; then at length their task was accomplished. \textcolor{benred8}{52}~On the twenty-fifth of Casleu, the ninth month, in the hundred and forty-eighth year, they rose before daybreak, \textcolor{benred8}{53}~and offered sacrifice, as the law bade, on the new altar they had set up. \textcolor{benred8}{54}~This was the very month, the very day, when it had been polluted by the Gentiles; now, on the same day of the same month, it was dedicated anew, with singing of hymns, and music of harp, zither and cymbals. \textcolor{benred8}{55}~Thereupon all the people fell down face to earth, to adore and praise, high as heaven, the author of their felicity; \textcolor{benred8}{56}~and for eight days together they celebrated the altar\textquotesingle s renewal, burned victim and brought welcome-offering with glad and grateful hearts.
\textcolor{benred8}{57}~They decked the front wall of the temple, at this time, with gold crowns and escutcheons, consecrated the gates and the priest\textquotesingle s lodging anew, and furnished it with doors; \textcolor{benred8}{58}~and all the while there was great rejoicing among the people; as for the taunts of the heathen, they were heard no more. \textcolor{benred8}{59}~No wonder if Judas and his brethren, with the whole assembly of Israel, made a decree that this feast should be kept year by year for eight days together, the feast-day of the altar\textquotesingle s dedication.\footnote[4]{Cf. Jn. 10.22.} Came that season, from the twenty-fifth day of Casleu onwards, all was to be rejoicing and holiday. \textcolor{benred8}{60}~At this time, too, they fortified the hill of Sion, with walls and strong towers all about; never more should Gentile feet profane it. \textcolor{benred8}{61}~Judas put a garrison there, and would have it strong enough to command Bethsura; a bulwark Israel must have against attack from the frontiers of Edom.
\begin{large}\begin{center}\textsc{Chapter v}\end{center}\end{large}
\lettrine[lines=2]{G}{reat} indignation had the Gentiles that lived round about, when they heard that altar and temple were standing as of old. \textcolor{benred8}{2}~Their first thought was to rid their own territory of Jacob\textquotesingle s breed, and all at once they set about to murder and harry them. \textcolor{benred8}{3}~So Judas must needs take arms against them, Esau\textquotesingle s race in Idumaea, and the men of Acrabathane, that were keeping Israelite folk under strict siege; and signally he defeated them. \textcolor{benred8}{4}~Nor might he overlook Ben\textquotesingle s tribe and the treachery they shewed, ever catching Israel at unawares by laying an ambush in his path. \textcolor{benred8}{5}~These he chased into their strongholds and besieged them there; laid them under a ban\footnote[1]{Cf. Jos. 6.17, 18 and other passages.} and burned the strongholds to the ground, with their defenders in them. \textcolor{benred8}{6}~Then he crossed over into Ammon, where he came upon strong resistance and a great muster of men, that had one Timotheus for their leader; \textcolor{benred8}{7}~often he engaged them, and as often put them to rout; when he had defeated them, \textcolor{benred8}{8}~and taken Gazer with its daughter townships, he marched back into Judaea.
\textcolor{benred8}{9}~But by this all the heathen folk in the country of Galaad were making common cause against the Israelites who dwelt there, eager to be rid of them. And these, taking refuge in the stronghold of Datheman, \textcolor{benred8}{10}~sent dispatches to Judas and his brethren. Here be all the neighbours, they wrote, banded together for our destruction. \textcolor{benred8}{11}~Even now, Timotheus at their head, they are setting about the reduction of this our fortress; \textcolor{benred8}{12}~come speedily to the rescue; they have taken cruel toll of our lives already. \textcolor{benred8}{13}~Slain, all those clansmen of ours that had their dwelling in the Tubin country, carried away, their wives, their children, and their goods; nigh upon a thousand warriors then and there have perished.
\textcolor{benred8}{14}~This letter was still in the reading, when all of a sudden came other envoys from Galilee, their garments rent about them; their message was, \textcolor{benred8}{15}~Ptolemais, Tyre and Sidon were up in arms together, and all Galilee was overrun with heathen folk, bent on massacre. \textcolor{benred8}{16}~Grave tidings, these, for Judas and his people; met they in high debate, and took counsel how they might best aid their brethren in peril of assault. \textcolor{benred8}{17}~And now Judas must share the command with his brother Simon; Pick thy men, said he, and make for Galilee, while Jonathan and I march into Galaad. \textcolor{benred8}{18}~Part of his army he left to defend Judaea, with Joseph son of Zachary and Azarias for its captains; \textcolor{benred8}{19}~Here is your charge, said he; and see to it that you do not embroil yourselves with the Gentiles while we are gone. \textcolor{benred8}{20}~To Simon and to Galilee three thousand men were allotted; to Judas and to Galaad eight thousand. \textcolor{benred8}{21}~As for Simon, when he reached Galilee, full many a battle he must fight with the Gentiles, that he drove ever before him, till he pursued them at last to the very gates of Ptolemais. \textcolor{benred8}{22}~Of the enemy, some three thousand fell, and his men had the spoiling of them; \textcolor{benred8}{23}~the Israelites that dwelt in Galilee and Arbata he took home with him, and their wives and children and all they had; great rejoicing there was when he brought them back safe to Judaea.
\textcolor{benred8}{24}~Meanwhile Judas Machabaeus and his brother Jonathan had crossed the Jordan, and marched for three days through the desert. \textcolor{benred8}{25}~There the Nabuthaeans came to meet them, and gave them friendly welcome, and told them of all that had befallen their brethren in the Galaad country; \textcolor{benred8}{26}~how there were many whom their fellow-citizens had brought to bay in such great fortified cities as Barasa, Bosor, Alima, Casphor, Mageth and Carnaim; \textcolor{benred8}{27}~besides many others cut off in the rest of the Galaadite towns. And to-morrow, he was told, the heathen mean to occupy these cities with their army, seizing upon the Israelites and making an end of them, all in one day\textquotesingle s work. \textcolor{benred8}{28}~Whereupon Judas and his men suddenly turned aside from their course into the desert of Bosor, and took the city; all its men-folk he put to the sword, and carried off the spoil of it, and burned it to the ground. \textcolor{benred8}{29}~At night-fall they continued their journey, and reached the Israelite stronghold.\footnote[2]{The word \textasciigrave Israelite\textquotesingle  is not in the original; but evidently the stronghold was either that mentioned in verse 11 above, or else one in which the Israelites of Bosor had taken refuge.} \textcolor{benred8}{30}~What a sight was this that met their eyes, when day broke! A great rabble of men past all counting, that brought up scaling-ladders and engines, as if they would take the stronghold by storm. \textcolor{benred8}{31}~Here was the battle fairly begun; the cry of them went up to heaven, loud as clarion-call, and a great cry, too, was raised within the city. \textcolor{benred8}{32}~Now, cried Judas to his men, now to fight for your brethren\textquotesingle s deliverance! \textcolor{benred8}{33}~And hard at the enemy\textquotesingle s heels he followed, with three companies of warriors that blew trumpets as they went, and cried aloud in prayer. \textcolor{benred8}{34}~The name of Machabaeus once heard, how fled Timotheus\textquotesingle  army at his approach! How grievous the blow that fell on them, when eight thousand fell in a single day! \textcolor{benred8}{35}~Once more Judas turned aside, to Maspha; took it by storm, slew men of it, took spoil of it, burned it to the ground; \textcolor{benred8}{36}~then on to seize Casbon, and Mageth, and Bosor, and the remaining cities of Galaad.\footnote[3]{There is some uncertainty about the names here; \textasciigrave Casbon\textquotesingle  is perhaps the \textasciigrave Casphor\textquotesingle  of verse 26, and \textasciigrave Bosor\textquotesingle  should perhaps be identified with the \textasciigrave Barasa\textquotesingle  (in the Greek, Bossora) of the same verse. Bosor has already been destroyed in verse 28.}
\textcolor{benred8}{37}~Yet, when all was done, Timotheus put another army into the field, and encamped close by Raphon, across the stream. \textcolor{benred8}{38}~What learned Judas from the scouts he had sent forward? Here were all the neighbouring tribes assembled in great force, \textcolor{benred8}{39}~with hired support from Arabia besides, camped beyond the stream ready to engage him; so out he marched to offer battle. \textcolor{benred8}{40}~Wait we, said Timotheus to his captains, till Judas and his army reach yonder stream. Cross he and challenge us, we may not speed; beyond doubt he has the mastery of us. \textcolor{benred8}{41}~Fear he the passage, and encamp on the further side, then cross we boldly, the day is ours. \textcolor{benred8}{42}~But Judas, when he drew near the ravine, had muster-masters in attendance by the stream, that were charged to let none linger behind, but send every man across into battle. \textcolor{benred8}{43}~So he crossed, challenging them, and all the army at his heels, and sure enough the Gentile host was routed at their coming; threw arms away, and sought refuge in the temple at Carnaim.\footnote[4]{vv. 40-43. It is not clear whether Timotheus was testing the courage of his opponents, or taking an omen from the course of events (cf. I Kg. 14.9, 10). Perhaps we are not meant to picture the crossing as taking place unopposed.} \textcolor{benred8}{44}~Upon taking the city, he burned its temple to the ground with all that were sheltered in it; so was Carnaim vanquished, and could make head against Juda no more.
\textcolor{benred8}{45}~And now Judas gathered all the Israelites in the Galaad country, high and low, with their wives and children, a whole army of them, to come back with him to Juda. \textcolor{benred8}{46}~They journeyed safely as far as Ephron, that was a great city and well fortified, the very gate of Juda; turn to right or left they might not, their road lay through the heart of it. \textcolor{benred8}{47}~And what must they do, the townspeople, but stand to the defence of it, and barricade the entrance with great boulders! Thereupon Judas made peaceful overtures to them; \textcolor{benred8}{48}~Grant us leave, said he, to make our way through your country to ours, nor any harm shall befall you; we ask but the right of passage, and on foot. But open the gates they would not; \textcolor{benred8}{49}~so Judas made a cry through the camp, every man should go to the assault,\footnote[5]{\textasciigrave Go to the assault\textquotesingle ; literally, according to the Greek, \textasciigrave to encamp\textquotesingle , but cf. the use of the corresponding Hebrew verb in Jos. 10.31 and elsewhere.} there where he stood; \textcolor{benred8}{50}~and go to the assault they did, the fighting men of his company. All day and all night they attacked the city, and Judas was given the mastery of it. \textcolor{benred8}{51}~Never a male creature there but was put to the sword; the city was plundered and pulled down; and so he passed on through the streets of it, all paved with dead men. \textcolor{benred8}{52}~Jordan they must still cross, there by the great plain that faces Bethsan; \textcolor{benred8}{53}~and to the last Judas went ever to and fro, rallying the stragglers and encouraging the people on their journey, till the land of Juda was reached. \textcolor{benred8}{54}~Glad and merry were men\textquotesingle s hearts as they climbed up Sion mountain, and there offered burnt-sacrifice in thanks for their safe home-coming, with never a life lost.\footnote[6]{Literally, \textasciigrave Because none of them had fallen until all returned safe and sound\textquotesingle , cf. note on Mt. 1.25.}
\textcolor{benred8}{55}~So fought Judas and Jonathan in Galaad, and their brother Simon in Galilee at the gates of Ptolemais; \textcolor{benred8}{56}~meanwhile, what of Joseph son of Zachary, and Azarias, that had charge of the garrison? News came to them of victories gained, and great deeds done, \textcolor{benred8}{57}~and nothing would serve but they must make a great name for themselves too, by offering battle to the Gentiles round about. \textcolor{benred8}{58}~So orders went out to the army, to march on Jamnia, \textcolor{benred8}{59}~where Gorgias and his men came out to meet them. \textcolor{benred8}{60}~Back fell Joseph and Azarias to the frontiers of Judaea in great disorder, with a loss to Israel of two thousand men; such defeat they brought on our arms, \textcolor{benred8}{61}~because they would not listen to Judas and his brethren, but must be great warriors like the rest. \textcolor{benred8}{62}~Not of that race they sprang that should afford Israel deliverance.
\textcolor{benred8}{63}~But as for Judas and his company,\footnote[7]{\textasciigrave Judas and his company\textquotesingle ; the Latin here has \textasciigrave the men of Juda\textquotesingle , which yields no good sense, \textasciigrave Juda\textquotesingle  and \textasciigrave Israel\textquotesingle  being convertible terms at this period. The Greek has, \textasciigrave the man Judas and his brethren\textquotesingle .} they were held high in honour, both among Israelite folk, and wherever the renown of them was heard; \textcolor{benred8}{64}~all flocked to greet them with cries of acclaim. \textcolor{benred8}{65}~But still he and his brethren would be on the march, reducing the men of Edom in the south country; on Hebron and its daughter townships the blow fell, neither wall nor tower of it but was burned to the ground. \textcolor{benred8}{66}~Then he moved camp, to march on Philistia, and would make his way through Samaria.\footnote[8]{\textasciigrave Samaria\textquotesingle  is probably a false reading for Maresa, which lay on the route between Edom and the Philistines.} \textcolor{benred8}{67}~Priests there were that took up arms and fell in battle that day, rashly desirous of a warrior\textquotesingle s renown.\footnote[9]{Most of the Greek manuscripts have a different and very curious reading, \textasciigrave Priests there were that fell in battle that day, because he (Judas?) desired to play the warrior, with which design he went into action unadvisedly\textquotesingle .} \textcolor{benred8}{68}~And now Judas turned aside to Azotus, in the country of the Philistines; altars he pulled down, images of their gods burned to ashes, gave up their cities to plunder, and so came back again to the land of Juda.
\begin{large}\begin{center}\textsc{Chapter vi}\end{center}\end{large}
\lettrine[lines=2]{K}{ing} Antiochus was still on his journey through the high countries, when he heard tell of a city in Persia called Elymais, renowned for its treasures of silver and gold; \textcolor{benred8}{2}~here was a temple of great magnificence, that had golden armour in it, breastplate and shield left there by Philip\textquotesingle s son, Alexander of Macedon, the first overlord of Greece. \textcolor{benred8}{3}~Thither he marched, intent on seizing the city and plundering it; but seize it he might not, because the townsfolk had news of his purpose, \textcolor{benred8}{4}~and came out to offer battle. So he was put to the rout, and must take himself back to Babylon, grievously disappointed.
\textcolor{benred8}{5}~And here, in the Persian country, a messenger reached him with tidings from Juda. Fled were his armies, \textcolor{benred8}{6}~and Lysias, that erstwhile marched out with so brave a retinue, had left the Jews masters of the field. Now they were strong and well-armed, such spoil they had taken from the armies they overthrew; \textcolor{benred8}{7}~gone was that defiling image he had set up over the altar at Jerusalem; high walls, as of old, protected the sanctuary; nay, they had made shift to fortify his own stronghold of Bethsura.
\textcolor{benred8}{8}~What news was this! The king was all bewilderment and consternation; he took to his bed, fallen into a decline for very sadness that his hopes had failed him. \textcolor{benred8}{9}~Long time he languished under the double burden of his grief, and knew at last he was a-dying. \textcolor{benred8}{10}~So he called his friends about him, and bade them farewell; Here is sleep quite gone from me, said he; so dazed is this heart of mine with doubt unresolved. \textcolor{benred8}{11}~Thus runs my thought: How comes it that I have fallen upon such evil times, such a flood of calamity as now engulfs me; I, that in the days of my greatness loved men well, and was well beloved? \textcolor{benred8}{12}~And now returns the memory of all the havoc I made in Jerusalem, spoil of gold and silver I robbed from it, doom of mine against the townsfolk, and for no fault. \textcolor{benred8}{13}~Past all doubt, here is the source of all those miseries that have come upon me; look you, how I die consumed of grief, in a strange land! \textcolor{benred8}{14}~Then he sent for Philip, one of his trusted friends, and gave all the kingdom into his charge; \textcolor{benred8}{15}~crown and robe and ring he delivered to him, bidding him seek out prince Antiochus, and bring him up as heir to the throne. \textcolor{benred8}{16}~Then and there died king Antiochus, in the hundred and forty-ninth year of the Grecian empire. \textcolor{benred8}{17}~And Lysias, hearing of his death, crowned this same prince Antiochus, that he had brought up from boyhood, giving him the name of Eupator.
\textcolor{benred8}{18}~Meanwhile, what of the Jews that dwelt about the holy place? Here was the garrison of the citadel hemming them in, seeking ever to do them injury, and to sustain the Gentile cause. \textcolor{benred8}{19}~So Judas was fain to make an end of it, and summoned the whole people to rally for the siege. \textcolor{benred8}{20}~Rally they did, and began the siege in the hundred and fiftieth year, with much contriving of catapults and engines. \textcolor{benred8}{21}~But some of the defenders slipped out; and these, with traitors of Israelite stock to support them, \textcolor{benred8}{22}~went off to gain the king\textquotesingle s audience. Wilt thou never bring redress, they asked, and do our brethren right? \textcolor{benred8}{23}~Jews are we, that resolved we would be loyal to thy father, his policy furthering, his will obeying. \textcolor{benred8}{24}~What came of it? Our own fellow Israelites would have no more of our company, slew all they could lay hands on, robbed us of our possessions. \textcolor{benred8}{25}~Not us only, but all the country about them, their violence threatens; \textcolor{benred8}{26}~even now they stand arrayed against the citadel of Jerusalem, ready to take it by storm, and have fortified Bethsura. \textcolor{benred8}{27}~Forestall their plans thou must, and speedily, or they will go further yet, and there will be no holding them.
\textcolor{benred8}{28}~Angered by these tidings, the king sent for all his trusted friends, for his army captains and his commanders of horse; \textcolor{benred8}{29}~mercenaries, too, were hired from foreign countries, and from the islands out at sea, \textcolor{benred8}{30}~till he could put a hundred thousand foot and twenty thousand horse into the field, besides thirty-two elephants, inured to war. \textcolor{benred8}{31}~Through Edom they marched, and invested Bethsura; long they held it besieged, and built engines to attack it, but these, by a brave sally, the defenders burned to ashes.
\textcolor{benred8}{32}~Meanwhile, Judas drew away from the citadel, and encamped at Bethzacharam, close to the king\textquotesingle s army. \textcolor{benred8}{33}~Ere dawn broke, the king was astir, and his men marching hot-foot towards Bethzacharam, where the armies made ready for battle, with a great blowing of trumpets. \textcolor{benred8}{34}~As for the elephants, they were blooded to battle with juice of grape and mulberry, \textcolor{benred8}{35}~and so divided here and there among the troops. A thousand foot-soldiers were assigned to each, in coat of mail and helmet of bronze; with each went five hundred picked horsemen; \textcolor{benred8}{36}~these were waiting ready for every beast at its station, and must go wherever it went, never leaving its side. \textcolor{benred8}{37}~On the back of every beast was a strong protecting tower of wood, cunningly fitted; and thirty-two valiant men were appointed to do battle from this height, over and above the Indian that was the beast\textquotesingle s driver.\footnote[1]{The meaning of the original is, that each elephant carried no less than thirty-two fighting men. The statement is perhaps due to some error in the copying of the Greek manuscripts.} \textcolor{benred8}{38}~The remainder of the cavalry were stationed on either wing, to daunt the oncoming host with a clamour of trumpets,\footnote[2]{There is no mention of trumpets in the Greek original. But the text is obscure, and perhaps corrupt.} and harass them as they stood tight packed in their ranks. \textcolor{benred8}{39}~Brightly the sun shone down on shield of gold, shield of bronze, till all the mountain-side gave back the glancing rays of them, and dazzled like points of fire. \textcolor{benred8}{40}~Part of the king\textquotesingle s army was drawn up on the heights, part on the level plain; warily they came on and in good order; \textcolor{benred8}{41}~and ever, as they went, murmur of voices, tramp of feet, and clash of arms daunted the country-side around them, so great yonder army was, and so valiant. \textcolor{benred8}{42}~But Judas and his men closed with them, and gave battle; and of the king\textquotesingle s soldiers, there were six hundred that fell.
\textcolor{benred8}{43}~What did Eleazar that day, the son of Sauran?\footnote[3]{In the Greek, \textasciigrave Eleazar Sauaran\textquotesingle , perhaps a corruption of \textasciigrave Abaron\textquotesingle  (2.5 above).} Here was one of the beasts that went decked in royal trappings, and towered high above the rest; There rides the king, thought he, \textcolor{benred8}{44}~and with that, he gave his life, to win deliverance for his country, and for himself imperishable renown. \textcolor{benred8}{45}~Bravely he ran up to it, there in the heart of the press, slaying to right and left of him, men falling on either side, \textcolor{benred8}{46}~till he could creep in between the very feet of the elephant; crouched there, and dispatched it, and so, crushed by its fall to earth, died where he lay.
\textcolor{benred8}{47}~But now, finding the royal forces so strong, and so determined in their attack, the Jews withdrew from the encounter. \textcolor{benred8}{48}~To Jerusalem the king\textquotesingle s men followed them, and now here was the king entrenched against Judaea and mount Sion itself. \textcolor{benred8}{49}~With the defenders of Bethsura he had made terms; yield up the city they must, so ill were they victualled for a siege, in a year when the land lay fallow;\footnote[4]{Cf. Lev. 25.4.} \textcolor{benred8}{50}~thus Bethsura was in the king\textquotesingle s hands, and he put a garrison there. \textcolor{benred8}{51}~But it was against the holy place itself that he turned his arms, and long he beleaguered it; what catapults he brought to bear on it, what engines! Flew fiery darts, flew stone and javelin and arrow from mangonel and arbalest, and the slings took their turn. \textcolor{benred8}{52}~As for the Jews, they met engine with engine, and fought on day after day; \textcolor{benred8}{53}~but the seventh year had come round, and what store was left in the city had been eaten up by the new citizens rescued from Gentile countries, so food was none to be had. \textcolor{benred8}{54}~Only a few defenders were left in the holy place now; the rest, overtaken by famine, had dispersed to their homes.
\textcolor{benred8}{55}~But Lysias could not wait; he had news from Antioch. That same Philip, whom king Antiochus, on his death-bed, had appointed to bring up the young prince as heir to the throne, \textcolor{benred8}{56}~was now returned at the head of his army from the land of the Medes and Persians, and would fain take charge of the realm. \textcolor{benred8}{57}~So Lysias must betake himself to the king and his generals, with such words as these: Our plight grows daily worse; scant food is left us, and here is a fortress well defended; all the business of the realm claims our care. \textcolor{benred8}{58}~What remains, but to make friendly advances, offer terms to the besieged and to all their countrymen? \textcolor{benred8}{59}~Give we leave they should follow their own customs as of old, which customs neglecting, we have brought all this ill-will and all this trouble upon us. \textcolor{benred8}{60}~King and chieftain fell in with his design; offer peace they did, and the offer was accepted. \textcolor{benred8}{61}~So, upon terms with the king and his generals, the Jews gave up their stronghold; \textcolor{benred8}{62}~and what must the king do, once he had set foot on mount Sion and discovered the strength of its defences, but break his oath, and have all the walls of it pulled down! \textcolor{benred8}{63}~Then, with all haste, he took leave of it, and returned to Antioch, where he found Philip in possession, and levied war on him, taking the city by storm.
\begin{large}\begin{center}\textsc{Chapter vii}\end{center}\end{large}
\lettrine[lines=2]{I}{t} was now, in the hundred and fifty-first year, that Demetrius, the son of Seleu-cus, escaped from Rome and landed with a small retinue at one of the sea-ports, where he was proclaimed king. \textcolor{benred8}{2}~No sooner had he set foot in the palace of his ancestors, than his men laid hold of Antiochus and Lysias, meaning to bring them into his presence. \textcolor{benred8}{3}~But he was warned of it, and gave it out, sight of them he would have none; \textcolor{benred8}{4}~so they were dispatched by the troops, and Demetrius established himself on the royal throne.
\textcolor{benred8}{5}~To him came certain Israelites, enemies of the law and of religion, with Alcimus at their head, a man who coveted the high-priestly office. \textcolor{benred8}{6}~And thus, in the royal presence, they defamed their own people: Here be Judas and his brethren have made away with all thy friends, and driven us out of our country! \textcolor{benred8}{7}~Do but send some trusted agent to survey the scene of it, the havoc this man has wrought upon our own persons and upon the king\textquotesingle s domain; ay, and to punish his partisans, with all who comfort them. \textcolor{benred8}{8}~The king\textquotesingle s choice fell on Bacchides, a courtier that was loyal to him, and had charge now of all the realm east of Euphrates. \textcolor{benred8}{9}~Of the havoc wrought by Judas he should be judge, and with him went the traitor Alcimus, now confirmed in the high priesthood; thus should the royal vengeance fall on Israel.
\textcolor{benred8}{10}~So they took the road, and reached the land of Juda with a great army at their heels. Envoys they sent out, to cheat Judas and his brethren with fair promises; \textcolor{benred8}{11}~but from these they got no hearing; the sight of such armed strength was enough. \textcolor{benred8}{12}~It was a company of scribes that went out to meet Alcimus and Bacchides, asking for honourable terms; \textcolor{benred8}{13}~of all Israel, the Assidaeans were foremost in demanding peace; \textcolor{benred8}{14}~Here is a priest of Aaron\textquotesingle s line, said they, in yonder company, fear we no treachery from him. \textcolor{benred8}{15}~Fair promises he made them, and swore they should take no harm, nor their friends neither; \textcolor{benred8}{16}~and they took him at his word. And what did he? A full sixty of them he seized and put to death in one day. Not idly the word was written, \textcolor{benred8}{17}~Bleeding corpses of thy true lovers they have strewn about on every side of Jerusalem, and there was none to bury the dead. \textcolor{benred8}{18}~After this, all alike dreaded the newcomers and shrank from them; here was neither trust nor troth, when covenant and sworn promise went for nothing. \textcolor{benred8}{19}~So Bacchides left Jerusalem and pitched his camp at Bethzecha, where he made search and laid hands on many that had deserted from his own army; some of the Jews he massacred besides, and had their bodies thrown into the Great Cistern; \textcolor{benred8}{20}~then he left the whole country in Alcimus\textquotesingle  charge, with troops to maintain him. So off went Bacchides to his master, \textcolor{benred8}{21}~and Alcimus remained to make the best of his high priesthood. \textcolor{benred8}{22}~Be sure all the malcontents in Judaea rallied to his side, and took possession of the country, to Israel\textquotesingle s great mischief.
\textcolor{benred8}{23}~Little it liked Judas, to see Alcimus and his crew mishandling the men of Israel as never the Gentiles had; \textcolor{benred8}{24}~from end to end of Juda he passed, executing vengeance on such as had left his cause, till they might take the field no longer. \textcolor{benred8}{25}~Everywhere Judas and his company had their way, and the sight of it warned Alcimus he was no match for them; so he, too, went back to the king, loud in his complaints. \textcolor{benred8}{26}~Thereupon the king sent out an army for the people\textquotesingle s undoing, with Nicanor at the head of it, that was one of his most notable princes, and had a grudge against Israel to satisfy. \textcolor{benred8}{27}~This Nicanor, reaching Jerusalem with a great array, made peaceful overtures to Judas and his brethren, but treacherously; \textcolor{benred8}{28}~Need is none there should be blows given between us, he said. Let me come with a handful of men, and parley we together under safe conduct. \textcolor{benred8}{29}~Come he did, and the greeting between them was friendly enough, but Judas was like to have been seized, then and there, by the enemy; \textcolor{benred8}{30}~and when he had proof of Nicanor\textquotesingle s treachery, he went in dread of him and would parley with him no longer.
\textcolor{benred8}{31}~Nicanor, then, his plot being now manifestly discovered, would take to the field; it was close to Capharsalama that he engaged Judas; \textcolor{benred8}{32}~and his army, routed with a loss of five thousand men, must needs take refuge in the Keep of David.\footnote[1]{According to Josephus, who usually follows the sacred narrative closely, it was Judas who was defeated and forced to take refuge in Jerusalem; this would accord better with what follows, and it seems possible that the text has been incorrectly preserved.} \textcolor{benred8}{33}~It was after this that Nicanor made his way to mount Sion, where some of the priests and elders came out to greet him in friendly fashion, and shew him how burnt-sacrifice was offered there on the king\textquotesingle s behalf. \textcolor{benred8}{34}~But nothing could they get from him but mockery and contempt; he did despite to their sacred persons, and sent them away with threats. \textcolor{benred8}{35}~In his anger, he swore to them nothing would serve but he should have Judas and Judas\textquotesingle  army at his mercy; if not, he would burn the temple down, as soon as ever he returned in safety. So, in high disdain, he left them; \textcolor{benred8}{36}~and the priests must take themselves back within the walls, where they stood before altar and temple, praying very mournfully. \textcolor{benred8}{37}~Lord, they said, thou hast chosen this house to be the shrine of thy name; here thy people should offer prayer, and sue for thy favour. \textcolor{benred8}{38}~Do thou avenge thyself on chieftain and army both; die they at the sword\textquotesingle s point! Wouldst thou forget their blasphemy; should they escape with their lives?
\textcolor{benred8}{39}~After this, Nicanor left Jerusalem, and pitched his camp at Bethoron, where he was met by a fresh army from Syria; \textcolor{benred8}{40}~Judas, in his camp at Adarsa, had but three thousand men. And this was the prayer Judas prayed: \textcolor{benred8}{41}~Time was, Lord, when Sennacherib\textquotesingle s men were loud in their blasphemy, and thy angel must go out to smite them down, a hundred and eighty thousand of them. \textcolor{benred8}{42}~This day a new enemy overwhelm with our onslaught, and let all the world know what comes of threatening thy holy place; for his ill-doing, ill requite him!
\textcolor{benred8}{43}~It was the thirteenth of Adar when the two armies met; sure enough, Nicanor\textquotesingle s army was overwhelmed, and himself the first to fall in the encounter; \textcolor{benred8}{44}~whereupon the rest, seeing their leader gone, cast weapons away and took to their heels. \textcolor{benred8}{45}~For a whole day the pursuit of them went on, all the way from Adazer to the approaches of Gazara, and ever there were trumpets sounding the hue and cry. \textcolor{benred8}{46}~Out came Jewish folk from all the villages round about, to head them off,\footnote[2]{\textasciigrave To head them off\textquotesingle ; the Latin has \textasciigrave and tossed them with the horn\textquotesingle , probably through a misunderstanding of the word used in the Greek text.} till at last they turned at bay and fell at the sword\textquotesingle s point all of them, never a man left. \textcolor{benred8}{47}~Spoil of them was plundered where they lay; as for Nicanor, the Jews cut off the head from his body, and that right hand he lifted up so defiantly, and took them away, to be hung up in full sight of Jerusalem. \textcolor{benred8}{48}~Glad men they were that day, and kept high festival, \textcolor{benred8}{49}~decreeing that never thenceforward should the thirteenth day of Adar go unobserved. \textcolor{benred8}{50}~And for a little while the land of Juda had peace.
\begin{large}\begin{center}\textsc{Chapter viii}\end{center}\end{large}
\lettrine[lines=2]{J}{udas} had heard tell of the Romans, and their renown. Here was a powerful nation, that would entertain overtures none the less from such as craved their friendship, plighting their word faithfully. A powerful nation indeed; \textcolor{benred8}{2}~what battles they had fought, what exploits achieved yonder among the Galatians, their conquered vassals now! \textcolor{benred8}{3}~In Spain, too, they had done great feats of arms; and at last, by policy and patient striving, won over the whole country, made themselves masters of all the silver and gold that was mined there. \textcolor{benred8}{4}~Came peoples from far away, kings from the furthest corners of earth, to offer battle, they were overwhelmed and signally defeated; those nearer at hand were content to pay yearly tribute. \textcolor{benred8}{5}~Had they not crushed and conquered Philip, and Perseus king of the Greeks, and all others that had levied war upon them? \textcolor{benred8}{6}~And what of Antiochus the Great, that ruled all Asia, and came against them with a hundred and twenty elephants, with horsemen and chariots, and a great array besides? The Romans overcame him, \textcolor{benred8}{7}~caught him alive, and demanded both from him and from his heirs rich tribute, and hostages, with other conditions of surrender; \textcolor{benred8}{8}~took away from him India, Media, and Lydia, that were his most cherished provinces, and gave them to king Eumenes instead. \textcolor{benred8}{9}~Later, word came that the men of Hellas were for marching in and making an end of them; what was the issue of it? \textcolor{benred8}{10}~One of the Roman generals was sent out to engage them; fell many in battle, wives and children were carried off into exile, goods plundered, the land conquered, its fortresses destroyed, and they are slaves to this day.
\textcolor{benred8}{11}~So it was with all the kingdoms and islands that defied their will; the Romans crushed them and took their lands away. \textcolor{benred8}{12}~But to their friends, that would live at peace with them, they were ever good friends in return. Kingdoms both far and near became their vassals, nor any that heard their name but feared it; \textcolor{benred8}{13}~helped they any man to a throne, the throne was his; their good will lost, his throne was lost too; so high was their renown.
\textcolor{benred8}{14}~Yet, with all this, was never one of them that wore crown, or went clad in purple for his own aggrandizement. \textcolor{benred8}{15}~A senate-house they would have, where a council of three hundred and twenty met day by day, providing ever for the good estate of the commonalty; \textcolor{benred8}{16}~and every year they would entrust one man with the rule and governance of their whole country, the rest obeying him, without any debate or contention moved.\footnote[1]{These verses record only the impression which had reached Judas; it is not necessarily accurate in all points.}
\textcolor{benred8}{17}~So now Judas made choice of two envoys, Eupolemus, son of John, son of Jacob, and Jason, son of Eleazar; to Rome they should go, and there make a treaty of good will and alliance. \textcolor{benred8}{18}~Rome\textquotesingle s task it should be to rid them of the Grecian yoke; from the Greeks it was plain they could expect nothing better than grinding slavery. \textcolor{benred8}{19}~So, after long journeying, to Rome they came, and were admitted to the senate house, where they gave their message as follows: \textcolor{benred8}{20}~We have been sent to you by Judas Machabaeus and his brethren, and by our countrymen at large, to make a treaty of alliance with you; fain would they be enrolled among your confederates and friends. \textcolor{benred8}{21}~This proposition liked the Romans well; \textcolor{benred8}{22}~and they wrote back to the Jews on tablets of bronze, that should be kept in Jerusalem to serve them for a memorial of treaty and alliance made, to this effect: \textcolor{benred8}{23}~Well speed they at all times, the Roman and the Jewish peoples, by sea and land alike; far removed from either be alarm of war, assault of the enemy! \textcolor{benred8}{24}~Yet if war befall, and threaten the Romans first, or any ally of theirs in any part of their dominions, \textcolor{benred8}{25}~such aid the Jewish people shall give as the occasion demands, ungrudgingly. \textcolor{benred8}{26}~For the needs of the enemy they shall nothing find or furnish, be it corn, or arms, or money, or ships, according to the agreement made at Rome; and they shall observe these undertakings with no thought of their own advantage. \textcolor{benred8}{27}~In like manner, if the Jews be first threatened, it shall be for the Romans to give aid as the occasion demands, most willingly; \textcolor{benred8}{28}~providing neither corn nor arms, money nor ships, to any that take part against them, according to the agreement made at Rome;\footnote[2]{The Latin has, \textasciigrave according to the agreement made by the Romans\textquotesingle .} and they shall observe these undertakings honourably. \textcolor{benred8}{29}~Upon these terms the Romans and the Jewish people are agreed; \textcolor{benred8}{30}~if hereafter it should be the will of both parties\footnote[3]{\textasciigrave Both parties\textquotesingle ; according to the Latin version, \textasciigrave either party\textquotesingle , but such a stipulation would make the whole treaty ineffectual.} to enlarge or to restrict them, they may do so at their discretion, and such enlargement or restriction shall have force accordingly. \textcolor{benred8}{31}~As for the wrong done by king Demetrius, we have sent him warning, What meanest thou, to burden with so heavy a yoke the Jewish people, our friends and allies? \textcolor{benred8}{32}~Let them complain of thee once more, and we will surely give them redress, by land and sea levying war against thee.
\begin{large}\begin{center}\textsc{Chapter ix}\end{center}\end{large}
\lettrine[lines=2]{W}{hile} this was afoot, news came to Demetrius that Nicanor and his men had perished in the encounter. But he would still have his way; Bacchides and Alcimus should be sent back to Judaea, and the northern command\footnote[1]{\textasciigrave The northern command\textquotesingle ; literally, \textasciigrave the right wing\textquotesingle . Presumably this means the right extremity from Demetrius\textquotesingle  point of view.} of his army with them. \textcolor{benred8}{2}~Marching out along the Galgala road, they encamped at Masaloth in Arbella; the town was surprised, and many of its inhabitants massacred. \textcolor{benred8}{3}~Then, in the first month of the hundred and fifty-second year, they began an attack on Jerusalem, \textcolor{benred8}{4}~moving their camp to Berea. It was a force of twenty thousand foot and two thousand horse; \textcolor{benred8}{5}~Judas, encamped at Laisa, had three thousand picked followers with him, \textcolor{benred8}{6}~but these were greatly daunted when they saw what heavy odds were against them, and began to desert their lines, till no more than eight hundred of them were left. \textcolor{benred8}{7}~One by one they slipped away, and raise fresh levies he might not, with the battle so hard upon his heels; what wonder if Judas lost heart, and was unmanned? \textcolor{benred8}{8}~Yet said he to the remnant that was left him, Up, go we to the attack, and try conclusions with the enemy! \textcolor{benred8}{9}~In vain they sought to dissuade him; Speed we may not, they said; let us save our skins now, we may yet join hands with our brethren, and do battle hereafter; why, we are but a handful! \textcolor{benred8}{10}~Nay, said Judas, that may I never do; what, shew them our backs? If our time is come, die we manfully in our brethren\textquotesingle s cause, nor suffer any foul blot to fall on our name!
\textcolor{benred8}{11}~By this, the opposing army had moved forward out of its lines, and stood fronting them; here were the two bodies of horse, the slingers and archers going on before the rest, and the choice troops that would bear the shock of the encounter; \textcolor{benred8}{12}~here was Bacchides himself, on the right wing. This side and that the phalanx drew nearer, with a great blowing of trumpets, \textcolor{benred8}{13}~and Judas\textquotesingle  men, they raised a great cry on their own part, till the earth rang again with the noise of the two armies. Thus begun, the battle went on from morning till dusk. \textcolor{benred8}{14}~On the right, where he saw Bacchides\textquotesingle  army was strongest, Judas made the attack, and all the most valiant of his men with him; \textcolor{benred8}{15}~broke their line, and chased them all the way to mount Azotus. \textcolor{benred8}{16}~But now those on the left, seeing their right wing routed, cut off Judas and his men from the rear; \textcolor{benred8}{17}~now indeed the battle grew fierce, and there were many fell wounded on either part, \textcolor{benred8}{18}~till at last Judas fell, and with that, all the rest took to their heels.
\textcolor{benred8}{19}~As for his body, his brothers Jonathan and Simon recovered it, and so buried him where his fathers were buried, in the city of Modin. \textcolor{benred8}{20}~Great lament all Israel made over him, and long they mourned him; \textcolor{benred8}{21}~Here is a great warrior fallen, they said, that once brought his people deliverance! \textcolor{benred8}{22}~What other battles Judas fought, deeds did, greatness achieved, you shall not find set down here; too long the record of them.
\textcolor{benred8}{23}~Once Judas was dead, there was no corner in Israel but treason began to shew its face there, and lawlessness to abound; \textcolor{benred8}{24}~under such leadership the whole country, at this time much distressed by famine, went over to Bacchides. \textcolor{benred8}{25}~Good care he took to choose out godless men, that should have the governance of his territory; \textcolor{benred8}{26}~and these raised a hue and cry after Judas\textquotesingle  partisans, haling them before Bacchides to be punished and used despitefully; \textcolor{benred8}{27}~never, since prophecy died out among them, had the men of Israel known such distress. \textcolor{benred8}{28}~And now all that had loved Judas rallied to Jonathan instead; \textcolor{benred8}{29}~Since thy brother\textquotesingle s death, they told him, none is left to take the field against our enemies as he did, this Bacchides and all else that bear a grudge against our race.\footnote[2]{The meaning of the Greek text is, \textasciigrave all those of our own race that bear a grudge against us\textquotesingle .} \textcolor{benred8}{30}~There is but one way of it; this day we have chosen thee to be our ruler, our chieftain, to fight our battles for us. \textcolor{benred8}{31}~So, from that day forward, Jonathan took command, in succession to his brother Judas. \textcolor{benred8}{32}~Bacchides no sooner heard of it than he marked him down for death; \textcolor{benred8}{33}~but of this Jonathan had warning, and took refuge, with his brother Simon and all his company, in the desert of Thecue. It was there, by Asphar pool, they halted; \textcolor{benred8}{34}~and it was there that Bacchides, well informed of their movements, crossed Jordan at the head of his army and came upon them, one sabbath day.
\textcolor{benred8}{35}~This was the manner of it. Jonathan had sent his brother John, that was in command of the camp followers, on an errand to his good friends the Nabuthaeans. They had brought a deal of their household stuff with them; would the Nabuthaeans take it into safe keeping? \textcolor{benred8}{36}~But, as they went, some of Jambri\textquotesingle s folk came out from Madaba, seized John and all that he had with him, and went off with them. \textcolor{benred8}{37}~Afterwards, Jonathan and Simon heard that the men of Jambri had a great wedding toward; they must bring home the bride from Nadabatha,\footnote[3]{\textasciigrave Nadabatha\textquotesingle ; the Latin version has \textasciigrave Madaba\textquotesingle , presumably a copyist\textquotesingle s error.} and with much pomp, because her father was a notable Chanaanite chief. \textcolor{benred8}{38}~So, to avenge the death of their brother John, they climbed the hill-side and lay in ambush there. \textcolor{benred8}{39}~What a sight was this met their eyes! All manner of rout and display; the bridegroom, his friends and his brethren, passing on their way to the trysting-place, with beating of drums, and making of music, and all manner of warlike array! \textcolor{benred8}{40}~Then rose they up from their ambush and laid about them, till many fell wounded, and the rest fled into the hills, leaving all their spoil behind them. \textcolor{benred8}{41}~So turned they wedding mirth into funeral dirge, \textcolor{benred8}{42}~to avenge the murder of their brother, and withdrew to the banks of Jordan again.\footnote[4]{vv. 35-42. This incident is given by Josephus as if it followed the events described in verses 43-49; but he has probably misunderstood the sequence of the narrative. In verse 35, the Latin version seems to suggest that Jonathan asked the Nabuthaeans for the loan of their equipment, but the account given in the Greek text is more natural.}
\textcolor{benred8}{43}~Hearing of these alarms, Bacchides marched down to Jordan bank one sabbath day, in great force. \textcolor{benred8}{44}~Up now! cried Jonathan to his men; engage our enemy we must. Gone is the vantage we had till now; \textcolor{benred8}{45}~here is armed force confronting us, and all around us is Jordan stream, Jordan banks full of marshes and thickets; escape is none. \textcolor{benred8}{46}~Cry we rather upon heaven, for deliverance out of the enemy\textquotesingle s hand. So the battle was joined; \textcolor{benred8}{47}~and here was Jonathan exerting all his strength to deal a blow at Bacchides, who declined the encounter! \textcolor{benred8}{48}~What did Jonathan then? With all his company, he leapt into Jordan. So now, to reach them, the enemy must swim for it across the stream.\footnote[5]{Literally, \textasciigrave and they swam across the Jordan to them\textquotesingle . The Greek text has \textasciigrave and they did not swim across the Jordan to them\textquotesingle . The meaning, in either case, can only be a matter of conjecture.} \textcolor{benred8}{49}~A thousand men of his following Bacchides lost that day, and was fain to return to Jerusalem.
\textcolor{benred8}{50}~After this, they took to fortifying the cities of Judaea with high walls and barred gates, making strongholds at Jericho, Ammaum, Bethoron, Bethel, Thamnata, Phara and Thopo; \textcolor{benred8}{51}~here garrisons were set, for the harrying of Israel. \textcolor{benred8}{52}~Bethsura, too, Bacchides fortified, and Gazara, and the Citadel itself, keeping all of them well manned and provisioned; \textcolor{benred8}{53}~ay, and the great men of all the country round must yield up their children as hostages, to be held in Jerusalem citadel for safe keeping. \textcolor{benred8}{54}~Then, in the second month of the hundred and fifty-third year, came an order from Alcimus, the dividing wall of the temple\textquotesingle s inner court should be dismantled. The Prophets\textquotesingle  Building\footnote[6]{The Prophet\textquotesingle s Building (literally, Work) was presumably the name given to some part of Zorobabel\textquotesingle s temple; no allusion is made to it elsewhere.} he had already cleared away, and begun the dismantling, \textcolor{benred8}{55}~when himself was smitten down, and all his plans interrupted. Dumbstricken and palsied, he never spoke again, even to dispose of his goods, \textcolor{benred8}{56}~but died there and then, in great torment.
\textcolor{benred8}{57}~Alcimus dead, Bacchides was for Judaea no longer; away he went to the king\textquotesingle s court, and for two years the land was at peace. \textcolor{benred8}{58}~But ere long there was conspiracy afoot among the godless party; here were Jonathan and his men living secure of their safety; let Bacchides come in again, he might seize them all, and make one night\textquotesingle s work of it. \textcolor{benred8}{59}~To Bacchides, then, they went, and imparted their scheme to him; \textcolor{benred8}{60}~whereupon he raised a great army for marching on Judaea, but first sent word privately to his partisans there, bidding them seize Jonathan and his company for themselves. Word went abroad, and the plan miscarried; \textcolor{benred8}{61}~it was Jonathan\footnote[7]{The name Jonathan is not given, but it seems the natural one to supply in the context. Josephus, perhaps through a misinterpretation, attributes the massacre to Bacchides.} seized fifty notables of Judaea, that were the authors of the conspiracy, and put them to death. \textcolor{benred8}{62}~Then, with Simon and the rest of his following, he removed to Bethbessen, out in the desert, and set about rebuilding it, to make a stronghold for them.
\textcolor{benred8}{63}~Bacchides had news of this; mustering his whole force, and sending word to his Jewish supporters, \textcolor{benred8}{64}~he marched in and pitched his camp so as to command Bethbessen. Long time he besieged it, and brought up engines against it; \textcolor{benred8}{65}~meanwhile, Jonathan had left his brother Simon in command of the city, and was roaming the country-side. When he came back, it was with a band of men at his heels; \textcolor{benred8}{66}~smote he Odares and his clan, smote he the men of Phaseron where they lay encamped; everywhere laid about him, and still gained strength.\footnote[8]{vv. 65, 66. The original here is strangely worded, and perhaps corrupt.} \textcolor{benred8}{67}~As for Simon and his company, they made a sally out of the town, and set fire to the engines; \textcolor{benred8}{68}~afterwards they engaged Bacchides himself, and worsted him, so that he must pay dearly for plot and tryst of his that came to nothing. \textcolor{benred8}{69}~So enraged was he with the malcontents whose counsel had brought him into Judaea, he put many of them to death, and was for marching home again with the rest of his following, \textcolor{benred8}{70}~when Jonathan, hearing of it, sent envoys to offer peace, and an exchange of prisoners. \textcolor{benred8}{71}~This offer he gladly accepted, and carried out the terms of it, giving his word he would do Jonathan no more injury as long as he lived, \textcolor{benred8}{72}~and restoring all the prisoners he had ever taken in the land of Juda. So he took himself back to his own country, and never came that way again. \textcolor{benred8}{73}~Israel had a respite from fighting at last, and Jonathan took up his dwelling at Machmas, whence he ruled the people thenceforward, ridding the land of godless folk altogether.
\begin{large}\begin{center}\textsc{Chapter x}\end{center}\end{large}
\lettrine[lines=2]{A}{nd} now, in the hundred and sixtieth year, came Alexander, a son of Antiochus the Illustrious, and took possession of Ptolemais, where he was received with royal honours. \textcolor{benred8}{2}~A great force king Demetrius levied, when he heard of it, and went out to give him battle; \textcolor{benred8}{3}~at the same time, he wrote to Jonathan, in such loving terms as should flatter his dignity. \textcolor{benred8}{4}~No time to be lost, thought he, in making friends with this man, before he takes to comforting Alexander against us; \textcolor{benred8}{5}~for wrong done to himself, and his brother, and all his race, he bears us a grudge yet. \textcolor{benred8}{6}~So he empowered Jonathan to muster an army, and to make weapons of war, as the ally of Syria; the hostages, too, in the citadel were to be given back to him.
\textcolor{benred8}{7}~When Jonathan came to Jerusalem, and read this letter aloud, not to the townsfolk only, but to the citadel garrison, \textcolor{benred8}{8}~great was the fear fell on all who listened; here was Jonathan commissioned to levy troops by the king\textquotesingle s own order! \textcolor{benred8}{9}~The hostages were surrendered without more ado, and given back to their parents; \textcolor{benred8}{10}~and he himself took up his quarters in Jerusalem, where he set about building up the city and repairing it. \textcolor{benred8}{11}~It was the walls needed rebuilding, so he told his workmen; on every side, the hill of Sion must be defended with hewn stone; and punctually they obeyed him. \textcolor{benred8}{12}~As for the alien folk that guarded the strongholds Bacchides had left, they fled incontinently; \textcolor{benred8}{13}~what matter if their posts were abandoned? They were for home. \textcolor{benred8}{14}~Only Bethsura was garrisoned now, and that by traitors to God\textquotesingle s law and commandment; it was all the refuge they had.
\textcolor{benred8}{15}~King Alexander heard of these overtures made by Demetrius; heard, too, the story of Jonathan and his brethren, battles fought, and deeds done, and labours endured. \textcolor{benred8}{16}~Why, said he, this man has not his match anywhere; time it is we should court his friendship and alliance. \textcolor{benred8}{17}~With that, he wrote him a letter, and these were the terms of it: \textcolor{benred8}{18}~King Alexander, to Jonathan his brother-prince, greeting! \textcolor{benred8}{19}~We have heard tell of thee, a man so valiant, and so well worthy of our friendship; \textcolor{benred8}{20}~in token whereof, we appoint thee high priest of thy own race henceforward, and to have the title of the King\textquotesingle s Friend. With that, he sent him a purple robe and a gold crown; Take ever our part, said he, and hold fast the bond of friendship. \textcolor{benred8}{21}~So, when the seventh month came round, in the hundred and sixtieth year, Jonathan clad himself with the sacred vesture at the feast of Tent-dwelling; an army he levied besides, and made weapons of war in great abundance.
\textcolor{benred8}{22}~Sick and sorry Demetrius was when he heard of these doings; \textcolor{benred8}{23}~Here is an ill day\textquotesingle s work, said he, to let Alexander forestall us in making alliance with the Jews, to his great comfort! \textcolor{benred8}{24}~From me, too, they shall have a message of entreaty, they shall have honours and gifts; the Jews shall be my good friends yet. \textcolor{benred8}{25}~And thus he wrote: King Demetrius, to the people of the Jews, greeting! \textcolor{benred8}{26}~Here is welcome news we have of you; right well you have kept troth with us, honouring the treaty when you might have taken part with our enemies. \textcolor{benred8}{27}~In that loyal mind continue, and your good offices shall not go unrewarded; \textcolor{benred8}{28}~much immunity you shall enjoy, much largesse receive.
\textcolor{benred8}{29}~By these presents, I exempt both you and all Jews from the poll-tax; salt-tax and coronation dues I remit and forgo, with my right to a third part of your seed-corn, \textcolor{benred8}{30}~and half your fruit-crop. From this day forward, now and for ever, I resign all this; from Juda and from the three cantons of Samaria and Galilee\footnote[1]{\textasciigrave Samaria and Galilee\textquotesingle , here evidently treated as a single unit; the three districts concerned had actually been Samaritan (verse 38).} lately added to it, there shall be no toll taken. \textcolor{benred8}{31}~For Jerusalem, it shall be a place set apart, a free city with its own confines, mistress of its own tithe and tribute; \textcolor{benred8}{32}~nor claim I any rights over the citadel there, I make it over to the high priest, to garrison it as he will. \textcolor{benred8}{33}~All persons of Jewish blood in all my realm that were taken away as prisoners from Juda shall now be set free gratuitously, and no distraint made on their revenues or cattle. \textcolor{benred8}{34}~Feast-day and new moon and sabbath, and all other such solemnities as are appointed to be observed, with the three days before and after the feast itself, shall be days of immunity and respite for all the Jews in my realm; \textcolor{benred8}{35}~nor any business done or debate moved to their detriment at such times. \textcolor{benred8}{36}~In the king\textquotesingle s army, Jews may be enrolled up to the number of thirty thousand, paid according to the common rate of the royal troops; and the same shall be free to serve in all the fortified towns of our empire. \textcolor{benred8}{37}~Jews may be employed besides in all positions of trust, and appointed governors,\footnote[2]{\textasciigrave And appointed governors\textquotesingle ; according to the Greek text, \textasciigrave and let their governors be men of their own race\textquotesingle .} yet live still by their own laws, that have royal sanction in the land of Juda. \textcolor{benred8}{38}~The three cantons taken from Samaria and added to Judaea shall be accounted part of Juda, under a single government, with no allegiance but to the high priest.
\textcolor{benred8}{39}~Ptolemais, with all the country that lies about it, I hereby convey as a free gift to the temple precincts at Jerusalem, to defray the temple expenses.\footnote[3]{A fine touch; cf. verse 1.} \textcolor{benred8}{40}~To this gift I add a sum of fifteen thousand silver sicles yearly, out of the royal dues that belong to me. \textcolor{benred8}{41}~With this sum, arrears shall be made good in payments for the temple building, withheld till now by such as had charge of the matter; \textcolor{benred8}{42}~and restitution made, to the priests now in office, for the five thousand sicles that were confiscated year by year from the temple treasury.\footnote[4]{vv. 41, 42. The Latin here differs from the Greek text, which is less intelligible.} \textcolor{benred8}{43}~Debtor to the king, whatever be the charge against him, that takes sanctuary in the temple or its precincts, shall be left at liberty, and no distraint made upon goods of his within these dominions. \textcolor{benred8}{44}~Payment shall be made besides from the royal treasury for the finishing and repairing of the temple fabric; \textcolor{benred8}{45}~as also for building up and making strong the walls of Jerusalem, and restoring the fortresses of Judaea.
\textcolor{benred8}{46}~But in vain were such promises made to Jonathan and the Jewish folk, nor credence found they any nor assent. Could they forget all the mischief Demetrius had done in Israel, all the tyranny they had endured? \textcolor{benred8}{47}~Alexander it was had all their good wishes; his was the first offer of terms that reached them, and all the while it was his cause they cherished. \textcolor{benred8}{48}~By this, Alexander had mustered a great force, and marched against Demetrius. \textcolor{benred8}{49}~When the two kings met, it was Demetrius\textquotesingle  men took to their heels, and Alexander gave chase, pressing them hard; \textcolor{benred8}{50}~fiercely the battle raged till sun-down, and before the day was over, Demetrius fell.
\textcolor{benred8}{51}~Hereupon Alexander sent an embassy to Ptolemy, king of Egypt, addressing him in these terms following. \textcolor{benred8}{52}~Take notice I have returned to my kingdom, and sit now on the throne of my fathers, in full possession of my princely rights. Would I regain Syria, needs must I should overthrow Demetrius; \textcolor{benred8}{53}~overthrow him I did, on field of battle, with all his army, and here I sit in his place. \textcolor{benred8}{54}~And should we not be upon terms of friendship, thou and I? Let me have thy daughter to wife; a niggardly wooer thou shalt not find me, nor she either. \textcolor{benred8}{55}~And what answer made king Ptolemy? An auspicious day, said he, this day of thy return to the land and throne of thy fathers! \textcolor{benred8}{56}~Boon thy letter asks of me thou shalt have; but first meet we together, face to face, yonder at Ptolemais; there will I pledge my word to the articles thou namest. \textcolor{benred8}{57}~So here was king Ptolemy come from Egypt, with his daughter Cleopatra, all the way to Ptolemais, in the hundred and sixty-second year; \textcolor{benred8}{58}~and there king Alexander met him and took his daughter Cleopatra to wife, and they held the wedding with great magnificence, as kings will.
\textcolor{benred8}{59}~King Alexander had sent word to Jonathan, he should come and keep tryst with him; \textcolor{benred8}{60}~so to Ptolemais Jonathan went with great state, and met the two kings there. Gifts a many he made them, of silver and gold and much else, and was high in favour with them. \textcolor{benred8}{61}~It chanced that certain Israelites, pestilent fellows of the traitorous party, came there to bring charges against him. But to these the king would not listen; \textcolor{benred8}{62}~he would have Jonathan change his garments, and go clad in purple, and when this was done, a seat he must have beside the king himself. \textcolor{benred8}{63}~Take him out into the heart of the city, Alexander said to his vassals, and there make proclamation, none may bring charge against him on any pretext, or in any fashion molest him. \textcolor{benred8}{64}~No thought had his accusers, when they heard such proclamation made, and saw Jonathan there dressed in purple, but to escape, one and all, as best they could; \textcolor{benred8}{65}~he himself was loaded with honours, enrolled among the king\textquotesingle s chief friends, and made a prince, with a share in the governance of the kingdom. \textcolor{benred8}{66}~So Jonathan made his way back to Jerusalem undisturbed, and well content.
\textcolor{benred8}{67}~Then, in the hundred and sixty-fifth year, came Demetrius, son of that other Demetrius, from the island of Crete, and landed in his native country; \textcolor{benred8}{68}~ill hearing indeed for Alexander, who returned at once to Antioch. \textcolor{benred8}{69}~Demetrius\footnote[5]{It seems possible that the word \textasciigrave Demetrius\textquotesingle  may have been inserted for the sake of clearness; Josephus treats Apollonius throughout as fighting on the side of Alexander (in spite of verse 88).} gave command of his army to Apollonius, that was governor of Coelesyria, and a great array it was he levied. From Jamnia, where he took up his quarters, this Apollonius sent word to the high priest Jonathan: \textcolor{benred8}{70}~What, wilt thou defy us, and all alone? Here am I mocked and flouted by the resistance offered me, up yonder in the hills! \textcolor{benred8}{71}~Nay, if such confidence thou hast in thy own resources, come down and meet us in the plain; try we conclusions there! Trust me, I am master of the field; \textcolor{benred8}{72}~what I am, what my troops are, thou shalt learn upon a little enquiry; stand thou canst not, they will tell thee, before onslaught of ours. Twice, on their native soil, thy fathers fled in disorder, \textcolor{benred8}{73}~and wilt thou make head against such an array of horse and foot, here in the plain, where rock is none, nor gravel-bed, to aid thy flight?
\textcolor{benred8}{74}~Roused by this challenge, Jonathan marched out from Jerusalem with a muster of ten thousand men; his brother Simon joined hands with him; \textcolor{benred8}{75}~and together they appeared before the gates of Joppe. Enter they might not, for Apollonius had a garrison there, but must needs attack it; \textcolor{benred8}{76}~whereupon the citizens took alarm, and themselves opened the gates. Thus came Joppe into the power of Jonathan; \textcolor{benred8}{77}~the news reached Apollonius, and he brought up three thousand horse, with a great array of men besides. \textcolor{benred8}{78}~To Azotus he marched, as if he meant to pass them by, but all the while he was luring them on into the plain;\footnote[6]{\textasciigrave All the time he was luring them on into the plain\textquotesingle ; this seems to be the meaning of the Greek text, although Joppe and Azotus were both on the sea-board, far away from any hill-country. The Latin has \textasciigrave immediately he went out into the plain\textquotesingle , which yields no satisfactory sense.} in horse lay his strength and his confidence. To Azotus Jonathan followed him, and battle was joined.
\textcolor{benred8}{79}~Apollonius, by a secret feint, had left a thousand horsemen encamped in their rear; \textcolor{benred8}{80}~so all at once Jonathan found himself cut off by an ambush. Round his army they rode, casting javelins into the ranks, from morning till night-fall; \textcolor{benred8}{81}~but ever it stood firm, at Jonathan\textquotesingle s bidding, till the horses were tired out at last. \textcolor{benred8}{82}~Then, the force of the cavalry once spent, out came Simon with his troops to attack the main body, which thereupon broke and fled. \textcolor{benred8}{83}~Scattered over the open country, in vain they rallied at Azotus, and took refuge in the precincts of their god Dagon; \textcolor{benred8}{84}~both Azotus and all the neighbouring cities Jonathan burnt and plundered, and Dagon\textquotesingle s temple, with all that took shelter there, was burnt with the rest. \textcolor{benred8}{85}~So perished, by sword and fire, some eight thousand men; \textcolor{benred8}{86}~as for Jonathan, he had no sooner encamped before Ascalon, than the townsfolk opened the gates to him, and gave him honourable welcome.
\textcolor{benred8}{87}~So Jonathan came back to Jerusalem, and the army behind him, laden with spoils. \textcolor{benred8}{88}~More than ever, when he heard of it, did king Alexander heap honours upon him; \textcolor{benred8}{89}~a buckle of gold he sent him, ever the gift kings make to men of blood royal, and Accaron, with all the countryside about it, granted him for his domain.
\begin{large}\begin{center}\textsc{Chapter xi}\end{center}\end{large}
\lettrine[lines=2]{A}{nd} now Ptolemy, king of Egypt, levied a great army, countless as sand on the beach, and a fleet besides; to win Alexander\textquotesingle s realm his treacherous design was, and add it to his own. \textcolor{benred8}{2}~To Syria he came, full of fair speeches, and all the towns opened their gates to welcome him; such welcome Alexander himself had prescribed; was not the king of Egypt his father-in-law? \textcolor{benred8}{3}~And never a town king Ptolemy entered, but he left a guard of soldiers there. \textcolor{benred8}{4}~When he reached Azotus, here was Dagon\textquotesingle s temple burnt, here was the town itself and all its neighbourhood in ruins; the dead lay unburied, where they fell in battle, or in heaps by the road-side. All this they shewed him, \textcolor{benred8}{5}~and told him, with malicious intent, how it was Jonathan\textquotesingle s doing; but no word said king Ptolemy. \textcolor{benred8}{6}~As for Jonathan, he went to meet the king at Joppe, with a deal of pomp; there they greeted one another, and passed the night, \textcolor{benred8}{7}~nor would Jonathan return to Jerusalem till he had escorted the king as far as the river called Eleutherus.
\textcolor{benred8}{8}~All the cities of the sea-coast, as far as maritime Seleucia, king Ptolemy occupied, and with no friendly purpose towards Alexander; \textcolor{benred8}{9}~it was to Demetrius he sent envoys instead. Come, said he, a pact between us! My daughter thou shalt have in Alexander\textquotesingle s place, and therewithal the throne of thy fathers; \textcolor{benred8}{10}~here is an ill son-in-law I have chosen, that went about but now to kill me! \textcolor{benred8}{11}~Thus, to find pretext for dethroning his rival, king Ptolemy defamed him; \textcolor{benred8}{12}~took his daughter away, and gave her to Demetrius. His estrangement from Alexander now come to an open breach, \textcolor{benred8}{13}~what must he do next but enter the city of Antioch, and there assume the double crown, as ruler of Egypt and Asia both? \textcolor{benred8}{14}~As for Alexander, that was then in Cilicia, quelling a revolt in those parts, \textcolor{benred8}{15}~he came out to do battle when the news reached him; but Ptolemy brought up his army, met him with a superior force, and routed him. \textcolor{benred8}{16}~Thus Egypt had the mastery; and when Alexander fled to Arabia for refuge, \textcolor{benred8}{17}~Zabdiel, an Arabian, cut off his head and sent it to the conqueror. \textcolor{benred8}{18}~Three days later, Ptolemy himself lay dead; whereupon the garrisons he had left in the towns were massacred by the citizens, \textcolor{benred8}{19}~and the royal power passed to Demetrius in this, the hundred and sixty-seventh year.
\textcolor{benred8}{20}~Now it was that Jonathan mustered the men of Judaea to deliver an attack on the Gentile citadel in Jerusalem; engines a many they brought against it. \textcolor{benred8}{21}~Nor wanted there Jews of the godless party, traitors to their own race, that went off and told Demetrius it was being attacked; \textcolor{benred8}{22}~the news greatly angered him, and he hastened to Ptolemais, bidding Jonathan raise the siege and come to meet him in conference without more ado. \textcolor{benred8}{23}~This message notwithstanding, Jonathan would have the siege go forward; certain elders of Israel, and certain of the priests, he chose out to bear him company, and so put his own life in peril, \textcolor{benred8}{24}~going off to meet the king at Ptolemais, with gold and silver and garments and other gifts in great number. He was received graciously enough; \textcolor{benred8}{25}~let his own traitorous fellow-countrymen bring what accusations they would, \textcolor{benred8}{26}~the king would not be behind his predecessors in making much of Jonathan, for all his courtiers to see. \textcolor{benred8}{27}~He was confirmed in the high priesthood, and what other high dignities he held aforetime, and declared besides the chief of the king\textquotesingle s friends.
\textcolor{benred8}{28}~And now Jonathan had a favour to ask; exemption from tribute for Judaea, and the three cantons, and Samaria with its neighbouring townships;\footnote[1]{\textasciigrave The three cantons, and Samaria with its neighbouring townships\textquotesingle ; some think this is a copyist\textquotesingle s error for \textasciigrave the three cantons which had belonged to Samaria, with their neighbouring townships\textquotesingle , cf. verse 34.} he promised in return a payment of three hundred talents. \textcolor{benred8}{29}~To this the king agreed, writing thus to Jonathan upon the matter raised: \textcolor{benred8}{30}~King Demetrius, to his brother prince Jonathan, and to the people of the Jews, greeting. \textcolor{benred8}{31}~We send you herewith, for your better information, a copy of the instructions we have given to our cousin Lasthenes in your regard. \textcolor{benred8}{32}~King Demetrius, to Lasthenes, his good father, greeting. \textcolor{benred8}{33}~Whereas the people of the Jews have ever been trusty friends to us, our pleasure it is to reward them for the loyalty they have shewn us. \textcolor{benred8}{34}~We therefore confirm them in the possession of all Judaea, the three cities of Ephraim, Lydda and Ramathan, that formerly belonged to Samaria, and all their neighbouring townships~\ldots\  to all those who do sacrifice at Jerusalem; instead of the yearly revenues hitherto set apart for the king from harvest and fruit-gathering.\footnote[2]{The sentence is obscure, and perhaps the text has been inaccurately transmitted. For \textasciigrave Ephraim\textquotesingle  the Greek has the form \textasciigrave Aphaerema\textquotesingle , which the Latin interprets as a common noun, \textasciigrave sequestration\textquotesingle .} \textcolor{benred8}{35}~Tithe and tribute that was ours we also remit to them; nor lay any claim to the salt-pits, or the crowns which from time to time were bestowed upon us. \textcolor{benred8}{36}~Of all this we give them a full discharge, that shall be valid in perpetuity. \textcolor{benred8}{37}~See to it that a copy of this decree shall be made, and handed over to Jonathan, who shall set it up in a public place on the holy mountain.
\textcolor{benred8}{38}~Here, then, was the whole realm at peace under Demetrius\textquotesingle  rule, nor any rival had he; what must he do but disband all his soldiers and send them home, except the foreign troops he had levied from the islands out at sea? Bitterly they hated him for it, the men who had served under his fathers; \textcolor{benred8}{39}~and there was one Tryphon that took good note of these discontents in the army. This Tryphon was formerly of Alexander\textquotesingle s faction, and now he had recourse to Emalchuel, the Arabian, that had care of Alexander\textquotesingle s son Antiochus. \textcolor{benred8}{40}~Much persuasion he used with him, to let Antiochus return to his father\textquotesingle s throne; much told him of Demetrius, and how the soldiers were disaffected against him.Time passed, and Tryphon was in Arabia still. \textcolor{benred8}{41}~Meanwhile, Jonathan was urgent with king Demetrius to withdraw the garrisons from Jerusalem citadel and the other strongholds, where they bore arms yet, and against Israel. \textcolor{benred8}{42}~Nay answered Demetrius, that I will do and more; great honours I have in store, for thee and for thy people both, when the time is ripe for it. \textcolor{benred8}{43}~For this present, it were well done to send troops for my own protection; here is all my army revolted from me! \textcolor{benred8}{44}~Three thousand picked men Jonathan dispatched to Antioch, to the king\textquotesingle s side, and right glad he was at their coming. \textcolor{benred8}{45}~What though the citizens, a hundred and twenty thousand strong, were banded together against his royal person, \textcolor{benred8}{46}~driving him to take refuge within the court, and occupying the city streets in warlike fashion? \textcolor{benred8}{47}~He had but to call the Jews to his aid, and they rallied at his summons; posted themselves here and there about the streets, \textcolor{benred8}{48}~and in one day slew a hundred thousand men, setting fire to the town besides. There was spoil enough for the winning, that day when they saved the king\textquotesingle s life. \textcolor{benred8}{49}~The townsfolk, when they saw how easily the Jews got the mastery of them, had no more stomach for fighting; they were loud in their entreaties: \textcolor{benred8}{50}~A truce! A truce! Havoc enough yonder Jews have made of us and of the city! \textcolor{benred8}{51}~And so, flinging away their weapons, they came to terms. Prince and people both had good proof, by now, of the Jews\textquotesingle  valour; back they went to Jerusalem high in repute among the Syrians, and laden with spoils.
\textcolor{benred8}{52}~Demetrius, now firmly established on the throne, his dominions all at peace, \textcolor{benred8}{53}~recked little enough of his promises; from Jonathan he was estranged altogether, left his services unrecompensed, and much mischief did him besides. \textcolor{benred8}{54}~It was now that Tryphon came back, and with him the young prince Antiochus, that took the style of king and had himself crowned; \textcolor{benred8}{55}~all the disbanded armies of Demetrius rallied to them, and turned upon their former master, who fled routed before them; \textcolor{benred8}{56}~Tryphon, meanwhile, got possession of the elephants, and Antioch fell into his hands. \textcolor{benred8}{57}~Thereupon came a letter from the young Antiochus to Jonathan, confirming him in the high priesthood, and in possession both of Judaea and of the three cantons; he was acclaimed as the king\textquotesingle s friend, \textcolor{benred8}{58}~and a present of golden cups sent for his use, with the right to drink out of gold ware, to dress in purple, and to carry the golden buckle. \textcolor{benred8}{59}~His brother Simon, too, was made lord of the sea-coast, from Tyre to the frontiers of Egypt.
\textcolor{benred8}{60}~And now Jonathan was on the march, across the river, patrolling the cities everywhere, with all the armies of Syria gathered to aid him.~\ldots\  He came to Ascalon, where the townsfolk welcomed him with all honour;\footnote[3]{It is hardly possible to give any satisfactory account of the text as it stands. Even if we take \textasciigrave beyond\textquotesingle  as referring to the Jewish side of the Euphrates, why should an expedition into Syria bring Jonathan down to Ascalon, some forty miles west of Jerusalem? Conceivably there has been some disturbance in the text, which may have read originally, \textasciigrave And he (Antiochus) began patrolling the cities across the river, with all the armies of Syria gathered to aid him; and Jonathan marched out and came to Ascalon\ldots\ \textquotesingle .} \textcolor{benred8}{61}~came to Gaza, where they shut the gates on him, and he must needs undertake the siege of it. But when he had spread fire and rapine through the country-side, \textcolor{benred8}{62}~the men of Gaza asked for terms, which he gave them, carrying off their sons as hostages to Jerusalem. Then he went on patrolling the country, all the way to Damascus. \textcolor{benred8}{63}~News reached him that the chiefs of Demetrius\textquotesingle  faction were making head at Cades, in Galilee, with a whole army to support them, and their design was to remove him from office. \textcolor{benred8}{64}~So he went to meet them, leaving his brother Simon in charge of Judaea.
\textcolor{benred8}{65}~As for Simon, he made an assault upon Bethsura, and kept it for a long while besieged, \textcolor{benred8}{66}~till at last it obtained terms of surrender; he rid the place of its defenders and took over the command of it, putting in a garrison of his own. \textcolor{benred8}{67}~Meanwhile, Jonathan was encamped by the waters of Genesar; here, on the plain of Asor, they were on the watch before day-break, \textcolor{benred8}{68}~when they saw the enemy\textquotesingle s force coming to meet them over the level plain. These had an ambush ready for him on the hill-side, and as he advanced to meet the main body, \textcolor{benred8}{69}~the men in ambush sprang up, and engaged him. \textcolor{benred8}{70}~At this, all Jonathan\textquotesingle s supporters took to their heels; none stood their ground but Mathathias son of Absalom and Judas son of Calphi, that had the marshalling of his men. \textcolor{benred8}{71}~What marvel if Jonathan tore his garments about him, and strewed earth on his head, and betook himself to prayer? \textcolor{benred8}{72}~Afterwards, he offered battle afresh, and routed his enemies; as the fight went on, \textcolor{benred8}{73}~his own men that had deserted their ranks rallied to him, and joined in the pursuit all the way to Cades, where they encamped once more. \textcolor{benred8}{74}~In that day\textquotesingle s fighting, three thousand of the Gentiles fell; and so Jonathan made his way back to Jerusalem.
\begin{large}\begin{center}\textsc{Chapter xii}\end{center}\end{large}
\lettrine[lines=2]{H}{ere} was a posture of affairs suited Jonathan well enough; yet would he send delegates to confirm and renew his alliance with the Romans; \textcolor{benred8}{2}~Lacedaemon, too, and other countries should have letters of the same tenour. \textcolor{benred8}{3}~To Rome, then, his messengers went, gained audience of the senate, and told how the high priest Jonathan and the Jewish people had sent them to renew their old treaty of friendship; \textcolor{benred8}{4}~and the Romans gave them such letters of recommendation to this country or that, as should bring them home to Juda under safe conduct.
\textcolor{benred8}{5}~The message Jonathan sent to the men of Sparta was in these terms following. \textcolor{benred8}{6}~The high priest Jonathan, with the elders and priests and all the people of the Jews, to their brethren the Spartans, greeting. \textcolor{benred8}{7}~Long since, your king Arius wrote to our own high priest, Onias, claiming kinship between us, as witness the copy here subjoined; \textcolor{benred8}{8}~an honourable welcome Onias gave to this messenger of yours, and accepted the proposal of friendly alliance. \textcolor{benred8}{9}~For ourselves, we have little need of such friendship; seek we comfort, it is in the sacred books committed to our charge. \textcolor{benred8}{10}~Yet we thought it best to treat with you for the renewal of this brotherly compact, before any estrangement should arise between us; your embassy to us is of long ago. \textcolor{benred8}{11}~Never feast-day passes, nor day apt for remembrance, but you are remembered, as brothers should be, in sacrifice and prayer we offer; \textcolor{benred8}{12}~renown of yours is pride of ours still. \textcolor{benred8}{13}~In wars and calamities much involved of late, powerful kings for our neighbours and our enemies, \textcolor{benred8}{14}~we would not embroil you, nor other allies of ours, in these quarrels. \textcolor{benred8}{15}~Now, by the grace of heaven, we are delivered; our enemies lie crushed; \textcolor{benred8}{16}~delegates of ours, Numenius son of Antiochus and Antipater son of Jason, are on their way to Rome, friendship and alliance of former days to confirm afresh; \textcolor{benred8}{17}~and should we send them with no errand to you, no greeting, no word from us of brotherhood revived? \textcolor{benred8}{18}~Pray you, send us fair answer in your turn.
\textcolor{benred8}{19}~And, for Arius\textquotesingle  letter to Onias, thus the copy of it ran, \textcolor{benred8}{20}~Arius, king of the Spartans, to the high priest Onias, greeting. \textcolor{benred8}{21}~Spartan and Jew, written record shews it, come of one blood, Abraham\textquotesingle s. \textcolor{benred8}{22}~Apprised of this, we would fain know how you do; pray tell us. \textcolor{benred8}{23}~And take this message in return, Cattle and whatever else is ours, is yours, and yours ours; of that, the bearer of this letter brings you assurance.
\textcolor{benred8}{24}~Then came news to Jonathan that the chiefs of Demetrius\textquotesingle  faction were returning to the attack, and in greater force than ever; \textcolor{benred8}{25}~so out he marched, and met them in the Amathite country; respite he would not give them, to invade his own. \textcolor{benred8}{26}~Spies of his went out into the enemy\textquotesingle s camp, and reported, all was ready for a night attack; \textcolor{benred8}{27}~so, when the sun was down, Jonathan would have his men keep watch, ready armed all night for battle, and posted sentries round his lines. \textcolor{benred8}{28}~The enemy, hearing of such preparedness on their part, took alarm and let cowardly counsels prevail;\footnote[1]{\textasciigrave Let cowardly counsels prevail\textquotesingle ; literally, \textasciigrave were dismayed in their hearts\textquotesingle , but the context shews that in fact they beat a retreat.} they were at pains to leave watchfires burning in their camp, \textcolor{benred8}{29}~so that Jonathan and his men, deceived by the glow of light, knew nothing of their plans till morning; \textcolor{benred8}{30}~and when he gave chase, it was too late to catch them; already they had crossed the river Eleutherus. \textcolor{benred8}{31}~Thereupon he turned his attack against the Zabadeans, an Arabian tribe, defeating them and taking spoils from them; \textcolor{benred8}{32}~and so, harnessing his waggons, pressed on to Damascus, patrolling all the country round about. \textcolor{benred8}{33}~Meanwhile, Simon had marched out to Ascalon and the neighbouring strongholds; thence he turned aside to Joppe, and took possession of it; \textcolor{benred8}{34}~rumour had reached him, the townsfolk would yield the citadel to Demetrius\textquotesingle  party, and he must have a garrison there of his own.
\textcolor{benred8}{35}~When Jonathan returned, he summoned the elders of the people, and took counsel with them, how best to raise strongholds in Judaea, \textcolor{benred8}{36}~and build up walls in Jerusalem itself. Height these must have, above all, between the Citadel and the rest of the town; he would have it cut off from the rest, standing by itself, with no opportunity to buy and sell. \textcolor{benred8}{37}~A great muster there was for the city\textquotesingle s rebuilding; and where the wall had tumbled down, over the ravine on the east, he made it good; it is the part called Caphetetha. \textcolor{benred8}{38}~Meanwhile, Simon rebuilt Adiada in the Sephela and fortified it; bolt and bar it should have thenceforward.
\textcolor{benred8}{39}~And what of Tryphon? Lordship of all Asia he coveted, and a royal crown; it should be Antiochus\textquotesingle  turn next. \textcolor{benred8}{40}~The danger was, Jonathan would refuse his assent, and resort to arms; Jonathan first he must seize and put to death. So he moved his quarters to Bethsan, \textcolor{benred8}{41}~where Jonathan came out to meet him with forty thousand men, picked warriors all of them, at his back. \textcolor{benred8}{42}~Here was a great retinue; and Tryphon, daunted by this show of force, \textcolor{benred8}{43}~was fain to give him an honourable welcome. He would admit Jonathan among his closest friends, and bestow gifts on him; let Jonathan give orders, and Tryphon\textquotesingle s soldiers would obey. \textcolor{benred8}{44}~Then he asked, What needs it, such a host of men should go campaigning, when threat of war is none? \textcolor{benred8}{45}~It were better to disband them, and choose out a few for thy own retinue. That done, bear me company to Ptolemais; city and strongholds and troops and officers I will hand over into thy charge and so get me gone home; it was on that errand I came.
\textcolor{benred8}{46}~What did Jonathan? He fell into the trap, sent his men back to Juda, \textcolor{benred8}{47}~and kept but three thousand under arms; of these, he left two thousand in Galilee, and took but a thousand in his company. \textcolor{benred8}{48}~No sooner had he entered Ptolemais than the townsfolk shut the gates behind him, secured his person, and put his retinue to the sword. \textcolor{benred8}{49}~Horse and foot Tryphon sent out to Galilee, to find the rest of his followers in the Great Plain, and make an end of them; \textcolor{benred8}{50}~but these, hearing that Jonathan and his men had been caught and murdered, resolved to put a bold front on it, and marched in battle array. \textcolor{benred8}{51}~Finding them ready to sell their lives dearly, their pursuers abandoned the chase, \textcolor{benred8}{52}~and all reached Juda safe and sound. For Jonathan and his companions they made great dole, and loudly all Israel echoed their lament. \textcolor{benred8}{53}~Neighbouring people was none but went about to overthrow them, and no wonder; \textcolor{benred8}{54}~their chieftain, their champion gone, now was the time to fall upon them, and rid earth of their memory.
\begin{large}\begin{center}\textsc{Chapter xiii}\end{center}\end{large}
\lettrine[lines=2]{A}{nd} what did Simon, when he heard that Tryphon had levied a strong force, for Juda\textquotesingle s invasion and overthrow? \textcolor{benred8}{2}~Here was all the people in a great taking of fear; so he made his way to Jerusalem and there gathered them to meet him. \textcolor{benred8}{3}~And thus, to put heart into them, he spoke: Need is none to tell you what battles we have fought, what dangers endured, I and my brethren and all my father\textquotesingle s kin, law and sanctuary to defend. \textcolor{benred8}{4}~In that cause, and for the love of Israel, my brothers have died, one and all, till I only am left; \textcolor{benred8}{5}~never be it said of me, in the hour of peril I held life dear, more precious than theirs! \textcolor{benred8}{6}~Nay, come the whole world against us, to glut its malice with our ruin, race and sanctuary, wives and children of ours shall find me their champion yet. \textcolor{benred8}{7}~At these words, the spirit of the whole people revived; \textcolor{benred8}{8}~loud came their answer, Brother of Judas and Jonathan, thine to lead us now! \textcolor{benred8}{9}~Thine to sustain our cause; and never word of thine shall go unheeded!
\textcolor{benred8}{10}~Thereupon, he summoned all the fighting men together, and pressed on to have the walls of Jerusalem finished, till it was fortified all about; \textcolor{benred8}{11}~and he sent Jonathan, son of Absalom, to Joppe, at the head of a force newly raised; the garrison was disbanded, and a new captain held it now. \textcolor{benred8}{12}~Meanwhile, Tryphon had left Ptolemais, with a great army at his heels, marching on Juda; and with him went Jonathan, his prisoner. \textcolor{benred8}{13}~He found Simon encamped at Addus, that looks out over the plain; \textcolor{benred8}{14}~here was Jonathan\textquotesingle s brother Simon taking his place, and offering battle. Envoys were sent out to make his excuses: \textcolor{benred8}{15}~Hold we the person of thy brother Jonathan, it is because he is in default to the royal treasury over his dealings with it. \textcolor{benred8}{16}~Thou hast but to send a hundred talents of silver, and his two sons, to be surety he will not play us false when we release him, and he is a free man. \textcolor{benred8}{17}~Well Simon knew it was treacherously spoken; yet he gave orders, both money and hostages should be surrendered. A bitter grudge Israel\textquotesingle s people would bear him, if they had cause to say, \textcolor{benred8}{18}~For want of money paid over and surety given, Jonathan must die! \textcolor{benred8}{19}~Sent they were, the boys and the money both, but all was treachery; Jonathan never came back.
\textcolor{benred8}{20}~And now Tryphon invaded Juda, bent on its undoing; his troops must fetch a compass by the road that leads round to Ador, and, march they where they would, Simon and his army were at their heels. \textcolor{benred8}{21}~Word came to Tryphon from the defenders of Jerusalem citadel, he should make his way across the desert without more ado, and bring them supplies; \textcolor{benred8}{22}~and that same night he had all his cavalry in readiness for the march, but there was a great fall of snow, and come he might not~\ldots\  into the country of Galaad.\footnote[1]{It seems clear from the context that some words have dropped out. The Greek text has, \textasciigrave and come he might not, because of the snow. So he moved camp and went into the country of Galaad\textquotesingle . But the passage may originally have indicated what reasons Tryphon had for abandoning his campaign in Judaea.} \textcolor{benred8}{23}~When he reached Bascaman, then and there he put Jonathan and his sons to death; \textcolor{benred8}{24}~and with that, he turned about, and went back to his own country.
\textcolor{benred8}{25}~There lay the bones of Simon\textquotesingle s brother Jonathan, till he sent to fetch them, and gave them burial at Modin, the city of his fathers. \textcolor{benred8}{26}~Loud lament all Israel made for him, and long they bemoaned him. \textcolor{benred8}{27}~Over the graves of his father and his brethren Simon raised a towering monument, of dressed stone behind and before; \textcolor{benred8}{28}~then, with father and mother and his four brethren in mind, he built seven pyramids, in rows; \textcolor{benred8}{29}~and all about were great columns, carved with armour and ships; an abiding memorial, and a land mark to mariners at sea. \textcolor{benred8}{30}~Such was the tomb he raised at Modin, and it may be seen to this day. \textcolor{benred8}{31}~Meanwhile, as they were journeying together, Tryphon murdered the young king Antiochus by artifice, \textcolor{benred8}{32}~and succeeded to his throne, wearing the crown of all Asia; great mischief it was he did to his country.
\textcolor{benred8}{33}~All the fortresses of Judaea Simon repaired, building them up with high tower and stout wall, with bolt and bar; and never a garrison but had provisions laid up in store. \textcolor{benred8}{34}~Then he chose out envoys and sent them to king Demetrius, praying that the land might enjoy immunity after the tyrannous actions of Tryphon.\footnote[2]{\textasciigrave After the tyrannous actions of Tryphon\textquotesingle ; literally (in the Greek text), \textasciigrave because all the actions of Tryphon were seizures\textquotesingle . The meaning is perhaps that Simon now recognized all the actions of Tryphon as usurpations of power; it is evident from what follows that he was not sure of his position; had the Jews, by their support of a pretender, forfeited the privileges granted in 11.33-37?} \textcolor{benred8}{35}~When king Demetrius answered the request, he wrote in these terms following. \textcolor{benred8}{36}~King Demetrius to the high priest Simon, the friend of kings, and to all the elders and people of the Jews, greeting. \textcolor{benred8}{37}~Crown of gold and robe of scarlet you sent us were faithfully delivered. Great favour we mean to shew you, by sending word to the king\textquotesingle s officers to respect the remissions granted you. \textcolor{benred8}{38}~The decrees we made concerning you are yet in force; and, for the strongholds you have built, they shall be yours. \textcolor{benred8}{39}~Fault of yours in the past, witting or unwitting, is condoned; coronation tax you owed, and all other tribute that was due from Jerusalem, is due no longer. \textcolor{benred8}{40}~Fit be they for such enrolment, Jews shall be enrolled in our armies, and ever between us and you let there be peace!
\textcolor{benred8}{41}~Thus, in the hundred and seventieth year, Israel was free of the Gentile yoke at last; \textcolor{benred8}{42}~and this style the people began to use, were it private bond or public instrument they indited, In the first year of Simon\textquotesingle s high priesthood, chief paramount and governor of the Jews.
\textcolor{benred8}{43}~Then it was that Simon marched on Gaza,\footnote[3]{Josephus is perhaps right in reading \textasciigrave Gazara\textquotesingle  (less than twenty miles from Jerusalem) instead of \textasciigrave Gaza\textquotesingle ; cf. 14.7.} and beleaguered it with his army; built engines, and forced an entry into one of the towers. \textcolor{benred8}{44}~Out into the streets they sallied, that manned the engine, and there was a fine commotion in the city; \textcolor{benred8}{45}~here were the townsfolk, with their wives and children, mounting the walls with their garments rent about them, and crying aloud, Simon should give them quarter; \textcolor{benred8}{46}~great were their fault, greater still his clemency! \textcolor{benred8}{47}~At that, Simon relented; harry them to the death he would not, but he drove them out of the city, and cleansed all the houses where idols had stood; then, with singing of psalms and giving of thanks, he made his entry; \textcolor{benred8}{48}~and now all defilement must be put away, and such citizens it must have as did what the law commanded. After that, he fortified it, and made his own dwelling there.
\textcolor{benred8}{49}~And what of the Gentiles that were left in Jerusalem citadel? Enter Jewish territory or leave it they might not, buy or sell they might not, so that they were hard put to it for food, and many died of famine. \textcolor{benred8}{50}~At last they cried out to Simon, he should give them quarter, and give them quarter he did, but drove them out, and cleansed the citadel of its pollution. \textcolor{benred8}{51}~On the twenty-third day of the second month, in the hundred and seventy-first year, in came the Jewish folk singing praise and bearing palm-branches, with music of harp, and cymbals, and zither, and chanting of psalms; of such ill neighbours Israel was now rid. \textcolor{benred8}{52}~Every year, Simon proclaimed, holiday should be kept at this time; \textcolor{benred8}{53}~and he fortified that part of the temple mountain which was close by the citadel; here he dwelt, and his followers with him. \textcolor{benred8}{54}~And now here was his son John grown into a brave warrior; him Simon put at the head of the whole army, with his quarters at Gazara.
\begin{large}\begin{center}\textsc{Chapter xiv}\end{center}\end{large}
\lettrine[lines=2]{T}{he} year following, what must Demetrius do but muster his army and betake himself to Media, where he would raise levies for the war against Tryphon? \textcolor{benred8}{2}~When Arsaces, king of the Medes and Persians, had news of it, he gave orders to one of his chieftains, the invader must be taken alive and brought into his presence. \textcolor{benred8}{3}~The order was obeyed; the Syrian king, routed and captured, was brought before Arsaces, who put him safely in prison.
\textcolor{benred8}{4}~Thus, during Simon\textquotesingle s days, the whole land of Juda was at peace. Ever his people\textquotesingle s good sought he, and ever by willing hearts was obeyed and honoured. \textcolor{benred8}{5}~With great state he took possession of Joppe as a harbour,\footnote[1]{This is perhaps the best interpretation of a difficult phrase, \textasciigrave With all his glory he received Joppe to be a harbour\textquotesingle . For the language used, cf. 10.58, II Mac. 5.20. Joppe had been captured by Jonathan (10.76), and remained at least nominally in Jewish possession (12.33; 13.11); Simon\textquotesingle s achievement, according to verse 34 below, was to fortify it.} and so found access to the islands out at sea. \textcolor{benred8}{6}~How wide spread he the frontiers of Israel, how firmly held its possessions, \textcolor{benred8}{7}~captured how many of its foes! Gazara and Bethsura he won, ay, and the Citadel itself, ridding it of all defilement; there was no resisting his power. \textcolor{benred8}{8}~In his day, every man farmed his own lands in security, soil of Juda yielded its crops, and the trees their fruit; \textcolor{benred8}{9}~sat old men in the market-place, busy over the common weal, and young men wore the livery of their glorious campaigning. \textcolor{benred8}{10}~Never a city but he furnished it with store of provisions; a bulwark each of them should be of sturdy defence. What wonder if the story of his renown was noised to the world\textquotesingle s end? \textcolor{benred8}{11}~Such peaceful times brought he to his country, when all Israel kept high holiday, \textcolor{benred8}{12}~every man with his own vine and fig-tree for shade, and enemy was none to daunt them; \textcolor{benred8}{13}~domestic malice undone, foreign tyranny shattered all around! \textcolor{benred8}{14}~Among his own folk, what comfort he gave the friendless, how scrutinized the law, what short work made of traitor and malcontent; \textcolor{benred8}{15}~how adorned the sanctuary, how increased the number of its treasures!
\textcolor{benred8}{16}~To Rome, to Sparta itself, came tidings of Jonathan\textquotesingle s death, and was heard right sorrowfully. \textcolor{benred8}{17}~When they learned that his brother Simon had been made high priest instead, master now of the land and all its cities, \textcolor{benred8}{18}~they wrote to him on tablets made of bronze, to renew the treaty of friendship they had with his brethren, Judas and Jonathan, before him;\footnote[2]{The implication appears to be that both Rome and Sparta sent answers in the manner described.} \textcolor{benred8}{19}~and their letters were read out before the whole assembly at Jerusalem. The Spartans wrote in these words following: \textcolor{benred8}{20}~The rulers and commonalty of Sparta, to the high priest Simon, the elders and priests and all the people of the Jews, greeting. \textcolor{benred8}{21}~Welcome news your ambassadors have brought us, of fame and credit and prosperity you enjoy. \textcolor{benred8}{22}~And their errand stands recorded in our public annals; how Numenius son of Antiochus and Antipater son of Jason came on the Jews\textquotesingle  part to renew our old treaty of friendship with you; \textcolor{benred8}{23}~how the people resolved to give them fair greeting, and to lay up a copy of their report in the public archives, that should preserve the memory of it among the Spartan people; and how an account of all this was sent to the high priest Simon.
\textcolor{benred8}{24}~Numenius was sent on a further mission to Rome, bearing a great target of gold, a thousand minas in weight, to renew the alliance there. And when all this reached the ears of the people~\ldots\ \footnote[3]{v. 24. It would be natural to assume that the end of this verse refers to the Roman people; and the Clementine Vulgate, without any manuscript authority, actually inserts the word \textasciigrave Roman\textquotesingle . But the proceedings of verses 25 sqq. are evidently those of the Jewish people. There is a want of sequence in the narrative as it stands, and it seems possible that verses 15-24 of chapter 15 originally stood here, and were accidentally misplaced; if so, \textasciigrave all this\textquotesingle  will refer to the Roman dispatches.}
\textcolor{benred8}{25}~\ldots\  Men began to ask, how they could shew their gratitude to Simon, and to his sons? \textcolor{benred8}{26}~Here was one that had restored the fortunes of his race, and rid Israel of its foes. So they gave him exemption from public burdens, and inscribed their decree on tablets of bronze, fastened to pillars which were set up on mount Sion.
\textcolor{benred8}{27}~And thus the inscription ran: On this eighteenth day of Elul, in the hundred and seventy-second year of the Greek empire, the third of Simon\textquotesingle s high priesthood, \textcolor{benred8}{28}~there was a high assembly held at Saramel\footnote[4]{This name is not found elsewhere, and probably conceals some error in the text.} of priests and people, clan-chiefs and elders of the whole nation, that had before them these considerations following. All through the long wars of our country, \textcolor{benred8}{29}~Simon and his brethren, sons of Mattathias, of Jarib\textquotesingle s clan, put their lives in peril, and fought for law and sanctuary against the common enemy, much glory winning for their own nation. \textcolor{benred8}{30}~When Jonathan, that had rallied the people and been their high priest, became a part of his race, \textcolor{benred8}{31}~enemies thought to invade the country and crush the power of it, violate its holy places; \textcolor{benred8}{32}~and Simon it was withstood them. Champion of his people\textquotesingle s cause, much he spent to arm its warriors, and furnish them with pay. \textcolor{benred8}{33}~Juda\textquotesingle s cities he fortified, and others besides; Bethsura on the frontiers, once a stronghold of the enemy, garrisoned now by Jews, \textcolor{benred8}{34}~Joppe on the sea-coast, and Gazara in the Azotus region; Gazara, too, once hostile, with Jewish troops manned he, and in each town made provision for repairs to be done.
\textcolor{benred8}{35}~The people, seeing him so loyal a lover of his country\textquotesingle s renown, made him their ruler and high priest; no less was due to such exploits, public service so faithfully done, such constant ambition for his people\textquotesingle s honour. \textcolor{benred8}{36}~In his days it was, and by his means, the land was rid at last of Gentile intruders; not least the garrison of David\textquotesingle s own Keep at Jerusalem, that by their sallying out profaned the sacred precincts, and much defiled their purity; \textcolor{benred8}{37}~a Jewish garrison he set there, to guard both city and country-side, and built Jerusalem walls yet higher. \textcolor{benred8}{38}~High priesthood of his, king Demetrius must needs acknowledge, \textcolor{benred8}{39}~bestowing on him the title of king\textquotesingle s friend, and loading him with honours. What could he do else? \textcolor{benred8}{40}~Here was Rome itself greeting the Jewish folk as allies, good friends, and kinsmen, welcoming the envoys of Simon with civic state. \textcolor{benred8}{41}~Here were the Jews, priests and people both, agreed that he should rule them, granting him the high priesthood\footnote[5]{The Machabaean high priesthood seems to have been regarded as an emergency dispensation. For the mention of the prophet, cf. 4.46.} by right inalienable, until true prophet they should have once more. \textcolor{benred8}{42}~Their ruler he should be, and guardian of their temple; appoint officer and magistrate, master of ordnance and captain of garrison, \textcolor{benred8}{43}~and have charge of the sanctuary besides. Him all must obey, in his name deeds be drawn up, all the country through; of purple and gold should be his vesture. \textcolor{benred8}{44}~Of the rest, both priests and people, none should retrench these privileges, nor gainsay Simon\textquotesingle s will, nor convoke assembly in the country without him; garment of purple, buckle of gold none should wear; \textcolor{benred8}{45}~nor any man defy or void this edict, but at his peril.
\textcolor{benred8}{46}~The people\textquotesingle s pleasure it was to ennoble Simon after this sort; \textcolor{benred8}{47}~and Simon, he would not say them nay; high priest, and of priests and people leader, governor and champion, he would be henceforward. \textcolor{benred8}{48}~So they had the decree inscribed on tablets of bronze, and set up plain to view in the temple precincts; \textcolor{benred8}{49}~and a copy of it they put by in the treasury, in the safe keeping of Simon and his heirs.\footnote[6]{vv. 45-49. We should perhaps take these verses as part of the decree itself; otherwise verse 48 seems a needless repetition of verse 26.}
\begin{large}\begin{center}\textsc{Chapter xv}\end{center}\end{large}
\lettrine[lines=2]{N}{ow} turn we to Demetrius\textquotesingle  other son, Antiochus.\footnote[1]{Antiochus was son to the Demetrius whose death is recorded in 10.50, brother to the Demetrius first mentioned in 10.67.} He it was directed a letter, from the islands over sea, to Simon, high priest and ruler of the Jews, and to the whole nation; \textcolor{benred8}{2}~and this was the tenour of it. King Antiochus, to the high priest Simon and to the people of the Jews, greeting. \textcolor{benred8}{3}~Here is the kingdom of my fathers overrun by ill folk; I mean to challenge them, and bring back the old ways. To this end, I have made a great levy of mercenaries, and built ships of war; \textcolor{benred8}{4}~passage I needs must have through yonder territory, ere I can take vengeance for lands of mine ravaged, cities of mine laid waste. \textcolor{benred8}{5}~This grant, then, I make thee; exemption, such as thou hadst, in the name of former kings, from public offerings and all other payment due to me; \textcolor{benred8}{6}~the right of minting money within thy own borders; \textcolor{benred8}{7}~for Jerusalem, enjoyment of her sacred liberties; of weapons thou hast made, strongholds thou hast built, the undisturbed possession. \textcolor{benred8}{8}~Never a claim the king has, or shall have hereafter, on his subjects, but to thee it is remitted; \textcolor{benred8}{9}~and, when the kingdom is ours, such honours we will bestow as shall make thee, and thy race, and its sanctuary, renowned all the world over.
\textcolor{benred8}{10}~So, in the hundred and seventy-fourth year, Antiochus returned to his native country, and the armies rallied to him, until Tryphon had but a small following left; \textcolor{benred8}{11}~all down the sea-coast he fled, with Antiochus at his heels, till he reached Dora; \textcolor{benred8}{12}~and ever he saw the toils closing round him, now his troops had played him false. \textcolor{benred8}{13}~With a hundred and twenty thousand foot, and eight thousand horse, Antiochus came to the gates of Dora \textcolor{benred8}{14}~and began the siege of it; his ships, too, blockaded the coast, so that it was cut off by land and sea alike; enter it none might, nor leave it~\ldots\ 
\textcolor{benred8}{15}~And now here were Numenius and his fellow envoys come back from Rome,\footnote[2]{v. 15-24. See note on 14.24.} with a copy of despatches sent out to kings and nations everywhere, and this was the tenour of them. \textcolor{benred8}{16}~Lucius, the Roman consul, to king Ptolemy, greeting. \textcolor{benred8}{17}~Envoys we have but now received in audience from a friendly country, to wit, Judaea; the people of the Jews, with their high priest Simon, had sent to renew their old treaty of alliance with us, \textcolor{benred8}{18}~and had made us a present besides, a golden target of a thousand minas weight. \textcolor{benred8}{19}~Agreed we then to warn kings and nations everywhere, they should not hurt or assault the Jewish people, its cities and country-side, nor comfort its enemies; \textcolor{benred8}{20}~and for the target of gold, our pleasure was to accept the gift of it. \textcolor{benred8}{21}~If then there be malcontents from Judaea sheltering among you, our bidding is you should hand them over to the high priest Simon, for such punishment as the Jewish law prescribes. \textcolor{benred8}{22}~Copies of this decree have been sent to Demetrius, Attalus, Ariarathes and Arsaces, \textcolor{benred8}{23}~and to these countries following: Lampsacus, Sparta, Delos, Myndos, Sicyon, Caria, Samos, Pamphylia, Lycia, Halicarnassus, Cos, Side, Arados, Rhodes, Phaselis, Gortyna, Cnidus, Cyprus and Cyrene. \textcolor{benred8}{24}~A further copy has been sent to the high priest Simon and to the Jewish people~\ldots\ 
\textcolor{benred8}{25}~Once again\footnote[3]{Some Greek manuscripts have \textasciigrave on the second day\textquotesingle , instead of \textasciigrave once again\textquotesingle , but this is probably a correction, designed to clear up a difficulty. Nothing has been said which implies that the siege described in verse 14 had come to an end; and it is not easy to account for the mention of a second siege, unless we suppose a gap in the manuscript which has been accidentally filled up by verses 15-24.} king Antiochus laid siege to Dora, bringing fresh force to bear, and devising fresh engines; and ever he kept Tryphon hemmed in, so that escape was none. \textcolor{benred8}{26}~Thereupon Simon despatched two thousand picked men to aid in the siege, with silver and gold and a deal of tackle besides; \textcolor{benred8}{27}~but accept them the king would not; all his promises were forgot, and Simon a stranger now. \textcolor{benred8}{28}~Athenobius it was, one of the king\textquotesingle s friends, that came to treat with him, and this was the message he bore: Cities of mine you hold, Joppe, and Gazara, and Jerusalem citadel; \textcolor{benred8}{29}~lands about them you have laid waste, and done Syria much mischief besides, encroaching everywhere on my domain. \textcolor{benred8}{30}~Needs must you should hand over cities you have occupied, revenues of Gentile lands you have detained, \textcolor{benred8}{31}~or else five hundred talents of silver in exchange, and five hundred more to compensate for damage done and revenue lost; if not, we will come and overpower you by force of arms.
\textcolor{benred8}{32}~So came Athenobius, the king\textquotesingle s friend, to Jerusalem, where he saw what state Simon kept, much display of gold and silver, and a great throng of attendants, till he was dazzled at the sight. Yet delivered he his errand; \textcolor{benred8}{33}~to which Simon made this answer: Other men\textquotesingle s fief seized we never, nor other men\textquotesingle s rights detain; here be lands that were our fathers\textquotesingle  once, by enemies of ours for some while wrongfully held; \textcolor{benred8}{34}~opportunity given us, should we not claim the patrimony we had lost? \textcolor{benred8}{35}~As for thy talk of Joppe and Gazara, these were cities did much mischief to people and land of ours; for the worth of them, thou shalt have a hundred talents if thou wilt. Never a word said Athenobius, \textcolor{benred8}{36}~but went back to the king very ill pleased, and told him what answer was given; of Simon\textquotesingle s court, too, and of all else he had seen.Antiochus was in a great taking of anger; \textcolor{benred8}{37}~here was Tryphon newly escaped by ship to Orthosias! \textcolor{benred8}{38}~He must needs leave the sea-coast in charge of Cendebaeus, with a strong command both of horse and foot, \textcolor{benred8}{39}~while himself gave Tryphon chase. This Cendebaeus had orders to advance and threaten Judaea; Gedor\footnote[4]{\textasciigrave Gedor\textquotesingle  is \textasciigrave Cedron\textquotesingle  in the Greek text, here and in verse 40 below, to correspond with verse 9 of the following chapter.} he should fortify, and there make himself fast, the better to levy war on Juda. \textcolor{benred8}{40}~So he marched away to Jamnia, and set about harassing the Jews; now it was an inroad, with prisoners carried away, now a massacre; and all the while he was fortifying Gedor. \textcolor{benred8}{41}~Cavalry he quartered there, and other troops besides, to sally out and patrol the roads into Judaea; the king would have it so.
\begin{large}\begin{center}\textsc{Chapter xvi}\end{center}\end{large}
\lettrine[lines=2]{I}{t} was not long before John came up from Gazara, to tell his father Simon how ill Cendebaeus was using their fellow-countrymen. \textcolor{benred8}{2}~And at that, Simon must have his two elder sons present, Judas and John both, and made the command over to them. Still young we were, he said, I and my brothers and my father\textquotesingle s kin, when we began that war on Israel\textquotesingle s enemies which is being fought yet; under our banners once and again came victory, and the day was saved for Israel. \textcolor{benred8}{3}~I am an old man now, and it is yours to do what I and brother of mine did; march out, fight in our people\textquotesingle s cause, and heaven\textquotesingle s aid be with you!
\textcolor{benred8}{4}~Twenty thousand warriors John\footnote[1]{The name John is not mentioned in the original, either here or in verses 6 and 7 below. But it is plain that either John or Judas is meant, and 13.54 seems to give the best grounds for a decision.} chose out from the rest, and cavalry to support them, and away they went to fight Cendebaeus. That night they spent at Modin, \textcolor{benred8}{5}~and on the morrow, when they left it for the valley, what a huge array was this, both of horse and foot, encountering them! And a mountain torrent flowed in between. \textcolor{benred8}{6}~When John brought his army to the opposite bank, and found his men had little stomach for the crossing, he made the passage first, leaving the rest to follow at his heels; \textcolor{benred8}{7}~then drew them up by companies, with the cavalry in between, so greatly did the enemy\textquotesingle s cavalry outnumber them. \textcolor{benred8}{8}~And now the sacred trumpets sounded the charge; fled Cendebaeus, fled his army at their onslaught, and many were left dead on the field; for the rest, they were fain to take refuge behind their walls again. \textcolor{benred8}{9}~John went in pursuit, for all his brother Judas had been wounded in the battle, and chased them as far as the walls of Cedron~\ldots\  which he had fortified.\footnote[2]{v. 9. It is grammatically impossible to make Cendebaeus the subject of the verb \textasciigrave he had fortified\textquotesingle , as the sentence stands. It seems likely that there is some slight error or omission in the manuscripts.} \textcolor{benred8}{10}~Nor might they find shelter in the strongholds of the Azotus territory; he burnt these to the ground; a toll of two thousand men he had taken before he returned victorious to Judaea.
\textcolor{benred8}{11}~Turn we now to Ptolemy, son of Abobus, that was in charge of all Jericho plain, and had a purse well lined with silver and gold; \textcolor{benred8}{12}~was he not the son-in-law of a high priest?\footnote[3]{\textasciigrave Of a high priest\textquotesingle ; Josephus understands this as referring to Simon himself. But it does not seem likely that the author should have suppressed his name in verse 12, only to mention it in verse 13; nor does he mention the circumstances of affinity as adding to the heinousness of the crime (cf. verse 17). Possibly some other name has dropped out, e.g. that of Alcimus.} \textcolor{benred8}{13}~But higher still his ambition ran; he would make himself master of the whole country; murder he plotted for Simon and his sons together. \textcolor{benred8}{14}~It was in Sabath, the eleventh month, of the hundred and seventy-seventh year, that Simon came down to Jericho, as ever he visited all the cities of Judaea in his great care for them; and his sons Mattathias and Judas went with him. \textcolor{benred8}{15}~And there, in a castle he had built for himself, Doch is the name of it, the son of Abobus gave them treacherous welcome. A great feast he made, but he had men waiting in readiness, \textcolor{benred8}{16}~and with these, when Simon and his sons had drunk deep, he took arms, broke into the banqueting-chamber, and slew both father and sons, with certain of their retinue. \textcolor{benred8}{17}~Never saw Israel so treacherous a deed, or good service so ill rewarded.
\textcolor{benred8}{18}~News of all this was sent by Ptolemy to the king, and in writing; his plea was, an army should be sent out in support of him, and the country, with all its cities and all the tribute that came from them, given into his charge. \textcolor{benred8}{19}~Others of his men he despatched to Gazara; John must be put to death, he wrote, and for the captains, they should have silver and gold and good recompense, would they but rally to his side; \textcolor{benred8}{20}~others again were to take possession of Jerusalem, and of the temple hill. \textcolor{benred8}{21}~But too late; a messenger had reached John at Gazara, telling him his father and brothers were dead, and himself too marked down for slaughter; \textcolor{benred8}{22}~whereupon he took alarm in good earnest; their murderous errand known, he seized his executioners and made an end of them.
\textcolor{benred8}{23}~What else John did, and how fought he, brave deeds done, and strong walls built, and all his history, \textcolor{benred8}{24}~you may read in the annals of his time, that were kept faithfully since the day when he succeeded his father as high priest.


\end{document}






