%TEXT FROM:
%http://www.newadvent.org/bible/1ma001.htm

\documentclass[10pt]{book} % use larger type; default would be 10pt

\usepackage[latin1]{luainputenc}

\usepackage[oldstyle]{libertine}
\usepackage{lettrine}
\usepackage{xcolor}

% dual language parallel text thing
\usepackage{reledmac}
\usepackage{reledpar}

% BLUES
\definecolor{benblue1}{HTML}{2B22C7}

% REDS
\definecolor{benred8}{HTML}{E82C00} %seems good, maybe the best?

% THREE COMMANDS TO CLEAN UP ALL THE ELEDPAR/ELEDMAC STUFF
\newcommand{\StartOfLatin}{
	\begin{pairs}
	\begin{Leftside}
	\firstlinenum{10000}\linenumincrement{10000}\beginnumbering\pstart
}

\newcommand{\StartOfEnglish}{
	\pend\endnumbering
	\end{Leftside}
	\begin{Rightside}

	\firstlinenum{10000}\linenumincrement{10000}\beginnumbering\pstart
}

\newcommand{\EndOfEnglish}{
	\pend\endnumbering
	\end{Rightside}
	\end{pairs}
	\Columns
}

\usepackage{fancyhdr}
\pagestyle{fancy}
\cfoot{\textcolor{benblue1}{\thepage}}
\renewcommand{\headrulewidth}{0pt}

%redefine 'plain' page styel to have blue page numbers too
\fancypagestyle{plain}{
	\chead{}
	\cfoot{\textcolor{benblue1}{\thepage}}

}

\begin{document}

\chapter*{Psalms}

\begin{large}\begin{center}\textsc{Chapter i}\end{center}\end{large}
\lettrine[lines=3]{B}{lessed} is the man who does not guide his steps by ill counsel, or turn aside where sinners walk, or, where scornful souls gather, sit down to rest; \textcolor{benred8}{2}~the man whose heart is set on the law of the Lord, on that law, day and night, his thoughts still dwell. \textcolor{benred8}{3}~He stands firm as a tree planted by running water, ready to yield its fruit when the season comes, not a leaf faded; all that he does will prosper. \textcolor{benred8}{4}~Not such, not such the wicked; the wicked are like chaff the wind sweeps away. \textcolor{benred8}{5}~Not for the wicked, when judgement comes, to rise up and plead their cause; sinners will have no part in the reunion of the just. \textcolor{benred8}{6}~They walk, the just, under the Lord\textquotesingle s protection; the path of the wicked, how soon is it lost to sight!
\begin{large}\begin{center}\textsc{Chapter ii}\end{center}\end{large}
\lettrine[lines=2]{W}{hat} means this turmoil among the nations? Why do the peoples cherish vain dreams? \textcolor{benred8}{2}~See how the kings of the earth stand in array, how its rulers make common cause, against the Lord, and against the King he has anointed, \textcolor{benred8}{3}~crying, Let us break away from their bondage, rid ourselves of the toils! \textcolor{benred8}{4}~He who dwells in heaven is laughing at their threats, the Lord makes light of them; \textcolor{benred8}{5}~and at last, in his displeasure, he will speak out, his anger quelling them: \textcolor{benred8}{6}~Here, on mount Sion, my sanctuary, I enthrone a king of my own choice.Mine to proclaim the Lord\textquotesingle s edict; \textcolor{benred8}{7}~how he told me, Thou art my son; I have begotten thee this day. \textcolor{benred8}{8}~Ask thy will of me, and thou shalt have the nations for thy patrimony; the very ends of the world for thy domain. \textcolor{benred8}{9}~Thou shalt herd them like sheep\footnote[1]{\textasciigrave Herd them like sheep\textquotesingle ; or, according to another reading, \textasciigrave shatter them\textquotesingle .} with a crook of iron, break them in pieces like earthenware. \textcolor{benred8}{10}~Princes, take warning; learn your lesson, you that rule the world. \textcolor{benred8}{11}~Tremble, and serve the Lord, rejoicing in his presence, but with awe in your hearts. \textcolor{benred8}{12}~Kiss the rod,\footnote[2]{The word given in the Hebrew text should mean either \textasciigrave purity\textquotesingle , or \textasciigrave the Son\textquotesingle .} do not brave his anger, and go astray from the sure path. \textcolor{benred8}{13}~When the fire of his vengeance blazes out suddenly, happy are they who find their refuge in him.
\begin{large}\begin{center}\textsc{Chapter iii}\end{center}\end{large}
\textcolor{benred8}{1}~(A psalm David wrote, when he fled before his son Absalom.)
\textcolor{benred8}{2}~See how they surround me, Lord, my adversaries, how many rise up in arms against me; \textcolor{benred8}{3}~everywhere voices taunting me, His God cannot save him now. \textcolor{benred8}{4}~Yet, Lord, thou art my champion, thou art the pride that keeps my head erect. \textcolor{benred8}{5}~I have but to cry out to the Lord, and my voice reaches his mountain sanctuary, and there finds hearing. \textcolor{benred8}{6}~Safe in God\textquotesingle s hand I lay down, and slept, and have awoken; \textcolor{benred8}{7}~and now, though thousands of the people set upon me from every side, I will not be afraid of them. Bestir thyself, Lord; my God, save me; \textcolor{benred8}{8}~thine to smite my enemies on the cheek, thine to break the fangs of malice. \textcolor{benred8}{9}~From the Lord all deliverance comes; let thy blessing, Lord, rest upon thy people.
\begin{large}\begin{center}\textsc{Chapter iv}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. On stringed instruments. A psalm. Of David.\footnote[1]{All these titles prefixed to various psalms are very obscure, and there is little agreement about their meaning except that they refer to the musical and liturgical setting.})
\textcolor{benred8}{2}~When I call on thy name, listen to me, O God, and grant redress; still, in time of trouble, thou hast brought me relief; have pity on me now, and hear my prayer. \textcolor{benred8}{3}~Great ones of the world, will your hearts always be hardened,\footnote[2]{Our Hebrew text has, \textasciigrave Will my honour always be turned into reproach?\textquotesingle } will you never cease setting your heart on shadows, following a lie? \textcolor{benred8}{4}~To the souls he loves, be sure the Lord shews wondrous favour; whenever I call on his name, the Lord will hear me. \textcolor{benred8}{5}~Tremble, and sin no more; take thought, as you lie awake, in the silence of your hearts. \textcolor{benred8}{6}~Offer sacrifices with due observance, and put your trust in the Lord. There are many that cry out for a sight of better times; \textcolor{benred8}{7}~do thou, then, Lord, shew us the sunshine of thy favour. \textcolor{benred8}{8}~Never did rich harvests of corn and wine bring gladness like the gladness thou puttest into my heart. \textcolor{benred8}{9}~Even as I lie down, sleep comes, and with sleep tranquillity; \textcolor{benred8}{10}~what need, Lord, of aught but thyself to bring me confidence?
\begin{large}\begin{center}\textsc{Chapter v}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. On the flute. A psalm. Of David.)
\textcolor{benred8}{2}~Lord, listen to my plea, let me not sigh in vain; \textcolor{benred8}{3}~pay heed to my cry of petition, my King, my God. \textcolor{benred8}{4}~To thee, Lord, my prayer goes up, early to win thy audience; \textcolor{benred8}{5}~early in the morning I lay my petition before thee and await thy pleasure. No evil thing claims thy divine assent; \textcolor{benred8}{6}~with thee baseness cannot dwell; nor rebellion hold its ground at thy coming. \textcolor{benred8}{7}~Thou hatest the wrongdoer, and wilt bring the liar to destruction; blood-thirsty and treacherous men the Lord holds in abhorrence. \textcolor{benred8}{8}~I, then, encompassed by thy mercy, will betake myself to thy house, and in reverence of thee bow down before thy sanctuary.
\textcolor{benred8}{9}~Lord, do thou lead me with faithful care; clear show the path, while I walk beset by enemies. \textcolor{benred8}{10}~In their speech no truth can be found; their hearts are all treachery, \textcolor{benred8}{11}~their mouths gaping tombs; flattering is ever on their lips. Thy scourge, O God! Cheat them of their hopes, cast them out in all their wickedness; have they not defied thee? \textcolor{benred8}{12}~But for all those who trust in thee there is joy and everlasting triumph; welcome protection they have from thee, true lovers of thy name. \textcolor{benred8}{13}~Lord, thou givest thy benediction to the just; thou dost throw thy loving-kindness about us like a shield.
\begin{large}\begin{center}\textsc{Chapter vi}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. On stringed instruments. Over the octave. A psalm. Of David.)
\textcolor{benred8}{2}~Lord, when thou dost reprove me, let it not be in anger; when thou dost chastise me, let it not be in displeasure. \textcolor{benred8}{3}~Lord, pity me; I have no strength left; Lord, heal me; my limbs tremble; \textcolor{benred8}{4}~my spirits are altogether broken; Lord, wilt thou never be content? \textcolor{benred8}{5}~Lord, turn back, and grant a wretched soul relief; as thou art ever merciful, save me. \textcolor{benred8}{6}~When death comes, there is no more remembering thee; none can praise thee in the tomb. \textcolor{benred8}{7}~I am spent with sighing; every night I lie weeping on my bed, till the tears drench my pillow. \textcolor{benred8}{8}~Grief has dimmed my eyes, faded their lustre now, so many are the adversaries that surround me. \textcolor{benred8}{9}~Depart from me, all you that traffic in iniquity; the Lord has heard my cry of distress. \textcolor{benred8}{10}~Here was a prayer divinely heard, a boon divinely granted. \textcolor{benred8}{11}~All my enemies will be abashed and terrified; taken aback, all in a moment, and put to shame.
\begin{large}\begin{center}\textsc{Chapter vii}\end{center}\end{large}
\textcolor{benred8}{1}~(A lament of David\textquotesingle s, which he sang to the Lord because of Chus, the Benjamite.)
\textcolor{benred8}{2}~O Lord my God, my confidence is in thee; save me from all my pursuers, and grant me deliverance: \textcolor{benred8}{3}~else must I fall a helpless prey to the lion, be torn in pieces, with none to bring me aid. \textcolor{benred8}{4}~O Lord my God, if I too have been at fault, if these hands are stained with guilt; \textcolor{benred8}{5}~if I have been a false friend, and not rather spared even those that wronged me, \textcolor{benred8}{6}~then indeed let some enemy overtake me with his relentless pursuit, trample me to earth, and level my pride with the dust! \textcolor{benred8}{7}~Lord, rise up in thy anger, countervail the malice of my enemies; bestir thyself, O Lord my God, in defence of the laws thou thyself hast given us. \textcolor{benred8}{8}~All the nations will gather about thee, if thou wilt come back to thy throne and rule them, \textcolor{benred8}{9}~the Lord judging the nations! Give me redress, Lord, in my uprightness, in all the innocence of my heart. \textcolor{benred8}{10}~Surely thou wilt put an end to the wrong-doing of the wicked, and prosper the innocent; no thought or desire of ours can escape the scrutiny of thy divine justice. \textcolor{benred8}{11}~From the Lord, refuge of true hearts, my protection comes. \textcolor{benred8}{12}~God judges ever true; day by day his indignation mounts up; \textcolor{benred8}{13}~if they do not repent, his sword will flash bright; he has bent and aimed his bow; \textcolor{benred8}{14}~deadly are the weapons he is preparing for them; he has barbed his arrows with fire. \textcolor{benred8}{15}~Here was a heart pregnant with malice, that conceived only spite, and gave birth only to empty promise! \textcolor{benred8}{16}~Here was one who dug a pit and sunk it deep, and fell into a snare of his own setting! \textcolor{benred8}{17}~All his spite will recoil on himself, all his violence will fall his own head. \textcolor{benred8}{18}~I will ever thank the Lord for his just retribution, singing praises to the name of the Lord, the most High.
\begin{large}\begin{center}\textsc{Chapter viii}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. To the mood of the song, The Wine-presses. A psalm. Of David.)
\textcolor{benred8}{2}~O Lord, our Master, how the majesty of thy name fills all the earth! Thy greatness is high above heaven itself. \textcolor{benred8}{3}~Thou hast made the lips of children, of infants at the breast,\footnote[1]{Some would interpret the Hebrew here, \textasciigrave thou has founded strength through the lips of children and infants at the breast\textquotesingle .} vocal with praise, to confound thy enemies; to silence malicious and revengeful tongues. \textcolor{benred8}{4}~I look up at those heavens of thine, the work of thy hands, at the moon and the stars, which thou hast set in their places; \textcolor{benred8}{5}~what is man that thou shouldst remember him? What is Adam\textquotesingle s breed, that it should claim thy care? \textcolor{benred8}{6}~Thou hast placed him only a little below the angels,\footnote[2]{The Hebrew can also be translated \textasciigrave below God\textquotesingle .} crowning him with glory and honour, \textcolor{benred8}{7}~and bidding him rule over the works of thy hands. \textcolor{benred8}{8}~Thou hast put them all under his dominion, the sheep and the cattle, and the wild beasts besides; \textcolor{benred8}{9}~the birds in the sky, and the fish in the sea, that travel by the sea\textquotesingle s paths. \textcolor{benred8}{10}~O Lord, our Master, how the majesty of thy name fills all the earth!
\begin{large}\begin{center}\textsc{Chapter ix}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. To the mood of the song, Mut Ladben. A psalm. Of David.)
\textcolor{benred8}{2}~Lord, I give thee all the thanks of my heart, recounting thy wonderful do-ings; \textcolor{benred8}{3}~glad and triumphant in thee, I will sing psalms to thy name, O God most high. \textcolor{benred8}{4}~See how my enemies turn back, how they faint and melt away at the sight of thee! \textcolor{benred8}{5}~Thou hast given me redress and maintained my cause; thou art there on thy throne, seeing justice done. \textcolor{benred8}{6}~Thou hast checked the heathen in their course; thou hast brought the wicked to nothing, blotting out their name for all time. \textcolor{benred8}{7}~Spent is the enemy\textquotesingle s power, doomed to everlasting ruin; the memory of them has died with the fall of their cities. \textcolor{benred8}{8}~But the Lord abides for ever on the throne of judgement he has prepared, \textcolor{benred8}{9}~still judging the world rightly, still awarding each people its due; \textcolor{benred8}{10}~the Lord is a stronghold to the oppressed, a stronghold in time of peril. \textcolor{benred8}{11}~Those who acknowledge thy name, Lord, can trust in thee; never was man forsaken that had recourse to thee.
\textcolor{benred8}{12}~Sing, then, to the Lord, who dwells in Sion, tell the Gentiles of his great deeds; \textcolor{benred8}{13}~how he, the avenger of blood, cares for the afflicted, does not forget them when they cry to him. \textcolor{benred8}{14}~Have pity on me, Lord, look upon all that I suffer at my enemies\textquotesingle  hands; \textcolor{benred8}{15}~thou who didst ever rescue me from the gate of death, to proclaim thy praises at the gate of thy loved Sion, \textcolor{benred8}{16}~to exult in thy saving power. The heathen have been caught in their own deadly devices; their feet have been trapped in the very toils they had laid; \textcolor{benred8}{17}~now it will be seen how the Lord defends the right, how the wicked contrive their own undoing. \textcolor{benred8}{18}~To the place of death the wicked must return, heathens that have no thought of God. \textcolor{benred8}{19}~He does not forget the helpless; their time will come; the patience of the afflicted will not go for nothing. \textcolor{benred8}{20}~Bestir thyself, Lord, let not human strength prevail; let the heathen stand upon their trial before thee; \textcolor{benred8}{21}~let the heathen, too, feel thy terrors, and learn they are but men.(HERE, IN THE HEBREW TEXT, PSALM 10 BEGINS.)
\textcolor{benred8}{22}~Lord,\footnote[1]{The numbering of the Psalms here corresponds with that of the Vulgate. In the Hebrew text, which is followed by some modern versions, verses 22-39 of this Psalm are treated as a separate psalm and numbered 10; while Psalms 146 and 147 are given as one psalm. Thus the Vulgate numbers 10 to 147 correspond to 11-147 of the alternative reckoning.} why dost thou stand far off? In days of affliction, why dost thou make no sign? \textcolor{benred8}{23}~The hearts of the oppressed burn within them, so triumphant is the schemer that has entrapped them; \textcolor{benred8}{24}~so proud of his wicked end achieved, still robbing men, blaspheming and despising the Lord. \textcolor{benred8}{25}~God there is none to punish me, the sinner thinks in his pride, and makes that thought his rule; \textcolor{benred8}{26}~still, as he goes on prospering, he banishes thy laws from his mind, and makes light of his enemies. \textcolor{benred8}{27}~Endless time, he thinks, cannot shake his untroubled existence. \textcolor{benred8}{28}~His mouth overflows with curses, and calumny, and deceit; his tongue is a storehouse of dissension and mischief. \textcolor{benred8}{29}~Ambushed he lies at the village gate, to kill unawares the man who never wronged him; \textcolor{benred8}{30}~his eyes are continually on his prey; like a lion in its lair, he watches from his hiding-place, to surprise his defenceless foe, safe in the net. \textcolor{benred8}{31}~So he catches him in the toils; stands there bowing and scraping,\footnote[2]{v. 31. This seems the most natural way of taking a much-disputed sentence.} till the prey falls by his onset. \textcolor{benred8}{32}~Why not? he thinks to himself, God has forgotten about it; God still turns his face away, and sees nothing.
\textcolor{benred8}{33}~O Lord God, bestir thyself, lift up thy hand; do not forget the helpless. \textcolor{benred8}{34}~Why is the sinner allowed to defy God, to think he will never exact punishment? \textcolor{benred8}{35}~But in truth thou seest it; thou hast eyes for misery and distress, and wilt take them into thy keeping. The destitute are cast on no care but thine; to thee only the orphan looks for redress. \textcolor{benred8}{36}~Break down the power of the wicked oppressor, punish his ill-doing, and let him be seen no more. \textcolor{benred8}{37}~The Lord will reign for ever and ever, while you, the heathen, will vanish from the land he loves. \textcolor{benred8}{38}~The sighing of the defenceless has found audience; thou wilt heed them and bring courage to their hearts, \textcolor{benred8}{39}~wilt give redress to the fatherless and the persecuted; mortal man shall make himself feared no longer.
\begin{large}\begin{center}\textsc{Chapter x}\end{center}\end{large}
(To the choir-master. Of David.)
\textcolor{benred8}{2}~My trust is in the Lord; how is it that you say to me, Escape, like a frightened sparrow, to the hill-side? \textcolor{benred8}{3}~Escape; the rebels have strung their bows, have arrows ready on the string, to shoot from their hiding-places at an unoffending heart; \textcolor{benred8}{4}~they have thrown down all thou hadst built; what hope, now, for the just man?
\textcolor{benred8}{5}~Is not the Lord in his holy shrine, the same Lord whose throne is in heaven, whose eye watches, whose glance can appraise, the deeds of men? \textcolor{benred8}{6}~Innocent or sinful, he reads every heart, and the friends of wrong-doing are his enemies. \textcolor{benred8}{7}~Pitilessly his weapons rain down upon the offenders; burning coals, and brimstone, and scorching wind; such is the draught he brews for them. \textcolor{benred8}{8}~The Lord is just, and just are the deeds he loves; none but upright souls shall enjoy his presence.
\begin{large}\begin{center}\textsc{Chapter xi}\end{center}\end{large}
(To the choir-master. Over the octave. A psalm. Of David.)
\textcolor{benred8}{2}~Lord, come to my rescue; piety is dead; in a base world, true hearts have grown rare. \textcolor{benred8}{3}~None but exchanges empty forms of speech with his neighbour; everywhere false hearts and treacherous lips. \textcolor{benred8}{4}~Those treacherous lips, that tongue with high-sounding phrases; Lord, rid the earth of them! \textcolor{benred8}{5}~With our tongues, they say, we can do great things; our lips are good friends to us; we own no master. \textcolor{benred8}{6}~Now, says the Lord, I will bestir myself, on behalf of the helpless who are so ill used, of the poor who cry out so bitterly; I will win them the redress they long for. \textcolor{benred8}{7}~The promises of the Lord are true metal, like silver that is tested in the crucible, the stains of earth gone, seven times refined. \textcolor{benred8}{8}~Yes, Lord, thou wilt watch over us, and keep us ever safe from these evil days.
\textcolor{benred8}{9}~See how the wicked come and go all around us, how they rise to greatness, this base breed of men!
\begin{large}\begin{center}\textsc{Chapter xii}\end{center}\end{large}
(To the choir-master. A psalm. Of David.)
\textcolor{benred8}{1}~Lord, must I still go all unremembered, must thy look still be turned away from me? \textcolor{benred8}{2}~Each day brings a fresh load of care, fresh misery to my heart; \textcolor{benred8}{3}~must I be ever the sport of my enemies? \textcolor{benred8}{4}~Look upon me, O Lord my God, and listen to me; give light to these eyes, before they close in death; \textcolor{benred8}{5}~do not let my enemies claim the mastery, my persecutors triumph over my fall! \textcolor{benred8}{6}~I cast myself on thy mercy; soon may this heart boast of redress granted, sing in praise of the Lord, my benefactor.
\begin{large}\begin{center}\textsc{Chapter xiii}\end{center}\end{large}
(To the choir-master. Of David.)
\textcolor{benred8}{1}~There is no God above us, is the fond thought of reckless hearts; warped natures everywhere and hateful lives! There is not an innocent man among them. \textcolor{benred8}{2}~The Lord looks down from heaven at the race of men, to find one soul that reflects, and makes God its aim; \textcolor{benred8}{3}~but no, all have missed the mark and rebelled against him; an innocent man is nowhere to be found. \textcolor{benred8}{4}~What, can they learn nothing, all these traffickers in iniquity, who feed themselves fat on this people of mine, as if it were bread for their eating, \textcolor{benred8}{5}~and never invoke the Lord\textquotesingle s name? What wonder if fear unmans them, when the Lord takes the part of the innocent? \textcolor{benred8}{6}~Easily you thought to outwit the friendless; see, the Lord is his refuge! \textcolor{benred8}{7}~Oh, that it might dawn over Sion, Israel\textquotesingle s deliverance! Day of gladness for Jacob, day of Israel\textquotesingle s triumph, when the Lord restores the fortunes of his own people.\footnote[1]{This psalm is a curious literary problem; the whole of it can be found elsewhere in the Bible. Verses 1 and 2, with the first sentence of verse 3, and verses 4-7, reappear in Psalm 52, with some variation in verse 6. The remainder of verse 3 is found in the Greek and Latin, but not in the Hebrew text. It corresponds exactly with Rom. 3.10-18, and appears to be a canto of Scriptural quotations taken from the following sources: Ps. 5.10, 139.4, 9.28, Is. 59.7, Ps. 35.2.}
\begin{large}\begin{center}\textsc{Chapter xiv}\end{center}\end{large}
(A psalm. Of David.)
\textcolor{benred8}{1}~Who is it, Lord, that will make his home in thy tabernacle, rest on the mountain where thy sanctuary is? \textcolor{benred8}{2}~One that guides his steps without fault, and gives to all their due; \textcolor{benred8}{3}~one whose heart is all honest purpose, who utters no treacherous word, never defrauds a friend, or slanders a neighbour. \textcolor{benred8}{4}~He scorns the reprobate, keeping his reverence for such as fear God, and is true, come what may, to his pledged word; \textcolor{benred8}{5}~lends without usury, and takes no bribe to condemn the innocent. He who so lives will stand firm for ever.
\begin{large}\begin{center}\textsc{Chapter xv}\end{center}\end{large}
(A miktam. Of David.)
\textcolor{benred8}{1}~Keep me safe, Lord; I put my trust in thee. \textcolor{benred8}{2}~The Lord, whom I own as my God, confess that in him is all my good! \textcolor{benred8}{3}~There are faithful souls in this land of his; wondrous delight he gives me in their companionship. \textcolor{benred8}{4}~What do they do but lay up fresh store of sorrows, that betake themselves to alien gods? Not with these will I pour out the blood of sacrifice; I will not take forbidden names on my lips. \textcolor{benred8}{5}~No, it is the Lord I claim for my prize, the Lord who fills my cup; thou, and no other, wilt assure my inheritance to me. \textcolor{benred8}{6}~Portion is none were more to my liking; welcome the lot\textquotesingle s choice!
\textcolor{benred8}{7}~Blessed be the Lord, who schools me; late into the night my inmost thoughts chasten me. \textcolor{benred8}{8}~Always I can keep the Lord within sight; always he is at my right hand, to make me stand firm. \textcolor{benred8}{9}~Glad and merry am I, heart and soul of me; my body, too, shall rest in confidence \textcolor{benred8}{10}~that thou wilt not leave my soul in the place of death, or allow thy faithful servant to see corruption. \textcolor{benred8}{11}~Thou wilt shew me the way of life, make me full of gladness in thy presence; at thy right hand are delights that will endure for ever.
\begin{large}\begin{center}\textsc{Chapter xvi}\end{center}\end{large}
(A prayer. Of David.)
\textcolor{benred8}{1}~Lord, to my just complaint give ear; do not spurn my cry for aid. Listen to this prayer of mine; they are no treacherous lips that make it. \textcolor{benred8}{2}~At thy judgement seat I claim award; unerring thy scrutiny. \textcolor{benred8}{3}~Wilt thou read my heart, drawing near in the darkness to test me as if by fire, thou wilt find no treachery in me. \textcolor{benred8}{4}~Never have these lips been led astray by man\textquotesingle s evil example; still to thy law\textquotesingle s pattern thy warnings kept me true; \textcolor{benred8}{5}~still in thy paths my steps were firmly planted, my feet did not stumble.
\textcolor{benred8}{6}~And now I cry to thee, the God who ever hearest me; turn thy ear towards me, and listen to my plea. \textcolor{benred8}{7}~Thy mercy, thy signal mercy shew; none ever sought sanctuary at thy right hand in vain. \textcolor{benred8}{8}~Protect me as thou wouldst the apple of thy own eye; hide me under the shelter of thy wings, \textcolor{benred8}{9}~safe from the evil-doers who wrong me.See how my enemies close about me mercilessly, \textcolor{benred8}{10}~their hearts shut to pity, a boast on their lips! \textcolor{benred8}{11}~Even now their stealthy tread closes in on me, as they watch their opportunity to bring me down; \textcolor{benred8}{12}~treacherous as a lion that is eager for its prey, a young lion that waits hidden in its lair.
\textcolor{benred8}{13}~Bestir thyself, Lord; forestall him and throw him to the ground; bare thy sword, and save me from the evil-doer, \textcolor{benred8}{14}~raise thy hand, to rescue me from the hands of mortal men! Mortal men indeed, that have all their portion here on earth; whose desires thou dost satisfy with treasures from thy store, so that their children, too, live in abundance, and leave riches for new heirs to enjoy. \textcolor{benred8}{15}~As for me, I will come with upright heart into thy presence, and when I wake up, I shall be well content with thy likeness.
\begin{large}\begin{center}\textsc{Chapter xvii}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. Of David, the servant of the Lord. He addressed to the Lord the words of this song, on the day when God delivered him from the hand of Saul, and from the hands of all his enemies; as follows:\footnote[1]{The whole of this psalm has been preserved for us independently among the records of King David\textquotesingle s reign, with slight variations (II Kg. 22).})
\textcolor{benred8}{2}~Shall I not love thee, Lord, my only defender? \textcolor{benred8}{3}~The Lord is my rock-fastness, my stronghold, my rescuer; to God, my hiding-place, I flee for safety; he is my shield, my weapon of deliverance, my refuge. \textcolor{benred8}{4}~Praised be the Lord! When I invoke his name, I am secure from my enemies. \textcolor{benred8}{5}~All about me surged the waves of death, deep flowed the perilous tide, to daunt me; \textcolor{benred8}{6}~the grave had caught me in its toils, deadly snares had trapped my feet. \textcolor{benred8}{7}~One cry to the Lord, in my affliction, one word of summons to my God, and he, from his sanctuary, listened to my voice; the complaint I made before him found a hearing.
\textcolor{benred8}{8}~Earth thereupon shivered and shook, the very foundations of the hills quailed and quaked at his anger; \textcolor{benred8}{9}~at the fiery smoke that breathed from his mouth and nostrils, kindling coals to flame. \textcolor{benred8}{10}~He bade heaven stoop, and came down to earth, with a dark cloud at his feet; \textcolor{benred8}{11}~he came, cherub-mounted, borne up on the wings of the wind, \textcolor{benred8}{12}~shrouded in darkness, canopied with black rain-storm and deep mist. \textcolor{benred8}{13}~Then, while coals were kindled at the brightness as he came, \textcolor{benred8}{14}~the Lord sent his thunder from heaven, the most High let his voice be heard. \textcolor{benred8}{15}~How they scattered when he rained down his arrows on them, how they fled in confusion before the volleys of his lightning! \textcolor{benred8}{16}~The sea\textquotesingle s bed came to light, the very foundations of the world were laid bare, when thou didst threaten them, Lord, when thou didst blow upon them with the breath of thy anger. \textcolor{benred8}{17}~Then he reached down from heaven, caught hold of me, rescued me from that flood, \textcolor{benred8}{18}~saved me from triumphant malice, from the enemies that held me at their mercy. \textcolor{benred8}{19}~Evil days, when they faced me at every turn! Yet the Lord stood by me, \textcolor{benred8}{20}~and brought me out into freedom again; his great love befriended me.
\textcolor{benred8}{21}~So, for my faithfulness, the Lord would requite me, as he sees me guiltless in act, he would make return. \textcolor{benred8}{22}~Have I not kept true to the Lord\textquotesingle s paths? Have I not been ever loyal to my God? \textcolor{benred8}{23}~No law of his, but I have kept it before my eyes; no task he laid upon me have I refused; \textcolor{benred8}{24}~ever stainless in his presence, ever watchful to keep myself clear of guilt, \textcolor{benred8}{25}~ever faithful, ever guiltless in act, the Lord has requited me. \textcolor{benred8}{26}~Lovingly wilt thou treat those who love thee, with the loyal keep troth; \textcolor{benred8}{27}~pure of heart the pure of heart shall find thee, the cunning thou wilt overreach. \textcolor{benred8}{28}~To humble folk thou wilt bring deliverance; the proud, with their haughty looks, thou wilt bring down to earth. \textcolor{benred8}{29}~It is thou, Lord, that keepest the lamp of my hopes still burning; shinest on the darkness about me, O my God. \textcolor{benred8}{30}~In thy strength I will engage a host of the enemy, in my God\textquotesingle s strength I shall leap over all their defences.
\textcolor{benred8}{31}~Such is my God, unsullied in his dealings; his promises are like metal tested in the fire; he is the sure defence of all who trust in him. \textcolor{benred8}{32}~Who but the Lord is God? What other refuge can there be, except our God? \textcolor{benred8}{33}~It is he that girds me with strength, bids me go on my way untroubled. \textcolor{benred8}{34}~He makes me sure-footed as the deer, and gives me the freedom of the hills; \textcolor{benred8}{35}~these hands, through him, are skilled in battle, these arms are a match for any bow of bronze. \textcolor{benred8}{36}~Thy saving power, Lord, is my defence, thy right hand supports me; thy tender care fosters me. \textcolor{benred8}{37}~Through thee, my steps are untrammelled as I go, my tread never falters; \textcolor{benred8}{38}~I can overtake the enemies I pursue, and never turn home till I have made an end of them; \textcolor{benred8}{39}~I can beat them to their knees, and hurl them down at my feet. \textcolor{benred8}{40}~Thou girdest me about with a warrior\textquotesingle s strength; whatever power challenges me, thou dost subdue before me, \textcolor{benred8}{41}~putting my enemies to flight, and throwing all their malice into confusion. \textcolor{benred8}{42}~Loudly they cry out to the Lord, bereft of aid, but he makes no answer to their cries. \textcolor{benred8}{43}~I can crush them to pieces, like the dust which the wind blows along; I can trample them down like mire from the streets.
\textcolor{benred8}{44}~Nor was it enough, of domestic broils to rid me; a world should be my vassal; \textcolor{benred8}{45}~new realms should pay me homage, quick to do my bidding. \textcolor{benred8}{46}~See where they come, the alien born, come slinking out of their strongholds, pale of cheek, to cringe before me. \textcolor{benred8}{47}~Blessed be the living Lord who is my refuge, praised be the God who delivers me! \textcolor{benred8}{48}~It is thou, my God, that bringest me redress, that bendest peoples to my will, that savest me from the spite of my enemies: \textcolor{benred8}{49}~so that I am high above the reach of their assaults, proof against their violence. \textcolor{benred8}{50}~Then, Lord, I will give thee thanks in the hearing of all the nations, singing in praise of thy name; \textcolor{benred8}{51}~victory thy mercy grants to the King thou hast anointed, to David, and David\textquotesingle s line for ever.
\begin{large}\begin{center}\textsc{Chapter xviii}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. A psalm. Of David.)
\textcolor{benred8}{2}~See how the skies proclaim God\textquotesingle s glory, how the vault of heaven betrays his craftsmanship! \textcolor{benred8}{3}~Each day echoes its secret to the next, each night passes on to the next its revelation of knowledge; \textcolor{benred8}{4}~no word, no accent of theirs that does not make itself heard,\footnote[1]{The meaning of the Hebrew text here is quite uncertain.} \textcolor{benred8}{5}~till their utterance fills every land, till their message reaches the ends of the world. \textcolor{benred8}{6}~In these, he has made a pavilion for the sun, which comes out as a bridegroom comes from his bed, and exults like some great runner who sees the track before him. \textcolor{benred8}{7}~Here, at one end of heaven, is its starting-place, and its course reaches to the other; none can escape its burning heat.
\textcolor{benred8}{8}~The Lord\textquotesingle s perfect law, how it brings the soul back to life; the Lord\textquotesingle s unchallengeable decrees, how they make the simple learned! \textcolor{benred8}{9}~How plain are the duties which the Lord enjoins, the treasure of man\textquotesingle s heart; how clear is the commandment the Lord gives, the enlightenment of man\textquotesingle s eyes! \textcolor{benred8}{10}~How sacred a thing is the fear of the Lord, which is binding for ever; how unerring are the awards which the Lord makes, one and all giving proof of their justice! \textcolor{benred8}{11}~All these are more precious than gold, than a hoard of pure gold, sweeter than the honey, dripping from its comb. \textcolor{benred8}{12}~By these I, thy servant, live, observing them how jealously! \textcolor{benred8}{13}~And yet, who knows his own frailties? If I have sinned unwittingly, do thou absolve me. \textcolor{benred8}{14}~Keep me ever thy own servant, far from pride; so long as this does not lord it over me, I will yet be without fault, I will yet be innocent of the great sin. \textcolor{benred8}{15}~Every word on my lips, every thought in my heart, what thou wouldst have it be, O Lord, my defender, my redeemer!
\begin{large}\begin{center}\textsc{Chapter xix}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. A psalm. Of David.)
\textcolor{benred8}{2}~The Lord listen to thee in thy time of need, the power of Jacob\textquotesingle s God be thy protection! \textcolor{benred8}{3}~May he send thee aid from his holy place, watch over thee, there on mount Sion; \textcolor{benred8}{4}~may he remember all thy offerings, and find savour in thy burnt-sacrifice. \textcolor{benred8}{5}~May he grant thee what thy heart desires, crown thy hopes with fulfilment. \textcolor{benred8}{6}~So may we rejoice at thy deliverance, rallied in the name of the Lord our God; \textcolor{benred8}{7}~abundantly may he grant thy prayer. Shall I doubt that the Lord protects the king he has anointed, will listen to him from his sanctuary in heaven? Is not his right hand strong to save? \textcolor{benred8}{8}~Let others talk of horses and chariots; our refuge is the name of the Lord our God. \textcolor{benred8}{9}~Stumbled and fallen they, while we stand firm on our feet. \textcolor{benred8}{10}~O Lord, save the king,\footnote[1]{Some would understand the Hebrew text as meaning, \textasciigrave O Lord, our King, save us\textquotesingle .} and hear us in the hour when we call upon thee.
\begin{large}\begin{center}\textsc{Chapter xx}\end{center}\end{large}
(To the choir-master. A psalm. Of David.)
\textcolor{benred8}{2}~Well may the king rejoice, Lord, in thy protection, well may he triumph in thy saving power! \textcolor{benred8}{3}~Never a wish in his heart hast thou disappointed, never a prayer on his lips denied. \textcolor{benred8}{4}~With happy auguries thou dost meet him on his way, dost set a crown of pure gold his head. \textcolor{benred8}{5}~Prays he for life? Long continuance of his reign thou dost grant him; to last unfailing till the end of time. \textcolor{benred8}{6}~Great is the renown thy protection has won for him; glory and high honour thou hast made his. \textcolor{benred8}{7}~An everlasting monument of thy goodness, comforted by the smile of thy favour, \textcolor{benred8}{8}~he stands firm, trusting in the Lord; the favour of the most High is with him.
\textcolor{benred8}{9}~Ay, but thy enemies\textemdash they shall feel thy power; that right hand will not leave their malice unpunished. \textcolor{benred8}{10}~At thy frown, they will wither away like grass in the oven; whirled away by the Lord\textquotesingle s anger, burnt up in its flames. \textcolor{benred8}{11}~Thou wilt rid the land of their breed, their race will vanish from the world of men. \textcolor{benred8}{12}~See how all their false designs against thee, all their plots come to nothing! \textcolor{benred8}{13}~Thou wilt rout them; bent is thy bow to meet their onslaught.\footnote[1]{Verses 9-13 are perhaps addressed to God, more probably to the psalmist himself.}
\textcolor{benred8}{14}~Stand high above us, Lord, in thy protecting strength; our song, our psalm, shall be of thy greatness.
\begin{large}\begin{center}\textsc{Chapter xxi}\end{center}\end{large}
(To the choir-master. Melody, The hind at dawn. A psalm. Of David.)
\textcolor{benred8}{2}~My God, my God, why hast thou forsaken me? Loudly I call, but my prayer cannot reach thee. \textcolor{benred8}{3}~Thou dost not answer, my God, when I cry out to thee day and night, thou dost not heed.\footnote[1]{The Hebrew text here has the difficult phrase \textasciigrave and (there is) no silence for me\textquotesingle .} \textcolor{benred8}{4}~Thou art there none the less, dwelling in the holy place; Israel\textquotesingle s ancient boast. \textcolor{benred8}{5}~It was in thee that our fathers trusted, and thou didst reward their trust by delivering them; \textcolor{benred8}{6}~they cried to thee, and rescue came; no need to be ashamed of such trust as theirs. \textcolor{benred8}{7}~But I, poor worm, have no manhood left; I am a by-word to all, the laughing-stock of the rabble. \textcolor{benred8}{8}~All those who catch sight of me fall to mocking; mouthing out insults, while they toss their heads in scorn, \textcolor{benred8}{9}~He committed himself to the Lord, why does not the Lord come to his rescue, and set his favourite free?
\textcolor{benred8}{10}~What hand but thine drew me out from my mother\textquotesingle s womb? Who else was my refuge when I hung at the breast? \textcolor{benred8}{11}~From the hour of my birth, thou art my guardian; since I left my mother\textquotesingle s womb, thou art my God! \textcolor{benred8}{12}~Do not leave me now, when trouble is close at hand; stand near, when I have none to help me. \textcolor{benred8}{13}~My enemies ring me round, packed close as a herd of oxen, strong as bulls from Basan; \textcolor{benred8}{14}~so might a lion threaten me with its jaws, roaring for its prey. \textcolor{benred8}{15}~I am spent as spilt water, all my bones out of joint, my heart turned to molten wax within me; \textcolor{benred8}{16}~parched is my throat, like clay in the baking, and my tongue sticks fast in my mouth; thou hast laid me in the dust, to die. \textcolor{benred8}{17}~Prowling about me like a pack of dogs, their wicked conspiracy hedges me in; they have torn holes in my hands and feet; \textcolor{benred8}{18}~I can count my bones one by one; and they stand there watching me, gazing at me in triumph. \textcolor{benred8}{19}~They divide my spoils among them, cast lots for my garments. \textcolor{benred8}{20}~Then, Lord, do not stand at a distance; if thou wouldst aid me, come speedily to my side. \textcolor{benred8}{21}~Only life is left me; save that from the sword, from the power of these dogs; \textcolor{benred8}{22}~rescue me from the very mouth of the lion, the very horns of the wild oxen that have brought me thus low.
\textcolor{benred8}{23}~Then I will proclaim thy renown to my brethren; where thy people gather, I will join in singing thy praise, \textcolor{benred8}{24}~Praise the Lord, all you that are his worshippers; honour to him from the sons of Jacob, reverence to him from Israel\textquotesingle s race! \textcolor{benred8}{25}~He has not scorned or slighted the appeal of the friendless, nor turned his face away from me; my cry for help did not go unheeded. \textcolor{benred8}{26}~Take what I owe thee, my song of praise before a great assembly. I will pay my vows to the Lord in the sight of his worshippers; \textcolor{benred8}{27}~the poor shall eat now, and have their fill, those who look for the Lord will cry out in praise of him, Refreshed be your hearts eternally! \textcolor{benred8}{28}~The furthest dwellers on earth will bethink themselves of the Lord, and come back to him; all the races of the heathen will worship before him; \textcolor{benred8}{29}~to the Lord royalty belongs, the whole world\textquotesingle s homage is his due. \textcolor{benred8}{30}~Him shall they worship, him only, that are laid to rest in the earth, even from their dust they shall adore. \textcolor{benred8}{31}~I, too, shall live on in his presence,\footnote[2]{\textasciigrave I, too, shall live on in his presence\textquotesingle ; according to the Hebrew text, this should be \textasciigrave and their souls have no life\textquotesingle , belonging to verse 30 as part of the description that verse gives of the unfortunate.} and beget children to serve him; these to a later age shall speak of the Lord\textquotesingle s name; \textcolor{benred8}{32}~these to a race that must yet be born shall tell the story of his faithfulness, Hear what the Lord did.
\begin{large}\begin{center}\textsc{Chapter xxii}\end{center}\end{large}
(A psalm. Of David.)
\textcolor{benred8}{1}~The Lord is my shepherd; how can I lack anything? \textcolor{benred8}{2}~He gives me a resting-place where there is green pasture, leads me out to the cool water\textquotesingle s brink, refreshed and content. \textcolor{benred8}{3}~As in honour pledged, by sure paths he leads me; \textcolor{benred8}{4}~dark be the valley about my path, hurt I fear none while he is with me; thy rod, thy crook are my comfort. \textcolor{benred8}{5}~Envious my foes watch, while thou dost spread a banquet for me; richly thou dost anoint my head with oil, well filled my cup. \textcolor{benred8}{6}~All my life thy loving favour pursues me; through the long years the Lord\textquotesingle s house shall be my dwelling-place.
\begin{large}\begin{center}\textsc{Chapter xxiii}\end{center}\end{large}
(Of David. A psalm.)
\textcolor{benred8}{1}~The Lord owns earth, and all earth\textquotesingle s fulness, the round world, and all its inhabitants. \textcolor{benred8}{2}~Who else has built it out from the sea, poised it on the hidden streams?
\textcolor{benred8}{3}~Who dares climb the mountain of the Lord, and appear in his sanctuary? \textcolor{benred8}{4}~The guiltless in act, the pure in heart; one who never set his heart on lying tales, or swore treacherously to his neighbour. \textcolor{benred8}{5}~His to receive a blessing from the Lord, mercy from God, his sure defender; \textcolor{benred8}{6}~his the true breed that still looks, still longs for the presence of the God of Jacob.
\textcolor{benred8}{7}~Swing back, doors, higher yet; reach higher, immemorial gates, to let the King enter in triumph! \textcolor{benred8}{8}~Who is this great King? Who but the Lord, mighty and strong, the Lord mighty in battle? \textcolor{benred8}{9}~Swing back, doors, higher yet; reach higher, immemorial gates, to let the King enter in triumph! \textcolor{benred8}{10}~Who is this great King? It is the Lord of Armies that comes here on his way triumphant.
\begin{large}\begin{center}\textsc{Chapter xxiv}\end{center}\end{large}
(Of David.)
\textcolor{benred8}{1}~All my heart goes out to thee, O Lord my God. \textcolor{benred8}{2}~Belie not the trust I have in thee, let not my enemies boast of my downfall. \textcolor{benred8}{3}~Can any that trust in thee be disappointed, as they are disappointed who lightly break their troth? \textcolor{benred8}{4}~Direct my way, Lord, as thou wilt, teach me thy own paths. \textcolor{benred8}{5}~Ever let thy truth guide and teach me, O God my deliverer, my abiding hope. \textcolor{benred8}{6}~Forget not, Lord, thy pity, thy mercies of long ago. \textcolor{benred8}{7}~Give heed no more to the sins and frailties of my youth, but think mercifully of me, as thou, Lord, art ever gracious. \textcolor{benred8}{8}~How gracious is the Lord, how faithful, guiding our strayed feet back to the path! \textcolor{benred8}{9}~In his own laws he will train the humble, in his own paths the humble he will guide. \textcolor{benred8}{10}~Jealous be thy keeping of covenant and ordinance, and the Lord\textquotesingle s dealings will be ever gracious, ever faithful with thee. \textcolor{benred8}{11}~Kindly be thy judgement of my sin, for thy own honour\textquotesingle s sake, my grievous sin.
\textcolor{benred8}{12}~Let a man but fear the Lord, what path to choose he doubts no longer. \textcolor{benred8}{13}~Much joy he shall have of his lands and to his heirs leave them. \textcolor{benred8}{14}~No stranger the Lord is, no secret his covenant, to his true worshippers. \textcolor{benred8}{15}~On the Lord I fix my eyes continually, trusting him to save my feet from the snare. \textcolor{benred8}{16}~Pity me, Lord, as thou seest me friendless and forlorn. \textcolor{benred8}{17}~Quit my heart of its burden, deliver me from my distress. \textcolor{benred8}{18}~Restless and forlorn, I claim thy pity, to my sins be merciful. \textcolor{benred8}{19}~See how many are my foes, and how bitter is the grudge they bear me. \textcolor{benred8}{20}~Take my soul into thy keeping; come to my rescue, do not let me be disappointed of my trust in thee. \textcolor{benred8}{21}~Uprightness and purity be my shield, as I wait patiently, Lord, for thy help. \textcolor{benred8}{22}~When wilt thou deliver Israel, my God, from all his troubles?
\begin{large}\begin{center}\textsc{Chapter xxv}\end{center}\end{large}
(Of David.)
\textcolor{benred8}{1}~Lord, be thou my judge; have I not guided my steps clear of wrong? Have I trusted in the Lord, only to stumble on my path? \textcolor{benred8}{2}~Test me, Lord, put me to the proof; assay my inmost desires and thoughts. \textcolor{benred8}{3}~Ever I keep thy mercies in mind, ever thy faithfulness bears me company. \textcolor{benred8}{4}~I have not consorted with false men, or joined in plotting evil; \textcolor{benred8}{5}~I have shunned the company of the wicked, never sat at my ease with sinners. \textcolor{benred8}{6}~With the pure in heart I will wash my hands clean, and take my place among them at thy altar, \textcolor{benred8}{7}~there making thy praises known, telling the story of all thy wonderful deeds. \textcolor{benred8}{8}~How well, Lord, I love the house where thou dwellest, the shrine of thy glory! \textcolor{benred8}{9}~Lord, never count this soul for lost with the wicked, this life among the blood-thirsty; \textcolor{benred8}{10}~hands ever stained with guilt, palms ever itching for a bribe! \textcolor{benred8}{11}~Be it mine to guide my steps clear of wrong; deliver me in thy mercy. \textcolor{benred8}{12}~On sure ground my feet are set; where his people gather I will join in blessing the Lord\textquotesingle s name.
\begin{large}\begin{center}\textsc{Chapter xxvi}\end{center}\end{large}
(Of David.)
\textcolor{benred8}{1}~The Lord is my light and my deliverance; whom have I to fear? The Lord watches over my life; whom shall I hold in dread? \textcolor{benred8}{2}~Vainly the malicious close about me, as if they would tear me in pieces, vainly my enemies threaten me; all at once they stumble and fall. \textcolor{benred8}{3}~Though a whole host were arrayed against me, my heart would be undaunted; though an armed onset should threaten me, still I would not lose my confidence. \textcolor{benred8}{4}~One request I have ever made of the Lord, let me claim it still, to dwell in the Lord\textquotesingle s house my whole life long, resting content in the Lord\textquotesingle s goodness, gazing at his temple. \textcolor{benred8}{5}~In his royal tent he hides me, in the inmost recess of his royal tent, safe from peril. \textcolor{benred8}{6}~On a rock fastness he lifts me high up; my head rises high above the enemies that encompass me. I will make an offering of triumphant music in this tabernacle of his, singing and praising the Lord.
\textcolor{benred8}{7}~Listen to my voice, Lord, when I cry to thee; hear and spare. \textcolor{benred8}{8}~True to my heart\textquotesingle s promise, I have eyes only for thee; I long, Lord, for thy presence. \textcolor{benred8}{9}~Do not hide thy face, do not turn away from thy servant in anger, but give me still thy aid; do not forsake me, do not neglect me, O God, my defender. \textcolor{benred8}{10}~Father and mother may neglect me, but the Lord takes me into his care. \textcolor{benred8}{11}~Lord, shew me the way thou hast chosen for me, guide me along the sure path, beset as I am with enemies; \textcolor{benred8}{12}~do not give me over to the will of my oppressors, when false witnesses stand up to accuse me, breathe out threats against me. \textcolor{benred8}{13}~My faith is, I will yet live to see the Lord\textquotesingle s mercies. \textcolor{benred8}{14}~Wait patiently for the Lord to help thee; be brave, and let thy heart take comfort; wait patiently for the Lord.
\begin{large}\begin{center}\textsc{Chapter xxvii}\end{center}\end{large}
(Of David.)
\textcolor{benred8}{1}~To thee, my Lord, my refuge, I cry aloud, do not leave my cry unanswered; listen to me, or I am no better than a dead man, sinking to the grave. \textcolor{benred8}{2}~Listen Lord, to my plea as I call upon thee, as I raise my hands in prayer towards thy holy temple. \textcolor{benred8}{3}~Do not summon me, with the wicked, before thy judgement-seat; with men who traffic in iniquity, men who talk of peace to their neighbours, while their hearts are full of malice. \textcolor{benred8}{4}~For them, the reward of their own acts, their own evil ways; as they did, be it done to them, in their own coin repaid. \textcolor{benred8}{5}~Of the Lord\textquotesingle s acts, the Lord\textquotesingle s ways, they took no heed; ruin be theirs, ruin irreparable.
\textcolor{benred8}{6}~Blessed be the Lord\textquotesingle s name, my plea is heard; \textcolor{benred8}{7}~the Lord is my strength and shield. Trusting in him, I found redress; there is triumph in my heart, on my lips the song of praise. \textcolor{benred8}{8}~The Lord defends his own people, protects the king he has anointed. \textcolor{benred8}{9}~Lord, save thy people, bless thy own chosen race; be their shepherd, evermore in thy arms upholding them.
\begin{large}\begin{center}\textsc{Chapter xxviii}\end{center}\end{large}
(A psalm. Of David.)
\textcolor{benred8}{1}~Sons of God, make your offering to the Lord; \textcolor{benred8}{2}~an offering to the Lord of honour and glory, an offering to the Lord of the glory that befits his name; worship the Lord, in holy vesture habited. \textcolor{benred8}{3}~The voice of the Lord is heard over the waters, when the glorious God thunders, the Lord, thundering over swollen waters; \textcolor{benred8}{4}~the Lord\textquotesingle s voice in its power, the Lord\textquotesingle s voice in its majesty. \textcolor{benred8}{5}~The Lord\textquotesingle s voice, that breaks the cedars; the Lord breaks the cedars of Lebanon; \textcolor{benred8}{6}~bids Lebanon and Sarion leap high as a bullock leaps, breed of the wild ox. \textcolor{benred8}{7}~The Lord\textquotesingle s voice kindles flashing fire; \textcolor{benred8}{8}~the Lord\textquotesingle s voice makes the wilderness rock; the Lord, rocking the wilderness of Cades. \textcolor{benred8}{9}~The Lord\textquotesingle s voice sets the oak-trees a-swaying, strips the deep forest bare. Meanwhile, in his sanctuary, there is no sound but tells of his glory. \textcolor{benred8}{10}~Out of a raging flood, the Lord makes a dwelling-place; the Lord sits enthroned as a king for ever. \textcolor{benred8}{11}~And this Lord will give strength to his people; the Lord will give his people his own blessing of peace.
\begin{large}\begin{center}\textsc{Chapter xxix}\end{center}\end{large}
\textcolor{benred8}{1}~(A psalm. A hymn for the feast of the temple\textquotesingle s dedication. Of David.)
\textcolor{benred8}{2}~Praise to thee, Lord, thou hast taken me under thy protection, and baulked my enemies of their will; \textcolor{benred8}{3}~I cried out to the Lord my God, and thou didst grant me recovery. \textcolor{benred8}{4}~So didst thou bring me back, Lord, from the place of shadows, rescue me from the very edge of the grave. \textcolor{benred8}{5}~Sing praise to the Lord, then, faithful souls, invoke his name with thankfulness. \textcolor{benred8}{6}~For a moment lasts his anger, for a life-time his love; sorrow is but the guest of a night, and joy comes in the morning.
\textcolor{benred8}{7}~I, too, had thought, in time of ease, Nothing can shake me now; \textcolor{benred8}{8}~such power and state, Lord, had thy mercy granted me. Then thou didst turn thy face away from me, and I was at peace no more. \textcolor{benred8}{9}~Lord, I was fain to plead with thee, cry upon God for pity: \textcolor{benred8}{10}~How will it profit thee to take my life? I can but go down into the grave; and will this dust give thanks to thee, or acknowledge, there, thy faithfulness? \textcolor{benred8}{11}~Listen, Lord, and spare; Lord, let thy aid befriend me. \textcolor{benred8}{12}~With that, thou didst turn my sadness into rejoicing, thou hast undone the sackcloth I wore, and girded me about with gladness. \textcolor{benred8}{13}~So may this heart never tire of singing praises; O Lord my God, I will give thanks to thee for ever.
\begin{large}\begin{center}\textsc{Chapter xxx}\end{center}\end{large}
(To the choir-master. A psalm. Of David.)
\textcolor{benred8}{2}~To thee, O Lord, I look for refuge, never let me be ashamed of my trust; in thy faithful care, deliver me. \textcolor{benred8}{3}~Grant me audience, and make haste to rescue me; my hill-fastness, my stronghold of defence, to save me from peril. \textcolor{benred8}{4}~Thou dost strengthen and defend me; thou, for thy own honour, dost guide and escort me; \textcolor{benred8}{5}~by thee protected, I shall escape from the snare that lies hidden in my path. \textcolor{benred8}{6}~Into thy hands I commend my spirit; thou, God ever faithful, wilt claim me for thyself. \textcolor{benred8}{7}~Let fools provoke thee by the worship of false gods; for me, no refuge but the Lord. \textcolor{benred8}{8}~I will triumph and exult in thy mercy; it was thou didst pity my weakness, and save me when I was hard bestead; \textcolor{benred8}{9}~before the enemy\textquotesingle s toils could close around me, the open plain lay at my feet.
\textcolor{benred8}{10}~And now, Lord, have compassion on my distress; vexation has dimmed my eyes, frets me away, soul and body. \textcolor{benred8}{11}~My life is all grief, my years are but sighs; for very misery,\footnote[1]{In the Hebrew text \textasciigrave through my sin\textquotesingle .} my strength ebbs away, my frame is wasted. \textcolor{benred8}{12}~Openly my foes deride me; even to my neighbours I am a thing of utter scorn; my friends are adread, and the passer-by shuns my contact; \textcolor{benred8}{13}~I am lost to memory, like a dead man, discarded like a broken pitcher. \textcolor{benred8}{14}~On every side their busy whispering comes to my ears; peril all around, so powerful the conspiracy that threatens my life. \textcolor{benred8}{15}~And still, Lord, my trust in thee is not shaken; still I cry, Thou art my God, \textcolor{benred8}{16}~my fate is in thy hand; save me from the enemy\textquotesingle s power, save me from my pursuers! \textcolor{benred8}{17}~Smile on thy servant once more, and deliver me in thy mercy; \textcolor{benred8}{18}~Lord, do not let me plead in vain. Disappoint the wicked of their hopes, hurl them down thwarted into the abyss; \textcolor{benred8}{19}~let silence fall on those treacherous lips, that spoke maliciously of the innocent in the days of their pride and scorn!
\textcolor{benred8}{20}~What treasures of loving-kindness, Lord, dost thou store up for the men who fear thee, rewarding their confidence for all the world to see! \textcolor{benred8}{21}~Thy presence is a sanctuary, to hide them away from the world\textquotesingle s malice; thy tabernacle a refuge from its noisy debate. \textcolor{benred8}{22}~Blessed be the Lord; so wondrous is his mercy, so strong the wall of his protection. \textcolor{benred8}{23}~I thought, bewildered, that thy watchful care had lost sight of me; but I cried out to thee, and thou thereupon didst listen to my plea. \textcolor{benred8}{24}~Love the Lord well, you who worship him; the Lord keeps faith with his servants, and repays the actions of the proud above measure. \textcolor{benred8}{25}~Take heart, keep high your courage, all you that wait patiently for the Lord.
\begin{large}\begin{center}\textsc{Chapter xxxi}\end{center}\end{large}
(Of David. A maskil.)
\textcolor{benred8}{1}~Blessed are they who have their faults forgiven, their transgressions buried deep; \textcolor{benred8}{2}~blessed is the man who is not guilty in the Lord\textquotesingle s reckoning, the heart that hides no treason. \textcolor{benred8}{3}~While I kept my own secret, evermore I went sighing, so wasted my frame away, \textcolor{benred8}{4}~bowed down day and night by thy chastisement; still my strength ebbed, faint as in mid-summer heat. \textcolor{benred8}{5}~At last I made my transgression known to thee, and hid my sin no longer; Fault of mine, said I, I here confess to the Lord; and with that, thou didst remit the guilt of my sin. \textcolor{benred8}{6}~Let every devout soul, then, turn to thee in prayer when hard times befall; rise the floods never so high, they shall have no power to reach it. \textcolor{benred8}{7}~Thou art my hiding-place, when I am sore bestead; songs of triumph are all about me, and thou my deliverer.
\textcolor{benred8}{8}~Friend, let me counsel thee, trace for thee the path thy feet should tread; let my prudence watch over thee. \textcolor{benred8}{9}~Do not be like the horse and the mule, senseless creatures which will not come near thee unless their spirit is tamed by bit and bridle. \textcolor{benred8}{10}~Again and again the sinner must feel the lash; he who trusts in the Lord finds nothing but mercy all around him. \textcolor{benred8}{11}~Just souls, be glad, and rejoice in the Lord; true hearts, make your boast in him.
\begin{large}\begin{center}\textsc{Chapter xxxii}\end{center}\end{large}
\textcolor{benred8}{1}~Triumph, just souls, in the Lord; true hearts, it is yours to praise him. \textcolor{benred8}{2}~Give thanks to the Lord with the viol\textquotesingle s music, praise him with a harp of ten strings. \textcolor{benred8}{3}~For him let a new song be sung; give him of your best, sound the harp and cry out lustily. \textcolor{benred8}{4}~The Lord\textquotesingle s word is true, he is faithful in all his dealings; \textcolor{benred8}{5}~faithfulness he loves, and the just award, the whole earth overflows with the Lord\textquotesingle s goodness. \textcolor{benred8}{6}~It was the Lord\textquotesingle s word that made the heavens, the breath of his lips that peopled them; \textcolor{benred8}{7}~he it is who stores up the waters of the sea as in a cistern, treasures up all its waves. \textcolor{benred8}{8}~Let the whole earth hold the Lord in dread, let all the inhabitants of the world stand in awe of him; \textcolor{benred8}{9}~he spoke, and they were made, he gave his command, and their frame was fashioned. \textcolor{benred8}{10}~At the Lord\textquotesingle s bidding, a nation\textquotesingle s purposes come to nothing, a people\textquotesingle s designs are thwarted; \textcolor{benred8}{11}~his own designs stand firm for ever; generation after generation, his will does not swerve.
\textcolor{benred8}{12}~Blessed the nation that calls the Lord its own God, the people he has chosen out to be his! \textcolor{benred8}{13}~Looking down from heaven, he watches all mankind, \textcolor{benred8}{14}~his dwelling-place has the whole world in view; \textcolor{benred8}{15}~he has fashioned each man\textquotesingle s nature, and weighs the actions of each. \textcolor{benred8}{16}~There is no protection for kings in powerful armies, for warriors in abundant prowess; \textcolor{benred8}{17}~nor shall horses bring thee the mastery, brute strength that cannot save. \textcolor{benred8}{18}~It is the Lord, watching over those who fear him and trust in his mercy, \textcolor{benred8}{19}~that will protect their lives, will feed them in time of famine. \textcolor{benred8}{20}~Patiently we wait for the Lord\textquotesingle s help; he is our strength and our shield; \textcolor{benred8}{21}~in him our hearts will find contentment, in his holy name we trust. \textcolor{benred8}{22}~Lord, let thy mercy rest upon us, who put all our confidence in thee.
\begin{large}\begin{center}\textsc{Chapter xxxiii}\end{center}\end{large}
\textcolor{benred8}{1}~(Of David, when he feigned madness at the court of Abimelech, so that Abimelech sent him away, and he escaped.\footnote[1]{See I Kg. 21.10 and following.})
\textcolor{benred8}{2}~At all times I will bless the Lord; his praise shall be on my lips continually. \textcolor{benred8}{3}~Be all my boasting in the Lord; listen to me, humble souls, and rejoice. \textcolor{benred8}{4}~Come, sing the Lord\textquotesingle s praise with me, let us extol his name together. \textcolor{benred8}{5}~Did I not look to the Lord, and find a hearing; did he not deliver me from all my terrors? \textcolor{benred8}{6}~Ever look to him, and in him find happiness; here is no room for downcast looks. \textcolor{benred8}{7}~Friendless folk may still call upon the Lord and gain his ear, and be rescued from all their afflictions. \textcolor{benred8}{8}~Guardian of those who fear the Lord, his angel encamps at their side, and brings deliverance. \textcolor{benred8}{9}~How gracious the Lord is! Taste and prove it; blessed is the man that learns to trust in him. \textcolor{benred8}{10}~It is for you, his chosen servants, to fear the Lord; those who fear him never go wanting. \textcolor{benred8}{11}~Justly do the proud fall into hunger and want;\footnote[2]{For \textasciigrave the proud\textquotesingle , the Hebrew text has \textasciigrave the young lions\textquotesingle .} blessing they lack not that look to him.
\textcolor{benred8}{12}~Know, then, my children, what the fear of the Lord is; come and listen to my teaching. \textcolor{benred8}{13}~Long life, and prosperous days, who would have these for the asking? \textcolor{benred8}{14}~My counsel is, keep thy tongue clear of harm, and thy lips free from every treacherous word. \textcolor{benred8}{15}~Naught of evil cherish thou, but rather do good; let peace be all thy quest and aim. \textcolor{benred8}{16}~On the upright the Lord\textquotesingle s eye ever looks favourably; his ears are open to their pleading. \textcolor{benred8}{17}~Perilous is his frown for the wrong-doers; he will soon make their name vanish from the earth. \textcolor{benred8}{18}~Roused by the cry of the innocent, the Lord sets them free from all their afflictions. \textcolor{benred8}{19}~So near is he to patient hearts, so ready to defend the humbled spirit. \textcolor{benred8}{20}~Though a hundred trials beset the innocent, the Lord will bring him safely through them all. \textcolor{benred8}{21}~Under the Lord\textquotesingle s keeping, every bone of his is safe; not one of them shall suffer harm. \textcolor{benred8}{22}~Villainy hastes to its own undoing; the enemies of innocence will bear their punishment.
\textcolor{benred8}{23}~The Lord will claim his servant as his own; they go unreproved that put their trust in him.
\begin{large}\begin{center}\textsc{Chapter xxxiv}\end{center}\end{large}
(Of David.)
\textcolor{benred8}{1}~Lord, espouse my quarrel; disarm the enemies who rise in arms against me; \textcolor{benred8}{2}~grip target and shield, bestir thyself in my defence. \textcolor{benred8}{3}~With poised lance, bar the way against my pursuers; whisper in my heart, I am here to save thee. \textcolor{benred8}{4}~For my mortal enemies, shame and disappointment; for my ill-wishers one and all, ruin and remorse; \textcolor{benred8}{5}~chaff before the wind, with the angel of the Lord to scatter them, \textcolor{benred8}{6}~benighted and bemired on their way, with the angel of the Lord in pursuit! \textcolor{benred8}{7}~Wantonly they have spread their nets for me, wantonly they have dug the cruel snare; \textcolor{benred8}{8}~now let sudden doom overtake them, their own nets enmesh, their own snare entrap them! \textcolor{benred8}{9}~Mine to triumph in the Lord, to boast of the aid he brings me; \textcolor{benred8}{10}~this be the cry of my whole being, There is none like thee, Lord; who else rescues the afflicted from the hand of tyranny, the poor, the destitute, from his oppressors? \textcolor{benred8}{11}~See how perjured witnesses have come forward, to browbeat me over charges of which I know nothing; \textcolor{benred8}{12}~how they have repaid my kindness with cruelty, and left me friendless! \textcolor{benred8}{13}~Time was, when these were sick;\footnote[1]{\textasciigrave When these were in sick\textquotesingle , as if King David were recalling his earlier kindness to these enemies of his.} what did I then? Sackcloth was my wear; rigorously I kept fast, prayed from my heart\textquotesingle s depths. \textcolor{benred8}{14}~I went my way sadly, as one that mourns for brother or friend, bowed with grief, as one that bewails a mother\textquotesingle s loss. \textcolor{benred8}{15}~And now it was my turn to reel under fortune\textquotesingle s blows; what did they? \textcolor{benred8}{16}~Gleeful they met, and plotted to attack me unawares; tore at me without ceasing, baited and mocked me, gnashing their teeth in hatred. \textcolor{benred8}{17}~Lord, wilt thou look on unheeding still? Wilt thou leave my life at the mercy of these roaring lions? \textcolor{benred8}{18}~Let me live to praise and thank thee before the multitude that throngs thy courts. \textcolor{benred8}{19}~No more the leer of triumph in the eye of yonder treacherous foe, that wantonly assails me!
\textcolor{benred8}{20}~In all their whispering, never a word of peace; they will still be plotting against the land\textquotesingle s repose. \textcolor{benred8}{21}~See how they mop and mow at me, crying out, Joy, joy that we should have lived to see this! \textcolor{benred8}{22}~Thou too, Lord, hast seen it, do not pass it by in silence; Lord, do not abandon me. \textcolor{benred8}{23}~Bestir thyself and take my part, give me speedy redress, my Lord and my God. \textcolor{benred8}{24}~As thou art just, O Lord my God, give thy award for me, never let them triumph over me; \textcolor{benred8}{25}~never let them think, All goes well, and boast that they have made a prey of me. \textcolor{benred8}{26}~Disappoint them, fill them with confusion, the men who delight in my misfortune; cover them with shame and self-reproach, the enemies that triumph over me. \textcolor{benred8}{27}~Joy and gladness be theirs, who applaud my innocence; Praise to the Lord! be their cry, the Lord who defends his servant. \textcolor{benred8}{28}~And all day long, for thy just awarding, this tongue shall make known thy praise.
\begin{large}\begin{center}\textsc{Chapter xxxv}\end{center}\end{large}
(To the choir-master. Of David, the Lord\textquotesingle s servant.)
\textcolor{benred8}{2}~Deep in his heart the sinner hears the whispering of evil, and loses sight of the fear of God; \textcolor{benred8}{3}~flatters himself with the thought that his misdoings go undiscovered, earn no reproof. \textcolor{benred8}{4}~No word on his lips but is cruel and false; never a wise thought, a kindly deed. \textcolor{benred8}{5}~He lies awake plotting mischief, and lends himself to every evil course, never weary of wrong-doing.
\textcolor{benred8}{6}~Lord, thy mercy is high as heaven; thy faithfulness reaches to the clouds; \textcolor{benred8}{7}~thy justice stands firm as the everlasting hills,\footnote[1]{Literally \textasciigrave the hills of God\textquotesingle , a Hebrew form of speech for what is magnificent in nature.} the wisdom of thy decrees is deep as the abyss. Lord, thou dost give protection to man and beast, \textcolor{benred8}{8}~so rich is thy divine mercy; under the shelter of those wings the frail children of earth will find confidence. \textcolor{benred8}{9}~With thy rich store thou wilt nourish them, bid them drink deep at thy fountain of contentment. \textcolor{benred8}{10}~In thee is the source of all life; thy brightness breaks on our eyes like dawn. \textcolor{benred8}{11}~Still let thy mercy dwell with those who acknowledge thee, thy favour with upright hearts; \textcolor{benred8}{12}~do not suffer the proud to trample on me, the wicked to dispossess me. \textcolor{benred8}{13}~See what a fall awaits the wrong-doers, how they are cast down to earth, and can keep their feet no more!
\begin{large}\begin{center}\textsc{Chapter xxxvi}\end{center}\end{large}
(Of David.)
\textcolor{benred8}{1}~Art thou impatient, friend, when the wicked thrive; dost thou envy the lot of evil-doers? \textcolor{benred8}{2}~they will soon fade like the grass, like the green leaf wither away. \textcolor{benred8}{3}~Be content to trust in the Lord and do good; live on thy land, and take thy ease, \textcolor{benred8}{4}~all thy longing fixed in the Lord; so he will give thee what thy heart desires. \textcolor{benred8}{5}~Commit thy life to the Lord, and trust in him; he will prosper thee, \textcolor{benred8}{6}~making thy honesty clear as the day, the justice of thy cause bright as the sun at noon. \textcolor{benred8}{7}~Dumb and patient, to the Lord\textquotesingle s mercy look thou, never fretting over the man that has his own way, and thrives by villainy. \textcolor{benred8}{8}~End thy complaints, forgo displeasure, do not fret thyself into an evil mood; \textcolor{benred8}{9}~the evil-minded will be dispossessed, and patient souls, that wait for the Lord, succeed them. \textcolor{benred8}{10}~Forbear yet a little, and the sinner will be seen no more; thou wilt search in vain to find him, \textcolor{benred8}{11}~while patient souls are the land\textquotesingle s heirs, enjoying great peace.
\textcolor{benred8}{12}~Gnashing his teeth with envy, the wrong-doer plots against the innocent, \textcolor{benred8}{13}~and cannot see his own turn coming; but the Lord sees it, and laughs at his malice. \textcolor{benred8}{14}~How they draw the sword, how they bend the bow, these sinners, to bring ruin on helpless poverty, to murder the upright; \textcolor{benred8}{15}~swords that will pierce their own hearts, bows that will break in pieces! \textcolor{benred8}{16}~Innocence, ill endowed, has the better of the wicked in their abundance; \textcolor{benred8}{17}~soon fails the strength of their arms, and still the Lord has the just in his keeping. \textcolor{benred8}{18}~Jealously the Lord watches over the lives of the guiltless, they will hold their lands for ever, \textcolor{benred8}{19}~undismayed by adversity, in time of famine well content. \textcolor{benred8}{20}~Knavery will yet come to an end; like the spring\textquotesingle s finery they will die, the Lord\textquotesingle s enemies, vanish away like smoke. \textcolor{benred8}{21}~Let the sinner borrow, and never repay, still the good man will be a generous giver;\footnote[1]{It is difficult to see the connexion between this verse and the next, unless we suppose that it indicates a reversal of fortunes, in which the wicked will always be defaulting borrowers, and good men will be in a position to help them.} \textcolor{benred8}{22}~win the Lord\textquotesingle s blessing, and the land is thine, his ban is death. \textcolor{benred8}{23}~Man\textquotesingle s feet stand firm, if the Lord is with him to prosper his journey; \textcolor{benred8}{24}~he may stumble but never fall, with the Lord\textquotesingle s hand in his. \textcolor{benred8}{25}~Now youth is past, and I have grown old; yet never did I see the good man forsaken, or his children begging their bread; \textcolor{benred8}{26}~still he lends without stint, and men call down blessings on his posterity. \textcolor{benred8}{27}~Offend no more, rather do good, and be at rest continually; \textcolor{benred8}{28}~the Lord is ever just, and will not abandon his faithful servants. Perish the sinner, forgotten be the name of the evil-doer, \textcolor{benred8}{29}~but these will hold their land, and live on it always at rest.
\textcolor{benred8}{30}~Right reason is on the good man\textquotesingle s lips, well weighed are all his counsels; \textcolor{benred8}{31}~his steps never falter, because the law of God rules in his heart. \textcolor{benred8}{32}~Sinners lie in wait, plotting against the life of the innocent; \textcolor{benred8}{33}~but the Lord will never leave him in their power, never find him guilty when he is arraigned. \textcolor{benred8}{34}~Trust the Lord, and follow the path he has chosen; so he will set thee up in possession of thy land, and thou wilt live to see the wicked come to ruin. \textcolor{benred8}{35}~Until yesterday, I saw the evil-doer throned high as the branching cedars; \textcolor{benred8}{36}~then, when I passed by, he was there no longer, and I looked in vain to find him. \textcolor{benred8}{37}~Virtuous men and innocent mark thou well; he that lives peaceably will leave a race behind him, \textcolor{benred8}{38}~while sinners are rooted out every one, and their graceless names forgotten. \textcolor{benred8}{39}~When affliction comes, the Lord is the refuge and defence of the innocent; \textcolor{benred8}{40}~the Lord will aid and deliver them, rescue and preserve them from the power of wickedness, because they put their trust in him.
\begin{large}\begin{center}\textsc{Chapter xxxvii}\end{center}\end{large}
\textcolor{benred8}{1}~(A psalm. Of David. For a memorial.)
\textcolor{benred8}{2}~Thy reproof, Lord, not thy vengeance; thy chastisement, not thy condemnation! \textcolor{benred8}{3}~Thy arrows pierce me, thy hand presses me hard; \textcolor{benred8}{4}~thy anger has driven away all health from my body, never a bone sound in it, so grievous are my sins. \textcolor{benred8}{5}~My own wrong-doing towers high above me, hangs on me like a heavy burden; \textcolor{benred8}{6}~my wounds fester and rankle, with my own folly to blame. \textcolor{benred8}{7}~Beaten down, bowed to the earth, I go mourning all day long, \textcolor{benred8}{8}~my whole frame afire, my whole body diseased; \textcolor{benred8}{9}~so spent, so crushed, I groan aloud in the weariness of my heart. \textcolor{benred8}{10}~Thou, Lord, knowest all my longings, no complaint of mine escapes thee; \textcolor{benred8}{11}~restless my heart, gone my strength; the very light that shone in my eyes is mine no longer.
\textcolor{benred8}{12}~Friends and neighbours that meet me keep their distance from a doomed man; old companions shun me. \textcolor{benred8}{13}~Ill-wishers that grudge me life itself lay snares about me, threaten me with ruin; relentlessly their malice plots against me. \textcolor{benred8}{14}~And I, all the while, am deaf to their threats, dumb before my accusers; \textcolor{benred8}{15}~mine the unheeding ear, and the tongue that utters no defence. \textcolor{benred8}{16}~On thee, Lord, my hopes are set; thou, O Lord my God, wilt listen to me. \textcolor{benred8}{17}~Such is the prayer I make, Do not let my enemies triumph over me, boast of my downfall. \textcolor{benred8}{18}~Fall full well I may; misery clouds my view; \textcolor{benred8}{19}~I am ever ready to publish my guilt, ever anxious over my sin. \textcolor{benred8}{20}~Unprovoked, their malice still prevails; so many that bear me a grudge so wantonly, \textcolor{benred8}{21}~rewarding good with evil, and for the very rightness of my cause assailing me. \textcolor{benred8}{22}~Do not fail me, O Lord my God, do not forsake me; \textcolor{benred8}{23}~hasten to my defence, O Lord, my only refuge.
\begin{large}\begin{center}\textsc{Chapter xxxviii}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master, Idithun. A psalm. Of David.)
\textcolor{benred8}{2}~It was my resolve to live watchfully, and never use my tongue amiss; still, while I was in the presence of sinners, I kept my mouth gagged, \textcolor{benred8}{3}~dumb and patient, impotent for good. But indignation came back, \textcolor{benred8}{4}~and my heart burned within me, the fire kindled by my thoughts, \textcolor{benred8}{5}~so that at last I kept silence no longer.Lord, warn me of my end, and how few my days are; teach me to know my own insufficiency. \textcolor{benred8}{6}~See how thou hast measured my years with a brief span, how my life is nothing in thy reckoning! Nay, what is any man living but a breath that passes? \textcolor{benred8}{7}~Truly man walks the world like a shadow; with what vain anxiety he hoards up riches, when he cannot tell who will have the counting of them! \textcolor{benred8}{8}~What hope then is mine, Lord? In thee alone I trust. \textcolor{benred8}{9}~Clear me of that manifold guilt which makes me the laughing-stock of fools, \textcolor{benred8}{10}~tongue-tied and uncomplaining, because I know that my troubles come from thee; \textcolor{benred8}{11}~spare me this punishment; I faint under thy powerful hand. \textcolor{benred8}{12}~When thou dost chasten man to punish his sins, gone is all he loved, as if the moth had fretted it away; a breath that passes, and no more. \textcolor{benred8}{13}~Listen, Lord, to my prayer, let my cry reach thy hearing, and my tears win answer. What am I in thy sight but a passer-by, a wanderer, as all my fathers were? \textcolor{benred8}{14}~Thy frown relax, give me some breath of comfort, before I go away and am known no more.
\begin{large}\begin{center}\textsc{Chapter xxxix}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. Of David. A psalm.)
\textcolor{benred8}{2}~Patiently I waited for the Lord\textquotesingle s help, and at last he turned his look towards me; \textcolor{benred8}{3}~he listened to my plea, drew me up out of a deadly pit, where the mire had settled deep, and gave me a foothold on the rock, with firm ground to tread. \textcolor{benred8}{4}~He has framed a new music on my lips, a song of praise to our God, to fill all that stand by with reverence, and with trust in the Lord. \textcolor{benred8}{5}~Happy is the man whose trust is there bestowed, who shuns the rites of strange gods, the lure of lies. \textcolor{benred8}{6}~O Lord my God, how long is the story of thy marvellous deeds! Was ever care like thine? How should I tell the tale of those mercies, past all numbering? \textcolor{benred8}{7}~No sacrifice, no offering was thy demand; enough that thou hast given me an ear ready to listen.\footnote[1]{The Septuagint Greek has \textasciigrave thou hast prepared a body for me\textquotesingle ; cf. Heb. 10.5.} Thou hast not found any pleasure in burnt-sacrifices, in sacrifices for sin. \textcolor{benred8}{8}~See then, I said, I am coming to fulfil what is written of me, where the book lies unrolled; \textcolor{benred8}{9}~to do thy will, O my God, is all my desire, to carry out that law of thine which is written in my heart. \textcolor{benred8}{10}~And I told the story of thy just dealings before a great throng; be witness, Lord, that I do not seal my lips. \textcolor{benred8}{11}~Thy just dealings are no secret hidden away in my heart; I boast of thy faithful protection, proclaim that mercy, that faithfulness of thine for all to hear it.
\textcolor{benred8}{12}~Lord, do not withhold thy pity from me; let thy mercy and faithfulness be now, as ever, my shield. \textcolor{benred8}{13}~I am beset with evils past numbering, overtaken by my sins; they fill my prospect, countless as the hairs on my head; my courage fails me. \textcolor{benred8}{14}~Deign, Lord, to set me free; Lord, give heed and help. \textcolor{benred8}{15}~Disappointment and shame be theirs, who lay plots against my life; may they slink away covered with confusion, who now rejoice over my downfall. \textcolor{benred8}{16}~Joy, joy! is their cry; dumb-stricken let them stand, their hopes belied. \textcolor{benred8}{17}~Rejoicing and triumph for all the souls that look to thee; Praise to the Lord, will ever be their song, who now long for thy aid. \textcolor{benred8}{18}~I, so helpless, so destitute, and the Lord is concerned for me! Thou art my champion and my refuge; do not linger, my God, do not linger on the way.
\begin{large}\begin{center}\textsc{Chapter xl}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. A psalm. Of David.)
\textcolor{benred8}{2}~Blessed is that man who takes thought for the poor and the destitute; the Lord will keep him safe in time of trouble. \textcolor{benred8}{3}~The Lord will watch over him, and give him long life and happiness on earth, and baulk his enemies of their will. \textcolor{benred8}{4}~The Lord will sustain him when he lies bed-ridden, turn all to health in his sickness.
\textcolor{benred8}{5}~Lord have mercy on me, is my prayer; bring healing to a soul that has sinned against thee. \textcolor{benred8}{6}~Bitterly my enemies taunt me; How long, they ask, ere he will die, and his name be forgotten? \textcolor{benred8}{7}~When one comes to visit me, he comes with smooth words, his heart full of malice, ready to go out and plot against me. \textcolor{benred8}{8}~There they stand, my enemies, talking of me in whispers, devising hurt; \textcolor{benred8}{9}~Here is a foul plague loosed on him; he will leave his bed no more. \textcolor{benred8}{10}~Why, the very man I trusted most, my own intimate friend, who shared my bread, has lifted his heel to trip me up.\footnote[1]{Jn. 13.18.} \textcolor{benred8}{11}~Lord, have mercy on me; give me back health, and let me requite them! \textcolor{benred8}{12}~Proof of thy favour, my enemies are baulked of their triumph; \textcolor{benred8}{13}~thou dost befriend my innocence; nevermore wilt thou banish me from thy presence.
\textcolor{benred8}{14}~Blessed be the Lord God of Israel, from the beginning to the end of time, Amen, Amen.
\begin{large}\begin{center}\textsc{Chapter xli}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. A maskil. Of the sons of Core.)
\textcolor{benred8}{2}~O God, my whole soul longs for thee, as a deer for running water; \textcolor{benred8}{3}~my whole soul thirsts for God, the living God; shall I never again make my pilgrimage into God\textquotesingle s presence? \textcolor{benred8}{4}~Morning and evening, my diet still of tears! Daily I must listen to the taunt, Where is thy God now? \textcolor{benred8}{5}~Memories come back to me yet, melting the heart; how once I would join with the throng, leading the way to God\textquotesingle s house, amid cries of joy and thanksgiving, and all the bustle of holiday. \textcolor{benred8}{6}~Soul, art thou still downcast? Wilt thou never be at peace? Wait for God\textquotesingle s help; I will not cease to cry out in thankfulness, My champion and my God.
\textcolor{benred8}{7}~In my sad mood I will think of thee, here in this land of Jordan and Hermon, here on Misar mountain. \textcolor{benred8}{8}~One depth makes answer to another\footnote[1]{These \textasciigrave depths\textquotesingle  are ordinarily understood as bein waterfalls; but the second part of the verse would seem to be a metaphor derived from storms at sea. Cf. Jon. 2.4.} amid the roar of the floods thou sendest; wave after wave, crest after crest overwhelms me. \textcolor{benred8}{9}~Would he but lighten the day with his mercy, what praise would I sing at evening to the Lord God who is life for me! \textcolor{benred8}{10}~Thou art my stronghold, I cry out to him still; hast thou never a thought for me? Must I go mourning, with enemies pressing me hard; \textcolor{benred8}{11}~racked by the ceaseless taunts of my persecutors, Where is thy God now? \textcolor{benred8}{12}~Soul, art thou still downcast? Wilt thou never be at peace? Wait for God\textquotesingle s help; I will not cease to cry out in thankfulness, My champion and my God.
\begin{large}\begin{center}\textsc{Chapter xlii}\end{center}\end{large}
\lettrine[lines=2]{O}{} God, sustain my cause; give me redress against a race that knows no piety; save me from a treacherous foe and cruel. \textcolor{benred8}{2}~Thou, O God, art all my strength; why hast thou cast me off? Must I go mourning, with enemies pressing me hard? \textcolor{benred8}{3}~The light of thy presence, the fulfilment of thy promise, let these be my escort, bringing me safe to thy holy mountain, to the tabernacle where thou dwellest. \textcolor{benred8}{4}~There I will go up to the altar of God, the giver of triumphant happiness; thou art my own God, with the harp I hymn thy praise. \textcolor{benred8}{5}~Soul, art thou still downcast? Wilt thou never be at peace? Wait for God\textquotesingle s help; I will not cease to cry out in thankfulness, My champion and my God.\footnote[1]{The repetition here of the refrain used in Psalm 41 makes it probable that these two were originally a single psalm.}
\begin{large}\begin{center}\textsc{Chapter xliii}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. Of the sons of Core. A maskil.)
\textcolor{benred8}{2}~O God, the tale has come to our ears\textemdash have not our fathers told it?\textemdash of the great things thou didst in their time, in days long ago; \textcolor{benred8}{3}~it was thy power that gave them a home by rooting out the heathen, crushing and dispossessing nations to make room for them. \textcolor{benred8}{4}~It was not by their own sword that our fathers won the land, it was not their own strength that brought them victory; it was the work of thy hand, thy strength; thy smile shone upon them, in proof of thy favour. \textcolor{benred8}{5}~I too have no King, no God, save thee; who else sent deliverance to Jacob? \textcolor{benred8}{6}~Through thee we routed our enemies; under thy protection we crushed their onslaught. \textcolor{benred8}{7}~Not in my bow I trusted, not to my sword I looked for safety; \textcolor{benred8}{8}~thine it was to save us from our enemies, and cover their malice with confusion. \textcolor{benred8}{9}~In God was ever our boast; his name we praise unceasingly.
\textcolor{benred8}{10}~And now? Now, O God, thou hast disowned us, and put us to shame, by refusing to go into battle with our armies. \textcolor{benred8}{11}~Thou dost put us to flight before our enemies; our ill-wishers plunder us as they will. \textcolor{benred8}{12}~Thou hast made us like sheep sold for food, scattered here and there among the heathen; \textcolor{benred8}{13}~thou hast bartered away thy people without profit, asking no rich amends for thy loss. \textcolor{benred8}{14}~Thou hast turned us into a laughing-stock for our neighbours, mocked and derided by all who dwell around; \textcolor{benred8}{15}~till the heathen make a by-word of us, and Gentiles toss their heads at us in scorn. \textcolor{benred8}{16}~Ever my disgrace confronts me; my cheeks are covered with blushes, \textcolor{benred8}{17}~as I hear nothing but reproach and reviling, see none but enemies, none but persecutors.
\textcolor{benred8}{18}~All this has come upon us, and it was not that we had forgotten thee. We have not been untrue to thy covenant, \textcolor{benred8}{19}~or withdrawn our hearts from thee, that we should let our steps wander away from thy paths. \textcolor{benred8}{20}~And all the while thou wouldst bring us low, anguish on every side, darkness hanging over us. \textcolor{benred8}{21}~If we had forgotten the name of our own God, and spread out our hands in prayer to the gods of the alien, \textcolor{benred8}{22}~would not he know of it? He can read the secrets of men\textquotesingle s hearts. \textcolor{benred8}{23}~No, it is for thy sake that we face death at every moment, reckoned no better than sheep marked down for slaughter. \textcolor{benred8}{23}~Bestir thyself, Lord, why dost thou sleep on? Awake, do not banish us from thy presence for ever. \textcolor{benred8}{24}~How canst thou turn thy face away, without a thought for our need and our affliction? \textcolor{benred8}{25}~Our pride is bowed in the dust; prostrate, we cannot lift ourselves from the ground. \textcolor{benred8}{26}~Arise, Lord, and help us; in thy mercy, claim us for thy own.
\begin{large}\begin{center}\textsc{Chapter xliv}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. Melody: The Lilies. Of the sons of Core. A maskil. A love-song.)
\textcolor{benred8}{2}~Joyful the thoughts that well up from my heart, the King\textquotesingle s honour for my theme; my tongue flows readily as the pen of a swift writer. \textcolor{benred8}{3}~Thine is more than mortal beauty, thy lips overflow with gracious utterance; the blessings God has granted thee can never fail. \textcolor{benred8}{4}~Gird on thy sword at thy side, great warrior, \textcolor{benred8}{5}~gird thyself with all thy majesty and all thy beauty; ride on triumphant, in the name of faithfulness and justice. Dread counsel thy own might shall give thee; \textcolor{benred8}{6}~so sharp are thy arrows, subduing nations to thy will, daunting the hearts of the king\textquotesingle s enemies. \textcolor{benred8}{7}~Thy throne, O God, endures for ever and ever, the sceptre of thy royalty is a rod that rules true;\footnote[1]{\textasciigrave Thy throne, O God\textquotesingle ; various other ways of taking these words have been suggested, as, for example, \textasciigrave God is (the support of) thy throne\textquotesingle . None of these has met with general acceptance. This verse and the following are applied to our Lord in the Epistle to the Hebrews (1.8).} \textcolor{benred8}{8}~thou hast been a friend to right, an enemy to wrong, and God, thy own God, has given thee an unction to bring thee pride beyond any of thy fellows. \textcolor{benred8}{9}~Thy garments are scented with myrrh, and aloes, and cassia; from ivory palaces there are harps sounding in thy honour. \textcolor{benred8}{10}~Daughters of kings come out to meet thee; at thy right hand stands the queen, in Ophir gold arrayed. \textcolor{benred8}{11}~(Listen, my daughter, and consider my words attentively; thou art to forget, henceforward, thy own nation, and the house of thy father; \textcolor{benred8}{12}~thy beauty, now, is all for the king\textquotesingle s delight; he is thy Lord, and worship belongs to him.) \textcolor{benred8}{13}~The people of Tyre, too, will have its presents to bring; the noblest of its citizens will be courting thy favour. \textcolor{benred8}{14}~She comes, the princess, all fair to see, her robe of golden cloth, \textcolor{benred8}{15}~a robe of rich embroidery, to meet the King. The maidens of her court follow her into thy presence, \textcolor{benred8}{16}~all rejoicing, all triumphant, as they enter the king\textquotesingle s palace! \textcolor{benred8}{17}~Thou shalt have sons worthy of thy own fathers, and divide a world between them for their domains. \textcolor{benred8}{18}~While time lasts, mine it is to keep thy name in remembrance; age after age, nations will do thee honour.
\begin{large}\begin{center}\textsc{Chapter xlv}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. Of the sons of Core. Melody: The Virgins. A song.)
\textcolor{benred8}{2}~God is our refuge and stronghold; sovereign aid he has brought us in the hour of peril. \textcolor{benred8}{3}~Not for us to be afraid, though earth should tumble about us, and the hills be carried away into the depths of the sea. \textcolor{benred8}{4}~See how its waters rage and roar, how the hills tremble before its might! \textcolor{benred8}{5}~The Lord of hosts is with us, the God of Jacob is our refuge.But the city of God, enriched with flowing waters, is the chosen sanctuary of the most High,\footnote[1]{There may be a reference here to II Par. 32.4.} \textcolor{benred8}{6}~God dwells within her, and she stands unmoved; with break of dawn\footnote[2]{Literally, \textasciigrave very early in the morning\textquotesingle ; which sense may be intended if those scholars are right who connect this psalm with the defeat of Sennacherib (cf. IV Kg. 19.35).} he will grant her deliverance. \textcolor{benred8}{7}~Nations may be in turmoil, and thrones totter, earth shrink away before his voice; \textcolor{benred8}{8}~but the Lord of hosts is with us, the God of Jacob is our refuge. \textcolor{benred8}{9}~Come near, and see God\textquotesingle s acts, his marvellous acts done on earth; \textcolor{benred8}{10}~how he puts an end to wars all over the world, the bow shivered, the lances shattered, the shields burnt to ashes! \textcolor{benred8}{11}~Wait quietly, and you shall have proof that I am God, claiming empire among the nations, claiming empire over the world. \textcolor{benred8}{12}~The Lord of hosts is with us, the God of Jacob is our refuge.
\begin{large}\begin{center}\textsc{Chapter xlvi}\end{center}\end{large}
(To the choir-master. Of the sons of Core. A psalm.)
\textcolor{benred8}{2}~Clap your hands, all you nations, in applause; acclaim your God with cries of rejoicing. \textcolor{benred8}{3}~The Lord is high above us, and worthy of dread; he is the sovereign Ruler of all the earth; \textcolor{benred8}{4}~he has tamed the nations to our will, bowed the Gentiles at our feet, \textcolor{benred8}{5}~claimed us for his own portion, Jacob the fair, the well beloved. \textcolor{benred8}{6}~God goes up, loud are the cries of victory; the Lord goes up, loudly the trumpets peal. \textcolor{benred8}{7}~A psalm, a psalm for our God, a psalm, a psalm for our King! \textcolor{benred8}{8}~God is King of all the earth; sound the hymn of praise! \textcolor{benred8}{9}~God reigns over the heathen, God sits enthroned in holiness. \textcolor{benred8}{10}~The rulers of the nations throw in their lot with us, that worship Abraham\textquotesingle s God; a God so high, he has all earth\textquotesingle s princes for his vassals.
\begin{large}\begin{center}\textsc{Chapter xlvii}\end{center}\end{large}
(A song. A psalm. Of the sons of Core.)
\textcolor{benred8}{2}~The Lord is great, great honour is his due, here in the city where he, our God, dwells. Fair rises the peak of his holy mountain, \textcolor{benred8}{3}~the pride of the whole world, and the true pole of earth, mount Sion, the city of the great King; \textcolor{benred8}{4}~within those walls, God has proved himself a sure defence. \textcolor{benred8}{5}~See, how the kings of the earth have made common cause, and met there in arms! \textcolor{benred8}{6}~At the sight of her all was bewilderment, and confusion, and dismay; \textcolor{benred8}{7}~fear took hold of them, sudden as the throes of a woman in travail; \textcolor{benred8}{8}~not more ruinously on ocean-going ships falls the east wind. \textcolor{benred8}{9}~Here, in this city of the Lord of hosts, the city of our own God, we have proved the tale long since told us, that God upholds her for all eternity; \textcolor{benred8}{10}~sheltered in thy temple, we give thanks for our deliverance. \textcolor{benred8}{11}~O God, wherever thy name is known on earth, thy praise is told, ever just in thy dealings; \textcolor{benred8}{12}~well may the hill of Sion rejoice, well may the townships of Juda triumph, at the decrees which thou, Lord, hast executed. \textcolor{benred8}{13}~Walk about Sion, make the round of her towers, and count the number of them; \textcolor{benred8}{14}~mark well the defences that are hers, pass all her strongholds in review; then give word to the next generation, \textcolor{benred8}{15}~Such is the God, who is our God for ever and ever; our Shepherd eternally.
\begin{large}\begin{center}\textsc{Chapter xlviii}\end{center}\end{large}
(To the choir-master. Of the sons of Core. A psalm.)
\textcolor{benred8}{2}~Listen, you nations far and wide; let all the world give hearing, \textcolor{benred8}{3}~poor clods of earth, and men nobly born, for rich and poor the same lesson. \textcolor{benred8}{4}~Here are wise words, thoughts of a discerning heart; \textcolor{benred8}{5}~mine to overhear mysteries, and reveal, with the harp\textquotesingle s music, things of deep import.
\textcolor{benred8}{6}~What need have I to be afraid in troubled times, when malice dogs my heels and overtakes me, \textcolor{benred8}{7}~malice of foes who trust in their own strength, and boast of their great possessions? \textcolor{benred8}{8}~No man can deliver himself from his human lot, paying a ransom-price to God; \textcolor{benred8}{9}~too great is the cost of a man\textquotesingle s soul; never will the means be his \textcolor{benred8}{10}~to prolong his days eternally and escape death. \textcolor{benred8}{11}~True it is, wise men die; but reckless fools perish no less; their riches will go to others, \textcolor{benred8}{12}~and the grave will be their everlasting home. Age after age, they will live on there, under the fields they once called their own. \textcolor{benred8}{13}~Short is man\textquotesingle s enjoyment of earthly goods; match him with the brute beasts, and he is no better than they.
\textcolor{benred8}{14}~Fatal path, that ensnares the reckless! Pitiful end of the men that love life! \textcolor{benred8}{15}~There they lie in the world beneath, huddled like sheep, with death for their shepherd, the just for their masters; soon, soon their image fades, the grave for its tenement. \textcolor{benred8}{16}~But my life God will rescue from the power of that lower darkness, a life that finds acceptance with him. \textcolor{benred8}{17}~Do not be disturbed, then, when a man grows rich, and there is no end to his household\textquotesingle s magnificence; \textcolor{benred8}{18}~he cannot take all that with him when he dies, magnificence will not follow him to the grave. \textcolor{benred8}{19}~While life lasts, he calls himself happy: None but will envy my success; \textcolor{benred8}{20}~but soon he will be made one with the line of his fathers, never again to see the light. \textcolor{benred8}{21}~Short is man\textquotesingle s careless enjoyment of earthly goods; match him with the brute beasts, and he is no better than they.
\begin{large}\begin{center}\textsc{Chapter xlix}\end{center}\end{large}
\lettrine[lines=2]{I}{t} is the Lord God that speaks; his message goes out to all the earth, from the sun\textquotesingle s rise to its setting. \textcolor{benred8}{2}~Out of Sion, in perfect beauty, \textcolor{benred8}{3}~God comes, revealed; he will keep silence no longer. Before him goes a raging fire; there is a whirling storm round about him. \textcolor{benred8}{4}~So, from on high, he summons heaven and earth to witness the judgement pronounced on his people: \textcolor{benred8}{5}~Muster in my presence my faithful servants, who honour my covenant still with sacrifice. \textcolor{benred8}{6}~The heavens themselves pronounce him just, God who is our judge.
\textcolor{benred8}{7}~Listen, my people, to these words of mine, listen, Israel, to the protestation I make thee; I, the God thou ownest as God. \textcolor{benred8}{8}~I do not find fault with thee over thy sacrifices; why, all day long thy burnt-offerings smoke before me. \textcolor{benred8}{9}~But the gifts I accept are not cattle from thy stock, or buck-goats from thy folds; \textcolor{benred8}{10}~I own already every wild beast in the forest, the hills are mine, and the herds that people them. \textcolor{benred8}{11}~There is no bird flies in heaven, no life stirs in the country-side, but I know of it. \textcolor{benred8}{12}~If I am hungry, I will not complain of it to thee, I, who am master of earth and all that earth contains. \textcolor{benred8}{13}~Wouldst thou have me eat bull\textquotesingle s flesh, and drink the blood of goats? \textcolor{benred8}{14}~The sacrifice thou must offer to God is a sacrifice of praise, so wilt thou perform thy vows to the most High. \textcolor{benred8}{15}~So, when thou criest to me in time of trouble, I will deliver thee; then thou shalt honour me as thou wilt.
\textcolor{benred8}{16}~But thus, to the sinner, God speaks: How is it that thou canst repeat my commandments by rote, and boast of my covenant with thee, \textcolor{benred8}{17}~and thou, all the while, hast no love for the amendment of thy ways, casting every warning of mine to the winds? \textcolor{benred8}{18}~Swift thou art to welcome the thief who crosses thy path, to throw in thy lot with the adulterers. \textcolor{benred8}{19}~Malice wells up from thy lips, and thy tongue is a ready engine of deceit; \textcolor{benred8}{20}~thou wilt sit there in conclave, speaking evil of thy brother, traducing thy own mother\textquotesingle s son. \textcolor{benred8}{21}~Such were thy ways, and should I make no sign? Should I let thee think I am such as thou? Here is thy reproof; here is thy indictment made plain to thee. \textcolor{benred8}{22}~Think well on this, you that forget God, or his hand will fall suddenly, and there will be no delivering you. \textcolor{benred8}{23}~He honours me truly, who offers me a sacrifice of praise; live aright, and you shall see the saving power of God.
\begin{large}\begin{center}\textsc{Chapter l}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. A psalm. Of David, \textcolor{benred8}{2}~when the prophet Nathan came to reproach him for his adultery with Bethsabee.)
\textcolor{benred8}{3}~Have mercy on me, O God, as thou art ever rich in mercy; in the abundance of thy compassion, blot out the record of my misdeeds. \textcolor{benred8}{4}~Wash me clean, cleaner yet, from my guilt, purge me of my sin, \textcolor{benred8}{5}~the guilt which I freely acknowledge, the sin which is never lost to my sight. \textcolor{benred8}{6}~Thee only my sins have offended; it is thy will I have disobeyed; thy sentence was deserved, and still when thou givest award thou hast right on thy side. \textcolor{benred8}{7}~For indeed, I was born in sin; guilt was with me already when my mother conceived me. \textcolor{benred8}{8}~But thou art a lover of faithfulness, and now, deep in my heart, thy wisdom has instructed me. \textcolor{benred8}{9}~Sprinkle me with a wand of hyssop, and I shall be clean; washed, I shall be whiter than snow; \textcolor{benred8}{10}~tidings send me of good news and rejoicing, and the body that lies in the dust shall thrill with pride.
\textcolor{benred8}{11}~Turn thy eyes away from my sins, blot out the record of my guilt; \textcolor{benred8}{12}~my God, bring a clean heart to birth within me; breathe new life, true life, into my being. \textcolor{benred8}{13}~Do not banish me from thy presence, do not take thy holy spirit away from me; \textcolor{benred8}{14}~give me back the comfort of thy saving power, and strengthen me in generous resolve. \textcolor{benred8}{15}~So will I teach the wicked to follow thy paths; sinners shall come back to thy obedience. \textcolor{benred8}{16}~My God, my divine Deliverer, save me from the guilt of bloodshed! This tongue shall boast of thy mercies; \textcolor{benred8}{17}~O Lord, thou wilt open my lips, and my mouth shall tell of thy praise. \textcolor{benred8}{18}~Thou hast no mind for sacrifice, burnt-offerings, if I brought them, thou wouldst refuse; \textcolor{benred8}{19}~here, O God, is my sacrifice, a broken spirit; a heart that is humbled and contrite thou, O God, wilt never disdain. \textcolor{benred8}{20}~Lord, in thy great love send prosperity to Sion, so that the walls of Jerusalem may rise again. \textcolor{benred8}{21}~Then indeed thou wilt take pleasure in solemn sacrifice, in gift and burnt-offering; then indeed bullocks will be laid upon thy altar.\footnote[1]{If this verse is part of the original psalm, it seems to imply that the sacrifice of beasts, though of no avail in reparation for grievous sin (verse 18), was a legitimate form of thanksgiving ceremony. (Cf. Ps. 49.15).}
\begin{large}\begin{center}\textsc{Chapter li}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. A maskil. Of David, \textcolor{benred8}{2}~when Doeg, the Edomite, came and told Saul, David went into Abimelech\textquotesingle s house.)
\textcolor{benred8}{3}~Wilt thou still take pride, infamous tyrant, in thy power to harm? \textcolor{benred8}{4}~Evermore thou settest thy heart on mischief; thy words are razor-edged, and thou a traitor. \textcolor{benred8}{5}~Cruelty, never kindness, is thy study, treason, never honest speech; \textcolor{benred8}{6}~well thy false lips love the word that brings men to ruin. \textcolor{benred8}{7}~And will not God destroy thee utterly, root thee up, drive thee from thy home, till thy stock is known among living men no more? \textcolor{benred8}{8}~Honest folk will watch, and wonder, and taunt him then; \textcolor{benred8}{9}~So much for the man who would have none of God\textquotesingle s help, but relied on his store of riches, and found his strength in knavery? \textcolor{benred8}{10}~And I? rooted like a fruitful olive-tree in the house of my God, I will trust for ever in his divine mercy; \textcolor{benred8}{11}~I will give thee eternal thanks for all thou hast done, and boast, as men should ever boast, of thy name, the faithful for my company.
\begin{large}\begin{center}\textsc{Chapter lii}\end{center}\end{large}
(To the choir-master. The Melody, Mahalat. A maskil. Of David.\footnote[1]{vv. 1-7. See notes on Ps. 13. In the part common to the two psalms, this one uses \textasciigrave God\textquotesingle  where the earlier psalm uses \textasciigrave the Lord\textquotesingle , but otherwise there is no substantial variation except in verse 6.})
\textcolor{benred8}{1}~There is no God above us, is the fond thought of reckless hearts. \textcolor{benred8}{2}~Warped natures everywhere and hateful lives, there is not an innocent man among them. \textcolor{benred8}{3}~God looks down from heaven at the race of men, to find one soul that reflects, and goes in search of him; \textcolor{benred8}{4}~but no, all have missed the mark and rebelled against him; an innocent man is nowhere to be found. \textcolor{benred8}{5}~What, can they learn nothing, all these traffickers in iniquity, who feed themselves fat on this people of mine, as if it were bread for their eating, \textcolor{benred8}{6}~and never invoke God\textquotesingle s name? What wonder if fear unmans them, where they have no cause for fear? Where are the foes that hemmed thee round?\footnote[2]{Some think that this verse was inserted to celebrate the destruction of Sennacherib\textquotesingle s army (IV Kg. 19.35).} God has scattered their bones far and wide, forgotten as his enemies must ever be. \textcolor{benred8}{7}~Oh, may Sion bring deliverance to Israel! Day of gladness for Jacob, day of Israel\textquotesingle s triumph, when God restores the fortunes of his own people.
\begin{large}\begin{center}\textsc{Chapter liii}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. For stringed instruments. A maskil. Of David, \textcolor{benred8}{2}~when the men of Ziph told Saul that David was in hiding among them.)
\textcolor{benred8}{3}~Lord, by the virtue of thy name deliver me, let thy sovereign power grant me redress; \textcolor{benred8}{4}~give a hearing, Lord, to my plea; let me speak, and know thou art listening. \textcolor{benred8}{5}~Scornful foes take arms against me, fierce foes that grudge me life itself, with no thought of God to check them. \textcolor{benred8}{6}~Ah, but God is here to help me; the Lord has my safety in his keeping. \textcolor{benred8}{7}~Let the blow recoil on my persecutors; ever faithful to thy word, do thou overthrow them. \textcolor{benred8}{8}~So will I joyfully offer thee sacrifice, and praise thy name, Lord, as praised it must ever be; \textcolor{benred8}{9}~who else has delivered me from all peril, and let me see the downfall of my enemies?
\begin{large}\begin{center}\textsc{Chapter liv}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. For stringed instruments. A maskil. Of David.)
\textcolor{benred8}{2}~Give audience to my prayer, O God; do not spurn this plea of mine; \textcolor{benred8}{3}~hear and grant relief. No rest I find in my distress, \textcolor{benred8}{4}~daunted ever by the hue and cry of godless enemies, whose malicious spite would compass my ruin. \textcolor{benred8}{5}~My heart is full of whirling thoughts; the fear of death stands over me; \textcolor{benred8}{6}~trembling and terrified, I see perils closing round me. \textcolor{benred8}{7}~Had I but wings, I cry, as a dove has wings, to fly away and find rest! \textcolor{benred8}{8}~Far would I flee; the wilderness should be my shelter, \textcolor{benred8}{9}~so I might find speedy refuge from the whirlwind and the storm.
\textcolor{benred8}{10}~Plunge them deep, Lord, in ruin, bring dissension into their councils! Do I not see, already, violence and sedition in the city? \textcolor{benred8}{11}~Day and night they make the round of its walls, and all the while there is wrong and oppression at the heart of it, \textcolor{benred8}{12}~its treacherous heart; cruelty and cunning walk ever in its streets. \textcolor{benred8}{13}~Had some enemy decried me, I could have borne it patiently; some open ill-wisher, I could have sheltered myself from his attack. \textcolor{benred8}{14}~But thou, my second self, my familiar friend! \textcolor{benred8}{15}~How pleasant was the companionship we shared, thou and I; how lovingly we walked as fellow pilgrims in the house of God! \textcolor{benred8}{16}~May death overtake them, may the abyss swallow them up alive, their homes, their hearts so tainted with evil!
\textcolor{benred8}{17}~Still I will call upon God, and the Lord will save; \textcolor{benred8}{18}~still at evening and morn and noon I will cry aloud and make my plea known; he will not be deaf to my appeal. \textcolor{benred8}{19}~He will win my soul peace, will rescue me from attack, when many take part against me. \textcolor{benred8}{20}~He, the God who reigned before time was, will listen to me, will bring them low. Never a change of heart, never the fear of God; \textcolor{benred8}{21}~not one but will turn against his friend, break his pledged word. \textcolor{benred8}{22}~Smooth as butter their looks, when their hearts are all hatred; soft as oil their speech, yet never was drawn sword so deadly.
\textcolor{benred8}{23}~Cast the burden of thy cares upon the Lord, and he will sustain thee; never will he let thee stumble, his servant if thou be. \textcolor{benred8}{24}~These, O God, thou wilt sink in a pit of ruin; the blood-thirsty, the treacherous, will not live out half their days; but I, Lord, will put my trust in thee.
\begin{large}\begin{center}\textsc{Chapter lv}\end{center}\end{large}
(To the choir-master. Melody: Yonat Elem Rehoqim. Of David. A miktam. When the Philistines had him in their hands at Gath.)
\textcolor{benred8}{2}~Have mercy on me, O God, downtrodden evermore by man\textquotesingle s cruel oppression; \textcolor{benred8}{3}~evermore my enemies tread me under foot, so many there are to make war upon me. \textcolor{benred8}{4}~In thee, the most High, I will put my trust when peril overtakes me. \textcolor{benred8}{5}~I claim God\textquotesingle s promise; my trust is in God, no mortal threat can daunt me. \textcolor{benred8}{6}~Evermore they traduce me, have no thought but for my undoing; \textcolor{benred8}{7}~conspire in secret, watching my steps as they plot against my life! \textcolor{benred8}{8}~O God, requite their treachery; trample on the Gentiles in thy anger. \textcolor{benred8}{9}~My wandering life none knows as thou; no tear of mine but thou dost hoard and record it. \textcolor{benred8}{10}~One cry raised to thee, and my enemies are driven back; shall I doubt God is on my side? \textcolor{benred8}{11}~I claim God\textquotesingle s promise; my trust is in God, man\textquotesingle s threats cannot daunt me. \textcolor{benred8}{12}~The vows which thou claimest from me, O God, my sacrifice of praise shall fulfil; \textcolor{benred8}{13}~hast thou not saved my life from every peril, my feet from every slip? \textcolor{benred8}{14}~And shall I not enjoy God\textquotesingle s favour, while the light of life is with me?
\begin{large}\begin{center}\textsc{Chapter lvi}\end{center}\end{large}
(To the choir-master. Melody: Do not destroy. Of David. A miktam. When he took refuge from Saul in a cave.)
\textcolor{benred8}{2}~Have mercy on me, O God, have mercy on me; here is a soul that puts its trust in thee, I will take refuge under the shelter of thy wings, till the storms pass by. \textcolor{benred8}{3}~I will cry out to the most high God, the God who has ever befriended me: \textcolor{benred8}{4}~may he send aid now from heaven to deliver me, and bring confusion on my oppressors; his mercy, his faithfulness be my speed! \textcolor{benred8}{5}~Fallen among lions I, that hungrily eat men\textquotesingle s flesh; here are envious teeth that bite deeper than spear or arrow, tongues sharp as any sword. \textcolor{benred8}{6}~O God, mount high above the heavens, till thy glory overshadows the whole earth. \textcolor{benred8}{7}~See where they have laid a snare for my feet, to bring me low, dug a pit in my path; may it be their own undoing! \textcolor{benred8}{8}~A true heart, my God, a heart true to thy service; its song, its music are for thee! \textcolor{benred8}{9}~Wake, my soul, wake, echoes of harp and viol; dawn shall find me watching; \textcolor{benred8}{10}~so will I give thee thanks, Lord, for all the world to hear it, sing psalms while the Gentiles listen, \textcolor{benred8}{11}~of thy mercy, high as heaven itself, of thy faithfulness, that reaches to the clouds. \textcolor{benred8}{12}~O God, mount high above the heavens, till thy glory overshadows the whole earth.
\begin{large}\begin{center}\textsc{Chapter lvii}\end{center}\end{large}
(To the choir-master. Melody: Do not destroy. Of David. A miktam.)
\textcolor{benred8}{2}~Nobles all, are they honest words you utter? And you, common folk, do you make just award? \textcolor{benred8}{3}~See how you devise treachery in your hearts, deal out to this land nothing but oppression! \textcolor{benred8}{4}~Sinners that left the womb only to go a-straying; renegades and liars their mothers bore them! \textcolor{benred8}{5}~They are venomous as serpents, as the asp that turns a deaf ear \textcolor{benred8}{6}~and will not listen to the snake-charmer\textquotesingle s music, skilful player though he be. \textcolor{benred8}{7}~My God, break their cruel fangs; Lord, shatter their jaws, strong as the jaws of lions. \textcolor{benred8}{8}~Like spilt water let them run to waste, shoot none but harmless arrows; \textcolor{benred8}{9}~melt into nothing, the snail\textquotesingle s way, perish like the untimely birth that sees never the light of the sun. \textcolor{benred8}{10}~Green stalks the whirlwind carries away, while yonder pot still waits for fuel! \textcolor{benred8}{11}~The innocent man will triumph at the sight of their punishment, as he dips his hands in the blood of the evil-doer; \textcolor{benred8}{12}~Sure enough, men will say, innocence has its reward, sure enough, there is a God who grants redress here on earth.
\begin{large}\begin{center}\textsc{Chapter lviii}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. Melody: Do not destroy. Of David. A miktam. When Saul had his house watched, so as to put him to death.)
\textcolor{benred8}{2}~O God, deliver me from my enemies, rescue me from their assaults; \textcolor{benred8}{3}~thwart their treacherous designs, disappoint their lust for blood. \textcolor{benred8}{4}~See how they plot against my life, how strong is their confederacy! \textcolor{benred8}{5}~Yet never, Lord, through any fault or offence of mine; unprovoked, they rush to the attack. \textcolor{benred8}{6}~Bestir thyself, come to my side and witness my wrongs. Lord of hosts, God of Israel, awake; a world chastise, and shew the treacherous no mercy. \textcolor{benred8}{7}~See how they come back at nightfall, like yelping dogs, to prowl about the city! \textcolor{benred8}{8}~Tongues that boast, lips that rail, as if none could hear them; \textcolor{benred8}{9}~and all the while thou, Lord, makest light of them, thou, in whose esteem all the nations are as nothing. \textcolor{benred8}{10}~To thee I look, the God who strengthens me, the God who watches over me; \textcolor{benred8}{11}~my God, and all my hope of mercy.With that divine aid, may I triumph over my enemies. \textcolor{benred8}{12}~Smite them down, my God, before they compass the overthrow of my people; let that power of thine overawe and crush them, my protector, my Master! \textcolor{benred8}{13}~Down with the guilty tongues, the boastful lips; let their own pride ensnare them, their cursing and their lies. \textcolor{benred8}{14}~Ruinous, ruinous be thy vengeance, overwhelming them; shew them that there is a God who rules over Jacob, rules over the utmost ends of the earth. \textcolor{benred8}{15}~Back come they at nightfall, like yelping dogs, and prowl about the city; \textcolor{benred8}{16}~far and wide they will roam in search of their prey, and snarl with rage when they go unfed at last.\footnote[1]{vv. 15, 16. The enemies of the Psalmist, who are patrolling the city to keep him in, will find that he has escaped, and will be scattered far and wide in their disappointed search for him.} \textcolor{benred8}{17}~And I, ere long, will be proclaiming thy greatness, will be triumphing in the mercy thou hast shewn me, thou, my stronghold and my refuge in my hour of peril. \textcolor{benred8}{18}~To thee I will sing, the God who strengthens me, the God who watches over me, my God, and all my hope of mercy.
\begin{large}\begin{center}\textsc{Chapter lix}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. Melody: Lily of the Law. A miktam. Of David. To be used for teaching. \textcolor{benred8}{2}~This was when he marched against the Syrians of Naharaim and Soba, and Joab on his way home defeated the men of Edom in the Valley of Salt, twelve thousand of them.)
\textcolor{benred8}{3}~Too long, O God, hast thou disowned us, and scattered our armies in flight; thy wrath forgo, and bring us back to our own. \textcolor{benred8}{4}~Heal the wounds of the land thou hast shaken and torn asunder, the land that trembles still. \textcolor{benred8}{5}~Heavy the burden thou didst lay on us; such a draught thou didst brew for us as made our senses reel. \textcolor{benred8}{6}~But now thou hast set up a standard to rally thy faithful servants, and to protect them from the archers\textquotesingle  onset; now bring aid to the men thou lovest, \textcolor{benred8}{7}~give our prayer answer, and lift thy right hand to save.
\textcolor{benred8}{8}~God\textquotesingle s word came to us from his sanctuary, In triumph I will divide up Sichem, and parcel out the valley of Tents; \textcolor{benred8}{9}~to me Galaad, to me Manasses belongs; Ephraim is my helmet, Juda the staff I bear. \textcolor{benred8}{10}~Now Moab, too, shall be my drudge; over Edom I will claim my right; I will lead the Philistines away in triumph.\footnote[1]{vv. 8-10. This would seem to be a quotation from some old prophecy, referring to the conquest of Chanaan. \textasciigrave Claim my right\textquotesingle  is literally \textasciigrave cast out my shoe\textquotesingle , cf. Ruth 4.7.} \textcolor{benred8}{11}~Such was the oracle;\footnote[2]{The words \textasciigrave such was the oracle\textquotesingle  are not expressed in the original.} but now who is to lead me on my march against this fortress, who is to find an entrance for me into Edom, \textcolor{benred8}{12}~when thou, O God, hast disowned us, and wilt not go into battle with our armies? \textcolor{benred8}{13}~It is thou that must deliver us from peril; vain is the help of man. \textcolor{benred8}{14}~Only through God can we fight victoriously; only he can trample our oppressors in the dust.
\begin{large}\begin{center}\textsc{Chapter lx}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. For stringed instruments. Of David.)
\textcolor{benred8}{2}~Listen, Lord, to this cry of appeal; do not let my prayer go unheeded, \textcolor{benred8}{3}~though it be from the ends of the earth that I call upon thee. When my heart misgives me, thou wilt set me high up on a rock, thou wilt bring me repose; \textcolor{benred8}{4}~thou, my only hope, my strong tower against the assault of my enemies. \textcolor{benred8}{5}~Oh let me dwell for ever in thy tabernacle, let me take refuge under the shelter of thy wings! \textcolor{benred8}{6}~Lord, thou hast listened to my prayer, a domain thou hast given me where thy name is held in awe. \textcolor{benred8}{7}~Year upon year do thou add to the king\textquotesingle s reign; while generations come and go, may his life still last. \textcolor{benred8}{8}~For ever may he reign under God\textquotesingle s favour; let mercy and faithfulness be his escort. \textcolor{benred8}{9}~Eternally I will sing thy praises, day after day perform my vows.
\begin{large}\begin{center}\textsc{Chapter lxi}\end{center}\end{large}
(To the choir-master. Melody: Idithun. A psalm. Of David.)
\textcolor{benred8}{2}~No rest has my soul but in God\textquotesingle s hands; to him I look for deliverance. \textcolor{benred8}{3}~I have no other stronghold, no other deliverer but him; safe in his protection, I fear no deadly fall. \textcolor{benred8}{4}~Still one man my enemies single out for their onslaught, not gaping hedge or ruinous wall more ripe for overthrow; \textcolor{benred8}{5}~from my safe fastness they would fain dislodge me, ready liars that speak me fair, but ever with a curse in their hearts. \textcolor{benred8}{6}~Yet even now, my soul, leave thyself in God\textquotesingle s hands; all my trust is in him. \textcolor{benred8}{7}~He is my stronghold and my deliverer, my protector, that makes me stand unmoved. \textcolor{benred8}{8}~God is all my defence and all my boast; my rock-fastness, my refuge is in God. \textcolor{benred8}{9}~Israelites, put ever your trust in him, and lay the homage of your hearts at his feet; God is our defence.
\textcolor{benred8}{10}~Man is a breath that passes; in Adam\textquotesingle s sons there is no trust; high in the scales they rise, weighed all together and lighter than a breath. \textcolor{benred8}{11}~Put your faith in extortion no more, boast no more of plunder; set never your heed on mounting store of riches. \textcolor{benred8}{12}~Not once, but twice I have heard God\textquotesingle s voice of warning; all power is God\textquotesingle s. \textcolor{benred8}{13}~To thee, Lord, mercy belongs; thou wilt repay every man the reward of his deeds.
\begin{large}\begin{center}\textsc{Chapter lxii}\end{center}\end{large}
(A psalm. Of David. This was when he was in the desert of Juda.)
\textcolor{benred8}{2}~O God, thou art my God; how eager my quest for thee, body athirst and soul longing for thee, \textcolor{benred8}{3}~like some parched wilderness, where stream is none! So in the holy place, I contemplate thee, ready for the revelation of thy greatness, thy glory. \textcolor{benred8}{4}~To win thy favour is dearer to me than life itself; my songs of praise can no more be withheld. \textcolor{benred8}{5}~So, all my life long, I will bless thee, holding up my hands in honour of thy name; \textcolor{benred8}{6}~my heart filled, as with some rich feast, my mouth, in joyful accents, singing thy praise. \textcolor{benred8}{7}~My thoughts shall go out to thee at dawn, as I lie awake remembering thee, \textcolor{benred8}{8}~and the protection thou hast given me. Gladly I take shelter under thy wings, \textcolor{benred8}{9}~cling close to thee, borne up by thy protecting hand. \textcolor{benred8}{10}~In vain do my enemies plot against my life, soon to be swallowed up in the depths of earth, \textcolor{benred8}{11}~a prey to the sword, carrion for jackals! \textcolor{benred8}{12}~The king shall triumph in God\textquotesingle s protection, blessed as they are ever blessed who take their vows in his name; silence shall fall on the lips that muttered treason.
\begin{large}\begin{center}\textsc{Chapter lxiii}\end{center}\end{large}
(To the choir-master. A psalm. Of David.)
\textcolor{benred8}{2}~O God, listen to my prayer when I plead with thee, save me from the threats of my mortal foe. \textcolor{benred8}{3}~Thine to defend me from this conspiracy of malice, this throng of evil-doers; \textcolor{benred8}{4}~the tongues that wound like a sharpened sword, the poisoned words aimed at me, like arrows. \textcolor{benred8}{5}~Stealthily they attack the innocent: \textcolor{benred8}{6}~suddenly, from a safe vantage-point, they wound him. See them pledged to an infamous resolve, plotting to lay snares for me, sure that they will go unseen! \textcolor{benred8}{7}~With what care they hatch their designs, planning treason double-dyed! Let the thoughts of man\textquotesingle s heart be deep as they will, \textcolor{benred8}{8}~yet God has arrows, too, to smite them with, sudden wounds to deal them; \textcolor{benred8}{9}~all their conspiring plays them false. Scornfully the onlookers shake their heads, \textcolor{benred8}{10}~awe-stricken every one; who but will acclaim God\textquotesingle s power, who but will ponder his great acts? \textcolor{benred8}{11}~Honest men will rejoice and put their trust in the Lord; upright hearts will not boast in vain.
\begin{large}\begin{center}\textsc{Chapter lxiv}\end{center}\end{large}
(To the choir-master. A psalm. Of David. A song.)
\textcolor{benred8}{2}~O God, thou shalt yet have praise in Sion; to thee let the vow be paid, \textcolor{benred8}{3}~hearer of prayer. To thee all mankind must look for pardon, \textcolor{benred8}{4}~weighed down by its sinfulness till thou dost forgive. \textcolor{benred8}{5}~Blessed the man on whom thy choice falls, whom thou takest to dwell with thee in thy own domain! Fill these hearts with love of thy house, with awe of thy holy temple. \textcolor{benred8}{6}~Wonderful the miracles thou shewest when in mercy thou dost hear us, O God our Saviour; at the bounds of earth, far over the seas, in thee we hope. \textcolor{benred8}{7}~What power girds thee about! In thy strength the mountains stand firm; \textcolor{benred8}{8}~thou dost calm the raging of the sea, raging sea-billows, ay, and the turmoil of angry nations. \textcolor{benred8}{9}~Thy portents strike terror at the world\textquotesingle s end, fill the lands of sunrise and sunset with rejoicing. \textcolor{benred8}{10}~And now thou hast brought relief to this land of ours, hast watered and greatly enriched it; deep flows the channel whence thy divine providence grants us food; long time thou dost prepare it, \textcolor{benred8}{11}~watering the furrow, loosening the clods, multiplying, with soft showers, the grain. \textcolor{benred8}{12}~Thy bounty it is that crowns the year; where thy feet have passed, the stream of plenty flows; \textcolor{benred8}{13}~flows through the desert pastures, till all the hill-sides are gaily clad, \textcolor{benred8}{14}~herds throng the fields, and the valleys stand deep in corn; the shout of joy everywhere, everywhere the hymn of praise.
\begin{large}\begin{center}\textsc{Chapter lxv}\end{center}\end{large}
(To the choir-master. A song. A psalm.)
\textcolor{benred8}{1}~Let the whole world keep holiday in God\textquotesingle s presence, \textcolor{benred8}{2}~sing praise to his name, pay homage to his glory! \textcolor{benred8}{3}~Cry out to God, What dread, Lord, thy acts inspire! How great is that might of thine, which makes thy enemies cringe before thee! \textcolor{benred8}{4}~Let the whole earth worship thee, sing of thee, sing praises to thy name. \textcolor{benred8}{5}~Come near, and see what God does, how wonderful he is in his dealings with human kind, \textcolor{benred8}{6}~how he turns the sea into land, and lets men cross a river dry-shod; ours to rejoice in his mercy. \textcolor{benred8}{7}~In that power of his he reigns for ever, and has eyes for what the Gentiles do; let rebellious souls tame their pride. \textcolor{benred8}{8}~Bless the name of our God, you Gentiles, echo the sound of his praise.
\textcolor{benred8}{9}~God\textquotesingle s will is we should live yet; he does not suffer our steps to falter. \textcolor{benred8}{10}~Yes, Lord, thou hast put us to the proof, tested us as men test silver in the fire; \textcolor{benred8}{11}~led us into a snare, and bowed our backs with trouble, \textcolor{benred8}{12}~while human masters rode us down; our way led through fire and water, yet in the end thou hast granted us relief. \textcolor{benred8}{13}~See, I come into thy house with burnt-offerings, to pay thee all the vows \textcolor{benred8}{14}~these lips have framed, this mouth has uttered, when trouble came upon me. \textcolor{benred8}{15}~Fat burnt-offerings of sheep shall be thine, and the smoke of ram\textquotesingle s flesh; bullocks and goats shall be thy sacrifice. \textcolor{benred8}{16}~Come and give ear, all you who worship God, while I tell of the great mercies he has shewn me; \textcolor{benred8}{17}~how this voice of mine cried out to him, this tongue did him honour. \textcolor{benred8}{18}~Would God listen to me, if my heart were set on wrong? \textcolor{benred8}{19}~And God has listened to me; given heed to my cry for succour. \textcolor{benred8}{20}~Blessed be God, who does not reject my prayer, does not withhold his mercy from me.
\begin{large}\begin{center}\textsc{Chapter lxvi}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. For stringed instruments. A psalm. A song.)
\textcolor{benred8}{2}~May God be merciful to us, and bless us; may he grant us the favour of his smile. \textcolor{benred8}{3}~Make known thy will, O God, wide as earth; make known among all nations thy saving power. \textcolor{benred8}{4}~Honour to thee, O God, from the nations, honour from all the nations! \textcolor{benred8}{5}~The Gentiles, too, may rejoice and be glad; a whole world abides thy judgement, and the Gentiles, too, obey on earth thy sovereignty. \textcolor{benred8}{6}~Honour to thee, O God, from the nations, honour from all the nations! \textcolor{benred8}{7}~The land has yielded its harvest; such bounty God, our own God, affords. \textcolor{benred8}{8}~God grant us ever his blessing, and may earth, far and wide, do him reverence.
\begin{large}\begin{center}\textsc{Chapter lxvii}\end{center}\end{large}
(To the choir-master. Of David. A psalm. A song.)
\textcolor{benred8}{2}~Let God bestir himself, needs must his foes be scattered, their malice take flight before his coming. \textcolor{benred8}{3}~Vanish the wicked at God\textquotesingle s presence as the smoke vanishes, as wax melts at the fire, \textcolor{benred8}{4}~while the just keep holiday and rejoice at the sight of him, glad and content. \textcolor{benred8}{5}~Sing, then, in God\textquotesingle s honour, praise his name with a psalm; a royal progress through the wilderness for the God whose name tells of omnipotence! Triumph in his presence; \textcolor{benred8}{6}~he is a father to the orphan, and gives the widow redress, this God who dwells apart in holiness. \textcolor{benred8}{7}~This is the God who makes a home for the outcast, restores the captives to a land of plenty, leaves none but the rebels to find their abode in the wilderness.
\textcolor{benred8}{8}~O God, when thou didst go forth at the head of thy people, on that royal progress of thine through the desert, \textcolor{benred8}{9}~how the earth trembled, how the sky broke at God\textquotesingle s coming, how even Sinai shook, when the God of Israel came! \textcolor{benred8}{10}~And on this thy own land, O God, thou sendest rain abundantly; all parched it lies, and thou dost bring it relief. \textcolor{benred8}{11}~Pasture-land of thy own flock; and shouldst thou not make bounteous provision here, O God, for thy pensioners?
\textcolor{benred8}{12}~Word has come from the Lord, good news borne on a multitude of lips: \textcolor{benred8}{13}~Routed their kings, routed the armies; they have left their spoils for housewives to carry away; \textcolor{benred8}{14}~never shone silver so bright on a dove\textquotesingle s feathers, never gold so fair on a dove\textquotesingle s wings; and you, all the while, resting quiet among the sheep-folds! \textcolor{benred8}{15}~White fell the snows on Salmon, when the Almighty put kings to rout.\footnote[1]{The early part of this psalm clearly refers to the victory of Barac over King Sisera; verses 8 and 9 being actually repeated from Jg. 5.4 and 5, and verse 14 from Jg. 5.16.}
\textcolor{benred8}{16}~Basan\textquotesingle s hills are high, Basan\textquotesingle s hills are rugged; \textcolor{benred8}{17}~must you turn your eyes, rugged hills, towards God\textquotesingle s mountain, and envy what you see? The mountain where God loves and will ever love to dwell. \textcolor{benred8}{18}~See where God comes, with chariots innumerable for his escort; thousands upon thousands; comes from Sinai to this his sanctuary. \textcolor{benred8}{19}~Thou dost mount up on high, thou dost capture thy spoil, and men must be thy tribute,\footnote[2]{S. Paul in Eph. 4.5, apparently using some other text, has \textasciigrave gave gifts to men\textquotesingle  instead of \textasciigrave men must be thy tribute\textquotesingle .} will they or no, yonder heathen must have the Lord God for their neighbour. \textcolor{benred8}{20}~Blessed be the Lord now and ever, the God who bears our burdens, and wins us the victory. \textcolor{benred8}{21}~Our God is a God of deliverance; the Lord is our Master, that saves men from peril of death. \textcolor{benred8}{22}~God will smite the heads of his enemies, smite the proud locks of the men who live at ease in their wickedness. \textcolor{benred8}{23}~I will restore my people, the Lord says; I will restore them to their land, from Basan, from the shore of the high seas. \textcolor{benred8}{24}~Soon the blood of thy enemies will stain thy feet, never a jackal that follows thee but shall lick its prey. \textcolor{benred8}{25}~Thou comest, O God, a mark for all eyes; he comes, my God and my king, to visit his sanctuary. \textcolor{benred8}{26}~Before him go the singers, and the minstrels follow, while the maids play on their tambours between: \textcolor{benred8}{27}~Give praise to the Lord God in this solemn assembly, sons of Israel! \textcolor{benred8}{28}~Here is Benjamin, youngest of the tribes, that marches in the van; here are the chieftains of Juda with their companies, chieftains, too, from Zabulon, chieftains from Nephtali.
\textcolor{benred8}{29}~Shew thy power, O God, shew thy Divine power, perfect thy own achievement for us; \textcolor{benred8}{30}~to honour thy temple at Jerusalem, kings shall bring gifts before thee. \textcolor{benred8}{31}~Tame the wild beasts of the marshes, fierce bulls that lord it over the peaceful herd of nations; down fall they, bringing silver pieces for their ransom. Scatter the nations that delight in war, \textcolor{benred8}{32}~till Egypt sends hither her princes, till Ethiopia makes her peace with God. \textcolor{benred8}{33}~Kingdoms of the earth, raise your voices in God\textquotesingle s honour, sing a psalm to the Lord; a psalm to God, \textcolor{benred8}{34}~who mounts on the heavens, the immemorial heavens, and utters his word in a voice of thunder. \textcolor{benred8}{35}~Pay honour to God, the God whose splendour rests over Israel, who holds dominion high among the clouds. \textcolor{benred8}{36}~Awe dwells about him in his holy place! The God of Israel gives his people strength and courage; blessed be God!
\begin{large}\begin{center}\textsc{Chapter lxviii}\end{center}\end{large}
(To the choir-master. Melody: The Lilies. Of David.)
\textcolor{benred8}{2}~O God, save me; see how the waters close about me, shoulder-high! \textcolor{benred8}{3}~I am like one who sticks fast in deep mire, with no ground under his feet, one who has ventured out into mid-ocean, to be drowned by the storm. \textcolor{benred8}{4}~Hoarse my throat with crying wearily for help; my eyes ache with looking up for mercy to my God. \textcolor{benred8}{5}~Countless as the hairs on my head are my wanton enemies, I am no match for the oppressors that wrong me. Should I make amends to them, I, that never robbed them? \textcolor{benred8}{6}~O God, thou knowest my rash doings, no fault of mine is hidden from thy sight. \textcolor{benred8}{7}~Master, Lord of hosts, shall ill fortune of mine bring shame to those who trust in thee, make men repent of looking for aid to thee, the God of Israel? \textcolor{benred8}{8}~It is for thy sake that I have met with reproach, that I have so often blushed with confusion, \textcolor{benred8}{9}~an outcast among my own brethren, a stranger to my own mother\textquotesingle s children. \textcolor{benred8}{10}~Was it not jealousy for the honour of thy house that consumed me; was it not uttered against thee, the reproach I bore?
\textcolor{benred8}{11}~What more could I do? I humbled myself before them by fasting; and that, too, was matter for finding fault; \textcolor{benred8}{12}~I dressed in sackcloth, and they made a by-word of me. \textcolor{benred8}{13}~Idlers in the market-place taunt me; the drunkards make a song of me over their wine. \textcolor{benred8}{14}~To thee, Lord, I make my prayer; never man more needed thy good will. Listen to me, O God, full of mercy as thou art, faithful as thou art to thy promise of aid. \textcolor{benred8}{15}~Save me from sinking in the mire, rescue me from my enemies, from the deep waters that surround me; \textcolor{benred8}{16}~let me not sink under the flood, swallowed up in its depths, and the well\textquotesingle s mouth close above me. \textcolor{benred8}{17}~Listen to me, Lord, of thy gracious mercy, look down upon me in the abundance of thy pity; \textcolor{benred8}{18}~do not turn thy face away from thy servant in this time of trouble, give a speedy answer to my prayer. \textcolor{benred8}{19}~Draw near in my distress, and grant deliverance; relieve me, so hard pressed by my enemies. \textcolor{benred8}{20}~Lord, thou knowest how they reproach me, how I blush with shame; \textcolor{benred8}{21}~thou seest how many are my persecutors. Heart-broken with that shame, I pine away, looking round for pity, where pity is none, for comfort, where there is no comfort to be found. \textcolor{benred8}{22}~They gave me gall to eat, and when I was thirsty they gave me vinegar to drink.\footnote[1]{Mt. 27.48.} \textcolor{benred8}{23}~Let their feast be turned into a trap, a net to catch them and theirs,\footnote[2]{Rom. 11.9.} \textcolor{benred8}{24}~ever the blind eye be theirs, ever the halting loin. \textcolor{benred8}{25}~Pour out thy anger upon them, let them be overtaken by the tide of thy vengeance; \textcolor{benred8}{26}~let their dwelling-place be deserted, their tents for ever uninhabited.\footnote[3]{Ac. 1.20.} \textcolor{benred8}{27}~Who is it they persecute? A man already afflicted by thee; hard was my hurt to bear, and these have added to it. \textcolor{benred8}{28}~Do thou add guilt to guilt in their reckoning; let them never claim thy acquittal; \textcolor{benred8}{29}~let their names be blotted out from the record of the living, and never be written among the just.
\textcolor{benred8}{30}~See how friendless I am, and how distressed! Let thy help, O God, sustain me. \textcolor{benred8}{31}~I will sing in praise of God\textquotesingle s name, herald it gratefully; \textcolor{benred8}{32}~a more acceptable sacrifice, this, to the Lord than any young bullock, for all its promise of horn and hoof. \textcolor{benred8}{33}~Here is a sight to make the afflicted rejoice; to cheer men\textquotesingle s spirits in their quest for God. \textcolor{benred8}{34}~The Lord listens to the prayer of the destitute; he does not forget his servants in their chains. \textcolor{benred8}{35}~To him be praise from sky, earth and sea, and from all the creatures that move about them. \textcolor{benred8}{36}~God will grant deliverance to Sion; the cities of Juda will rise from their ruins, inhabited now and held firmly in possession, \textcolor{benred8}{37}~an inheritance for the race that serves him, a home for all true lovers of his name.
\begin{large}\begin{center}\textsc{Chapter lxix}\end{center}\end{large}
(To the choir-master. Of David. A commemoration.)\footnote[1]{This psalm is nearly a repetition of Ps. 39.14-18.}
\textcolor{benred8}{2}~Deign, O God, to set me free; Lord, make haste to help me. \textcolor{benred8}{3}~Disappoint them, put them to the blush, the enemies who plot against my life! \textcolor{benred8}{4}~Baffled let them go their way, that rejoice at my ill fortune; slink away in confusion, that crowed over me so loud! \textcolor{benred8}{5}~Triumphant joy be theirs, who long for thee; Praise to the Lord, be ever their song, who look eagerly now for thy succour. \textcolor{benred8}{6}~Thou seest me helpless and destitute; my God, help me. Thou art my champion and my deliverer; Lord, do not delay thy coming.
\begin{large}\begin{center}\textsc{Chapter lxx}\end{center}\end{large}
\lettrine[lines=2]{T}{o} thee, O God, I turn for succour; may I never be disappointed! \textcolor{benred8}{2}~Rescue and deliver me, faithful as thou art; listen to my cry for succour. \textcolor{benred8}{3}~Let me find in thee a rock-fastness, a citadel of defence; I have no other stronghold, no other refuge, but thee. \textcolor{benred8}{4}~Rescue me, O God, from the power of the wicked, from the grasp of lawlessness and oppression; \textcolor{benred8}{5}~thou, my God and Master, the hope and confidence of my youth. \textcolor{benred8}{6}~Thou hast upheld me from birth, thou hast guarded me ever since I left my mother\textquotesingle s womb; ever in thee was my trust. \textcolor{benred8}{7}~Men stare at me now as a strange portent,\footnote[1]{It is not clear in what sense the Psalmist was a portent or sign to his contemporaries, but the context suggests that he may be referring to his great age, and God\textquotesingle s long preservation of him.} so signal the protection thou hast given me. \textcolor{benred8}{8}~And evermore praise was on my lips, my constant theme thy glory.
\textcolor{benred8}{9}~Do not cast me off now, in my old age; slowly my strength ebbs, do not thou forsake me. \textcolor{benred8}{10}~A mark thou seest me for envious eyes and tongues; they conspire together, \textcolor{benred8}{11}~and whisper, God has abandoned him; now is the time to overtake and seize him; no one can bring him rescue now. \textcolor{benred8}{12}~O God, do not keep thy distance from me; hasten, my God, to aid me. \textcolor{benred8}{13}~Defeat their plot against my life, and bring it to nothing; cover my ill-wishers with confusion and shame.
\textcolor{benred8}{14}~Still will I hope on, praising thee ever more and more. \textcolor{benred8}{15}~Day in, day out, these lips shall tell of thy faithfulness, of thy saving power, and find no end to them; \textcolor{benred8}{16}~thy divine wonders, thy matchless justice, Lord, they shall proclaim. \textcolor{benred8}{17}~It is thou, O God, that hast inspired me ever since the days of my youth, and still I am found telling the tale of thy wonders. \textcolor{benred8}{18}~O God, do not fail me, now when I am old and grey-headed, till I have made known the proofs of thy power to this, to all the generations that will follow; \textcolor{benred8}{19}~that faithfulness of thine which reaches up, O God, to the heavens, so signally made manifest. There is none like thee, O God, none like thee. \textcolor{benred8}{20}~Ah, how often thou hast burdened me with bitter trouble! And still thou wouldst relent, and give me back life, and bring me up again from the very depths of the earth. \textcolor{benred8}{21}~Lift my head high; turn back, and comfort me. \textcolor{benred8}{22}~So true to thy word, and shall I not give thee thanks with psalm-music, praise thee on the harp, O God, the Holy One of Israel? \textcolor{benred8}{23}~Gladly these lips will sing of thee, this heart, which owes thee its deliverance. \textcolor{benred8}{24}~Day in, day out, I will repeat the story of thy faithfulness, what shame fell, what confusion, on the men who sought to wrong me.
\begin{large}\begin{center}\textsc{Chapter lxxi}\end{center}\end{large}
(Of Solomon.)
\textcolor{benred8}{2}~Grant to the king, O God, thy own skill in judgement; the inheritor of a throne, may he be just, as thou art just; may he give thy people right awards, and to thy poor, redress. \textcolor{benred8}{3}~Such be the harvest his subjects shall reap, peace on every mountain, justice on every hill-side. \textcolor{benred8}{4}~Watch and ward he will keep over the friendless, protect the children of the poor, and crush the oppressor. \textcolor{benred8}{5}~Ageless as sun or moon he shall endure; \textcolor{benred8}{6}~kindly as the rain that drops on the meadow grass, as the showers that water the earth. \textcolor{benred8}{7}~Justice in his days shall thrive, and the blessings of peace; and may those days last till the moon shines no more.
\textcolor{benred8}{8}~From sea to sea, from the great river to the ends of earth, his sway shall reach. \textcolor{benred8}{9}~In his presence rebels shall bend the knee, all his enemies will be humbled in the dust; \textcolor{benred8}{10}~gifts shall flow in from the lords of Tharsis and the islanders, tribute from the kings of Arabia and of Saba; \textcolor{benred8}{11}~all kings must needs bring their homage, all nations serve him. \textcolor{benred8}{12}~He will give the poor redress when they cry to him, destitute folk, with none to befriend them; \textcolor{benred8}{13}~in their need and helplessness, they shall have his compassion. Their lives he will take into his keeping, \textcolor{benred8}{14}~set them free from the power of wrong and oppression, dearly avenge their blood. \textcolor{benred8}{15}~Long life shall be his, and gold from Arabia shall be given him; men will pray for him continually, bless his name evermore. \textcolor{benred8}{16}~The land shall have good store of corn, high up the hill-sides, rustling like the woods of Lebanon; shall multiply its citizens like grass on the ground. \textcolor{benred8}{17}~For ever let his name be used in blessing, a name to endure while the sun gives light; in him all the tribes of the earth shall be enriched, all the nations shall extol him.
\textcolor{benred8}{18}~Blessed be the Lord God of Israel, who does wonderful deeds as none else, \textcolor{benred8}{19}~and blessed for ever be his glorious name; all the earth shall be filled with his glory, Amen, Amen.
\begin{large}\begin{center}\textsc{Chapter lxxii}\end{center}\end{large}
(A psalm. Of Asaph.)
\textcolor{benred8}{1}~What bounty God shews, what divine bounty, to the upright, to the pure of heart! \textcolor{benred8}{2}~Yet I was near losing my foothold, felt the ground sink under my steps, \textcolor{benred8}{3}~such heart-burning had I at seeing the good fortune of sinners that defy his law; \textcolor{benred8}{4}~for them, never a pang; healthy and sleek their bodies shew. \textcolor{benred8}{5}~Not for these to share man\textquotesingle s common lot of trouble; the plagues which afflict human kind still pass them by. \textcolor{benred8}{6}~No wonder if pride clings to them like a necklace, if they flaunt, like fine clothes, their wrong-doing. \textcolor{benred8}{7}~From those pampered hearts what malice proceeds, what vile schemes are hatched! \textcolor{benred8}{8}~Ever jeering, ever talking maliciously, throned on high they preach injustice; \textcolor{benred8}{9}~their clamour reaches heaven, and their false tales win currency on earth.
\textcolor{benred8}{10}~Enviously the men of my own race look on, to see them draining life\textquotesingle s cup to the full; \textcolor{benred8}{11}~Can God, they ask, be aware of this? Does the most High know of all that passes? \textcolor{benred8}{12}~Look at these sinners, how they live at peace, how they rise to greatness! \textcolor{benred8}{13}~Why then, thought I, it is to no purpose that I have kept my heart true, and washed my hands clean in pureness of living; \textcolor{benred8}{14}~still, all the while, I am plagued for it, and no morning comes but my scourging is renewed. \textcolor{benred8}{15}~Was I to share their thoughts? Nay, that were to put the whole company of thy children in the wrong. \textcolor{benred8}{16}~I set myself to read the riddle, but it proved a hard search, \textcolor{benred8}{17}~until I betook myself to God\textquotesingle s sanctuary, and considered, there, what becomes of such men at last. \textcolor{benred8}{18}~The truth is, thou art making a slippery path for their feet, ready to plunge them in ruin; \textcolor{benred8}{19}~in a moment they are fallen, in a storm of terrors vanished and gone. \textcolor{benred8}{20}~And thou, Lord, dost rise up and brush aside all their imaginings, as a waking man his dream.
\textcolor{benred8}{21}~What if my mind was full of bitterness, what if I was pierced to the heart? \textcolor{benred8}{22}~I was all dumbness, I was all ignorance, \textcolor{benred8}{23}~standing there like a brute beast in thy presence. Yet ever thou art at my side, \textcolor{benred8}{24}~ever holdest me by my right hand. Thine to guide me with thy counsel, thine to welcome me into glory at last. \textcolor{benred8}{25}~What else does heaven hold for me, but thyself? What charm for me has earth, here at thy side? \textcolor{benred8}{26}~What though flesh of mine, heart of mine, should waste away? Still God will be my heart\textquotesingle s stronghold, eternally my inheritance. \textcolor{benred8}{27}~Lost those others may be, who desert thy cause, lost are all those who break their troth with thee; \textcolor{benred8}{28}~I know no other content but clinging to God, putting my trust in the Lord, my Master; within the gates of royal Sion I will be the herald of thy praise.
\begin{large}\begin{center}\textsc{Chapter lxxiii}\end{center}\end{large}
(A maskil. Of Asaph.)O God, hast thou altogether abandoned us? Sheep of thy own pasturing, must we feel the fires of thy vengeance?
\textcolor{benred8}{1}~O God, hast thou altogether abandoned us? Sheep of thy own pasturing, must we feel the fires of thy vengeance? \textcolor{benred8}{2}~Bethink thee of the company thou hast gathered, long ago; of the tribe thou hast chosen to be thy domain; of mount Sion, where thou hast thy dwelling-place. \textcolor{benred8}{3}~Hither turn thy steps, where all is ruin irretrievable; see what havoc thy enemies have wrought in the holy place, \textcolor{benred8}{4}~how their malice has raged in thy very precincts,\footnote[1]{Literally, \textasciigrave thy solemn feast\textquotesingle .} setting up its emblems \textcolor{benred8}{5}~for a trophy of conquest. Blow after blow, like woodmen in the forest, \textcolor{benred8}{6}~they have plied their axes, brought it down, with pick and mallet, to the ground. \textcolor{benred8}{7}~They have set fire to thy sanctuary, sullied the dwelling-place of thy glory in the dust. \textcolor{benred8}{8}~They think to destroy us like one man, sweep away every shrine of God in the land. \textcolor{benred8}{9}~Our own emblems are nowhere to be seen; there are no prophets left now, none can tell how long we must endure. \textcolor{benred8}{10}~O God, shall our enemy taunt us everlastingly, shall blasphemy still defy thy name? \textcolor{benred8}{11}~Why dost thou withhold thy hand? That right hand of thine, must it always lie idle in thy bosom?
\textcolor{benred8}{12}~Mine is a King who reigned before time was; here on earth he was the means to bring deliverance. \textcolor{benred8}{13}~What power but thine could sunder the shifting sea, crush the power of the monster beneath its waters; \textcolor{benred8}{14}~shatter Leviathan\textquotesingle s power, and give him up as prey to the sea-beasts?\footnote[2]{vv. 12-15. Some think these verses refer to the escape of Israel from Egypt and the miracles done in the wilderness, Leviathan (perhaps the crocodile) being used as a symbol of Egypt. Others think the whole is an allegorical account of the Creation.} \textcolor{benred8}{15}~Thou didst open up fountains and streams of water; thou, too, madest the swollen rivers run dry. \textcolor{benred8}{16}~Thine is the day, thine the night; moon and sun are of thy appointment; \textcolor{benred8}{17}~thou hast fixed all the bounds of earth, madest the summer, madest the cool of the year.
\textcolor{benred8}{18}~Wilt thou take no heed, when thy enemies taunt thee, and in their recklessness set the divine name at defiance? \textcolor{benred8}{19}~Must the dove be the vulture\textquotesingle s prey? Souls unbefriended, but for thee, wilt thou leave us quite forgotten? \textcolor{benred8}{20}~Bethink thee of thy covenant; everywhere oppression lurks, or walks openly through the fields. \textcolor{benred8}{21}~Do not let the humble go away disappointed; teach the poor and the helpless to exalt thy name. \textcolor{benred8}{22}~Bestir thyself, O God, to vindicate thy own cause; do not forget the taunts which reckless men hurl at thee, day after day; \textcolor{benred8}{23}~do not overlook them, the triumphant shouts of thy enemies, the ever growing clamour that here defies thee.
\begin{large}\begin{center}\textsc{Chapter lxxiv}\end{center}\end{large}
(To the choir-master. Melody: Do not destroy. A psalm. Of Asaph. A song.)
\textcolor{benred8}{2}~We praise thee, O God, and, praising thee, call upon thy name, tell the story of thy wondrous deeds. \textcolor{benred8}{3}~When the time is ripe, I will judge strictly; \textcolor{benred8}{4}~earth rocks to its fall, and all that dwell on it; I alone support its fabric.\footnote[1]{vv. 3, 4. It is not clear whether the speaker is God or man, nor whether he is still the speaker in verses 5 sqq. The two verses are ordinarily interpreted as God\textquotesingle s own utterance, but a comparison of verse 10 below suggests that an earthly judge may be intended.} \textcolor{benred8}{5}~Rebel no more, I cry to the rebels, Abate your pride, to the transgressors; \textcolor{benred8}{6}~would they match themselves against the most High, hurl defiance at God? \textcolor{benred8}{7}~Look east, look west, it will avail you nothing; no help comes from the desert, or the high hills; \textcolor{benred8}{8}~it is God who rules all, humbling one man and exalting another. \textcolor{benred8}{9}~In the Lord\textquotesingle s hand foams a full cup of spiced wine; he holds it to men\textquotesingle s lips, that must empty it to the dregs, sinners everywhere must drink them. \textcolor{benred8}{10}~Evermore will I triumph, singing praises to the God of Jacob; \textcolor{benred8}{11}~mine to crush the pride of every sinner, and raise high the courage of the just.
\begin{large}\begin{center}\textsc{Chapter lxxv}\end{center}\end{large}
(To the choir-master. For stringed instruments. A psalm. Of Asaph. A song.)
\textcolor{benred8}{2}~It is in Juda God makes himself known, in Israel that his name is extolled; \textcolor{benred8}{3}~there, in Salem, he makes his abode, dwells in Sion. \textcolor{benred8}{4}~It was there he broke the archers\textquotesingle  volleys, broke shield, and sword, and battle array. \textcolor{benred8}{5}~How princely was thy dawning over the everlasting hills! \textcolor{benred8}{6}~Brave hearts, foiled of their purpose, they slept their long sleep; lay there, no hand stirring, the warriors in their pride; \textcolor{benred8}{7}~lay there, chariots and horsemen, overthrown, God of Jacob, at thy word of rebuke.
\textcolor{benred8}{8}~Who can resist thee, so terrible, so sudden in thy anger? \textcolor{benred8}{9}~Loud rings in heaven the doom thou utterest; earth trembles and is silent \textcolor{benred8}{10}~when God rouses himself to execute his sentence, giving redress to those who are scorned on earth. \textcolor{benred8}{11}~Its madness tamed, Edom shall do thee honour; to thee the spared remnant of Emath shall keep holiday. \textcolor{benred8}{12}~To the Lord your God let your vows be made and paid; bring gifts from every side to God, the terrible; \textcolor{benred8}{13}~he it is that cows the hearts of princes, feared among all the kings of the earth.
\begin{large}\begin{center}\textsc{Chapter lxxvi}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. Melody: Idithun. Of Asaph. A psalm.)
\textcolor{benred8}{2}~A cry to my God in loud appeal, a cry to my God, to win his hearing! \textcolor{benred8}{3}~To the Lord I look when distress comes upon me; in his presence I lift up my hands amid the darkness, never wearied; grief like mine there is no comforting. \textcolor{benred8}{4}~Of God I bethink me, yet sighing still; of God I muse, yet ever faint-hearted. \textcolor{benred8}{5}~Sleepless that thought holds me, yet bewildered and dumb. \textcolor{benred8}{6}~I reflect upon days long past, the immemorial years possess my mind; \textcolor{benred8}{7}~deep musings occupy my thoughts at midnight, never will my mind be at rest. \textcolor{benred8}{8}~Can it be that God will always leave us forsaken, will never shew us again his old kindness? \textcolor{benred8}{9}~Can his favour desert us altogether, his promise be set aside eternally? \textcolor{benred8}{10}~Can God forget to be gracious, can anger move him to withhold his mercy? \textcolor{benred8}{11}~For me, I tell myself, this sorrow was reserved; the most High has altered the fashion of his dealings with men.
\textcolor{benred8}{12}~To remember all the Lord has done, to recall those wonderful acts of thine, long ago! \textcolor{benred8}{13}~To ponder over all thy doings, pass thy wonders in review! \textcolor{benred8}{14}~Thy path, O God, is hedged about with holiness; what god is great as our God is great? \textcolor{benred8}{15}~Thy own wonderful acts acclaim thy Deity; even to the Gentiles thou wouldst make thy power known, \textcolor{benred8}{16}~by forcing them to set free thy people, the sons of Jacob and of Joseph. \textcolor{benred8}{17}~The waters saw thee, O God, the waters trembled at the sight of thee, moved to their inmost depths; \textcolor{benred8}{18}~how the waves roared, how the clouds volleyed rain, what echoes from their midst! To and fro thy arrows passed, \textcolor{benred8}{19}~thy crackling thunders rolled, till all the world shone with thy lightning, and the troubled earth shook. \textcolor{benred8}{20}~Thy way led through the sea, the deep tide made a road for thee, and none may read the traces of thy passage, \textcolor{benred8}{21}~where thou, with Moses and Aaron for thy shepherds, didst bring thy people out on their journey.
\begin{large}\begin{center}\textsc{Chapter lxxvii}\end{center}\end{large}
(A maskil. Of Asaph.)
\textcolor{benred8}{1}~Listen, my people, to this testament of mine, do not turn a deaf ear to the words I utter; \textcolor{benred8}{2}~I speak to you with mysteries for my theme, read the riddles of long ago. \textcolor{benred8}{3}~It is a story often heard, well known among us; have not our fathers told it to us? \textcolor{benred8}{4}~And shall we keep it back from their children, from the generation which follows? Speak we of God\textquotesingle s praise, of his great power, of the wonderful deeds he did. \textcolor{benred8}{5}~He gave Jacob a rule to live by, framed for Israel a law, commanding our fathers to hand on the message, \textcolor{benred8}{6}~so that a new generation might learn it; sons would be born to take their place, and teach it to their own sons after them. \textcolor{benred8}{7}~They were to put their trust in God, ever remembering his divine dealings with them, ever loyal to his commands; \textcolor{benred8}{8}~they were not to be like their fathers, a stubborn and defiant breed, a generation of false aims, of a spirit that broke faith with God.
\textcolor{benred8}{9}~So it was that the sons of Ephraim, bow in hand,\footnote[1]{\textasciigrave Ephraim\textquotesingle  is probably used here to represent the people of Israel, as often in the prophets, who contrast it with \textasciigrave Juda\textquotesingle  (cf. verse 67). The reference may be to the refusal to invade Palestine (Num. 14), but the chronology of this psalm is difficult to disentangle.} were routed in the day of battle. \textcolor{benred8}{10}~They were false to God\textquotesingle s covenant, refused to follow his law, \textcolor{benred8}{11}~as if they had forgotten all his mercies, all those wonderful deeds of his they had witnessed. \textcolor{benred8}{12}~Had not their fathers seen wonders enough in Egypt, on the plains of Tanis, \textcolor{benred8}{13}~when he parted the sea to let them pass through it, making its waters stand firm as a mound of earth; \textcolor{benred8}{14}~when he led them with a cloud by day, with glowing fire all through the night? \textcolor{benred8}{15}~He pierced the rock, too, in the desert, and slaked their thirst as if from some deep pool, \textcolor{benred8}{16}~bidding the very stones yield water, till fountains gushed from them, abundant as rivers.
\textcolor{benred8}{17}~And still they went on offending him, there in the wilderness, rebelling against the most High, \textcolor{benred8}{18}~challenging God in their thoughts to give them the food they craved for. \textcolor{benred8}{19}~Defiantly they asked, Can God spread a table for us in the wilderness? \textcolor{benred8}{20}~True, he smote the rock, and made water flow from it, till the stream ran in flood, but can he give bread too, and provide meat for his people? \textcolor{benred8}{21}~All this the Lord heard, and his indignation blazed out; its mounting fires Jacob had fed, its fury must break on Israel. \textcolor{benred8}{22}~What, had they no faith in God, no trust in his power to save? \textcolor{benred8}{23}~He laid his command upon the clouds above them, threw open the doors of heaven, \textcolor{benred8}{24}~and rained down manna for them to eat. The bread of heaven was his gift to them; \textcolor{benred8}{25}~man should eat the food of angels,\footnote[2]{Literally \textasciigrave of the strong\textquotesingle ; but the Angels are probably meant; cf.Wisd.16.10.} and so their want should be supplied abundantly. \textcolor{benred8}{26}~Next, he summoned his east wind from the sky: it was his power brought the southern gale, \textcolor{benred8}{27}~raining down meat on them thick as dust, birds on the wing, plentiful as the sea-sand. \textcolor{benred8}{28}~Into their very camp it fell, close about their tents; \textcolor{benred8}{29}~and they ate, and took their fill. All they asked, he granted them; \textcolor{benred8}{30}~and now, their craving still unsatisfied, while the food was yet in their mouths, \textcolor{benred8}{31}~God\textquotesingle s anger against them reached its height, and slew their lordliest, brought them low, all the flower of Israel.
\textcolor{benred8}{32}~Yet, with all this, they continued to offend him; all his wonderful deeds left them faithless still. \textcolor{benred8}{33}~And ever he took away their lives untimely, hurried their days to an end. \textcolor{benred8}{34}~When he threatened them with death, they would search after him, feel their need of God once more; \textcolor{benred8}{35}~they would remind themselves that it was God who had protected them, his almighty power that had delivered them. \textcolor{benred8}{36}~But still they were lying lips, they were false tongues that spoke to him; \textcolor{benred8}{37}~their hearts were not true to him, no loyalty bound them to his covenant. \textcolor{benred8}{38}~Yet, such is his mercy, he would still pardon their faults, and spare them from destruction; again and again he curbed his indignation, to his vengeance would not give place. \textcolor{benred8}{39}~He would not forget that they were flesh and blood, no better than a breath of wind, that passes by and never returns. \textcolor{benred8}{40}~How often the desert saw them in revolt against him, how often, in those solitudes, they defied his anger! \textcolor{benred8}{41}~Always new challenges to God\textquotesingle s power, new rebellions against the Holy One of Israel.
\textcolor{benred8}{42}~Had they forgotten all he did for them, that day when he set them free from the power of their oppressor, \textcolor{benred8}{43}~all those miracles among the men of Egypt, those portents in the plain of Tanis, \textcolor{benred8}{44}~when he turned all their streams, all their channels into blood, so that they could not drink? \textcolor{benred8}{45}~He sent out flies, to their ruin, frogs to bring devastation on them, \textcolor{benred8}{46}~gave all their harvest over to the caterpillar, their tillage to the locust, \textcolor{benred8}{47}~sent hail on their vineyards, frost on their mulberry-trees, \textcolor{benred8}{48}~let the hail have its way with their cattle, the lightning with their flocks. \textcolor{benred8}{49}~He let his anger loose on them in all its vehemence; what rage, what fury, what havoc, as the angels of destruction thronged about them! \textcolor{benred8}{50}~So, the way made ready for his vengeance, he took toll of their lives, doomed even their cattle to the pestilence; \textcolor{benred8}{51}~on every first-born creature in Egypt, on the first-fruits of increase in all the dwellings of Cham, his stroke fell. \textcolor{benred8}{52}~Then, like a shepherd, he set his own people on their way, led them, his own flock, through the wilderness; \textcolor{benred8}{53}~guided them in safety, free from all alarm, while the sea closed over their enemy. \textcolor{benred8}{54}~So he brought them to that holy land of his, the mountain slopes he took, with his own right hand for title; so he drove out the heathen at their onset, parcelled out the land to them by lot, to each his own inheritance, \textcolor{benred8}{55}~bidding the tribes of Israel dwell where the heathen had dwelt before them.
\textcolor{benred8}{56}~These were the men who defied the most high God, and rebelled against him; would not observe his decrees, \textcolor{benred8}{57}~but turned away and broke faith with him as their fathers had done, like a bow that plays the archer false; \textcolor{benred8}{58}~made mountain shrines, to court his anger, carved images, to awake his jealousy! \textcolor{benred8}{59}~The Lord heard the bruit of it, and burned with anger, cast Israel away in bitter scorn; \textcolor{benred8}{60}~he forsook his tabernacle in Silo, that tabernacle where once he dwelt among men. \textcolor{benred8}{61}~Plunder, now, in the enemy\textquotesingle s hands, the ark that is shrine of his strength and majesty; \textcolor{benred8}{62}~he would leave his people at the mercy of the sword, disdain his own inheritance. \textcolor{benred8}{63}~Their young men fed the flames, and the maidens must go unwed; \textcolor{benred8}{64}~their priests fell by the sword, and never a widow left to mourn for them.
\textcolor{benred8}{65}~Then suddenly, like a man that wakes up from sleep, like some warrior that lay, till now, bemused with wine, the Lord roused himself; \textcolor{benred8}{66}~he smote his enemies as they turned to flee,\footnote[3]{Or possibly \textasciigrave in their buttocks\textquotesingle  (I Kg. 5.9).} branded them for ever with shame. \textcolor{benred8}{67}~But he refused, now, to make his dwelling with Joseph, it was not the tribe of Ephraim he would choose; \textcolor{benred8}{68}~he chose the tribe of Juda, and the hill of Sion, there to bestow his love. \textcolor{benred8}{69}~And there he built his sanctuary, immovable as heaven or earth, his own unchanging handiwork. \textcolor{benred8}{70}~He chose David, too, for his servant; took him away from herding the sheep; bade him leave off following the ewes that were in milk, \textcolor{benred8}{71}~and be the shepherd of Jacob\textquotesingle s sons, his own people, of Israel, his own domain. \textcolor{benred8}{72}~His was the loyal heart that should tend them, his the skilful hand that should be their guide.
\begin{large}\begin{center}\textsc{Chapter lxxviii}\end{center}\end{large}
(A psalm. Of Asaph.)
\textcolor{benred8}{1}~O God, the heathen have broken into thy inheritance; they have profaned the temple, thy sanctuary, and left Jerusalem in ruins. \textcolor{benred8}{2}~They have thrown the corpses of thy servants to feed all the birds of heaven; wild beasts prey on the carrion of the just; \textcolor{benred8}{3}~blood has flowed like water on every side of Jerusalem, and there was none to bury the dead. \textcolor{benred8}{4}~What a triumph was this for the nations that dwell around us; how have our neighbours mocked and derided us! \textcolor{benred8}{5}~Lord, must we always taste thy vengeance, must thy jealous anger still burn unquenched? \textcolor{benred8}{6}~Pour out this indignation of thine upon the nations that do not acknowledge thee, on the kingdoms that never invoke thy name; \textcolor{benred8}{7}~see how they have made Jacob their prey, and left his dwelling-place in ruins! \textcolor{benred8}{8}~Forget the long record of our sins, and haste in mercy to our side; never was need so sore as this.
\textcolor{benred8}{9}~O God, our Saviour, help us; deliver us, Lord, for the glory of thy name, pardon our sins for the sake of thy own renown! \textcolor{benred8}{10}~Shall the heathen ask, What has become of their God? Shall our eyes never witness thy vengeance upon the Gentiles, that open vengeance thou wilt take for thy servants\textquotesingle  blood? \textcolor{benred8}{11}~Could but the groaning of the captive reach thy presence! Thy arm has not lost its strength; from our bonds deliver us, a race doomed to die. \textcolor{benred8}{12}~Pour out seven-fold retribution into the laps of our neighbours, for all the insults, Lord, which they have put upon thee; \textcolor{benred8}{13}~and we, thy own people, sheep of thy pasturing, will give thee thanks for ever, echo, from one generation to the next, the story of thy renown.
\begin{large}\begin{center}\textsc{Chapter lxxix}\end{center}\end{large}
(To the choir-master. Melody: The Lily of the Law. Of Asaph. A psalm.)
\textcolor{benred8}{2}~Give audience, thou that art the guide of Israel, that leadest Joseph with a shepherd\textquotesingle s care. Thou who art enthroned above the Cherubim, \textcolor{benred8}{3}~reveal thyself to Ephraim, Benjamin, and Manasses; exert thy sovereign strength, and come to our aid. \textcolor{benred8}{4}~O God, restore us to our own; smile upon us, and we shall find deliverance. \textcolor{benred8}{5}~Lord God of hosts, wilt thou always turn away in anger from thy servants\textquotesingle  prayer; \textcolor{benred8}{6}~daily wilt thou allot us, for food, for drink, only the full measure of our tears? \textcolor{benred8}{7}~Thou hast made us a coveted prize to our neighbours, enemies mock at our ill fortune! \textcolor{benred8}{8}~O God of hosts, restore us to our own; smile upon us, and we shall find deliverance. \textcolor{benred8}{9}~Long ago, thou didst bring a vine out of Egypt, rooting out the heathen to plant it here; \textcolor{benred8}{10}~thou didst prepare the way for its spreading, and it took root where thou hadst planted it, filled the whole land. \textcolor{benred8}{11}~How it overshadowed the hills, how the cedars, divinely tall, were overtopped by its branches! \textcolor{benred8}{12}~It spread out its tendrils to the sea, its shoots as far as the great river. \textcolor{benred8}{13}~Why is it that in these days thou hast levelled its wall, for every passer-by to rob it of its fruit? \textcolor{benred8}{14}~See how the wild boar ravages it, how it gives pasture to every beast that roams! \textcolor{benred8}{15}~God of hosts, relent, look down from heaven, look to this vine, that needs thy care. \textcolor{benred8}{16}~Revive the stock which thy own hand has planted, branches that by thee throve, and throve for thee. \textcolor{benred8}{17}~Death be in thy frown for the men that have cut it down and burned it. \textcolor{benred8}{18}~Thy chosen friends, a race\footnote[1]{Literally, \textasciigrave the man of thy right hand, the son of man whom thou hast strengthened for thyself\textquotesingle . This probably refers to Israel personified.} by thee thriving, and thriving for thee, O let thy hand protect them still! \textcolor{benred8}{19}~Henceforth we will never forsake thee; grant us life, and we will live only to invoke thy name. \textcolor{benred8}{20}~Lord God of hosts, restore us to our own; smile upon us, and we shall find deliverance.
\begin{large}\begin{center}\textsc{Chapter lxxx}\end{center}\end{large}
(To the choir-master. Melody: The Winepresses. Of Asaph.)
\textcolor{benred8}{2}~Rejoice we all in honour of the God who aids us; cry out with gladness to the God of Jacob. \textcolor{benred8}{3}~Ring psaltery, and tambour, beat the harp, sweetly sounding, and the zither! \textcolor{benred8}{4}~A new month, and a full moon; blow the trumpet loud, to grace our festival! \textcolor{benred8}{5}~Duty demands it of Israel; the God of Jacob has decreed it, \textcolor{benred8}{6}~made it a law for Joseph, since the day he left Egypt, and gained the further shore. In a tongue unknown the message came to me; \textcolor{benred8}{7}~I have eased his shoulder of the burden, freed his hands from the slavery of the hod! \textcolor{benred8}{8}~Such deliverance I brought, when thou didst cry out to me in thy misery; gave thee audience under a canopy of cloud, and tested thee at the Waters of Rebellion.
\textcolor{benred8}{9}~Give heed, my people, to this warning of mine; Israel, wouldst thou but listen! \textcolor{benred8}{10}~Then let no strange worship find a home with thee; never let thy knees be bowed to an alien God; \textcolor{benred8}{11}~am not I the Lord thy God, I, who rescued thee from Egypt? Open thy mouth wide, and thou shalt have thy fill. \textcolor{benred8}{12}~So I spoke, but my people would not listen; Israel went on unheeding, \textcolor{benred8}{13}~till I was fain to give their hard hearts free play, let them follow their own devices. \textcolor{benred8}{14}~Ah, if my people did but listen to me! Did Israel but take me for their guide! \textcolor{benred8}{15}~How lightly, then, would I bring their enemies low, smite down their persecutors! \textcolor{benred8}{16}~The very men that were once the Lord\textquotesingle s enemies would be cringing at his feet; such, for ever, should be their destiny; \textcolor{benred8}{17}~Israel should have full ears of wheat to nourish them, and honey dripping from the rock to their heart\textquotesingle s content.
\begin{large}\begin{center}\textsc{Chapter lxxxi}\end{center}\end{large}
(A psalm. Of Asaph.)
\textcolor{benred8}{1}~See, where he stands, the Ruler of all, among the rulers\footnote[1]{Literally \textasciigrave gods\textquotesingle  as in verse 6, the Divine name being specially applied in this way to judges (cf. Ex. 21.6).} assembled, comes for-ward to pronounce judgement on the rulers themselves! \textcolor{benred8}{2}~Will you never cease perverting justice, espousing the cause of the wicked? \textcolor{benred8}{3}~Come, give redress to the poor and the friendless, do right to the afflicted and the destitute; \textcolor{benred8}{4}~to you need and poverty look for deliverance, rescue them from the hand of the wickedness. \textcolor{benred8}{5}~But no, ignorant and unperceiving, they grope their way in darkness; see how unstable are the props of earth! \textcolor{benred8}{6}~Gods you are, I myself have declared it; favoured children, every one of you, of the most High;\footnote[2]{Jn. 10.34.} \textcolor{benred8}{7}~yet the doom of mortals awaits you, you shall fall with the fall of human princes. \textcolor{benred8}{8}~Bestir thyself, Lord, bring the world to judgement; all the nations are thy own domain.
\begin{large}\begin{center}\textsc{Chapter lxxxii}\end{center}\end{large}
(A song. A psalm. Of Asaph.)
\textcolor{benred8}{2}~Be silent, Lord, no longer. O God, do not keep still now, do not hold back now! \textcolor{benred8}{3}~What turmoil among thy enemies; how their malice lifts its head! \textcolor{benred8}{4}~Busily they plot against thy people, compass the ruin of the men thou hast in thy keeping. \textcolor{benred8}{5}~Come, they whisper, let us put an end to their sovereignty, so that the very name of Israel will be remembered no more. \textcolor{benred8}{6}~All are agreed, all alike are ranged in confederacy against thee; \textcolor{benred8}{7}~here Edom lies encamped, there Ismael; Moab, too, and the Agarenes; \textcolor{benred8}{8}~Gebal, Ammon and Amelec, the Philistines, and the folk that dwell at Tyre. \textcolor{benred8}{9}~Even Assyria has made common cause with them, lends her aid to these children of Lot.\footnote[1]{vv. 7-9. It is not certain what was the occasion of this psalm. The names given suggest a simultaneous attack on the Jews by all their neighbours; the children of Lot are the Moabites and the Ammonites (Gen. 19.37).}
\textcolor{benred8}{10}~Do to these what thou didst to Madian, to Sisara and Jabin at the brook of Cison; \textcolor{benred8}{11}~the men who died at Endor, rotted there like dung on the ground.\footnote[2]{vv. 10-12. See Jg. 4; 7.25; 8.} \textcolor{benred8}{12}~May their princes fare as Oreb fared, and Zeb; may the doom of Zebee and Salmana be the doom of all their chieftains. \textcolor{benred8}{13}~And did they think to make God\textquotesingle s chosen portion their spoil? \textcolor{benred8}{14}~My God, send them whirling this way and that, like leaves, like straws before the wind. \textcolor{benred8}{15}~See how the fire burns up the forest, how its flames scorch the mountain-side! \textcolor{benred8}{16}~So let the fury of thy onset rout them, thy fury dismay them. \textcolor{benred8}{17}~Let their cheeks blush crimson with shame, Lord, till they come to sue for thy favour; \textcolor{benred8}{18}~confusion and dismay be theirs for ever, for ever let them be abashed and brought to nothing, \textcolor{benred8}{19}~till they, too, know the meaning of the divine name, acknowledge thee as the most high God, the Overlord of earth.
\begin{large}\begin{center}\textsc{Chapter lxxxiii}\end{center}\end{large}
(To the choir-master. Melody: The Winepresses. Of the sons of Core. A psalm.)
\textcolor{benred8}{2}~Lord of hosts, how I love thy dwelling-place! \textcolor{benred8}{3}~For the courts of the Lord\textquotesingle s house, my soul faints with longing. The living God! at his name my heart, my whole being thrills with joy. \textcolor{benred8}{4}~Where else should the sparrow find a home, the swallow a nest for her brood, but at thy altar, Lord of hosts, my king and my God? \textcolor{benred8}{5}~How blessed, Lord, are those who dwell in thy house! They will be ever praising thee. \textcolor{benred8}{6}~How blessed is the man who finds his strength in thee! Where there are hearts set on pilgrimage, \textcolor{benred8}{7}~the parched ravine turns into a water-course at their coming, new-clad by the bounty of returning rain. \textcolor{benred8}{8}~So, at each stage refreshed, they will reach Sion, and have sight there of the God who is above all gods.
\textcolor{benred8}{9}~Lord of hosts, listen to my prayer; God of Israel, grant me audience! \textcolor{benred8}{10}~God, ever our protector, do not disregard us now; look favourably upon him whom thou hast anointed!\footnote[1]{It is not certain whether King David, or the people of Israel, is here referred to as the anointed (or Christ) of God.} \textcolor{benred8}{11}~Willingly would I give a thousand of my days for one spent in thy courts! Willingly reach but the threshold of my God\textquotesingle s house, so I might dwell no more in the abode of sinners! \textcolor{benred8}{12}~Sun to enlighten, shield to protect us, the Lord God has favour, has honour to bestow. \textcolor{benred8}{13}~To innocent lives he will never refuse his bounty; Lord of hosts, blessed is the man who puts his confidence in thee.
\begin{large}\begin{center}\textsc{Chapter lxxxiv}\end{center}\end{large}
(To the choir-master. Of the sons of Core. A psalm.)
\textcolor{benred8}{2}~What blessings, Lord, thou hast granted to this land of thine, restoring Jacob\textquotesingle s fortunes, \textcolor{benred8}{3}~pardoning thy people\textquotesingle s guilt, burying away the record of their sins, \textcolor{benred8}{4}~all thy anger calmed, thy fierce displeasure forgotten! \textcolor{benred8}{5}~And now, God of our deliverance, do thou restore us; no longer let us see thy frown. \textcolor{benred8}{6}~Wouldst thou always be indignant with us? Must thy resentment smoulder on, age after age? \textcolor{benred8}{7}~Wilt thou never relent, O God, and give fresh life, to rejoice the spirits of thy people? \textcolor{benred8}{8}~Shew us thy mercy, Lord; grant us thy deliverance!
\textcolor{benred8}{9}~Let me listen, now, to the voice of the Lord God; it is a message of peace he sends to his people; to his loyal servants, that come back, now, with all their heart to him. \textcolor{benred8}{10}~For us, his worshippers, deliverance is close at hand; in this land of ours, the divine glory is to find a home. \textcolor{benred8}{11}~See, where mercy and faithfulness meet in one; how justice and peace are united in one embrace! \textcolor{benred8}{12}~Faithfulness grows up out of the earth, and from heaven, redress looks down.\footnote[1]{vv. 11, 12. These verses perhaps imply that God\textquotesingle s fidelity to his promises and God\textquotesingle s mercy have combined to restore the Jewish people; that his justice, satisfied with the expiation of their sins, no longer grudges them peace and well-being; that loyalty to the old covenant is once more a native growth in the land of Israel, and that God looks down to reward it. But the picture may be a more general one.} \textcolor{benred8}{13}~The Lord, now, will grant us his blessing, to make our land yield its harvest; \textcolor{benred8}{14}~justice will go on before him, deliverance follow where his feet tread.
\begin{large}\begin{center}\textsc{Chapter lxxxv}\end{center}\end{large}
(A prayer. Of David.)
\textcolor{benred8}{1}~Turn thy ear, Lord, and listen to me in my helplessness and my need. \textcolor{benred8}{2}~Protect a life dedicated to thyself; rescue a servant of thine that puts his trust in thee. \textcolor{benred8}{3}~In thee, my own God; have mercy, O Lord, for mercy I plead continually; \textcolor{benred8}{4}~comfort thy servant\textquotesingle s heart, this heart that aspires, Lord, to thee. \textcolor{benred8}{5}~Who is so kind and forgiving, Lord, as thou art, who so rich in mercy to all who invoke him? \textcolor{benred8}{6}~Give a hearing, then, Lord, to my prayer; listen to my plea \textcolor{benred8}{7}~when I cry out to thee in a time of sore distress, counting on thy audience. \textcolor{benred8}{8}~There is none like thee, Lord, among the gods; none can do as thou doest. \textcolor{benred8}{9}~Lord, all the nations thou hast made must needs come and worship thee, honouring thy name, \textcolor{benred8}{10}~so great thou art, so marvellous in thy doings, thou who alone art God.
\textcolor{benred8}{11}~Guide me, Lord, thy own way, thy faithful care my escort; be all my heart\textquotesingle s direction reverence for thy name. \textcolor{benred8}{12}~O Lord my God, with all my heart I will praise thee, eternally hold thy name in honour \textcolor{benred8}{13}~for the greatness of the mercy thou hast shewed me, in rescuing me thus from the lowest depths of hell. \textcolor{benred8}{14}~And now, O God, see how scornful foes have set upon me, how their dread conspiracy threatens my life, with no thought of thee to restrain it! \textcolor{benred8}{15}~But thou, Lord, art a Lord of mercy and pity, patient, full of compassion, true to thy promise. \textcolor{benred8}{16}~Look upon me and be merciful to me; rescue, with thy sovereign aid, one whose mother bore him to thy service! \textcolor{benred8}{17}~Shew me some token of thy favour; let my enemies see, abashed, how thou, Lord, dost help me, how thou, Lord, dost comfort me.
\begin{large}\begin{center}\textsc{Chapter lxxxvi}\end{center}\end{large}
(Of the sons of Core. A psalm. A song.)
\textcolor{benred8}{1}~His own building amidst the inviolate hills, \textcolor{benred8}{2}~dearer to the Lord are Sion walls than any other home in Israel. \textcolor{benred8}{3}~How high a boast, city of God, is made for thee, \textcolor{benred8}{4}~Mine it is to reckon the folk of Egypt, of Babylon, too, among my citizens! Philistines, Tyrians, Ethiopians, all must claim Sion as their birthplace; \textcolor{benred8}{5}~None was ever born, the proverb shall run, that did not take his birth from her; it was the most High, none other, that founded her. \textcolor{benred8}{6}~This was their birthplace, the Lord shall write over the muster-roll of the nations; \textcolor{benred8}{7}~nor any but shall tell her praises with song and dance, each claiming from her its only origin.\footnote[1]{The Latin version here rendered involves several corrections of the Hebrew text, which is very obscure.}
\begin{large}\begin{center}\textsc{Chapter lxxxvii}\end{center}\end{large}
(A song. A psalm. Of the sons of Core. To the choir-master. Melody: Mahalat. For singing. A maskil. Of Heman the Ezrahite.)
\textcolor{benred8}{2}~Lord God, day and night I cry bitterly to thee; \textcolor{benred8}{3}~let my prayer reach thy pre-sence, give audience to my entreaty, \textcolor{benred8}{4}~for indeed my heart is full of trouble. My life sinks ever closer to the grave; \textcolor{benred8}{5}~I count as one of those who go down into the abyss, like one powerless. \textcolor{benred8}{6}~As well lie among the dead, men laid low in the grave, men thou rememberest no longer, cast away, now, from thy protecting hand. \textcolor{benred8}{7}~Such is the place where thou hast laid me, in a deep pit where the dark waters swirl; \textcolor{benred8}{8}~heavily thy anger weighs down on me, and thou dost overwhelm me with its full flood. \textcolor{benred8}{9}~Thou hast estranged all my acquaintance from me, so that they treat me as a thing accursed; I lie in a prison whence there is no escape, \textcolor{benred8}{10}~my eyes grow dim with tears. On thee I call, to thee stretch out my hands, each day that passes.
\textcolor{benred8}{11}~Not for the dead thy wonderful power is shewn; not for pale shadows to return and give thee thanks. \textcolor{benred8}{12}~There in the grave, how shall they recount thy mercies; how shall they tell of thy faithfulness, now that life is gone? \textcolor{benred8}{13}~How can there be talk of thy marvels in a world of darkness, of thy favour in a land where all is forgotten? \textcolor{benred8}{14}~To prayer, Lord, fall I lustily; it shall reach thee, while there is yet time. \textcolor{benred8}{15}~Why dost thou reject my plea, Lord, and turn thy face away from me? \textcolor{benred8}{16}~Ever since youth, misery and mortal sickness have been my lot; wearily I have borne thy visitations; \textcolor{benred8}{17}~I am overwhelmed with thy anger, dismayed by thy threats, \textcolor{benred8}{18}~that still cut me off like a flood, all at once surrounding me. \textcolor{benred8}{19}~Friends and neighbours gone, a world of shadows is all my company.
\begin{large}\begin{center}\textsc{Chapter lxxxviii}\end{center}\end{large}
(A maskil. Of Ethan the Ezrahite.)
\textcolor{benred8}{2}~Here is a song to put the Lord\textquotesingle s mercies on record for ever; ages will pass, and still these words of mine shall proclaim thy faithfulness. \textcolor{benred8}{3}~Charter of everlasting mercy thy own lips have given; there, in the heavens, thy faithful promise rests: \textcolor{benred8}{4}~I have made a sworn covenant with my chosen servant David: \textcolor{benred8}{5}~To all time I will make thy posterity continue, age after age I will bid thy throne endure.
\textcolor{benred8}{6}~And are not those heavens, Lord, witnesses of thy wonderful power, of thy faithfulness, before the court of the holy ones? \textcolor{benred8}{7}~Who is there above the clouds to rival the Lord; where is the Lord\textquotesingle s like among all the sons of God? \textcolor{benred8}{8}~How is God feared, in that assembly of the holy ones; how great he is, how high in reverence above all that stand about him! \textcolor{benred8}{9}~Lord God of hosts, who can compare with thee; in the power, Lord, that is thine, in the faithfulness that everywhere attends thee? \textcolor{benred8}{10}~It is thou that dost curb the pride of the sea, and calm the tumult of its waves; \textcolor{benred8}{11}~wounded lies Rahab\footnote[1]{\textasciigrave Insolence\textquotesingle  in the original is \textasciigrave Rahab\textquotesingle , either referring to Egypt (as in 86.4), or personifying the principle of Chaos destroyed by the Creation (73.13).} at thy feet, by the strong arm that has routed thy enemies. \textcolor{benred8}{12}~Thine are the heavens, thine the earth; author, thou, of the world and all it holds. \textcolor{benred8}{13}~The north wind and the south are of thy fashioning; thy name wakes the glad echoes of Thabor and Hermon. \textcolor{benred8}{14}~God of the strong arm, the sure, the uplifted hand, \textcolor{benred8}{15}~right and justice are the pillars of thy throne; mercy and faithfulness the heralds of thy coming.
\textcolor{benred8}{16}~Happy is the people that knows well the shout of praise, that lives, Lord, in the smile of thy protection! \textcolor{benred8}{17}~Evermore they take pride in thy name, rejoice over thy just dealings. \textcolor{benred8}{18}~What else but thy glory inspires their strength? What else but thy favour bids us lift our heads? \textcolor{benred8}{19}~From the Lord, the Holy One of Israel, that royal protection comes which is our shield.
\textcolor{benred8}{20}~Long ago, in a vision, thou didst make a promise to thy faithful servants. Thou saidst, I have crowned you a warrior king, chosen out among the common folk a man to honour. \textcolor{benred8}{21}~Here was my servant David; on him my consecrating oil has been poured. \textcolor{benred8}{22}~My hand shall never leave him unprotected, my arm shall give him courage; \textcolor{benred8}{23}~no enemy shall take him unawares, no envious rival have power, henceforth, to crush him; \textcolor{benred8}{24}~beaten down, every foe, at his onset, baffled, all their ill will. \textcolor{benred8}{25}~My faithfulness and mercy shall go with him; by my favour he shall rise to pre-eminence. \textcolor{benred8}{26}~I will make his power rest on the sea; to the streams of the great river his hand shall reach out.\footnote[2]{\textasciigrave The great river\textquotesingle  is Euphrates, as in 71.8, 79.12.} \textcolor{benred8}{27}~Thou art my Father, he will cry out to me, thou art my God, my stronghold and my refuge; \textcolor{benred8}{28}~and I will acknowledge him as my first-born, overlord to all the kings of earth. \textcolor{benred8}{29}~I will continue my favour towards him for ever, my covenant with him shall remain unbroken; \textcolor{benred8}{30}~I will give him a posterity that never fails, a throne enduring as heaven itself. \textcolor{benred8}{31}~Do his children forsake my law, to follow paths not mine; \textcolor{benred8}{32}~do they violate my decrees, leave my will undone? \textcolor{benred8}{33}~Then they shall feel the rod for their transgressions, I will scourge them for their sin, \textcolor{benred8}{34}~but I will not cancel my gracious promise to him; \textcolor{benred8}{35}~never will I be guilty of unfaithfulness, never will I violate my covenant, or alter the decree once spoken. \textcolor{benred8}{36}~Pledged stands my inviolable word, I will never be false to David; \textcolor{benred8}{37}~his posterity shall continue for ever, \textcolor{benred8}{38}~his royalty, too, shall last on in my presence like the sun; like the moon\textquotesingle s eternal orb, that bears witness in heaven unalterable.
\textcolor{benred8}{39}~And now? Now thou hast only loathing and scorn for us; heavy thy hand falls on him thou hast anointed. \textcolor{benred8}{40}~Spurned lies the covenant thou didst make with thy servant, thou hast dishonoured his royalty in the dust, \textcolor{benred8}{41}~broken down all the walls about him, and made a ruin of his stronghold, \textcolor{benred8}{42}~till he is plundered by every passer-by, a laughing-stock to all his neighbours. \textcolor{benred8}{43}~Thou hast granted aid to the attacking armies, triumph to all his enemies, \textcolor{benred8}{44}~foiling the thrust of his sword, and denying him thy succour in battle. \textcolor{benred8}{45}~Thou hast robbed him of the bright glory\footnote[3]{Literally, \textasciigrave purification\textquotesingle .} that once was his; thou hast cast down his throne to earth, \textcolor{benred8}{46}~cut his manhood short before its time; confusion overwhelms him.
\textcolor{benred8}{47}~Lord, wilt thou always turn thy face away so obdurately, will the flame of thy anger never be quenched? \textcolor{benred8}{48}~Remember how frail a thing I am, how brief a destiny thou hast granted to all Adam\textquotesingle s sons. \textcolor{benred8}{49}~Where is the man that can live on, and leave death untasted; can ransom his life from the power of the world beneath? \textcolor{benred8}{50}~Lord, where are those mercies of an earlier time, promised so faithfully to David? \textcolor{benred8}{51}~Remember how a world\textquotesingle s taunts assail thy people, and this one heart must bear them all; \textcolor{benred8}{52}~shall they hurl taunts, Lord, these, thy enemies, after the man thou thyself hast anointed?
\textcolor{benred8}{53}~Blessed be the Lord for ever. Amen, Amen.
\begin{large}\begin{center}\textsc{Chapter lxxxix}\end{center}\end{large}
(A prayer. Of Moses, the man of God.)
\textcolor{benred8}{1}~Lord, thou hast been our refuge from generation to generation. \textcolor{benred8}{2}~Before the hills came to birth, before the whole frame of the world was engendered, from eternity to eternity, O God, thou art. \textcolor{benred8}{3}~And wilt thou bring man to dust again, that thou sayest, Return, children of Adam, to what you were? \textcolor{benred8}{4}~In thy sight, a thousand years are but as yesterday, that has come and gone, or as one of the night-watches. \textcolor{benred8}{5}~Swiftly thou bearest our lives away, as a waking dream, \textcolor{benred8}{6}~or the green grass that blooms fresh with the morning; night finds it faded and dead. \textcolor{benred8}{7}~Still thy anger takes toll of us, thy displeasure denies us rest, \textcolor{benred8}{8}~so jealous thy scrutiny of our wrong-doing, so clear our hidden sins shew in the light of thy presence. \textcolor{benred8}{9}~Day after day vanishes, and still thy anger lasts; swift as a breath our lives pass away. \textcolor{benred8}{10}~What is our span of days? Seventy years it lasts, eighty years, if lusty folk we be; for the more part, toil and frustration; years that vanish in a moment, and we are gone. \textcolor{benred8}{11}~Alas, that so few heed thy vengeance, \textcolor{benred8}{12}~measure thy anger by the reverence we owe thee! Teach us to count every passing day, till our hearts find wisdom.
\textcolor{benred8}{13}~Relent, Lord; must it be for ever? Be gracious to thy servants. \textcolor{benred8}{14}~For us thy timely mercies, for us abiding happiness and content; \textcolor{benred8}{15}~happiness that shall atone for the time when thou didst afflict us, for the long years of ill fortune. \textcolor{benred8}{16}~Let these eyes see thy purpose accomplished, to our own sons reveal thy glory; \textcolor{benred8}{17}~the favour of the Lord our God smile on us! Prosper our doings, Lord, prosper our doings yet.
\begin{large}\begin{center}\textsc{Chapter xc}\end{center}\end{large}
\lettrine[lines=2]{C}{ontent} if thou be to live with the most High for thy defence, under his Almighty shadow abiding still, \textcolor{benred8}{2}~him thy refuge, him thy stronghold thou mayst call, thy own God, in whom is all thy trust. \textcolor{benred8}{3}~He it is will rescue thee from every treacherous lure, every destroying plague. \textcolor{benred8}{4}~His wings for refuge, nestle thou shalt under his care, \textcolor{benred8}{5}~his faithfulness thy watch and ward. Nothing shalt thou have to fear from nightly terrors, \textcolor{benred8}{6}~from the arrow that flies by day-light, from pestilence that walks to and fro in the darkness, from the death that wastes under the noon. \textcolor{benred8}{7}~Though a thousand fall at thy side, ten thousand at thy right hand, it shall never come next or near thee; \textcolor{benred8}{8}~rather, thy eyes shall look about thee, and see the reward of sinners.
\textcolor{benred8}{9}~He, the Lord, is thy refuge; thou hast found a stronghold in the most High. \textcolor{benred8}{10}~There is no harm that can befall thee, no plague that shall come near thy dwelling. \textcolor{benred8}{11}~He has given charge to his angels concerning thee, to watch over thee wheresoever thou goest; \textcolor{benred8}{12}~they will hold thee up with their hands lest thou shouldst chance to trip on a stone. \textcolor{benred8}{13}~Thou shalt tread safely on asp and adder, crush lion and serpent under thy feet.
\textcolor{benred8}{14}~He trusts in me, mine it is to rescue him; he acknowledges my name, from me he shall have protection; \textcolor{benred8}{15}~when he calls upon me, I will listen, in affliction I am at his side, to bring him safety and honour. \textcolor{benred8}{16}~Length of days he shall have to content him, and find in me deliverance.
\begin{large}\begin{center}\textsc{Chapter xci}\end{center}\end{large}
(A psalm. A song. On the sabbath day.)
\textcolor{benred8}{2}~Sweet it is to praise the Lord, to sing, most high God, in honour of thy name; \textcolor{benred8}{3}~to proclaim thy mercy and faithfulness at daybreak and at the fall of night. \textcolor{benred8}{4}~Here is a theme for ten-stringed harp and viol, for music of voice and zither; \textcolor{benred8}{5}~so delightsome, Lord, is all thou doest, so thrills my heart at the sight of all thou hast made. \textcolor{benred8}{6}~How magnificent is thy creation, Lord, how unfathomable are thy purposes! \textcolor{benred8}{7}~And still, too dull to learn, too slow to grasp his lesson, the wrong-doer goes on in his busy wickedness. \textcolor{benred8}{8}~Still he thrives, makes a brave show like the grass in spring, yet is he doomed to perish eternally, \textcolor{benred8}{9}~whilst thou, Lord, art for ever exalted on high. \textcolor{benred8}{10}~Vanished away thy enemies, Lord, vanished away, and all their busy wickedness scattered to the winds!
\textcolor{benred8}{11}~Strength thy power gives me, that gives strength to the wild oxen; refreshes me as with the touch of pure oil. \textcolor{benred8}{12}~Blessed are these eyes with the sight of my enemies\textquotesingle  downfall, these ears with the tidings of insolent malice defeated. \textcolor{benred8}{13}~The innocent man will flourish as the palm-tree flourishes; he will grow to greatness as the cedars grow on Lebanon. \textcolor{benred8}{14}~Planted in the temple of the Lord, growing up in the very courts of our God\textquotesingle s house, \textcolor{benred8}{15}~the innocent will flourish in a green old age, all freshness and vigour still; \textcolor{benred8}{16}~theirs to proclaim how just is the Lord my refuge, his dealings how clear of wrong.
\begin{large}\begin{center}\textsc{Chapter xcii}\end{center}\end{large}
\lettrine[lines=2]{T}{he} Lord reigns as king, robed in majesty; royalty the Lord has for robe and girdle. He it was that founded the solid earth, to abide immovable. \textcolor{benred8}{2}~Firm stood thy throne ere ever the world began; from all eternity, thou art. \textcolor{benred8}{3}~Loud the rivers echo, Lord, loud the rivers echo, crashing down in flood. \textcolor{benred8}{4}~Magnificent the roar of eddying waters; magnificent the sea\textquotesingle s rage; magnificent above these, the Lord reigns in heaven. \textcolor{benred8}{5}~How faithful, Lord, are thy promises! Holy is thy house, and must needs be holy until the end of time.
\begin{large}\begin{center}\textsc{Chapter xciii}\end{center}\end{large}
\lettrine[lines=2]{I}{n} thy divine vengeance, Lord, in thy divine vengeance stand revealed! \textcolor{benred8}{2}~Judge of the world, mount thy throne, and give the proud their deserts! \textcolor{benred8}{3}~Must it be the sinners still, Lord, the sinners still that triumph? \textcolor{benred8}{4}~Shall there be no end to the prating, the rebellious talk, the boastfulness of wrong-doers? \textcolor{benred8}{5}~See, Lord, how they crush down thy people, afflict the land of thy choice, \textcolor{benred8}{6}~murder the widow and the stranger, slay the orphan! \textcolor{benred8}{7}~And they think, The Lord will never see it, the God of Israel pays no heed. \textcolor{benred8}{8}~Pay heed, rather, yourselves, dull hearts that count among my people; fools, learn your lesson ere it is too late. \textcolor{benred8}{9}~Is he deaf, the God who implanted hearing in us; is he blind, the God who gave us eyes to see? \textcolor{benred8}{10}~He who gives nations their schooling, who taught man all that man knows, will he not call you to account? \textcolor{benred8}{11}~The Lord looks into men\textquotesingle s hearts, and finds there illusion.
\textcolor{benred8}{12}~Happy, Lord, is the man whom thou dost chasten, reading him the lesson of thy law! \textcolor{benred8}{13}~For him, thou wilt lighten the time of adversity, digging a pit all the while to entrap the sinner. \textcolor{benred8}{14}~God will not abandon his people, will not desert his chosen land; \textcolor{benred8}{15}~ere long his justice will reappear in judgement, claiming all upright hearts for its own. \textcolor{benred8}{16}~Who takes my part against the oppressor? Who rallies to my side against the wrong-doers? \textcolor{benred8}{17}~It is the Lord that helps me; but for that, the grave would soon be my resting-place. \textcolor{benred8}{18}~Still, when my foothold seems lost, thy mercy, Lord, holds me up; \textcolor{benred8}{19}~amid all the thronging cares that fill my heart, my soul finds comfort in thy consolation. \textcolor{benred8}{20}~What part have these unjust judges with thee, that make mischief in the name of law? \textcolor{benred8}{21}~Let them harry the just as they will, pass sentence of death upon the innocent, \textcolor{benred8}{22}~the Lord will be my defence, in my God I shall find a rock-fastness still. \textcolor{benred8}{23}~He will punish the wrong, destroy them in their wickedness; doubt not the Lord our God will destroy them.
\begin{large}\begin{center}\textsc{Chapter xciv}\end{center}\end{large}
\textcolor{benred8}{1}~Come, friends, rejoice we in the Lord\textquotesingle s honour; cry we out merrily to God, our strength and deliverer; \textcolor{benred8}{2}~with praises court his presence, singing a joyful psalm! \textcolor{benred8}{3}~A high God is the Lord, a king high above all the gods;\footnote[1]{The Gallican Psalter, which is used in the liturgical recitation of the psalm, adds at the end of this verse \textasciigrave God will not abandon his people\textquotesingle  (cf. 93.14).} \textcolor{benred8}{4}~beneath his hand lie the depths of earth, his are the mountain peaks; \textcolor{benred8}{5}~his the ocean, for who but he created it? What other power fashioned the dry land? \textcolor{benred8}{6}~Come in, then, fall we down in worship, bowing the knee before God who made us. \textcolor{benred8}{7}~Who but the Lord is our God? And what are we, but folk of his pasturing, sheep that follow his beckoning hand?
\textcolor{benred8}{8}~Would you but listen to his voice to-day! Do not harden your hearts, \textcolor{benred8}{9}~as they were hardened once at Meriba, at Massa in the wilderness.\footnote[2]{Ex. 17.7.} Your fathers put me to the test, challenged me, as if they lacked proof of my power, \textcolor{benred8}{10}~for forty years together; from that generation I turned away in loathing;\footnote[3]{The Gallican Psalter has, \textasciigrave When I lived close to that generation\textquotesingle .} These, I said, are ever wayward hearts, \textcolor{benred8}{11}~these have never learned to obey me. And I took an oath in anger, They shall never attain my rest.\footnote[4]{vv. 8-11. See Heb. 3.7-19.}
\begin{large}\begin{center}\textsc{Chapter xcv}\end{center}\end{large}
\lettrine[lines=2]{S}{ing} the Lord a new song;\footnote[1]{This psalm is to be found in a slightly altered form in I Par. 16 (verses 23 and following).} in the Lord\textquotesingle s honour, let the whole earth make melody! \textcolor{benred8}{2}~Sing to the Lord, and bless his name; never cease to bear record of his power to save. \textcolor{benred8}{3}~Publish his glory among the heathen; his wonderful acts for all the world to hear. \textcolor{benred8}{4}~How great is the Lord, how worthy of honour! What other god is to be feared as he? \textcolor{benred8}{5}~They are but fancied gods the heathen call divine; the Lord, not they, made the heavens. \textcolor{benred8}{6}~Honour and beauty are his escort; worship and magnificence the attendants of his shrine.
\textcolor{benred8}{7}~Tribes of the heathen, make your offering to the Lord, an offering to the Lord of glory and homage, \textcolor{benred8}{8}~an offering of glory to the Lord\textquotesingle s name; bring sacrifice, come into his courts, \textcolor{benred8}{9}~worship the Lord in holy array. Before the Lord\textquotesingle s presence let the whole earth bow in reverence; \textcolor{benred8}{10}~tell the heathen, The Lord is king now, he has put the world in order, never to be thrown into confusion more; he gives the nations a just award. \textcolor{benred8}{11}~Rejoice, heaven, and let earth be glad; let the sea, and all the sea contains, give thunderous applause. \textcolor{benred8}{12}~The fields, and all the burden they bear, full of expectancy; no tree in the forest but will rejoice to greet its Lord\textquotesingle s coming. \textcolor{benred8}{13}~He comes to rule the earth; brings the world justice, to every race of men its promised award.
\begin{large}\begin{center}\textsc{Chapter xcvi}\end{center}\end{large}
\lettrine[lines=2]{T}{he} Lord reigns as king; let earth be glad of it, let the isles, the many isles, rejoice! \textcolor{benred8}{2}~See where he sits, clouds and darkness about him, justice and right the pillars of his throne; \textcolor{benred8}{3}~see where he comes, fire sweeping on before him, burning up his enemies all around. \textcolor{benred8}{4}~In the flash of his lightning, how shines the world revealed, how earth trembles at the sight! \textcolor{benred8}{5}~The hills melt like wax at the presence of the Lord; his presence, whom all the earth obeys. \textcolor{benred8}{6}~The very heavens proclaim his faithfulness; no nation but has sight of his glory. \textcolor{benred8}{7}~Shame upon the men that worship carved images, and make their boast of false gods! him only all the powers of heaven, prostrate, adore.
\textcolor{benred8}{8}~Glad news for Sion, rejoicing for Juda\textquotesingle s townships, when thy judgements, Lord, are made known; \textcolor{benred8}{9}~art thou not sovereign Lord of earth, beyond measure exalted above all gods? \textcolor{benred8}{10}~They are the Lord\textquotesingle s friends, who were never friends to wrong; souls that are true to him he guards ever, rescues them from the power of evil-doers. \textcolor{benred8}{11}~Dawn of hope for the innocent, dawn of gladness for honest hearts! \textcolor{benred8}{12}~Rejoice and triumph, just souls, in the Lord, of his holy name publish everywhere the renown.
\begin{large}\begin{center}\textsc{Chapter xcvii}\end{center}\end{large}
(A psalm.)
\textcolor{benred8}{1}~Sing the Lord a new song, a song of wonder at his doings; how his own right hand, his own holy arm, brought him victory. \textcolor{benred8}{2}~The Lord has given proof of his saving power, has vindicated his just dealings, for all the nations to see; \textcolor{benred8}{3}~has remembered his gracious promise, and kept faith with the house of Israel; no corner of the world but has witnessed how our God can save. \textcolor{benred8}{4}~In God\textquotesingle s honour let all the earth keep holiday; let all be mirth and rejoicing and festal melody! \textcolor{benred8}{5}~Praise the Lord with the harp, with harp and psaltery\textquotesingle s music; \textcolor{benred8}{6}~with trumpets of metal, and the music of the braying horn! Keep holiday in the presence of the Lord, our King; \textcolor{benred8}{7}~the sea astir, and all that the sea holds, the world astir, and all that dwell on it; \textcolor{benred8}{8}~the rivers echoing their applause, the hills, too, rejoicing to see the Lord come. \textcolor{benred8}{9}~He comes to judge the earth; brings the world justice, to every race of men its due award.
\begin{large}\begin{center}\textsc{Chapter xcviii}\end{center}\end{large}
\lettrine[lines=2]{T}{he} Lord is king, the nations are adread; he is throned above the Cherubim, and earth trembles before him. \textcolor{benred8}{2}~Great is the Lord who dwells in Sion, sovereign ruler of all peoples! \textcolor{benred8}{3}~Let them all praise that great name of thine, a name terrible and holy. \textcolor{benred8}{4}~He reigns in might, that right loves, to all assuring redress, giving the sons of Jacob doom and award. \textcolor{benred8}{5}~Praise, then, the Lord our God, and bow down before his footstool; that, too, is holy. \textcolor{benred8}{6}~Remember Moses and Aaron, and all those priests of his, Samuel and those others who called on his name, how the Lord listened when they called upon him. \textcolor{benred8}{7}~His voice came to them from the pillar of cloud; so it was they heard the decrees, the command he gave them. \textcolor{benred8}{8}~And thou, O Lord our God, didst listen to them, and they found thee a God of pardon; yet every fault of theirs thou wert quick to punish. \textcolor{benred8}{9}~Praise the Lord our God, and do worship on the holy mountain where he dwells; the Lord our God is holy.
\begin{large}\begin{center}\textsc{Chapter xcix}\end{center}\end{large}
(A psalm. For thanksgiving.)
\textcolor{benred8}{2}~Let the whole earth keep holiday in God\textquotesingle s honour; pay to the Lord the homage of your rejoicing, appear in his presence with glad hearts. \textcolor{benred8}{3}~Learn that it is the Lord, no other, who is God; his we are, he it was that made us; we are his own people, sheep of his own pasturing. \textcolor{benred8}{4}~Pass through these gates, enter these courts of his, with hymns of praise, give him thanks, and bless his name. \textcolor{benred8}{5}~Gracious is the Lord, everlasting his mercy; age after age, he is faithful to his promise still.
\begin{large}\begin{center}\textsc{Chapter c}\end{center}\end{large}
(Of David. A psalm.)
\textcolor{benred8}{1}~Of mercy and of justice my song shall be; a psalm in thy honour, Lord, \textcolor{benred8}{2}~from one that would guide his steps ever more perfectly. Ah, when wilt thou grant me thy presence?\footnote[1]{This is thought by some commentators to mean the coming of the Ark to Jerusalem (II Kg. 6).} Here in my house I would live with stainless heart; \textcolor{benred8}{3}~no ill purpose clouding my view, the transgressors of the law my enemies. \textcolor{benred8}{4}~None such will I have at my side; here treachery shall find no place, knavery no countenance; \textcolor{benred8}{5}~of whispered calumny, death shall be the reward; on scornful looks and proud thoughts I will have no mercy. \textcolor{benred8}{6}~To plain, honest folk in the land I will look for my company; my servants shall be such as follow the path of innocence. \textcolor{benred8}{7}~No welcome here for schemers, no standing in my presence for men who talk deceitfully. \textcolor{benred8}{8}~Mine, as the days pass, to root out from the land every guilty soul, till I purge the Lord\textquotesingle s city of all evil-doing.
\begin{large}\begin{center}\textsc{Chapter ci}\end{center}\end{large}
(A prayer for the friendless man, when he is troubled, and is pouring out his griefs before the Lord.)
\textcolor{benred8}{2}~O Lord, hear my prayer, and let my cry come unto thee. \textcolor{benred8}{3}~Do not turn thy face away from me, but lend me thy ear in time of affliction; give me swift audience whenever I call upon thee. \textcolor{benred8}{4}~See how this life of mine passes away like smoke, how this frame wastes like a tinder! \textcolor{benred8}{5}~Drained of strength, like grass the sun scorches, I leave my food untasted, forgotten; \textcolor{benred8}{6}~I am spent with sighing, till my skin clings to my bones. \textcolor{benred8}{7}~I am no better than a pelican out in the desert, an owl on some ruined dwelling; \textcolor{benred8}{8}~I keep mournful watch, lonely as a single sparrow on the house top. \textcolor{benred8}{9}~Still my enemies taunt me, in their mad rage make a by-word of me. \textcolor{benred8}{10}~Ashes are all my food, I drink nothing but what comes to me mingled with my tears; \textcolor{benred8}{11}~I shrink before thy vengeful anger, so low thou hast brought me, who didst once lift me so high. \textcolor{benred8}{12}~Like a tapering shadow my days dwindle, wasting away, like grass in the sun!
\textcolor{benred8}{13}~Lord, thou endurest for ever, thy name, age after age, is not forgotten; \textcolor{benred8}{14}~surely thou wilt bestir thyself, and give Sion redress! It is time, now, to take pity on her, the hour has come. \textcolor{benred8}{15}~See how thy servants love her even in ruin, how they water her dust with their tears! \textcolor{benred8}{16}~Will not the heathen learn reverence, Lord, for thy glorious name, all those monarchs of the earth, \textcolor{benred8}{17}~when they hear that the Lord has built Sion anew; that he has revealed himself there in glory, \textcolor{benred8}{18}~has given heed to the prayer of the afflicted, neglects their appeal no more? \textcolor{benred8}{19}~Such legend inscribe we for a later age to read it; a new people will arise, to praise the Lord; \textcolor{benred8}{20}~the Lord, who looks down from his sanctuary on high, viewing earth from heaven, \textcolor{benred8}{21}~who has listened to the groans of the prisoners, delivered a race that was doomed to die. \textcolor{benred8}{22}~There will be talk of the Lord\textquotesingle s name in Sion, of his praise in Jerusalem, \textcolor{benred8}{23}~when peoples and kings meet there to pay him their homage.
\textcolor{benred8}{24}~Here, on my journey, he has brought my strength to an end, cut short my days. \textcolor{benred8}{25}~What, my God, wilt thou snatch me away, my life half done? Age after age thy years endure; \textcolor{benred8}{26}~it was thou, Lord, that didst lay the foundations of earth when time began, it was thy hand that built the heavens. \textcolor{benred8}{27}~They will perish, but thou wilt remain; they will all be like a cloak that grows threadbare, and thou wilt lay them aside like a garment, and exchange them for new; \textcolor{benred8}{28}~thou art unchanging, thy years can never fail. \textcolor{benred8}{29}~The posterity of thy servants shall yet hold their lands in peace, their race shall live on in thy keeping.
\begin{large}\begin{center}\textsc{Chapter cii}\end{center}\end{large}
(Of David.)
\textcolor{benred8}{1}~Bless the Lord, my soul, unite, all my powers, to bless that holy name. \textcolor{benred8}{2}~Bless the Lord, my soul, remembering all he has done for thee, \textcolor{benred8}{3}~how he pardons all thy sins, heals all thy mortal ills, \textcolor{benred8}{4}~rescues thy life from deadly peril, crowns thee with the blessings of his mercy; \textcolor{benred8}{5}~how he contents all thy desire for good, restores thy youth, as the eagle\textquotesingle s plumage is restored. \textcolor{benred8}{6}~The Lord\textquotesingle s acts are acts of justice, every wronged soul he offers redress. \textcolor{benred8}{7}~The Lord, who told Moses his secrets, who shewed the sons of Israel his power!
\textcolor{benred8}{8}~How pitying and gracious the Lord is, how patient, how rich in mercy! \textcolor{benred8}{9}~He will not always be finding fault, his frown does not last for ever; \textcolor{benred8}{10}~he does not treat us as our sins deserve, does not exact the penalty of our wrong-doing. \textcolor{benred8}{11}~High as heaven above the earth towers his mercy for the men that fear him; \textcolor{benred8}{12}~far as the east is from the west, he clears away our guilt from us. \textcolor{benred8}{13}~For his own worshippers, the Lord has a father\textquotesingle s pity; \textcolor{benred8}{14}~does he not know the stuff of which we are made, can he forget that we are only dust? \textcolor{benred8}{15}~Man\textquotesingle s life is like the grass, he blooms and dies like a flower in the fields; \textcolor{benred8}{16}~once the hot wind has passed over, it has gone, forgotten by the place where it grew. \textcolor{benred8}{17}~But the Lord\textquotesingle s worshippers know no beginning or end of his mercy; he will keep faith with their children\textquotesingle s children, \textcolor{benred8}{18}~do they but hold fast by his covenant, and live mindful of his law. \textcolor{benred8}{19}~The Lord has set up his throne in heaven, rules with universal sway. \textcolor{benred8}{20}~Bless the Lord, all you angels of his; angels of sovereign strength, that carry out his commandment, attentive to the word he utters; \textcolor{benred8}{21}~bless the Lord, all you hosts of his, the servants that perform his will; \textcolor{benred8}{22}~bless the Lord, all you creatures of his, in every corner of his dominion; and thou, my soul, bless the Lord.
\begin{large}\begin{center}\textsc{Chapter ciii}\end{center}\end{large}
\lettrine[lines=2]{B}{less} the Lord, my soul; O Lord my God, what magnificence is thine! Glory and beauty are thy clothing. \textcolor{benred8}{2}~The light is a garment thou dost wrap about thee, the heavens a curtain thy hand unfolds. \textcolor{benred8}{3}~The waters of heaven are thy ante-chamber, the clouds thy chariot; on the wings of the wind thou dost come and go. \textcolor{benred8}{4}~Thou wilt have thy angels be like the winds,\footnote[1]{Some would render \textasciigrave Thou makest the winds thy messengers\textquotesingle , but the second half of the verse is against this.} the servants that wait on thee like a flame of fire.
\textcolor{benred8}{5}~The earth thou hast planted on its own firm base, undisturbed for all time. \textcolor{benred8}{6}~The deep once covered it, like a cloak; the waters stood high above the mountains,\footnote[2]{See Gen. 1.9.} \textcolor{benred8}{7}~then cowered before thy rebuking word, fled away at thy voice of thunder, \textcolor{benred8}{8}~leaving the mountain heights to rise, the valleys to sink into their appointed place! \textcolor{benred8}{9}~And to these waters thou hast given a frontier they may not pass; never must they flow back, and cover the earth again. \textcolor{benred8}{10}~Yet there shall be torrents flooding the glens, water-courses among the hills \textcolor{benred8}{11}~that give drink to every wild beast; here the wild asses may slake their thirst. \textcolor{benred8}{12}~The birds of heaven, too, will roost beside them; vocal is every bough with their music.
\textcolor{benred8}{13}~From thy high dwelling-place thou dost send rain upon the hills; thy hand gives earth all her plenty. \textcolor{benred8}{14}~Grass must grow for the cattle; for man, too, she must put forth her shoots, if he is to bring corn out from her bosom; \textcolor{benred8}{15}~if there is to be wine that will rejoice man\textquotesingle s heart, oil to make his face shine, and bread that will keep man\textquotesingle s strength from failing. \textcolor{benred8}{16}~Moisture there must be for the forest trees, for the cedars of Lebanon, trees of the Lord\textquotesingle s own planting. \textcolor{benred8}{17}~Here it is the birds build their nests; the stork makes its home in the fir-branches; \textcolor{benred8}{18}~finds refuge there such as the goats find in the high hills, the coney in its cave.
\textcolor{benred8}{19}~He has given us the moon for our calendar; the sun knows well the hour of his setting. \textcolor{benred8}{20}~Thou dost decree darkness, and the night falls; in the night all the forest is astir with prowling beasts; \textcolor{benred8}{21}~the young lions go roaring after their prey, God\textquotesingle s pensioners, asking for their food. \textcolor{benred8}{22}~Then the sun rises, and they slink away to lie down in their dens, \textcolor{benred8}{23}~while man goes abroad to toil and drudge till the evening. \textcolor{benred8}{24}~What diversity, Lord, in thy creatures! What wisdom has designed them all! There is nothing on earth but gives proof of thy creative power.
\textcolor{benred8}{25}~There lies the vast ocean, stretching wide on every hand; this, too, is peopled with living things past number, great creatures and small; \textcolor{benred8}{26}~the ships pass them on their course. Leviathan himself is among them; him, too, thou hast created to roam there at his pleasure.\footnote[3]{It is possible to read another sense; namely, that God has made the whale (or whatever creature Leviathan represents) as a jest for his own enjoyment.} \textcolor{benred8}{27}~And all look to thee to send them their food at the appointed time; \textcolor{benred8}{28}~it is through thy gift they find it, thy hand opens, and all are filled with content. \textcolor{benred8}{29}~But see, thou hidest thy face, and they are dismayed; thou takest their life from them, and they breathe no more, go back to the dust they came from. \textcolor{benred8}{30}~Then thou sendest forth thy spirit, and there is fresh creation; thou dost repeople the face of earth.
\textcolor{benred8}{31}~Glory be to the Lord for ever; still let him take delight in his creatures. \textcolor{benred8}{32}~One glance from him makes earth tremble; at his touch, the mountains are wreathed in smoke. \textcolor{benred8}{33}~While life lasts, I will sing in the Lord\textquotesingle s honour; my praise shall be his while I have breath to praise him; \textcolor{benred8}{34}~oh, may this prayer with him find acceptance, in whom is all my content! \textcolor{benred8}{35}~Perish all sinners from the land, let the wrong-doers be forgotten! But thou, my soul, bless the Lord. Alleluia.
\begin{large}\begin{center}\textsc{Chapter civ}\end{center}\end{large}
\lettrine[lines=2]{P}{raise} the Lord, and call upon his name; tell the story of his doings for all the nations to hear; \textcolor{benred8}{2}~greet him with song and psalm, recount his acts of miracle. \textcolor{benred8}{3}~Triumph in that holy name; let every heart that longs for the Lord rejoice. \textcolor{benred8}{4}~On the Lord, on the Lord\textquotesingle s greatness still let your hearts dwell, on the Lord\textquotesingle s presence be your hearts set. \textcolor{benred8}{5}~Remember the marvellous acts he did, his miracles, his sentences of doom; \textcolor{benred8}{6}~are you not the posterity of Abraham, his own servant, sons of that Jacob on whom his choice fell? \textcolor{benred8}{7}~And he, the Lord, is our own God, wide though his writ runs through all the world. \textcolor{benred8}{8}~He keeps in everlasting memory that covenant of his, that promise which a thousand ages might not cancel. \textcolor{benred8}{9}~He gave Abraham a promise, bound himself to Isaac by an oath; \textcolor{benred8}{10}~by that law Jacob should live, his Israel, bound to him with an eternal covenant. \textcolor{benred8}{11}~To thee, he said, I will give the land of Chanaan, a portion allotted to thee and thine. \textcolor{benred8}{12}~So few they were in number, only a handful, living there as strangers! \textcolor{benred8}{13}~And ever they passed on from country to country, the guests of king or people; \textcolor{benred8}{14}~but he suffered none to harm them;\footnote[1]{See Gen. 12.17.} to kings themselves the warning came; \textcolor{benred8}{15}~Lay no hand on them, never hurt them, servants anointed and true spokesmen of mine.
\textcolor{benred8}{16}~And now he brought famine on the land, cutting off all their supply of bread. \textcolor{benred8}{17}~But he had sent an envoy to prepare the way for them, that very Joseph, who was sold as a slave. \textcolor{benred8}{18}~Fetters held his feet, the yoke galled his neck, \textcolor{benred8}{19}~but he proved a true prophet at last, the Lord\textquotesingle s accomplished word to vindicate him.\footnote[2]{See Gen. 40, 41.} \textcolor{benred8}{20}~Then the king sent to release him; the proud ruler of many peoples set him free, \textcolor{benred8}{21}~and appointed him master of his household, lord of all the possessions that were his. \textcolor{benred8}{22}~Joseph should teach his courtiers to be as Joseph was, should train his aged counsellors in wisdom. \textcolor{benred8}{23}~So it was that Israel came into Egypt, that Jacob dwelt as an alien in the country of Cham.
\textcolor{benred8}{24}~Time passed, and he gave his people great increase of numbers, till it outmatched its rivals. \textcolor{benred8}{25}~And in these he wrought a change of heart; they grew weary of his people\textquotesingle s presence, devised ruin for his worshippers. \textcolor{benred8}{26}~And now he sent his servant Moses, and Aaron, the man of his choice, \textcolor{benred8}{27}~to bring about those signs, those miracles of his which the country of Cham would witness. \textcolor{benred8}{28}~Dark night he sent to benight them, and still his warnings went unheeded.\footnote[3]{The Hebrew text gives \textasciigrave And they?(Moses and Aaron) heeded his words\textquotesingle .} \textcolor{benred8}{29}~He turned their supply of water into blood, killing all the fish; \textcolor{benred8}{30}~frogs swarmed out of the ground, even in their royal palaces; \textcolor{benred8}{31}~at his word, flies attacked them, and gnats all their land over; \textcolor{benred8}{32}~hail was the rain he gave them, and it brought fire that burned up their countryside; \textcolor{benred8}{33}~he shattered their vines and fig-trees, broke down all the wood that grew in their domains. \textcolor{benred8}{34}~He gave the word, and locusts came, grasshoppers, too, past all numbering, \textcolor{benred8}{35}~eating up all the grass they had, eating up all the crops their land yielded. \textcolor{benred8}{36}~Then, his hand fell upon Egypt\textquotesingle s first-born, on the first-fruits of all they had engendered; \textcolor{benred8}{37}~and so he brought his people out, enriched with silver and gold, no foot that stumbled among all their tribes.
\textcolor{benred8}{38}~Glad indeed was Egypt at their going, such fear of them had overtaken it. \textcolor{benred8}{39}~He spread out a cloud to cover them, that turned to fire in the darkness, lighting their journey. \textcolor{benred8}{40}~Quails came, when they asked for food; he satisfied their desire, too, with bread from heaven, \textcolor{benred8}{41}~and pierced the rock so that water flowed down, running streams in the wilderness. \textcolor{benred8}{42}~So well did he remember that holy promise of his, made to his servant Abraham; \textcolor{benred8}{43}~in joy and triumph he led them out, his chosen people, \textcolor{benred8}{44}~and gave them the lands of the heathen for their own. \textcolor{benred8}{45}~There, on soil Gentile hands had tilled, his commandments should be kept sacred, his law should reign. Alleluia.
\begin{large}\begin{center}\textsc{Chapter cv}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Praise the Lord, the Lord is gracious; his mercy endures for ever; \textcolor{benred8}{2}~what tongue can recount all the great deeds of the Lord, can echo all his praise? \textcolor{benred8}{3}~Blessed are they who abide ever by his decrees, ever do the right! \textcolor{benred8}{4}~Remember me, Lord, with loving thoughts towards thy people, come and strengthen me with thy aid, \textcolor{benred8}{5}~to witness the prosperity of thy chosen servants, to rejoice with thy people that rejoices, to share the glory of thy own domain.
\textcolor{benred8}{6}~We have taken part in our fathers\textquotesingle  sins; we are guilty men, rebels against thee. \textcolor{benred8}{7}~So it was with our fathers in Egypt; unremarked, thy wonderful doings, unremembered, thy abundant mercies; even at the Red Sea they must prove rebellious. \textcolor{benred8}{8}~Yet, for his own honour, to make known his power, he delivered them, \textcolor{benred8}{9}~checking the Red Sea, so that it dried up, and leading them through its depths as safely as if they trod the desert sands. \textcolor{benred8}{10}~From a cruel tyrant\textquotesingle s grasp he rescued them, claimed them for his own; \textcolor{benred8}{11}~and the water overwhelmed their pursuers, till not one of them was left. \textcolor{benred8}{12}~They believed, then, in his promises, sang songs, then, in his honour, \textcolor{benred8}{13}~but soon they forgot what he had done, and could not wait upon his will. \textcolor{benred8}{14}~They must needs give way to their cravings in the wilderness, challenge God\textquotesingle s power, there in the desert, \textcolor{benred8}{15}~till he granted their will, then sent a wasting sickness to plague them. \textcolor{benred8}{16}~Faction raised its head in the camp against Moses, against Aaron, the Lord\textquotesingle s chosen priest; \textcolor{benred8}{17}~and now earth gaped, swallowing up Dathan, overwhelming Abiron and his conspiracy; \textcolor{benred8}{18}~fire broke out in their company, and the rebels perished by its flames. \textcolor{benred8}{19}~They made a calf, too, at Horeb, casting a golden image and worshipping it, \textcolor{benred8}{20}~as if they would exchange the glory that dwelt among them for the semblance of a bullock at grass. \textcolor{benred8}{21}~So little they remembered the God who had delivered them, those portents of his in Egypt, \textcolor{benred8}{22}~strange things seen in the land of Cham, terrible things down by the Red Sea! \textcolor{benred8}{23}~What wonder if he threatened to make an end of them? But Moses, the man of his choice, stood in the breach to confront his anger, to ward off destruction.
\textcolor{benred8}{24}~And now they poured scorn on the land of their desire, distrusting his promise; \textcolor{benred8}{25}~and the camp was all disaffection. So the Lord, finding they would not listen to his voice, \textcolor{benred8}{26}~lifted his hand and threatened to smite them down, there in the wilderness; \textcolor{benred8}{27}~they should be lost among the peoples, scattered wide through the world. \textcolor{benred8}{28}~They dedicated themselves to Beelphegor, in honour of the dead gods sat down to feast; \textcolor{benred8}{29}~till their wicked ways roused God\textquotesingle s anger, and a plague fell upon them. \textcolor{benred8}{30}~Nor might the destruction cease, till Phinees rose up and made amends, \textcolor{benred8}{31}~winning himself such title to God\textquotesingle s favour as shall be remembered, age after age, eternally. \textcolor{benred8}{32}~They provoked his anger, too, at the waters of Meriba, so that Moses was punished for their sake; \textcolor{benred8}{33}~because, in his heart\textquotesingle s bitterness, he broke out into open complaint.
\textcolor{benred8}{34}~Not theirs to root out the heathen, as the Lord had bidden them; \textcolor{benred8}{35}~they mingled with the heathen instead, and learned their ways; \textcolor{benred8}{36}~worshipping carved images, to their own undoing, \textcolor{benred8}{37}~sacrificing their sons and daughters in honour of devils. \textcolor{benred8}{38}~Innocent blood, the blood of their own sons and daughters, was poured out in worship to the idols of Chanaan; with blood the whole land was polluted, \textcolor{benred8}{39}~so heinous the guilt of its people, so wanton their ways. \textcolor{benred8}{40}~Then God\textquotesingle s anger blazed up against his people, his chosen race became abominable to him, \textcolor{benred8}{41}~and he handed them over to the Gentiles; despised slaves, \textcolor{benred8}{42}~they were oppressed by their enemies, bowed down under the yoke. \textcolor{benred8}{43}~Again and again he brought them deliverance, but ever there were fresh shifts to provoke him, there was fresh guilt to drag them in the dust. \textcolor{benred8}{44}~And still, when he saw their distress, when he heard their appeals to him, \textcolor{benred8}{45}~the thought him of his covenant availed them; in his great mercy he would relent; \textcolor{benred8}{46}~their very captors should be moved to pity.
\textcolor{benred8}{47}~Deliver us, O Lord our God, and gather us again, scattered as we are among the heathen, to praise thy holy name, to triumph in thy renown. \textcolor{benred8}{48}~Blessed be the God of Israel from all eternity to all eternity; let all the people cry, Amen, Alleluia.
\begin{large}\begin{center}\textsc{Chapter cvi}\end{center}\end{large}
\lettrine[lines=2]{P}{raise} the Lord, the Lord is gracious; his mercy endures for ever; \textcolor{benred8}{2}~be this the cry of men the Lord has rescued, rescued them from the enemy\textquotesingle s hand, and gathered them in \textcolor{benred8}{3}~from sunrising and sunset, from the north country and the south.
\textcolor{benred8}{4}~Some have wandered in parched deserts, missing the way to the city that was their home, \textcolor{benred8}{5}~hungry and thirsty, so that their spirits died within them. \textcolor{benred8}{6}~So they cried out to the Lord in their trouble, and he relieved their distress, \textcolor{benred8}{7}~guiding them surely to the place where they should find a home. \textcolor{benred8}{8}~Praise they the Lord in his mercies, in his wondrous dealings with mortal men; \textcolor{benred8}{9}~poor souls that were thirsty, contented now, poor souls that were hungry, satisfied now with all good.
\textcolor{benred8}{10}~Some lay where darkness overshadowed them, helpless in bonds of iron; \textcolor{benred8}{11}~their punishment for rebelling against God\textquotesingle s decrees, for thwarting the will of the most High. \textcolor{benred8}{12}~Their hearts bowed down with sorrow, none else to aid their faltering steps, \textcolor{benred8}{13}~they cried out to the Lord in their trouble, and he relieved their distress, \textcolor{benred8}{14}~rescuing them from darkness, from the shadows, tearing their chains asunder. \textcolor{benred8}{15}~Praise they the Lord in his mercies, in his wondrous dealings with mortal men; \textcolor{benred8}{16}~the Lord who has shattered the gates of brass, riven the bonds of iron.
\textcolor{benred8}{17}~Some for their own fault must needs be humbled; for their guilt \textcolor{benred8}{18}~they lay sick, with no stomach for food, close to death\textquotesingle s door. \textcolor{benred8}{19}~So they cried out to the Lord in their trouble, and he relieved their distress, \textcolor{benred8}{20}~uttered the word of healing, and saved them from their peril. \textcolor{benred8}{21}~Praise they the Lord in his mercies, in his wondrous dealings with mortal men; \textcolor{benred8}{22}~theirs to offer him sacrifice in thanksgiving, and proclaim joyfully what he has done for them.
\textcolor{benred8}{23}~Some there were that ventured abroad in ships, trafficking over the high seas; \textcolor{benred8}{24}~these are men that have witnessed the Lord\textquotesingle s doings, his wonderful doings amid the deep. \textcolor{benred8}{25}~At his word the stormy wind rose, churning up its waves; \textcolor{benred8}{26}~high up towards heaven they were carried, then sank into the trough, with spirits fainting at their peril; \textcolor{benred8}{27}~see them reeling and staggering to and fro as a drunkard does, all their seamanship forgotten! \textcolor{benred8}{28}~So they cried out to the Lord in their trouble, and he relieved their distress, \textcolor{benred8}{29}~stilling the storm into a whisper, till all its waves were quiet. \textcolor{benred8}{30}~Glad hearts were theirs, when calm fell about them, and he brought them to the haven where they longed to be. \textcolor{benred8}{31}~Praise they the Lord in his mercies, in his wondrous dealings with mortal men; \textcolor{benred8}{32}~let them extol his name, where the people gather, glorify him where the elders sit in council.
\textcolor{benred8}{33}~Here, he changes rivers into desert sand, wells into dry ground; \textcolor{benred8}{34}~land that once was fruitful into a salty marsh, to punish its people\textquotesingle s guilt. \textcolor{benred8}{35}~There, he turns the wilderness into pools of water, desert ground into springs; \textcolor{benred8}{36}~and establishes hungry folk there, so that they build themselves a city to dwell in, \textcolor{benred8}{37}~sow fields, and plant vineyards, and reap the harvest; \textcolor{benred8}{38}~he blesses them, so that their numbers increase beyond measure, and to their cattle grants increase. \textcolor{benred8}{39}~Once, they were but few, worn down by stress of need and ill fortune; \textcolor{benred8}{40}~but now the same power that shames proud chieftains, and keeps them wandering in a pathless desert, \textcolor{benred8}{41}~has rescued the poor from need, their households thrive like their own flocks. \textcolor{benred8}{42}~Honest men will rejoice to witness it, and malice will stand dumb with confusion. \textcolor{benred8}{43}~Heed it well, if thou wouldst be wise; be these thy study, the mercies of the Lord.
\begin{large}\begin{center}\textsc{Chapter cvii}\end{center}\end{large}
(A song. A psalm. Of David.)
\textcolor{benred8}{2}~A true heart, my God, a heart true to thy service; I will sing of thee and praise thee. \textcolor{benred8}{3}~Wake, my heart, wake, echoes of harp and viol; dawn shall find me watching. \textcolor{benred8}{4}~Let me give thanks, Lord, for all the world to hear it, sing psalms while the Gentiles listen, \textcolor{benred8}{5}~of thy mercy, high above heaven itself, of thy faithfulness, that reaches the clouds! \textcolor{benred8}{6}~O God, mount high above the heavens\footnote[1]{vv. 2-6. See Ps. 56.8-12.} till thy glory overshadows the whole earth. \textcolor{benred8}{7}~Now bring aid to the men thou lovest, give our prayer answer, and lift thy right hand to save. \textcolor{benred8}{8}~God\textquotesingle s word came to us from his sanctuary: In triumph I will divide up Sichem, and parcel out the valley of Tents; \textcolor{benred8}{9}~to me Galaad, to me Manasses belongs; Ephraim is my helmet, Juda the staff I bear. \textcolor{benred8}{10}~Now Moab, too, shall be my drudge; over Edom I will claim my right; I will lead the Philistines away in triumph. \textcolor{benred8}{11}~Such was the oracle; but now who is to lead me on my march against this fortress, who is to find an entrance for me into Edom, \textcolor{benred8}{12}~when thou, O God, hast disowned us, and wilt not go into battle with our armies? \textcolor{benred8}{13}~It is thou that must deliver us from peril; vain is the help of man. \textcolor{benred8}{14}~Only through God can we fight victoriously; only he can trample our oppressors in the dust.\footnote[2]{See Ps. 59.6-14.}
\begin{large}\begin{center}\textsc{Chapter cviii}\end{center}\end{large}
\textcolor{benred8}{1}~(To the choir-master. Of David. A psalm.)
\textcolor{benred8}{2}~God that guardest my renown, do not leave me unbefriended; there are malicious lips, treacherous lips, that decry me; \textcolor{benred8}{3}~whispering against me, hedging me about with a conspiracy of hatred, in unprovoked attack. \textcolor{benred8}{4}~On their side, all calumny in return for love, on mine all prayer; \textcolor{benred8}{5}~kindness is repaid with injury, love with ill will.
\textcolor{benred8}{6}~An ill master let him have,\footnote[1]{Verses 6-19 are perhaps spoken, not by the Psalmist, but by his detractors.} and an accuser ready at his side; \textcolor{benred8}{7}~let him leave the court of judgement a doomed man, pleading with heaven in vain. \textcolor{benred8}{8}~Swiftly let his days come to an end, and his office be entrusted to another;\footnote[2]{Ac. 1.20.} \textcolor{benred8}{9}~orphancy for the children, widowhood for the wife! \textcolor{benred8}{10}~Driven from a ruined home, to and fro let his children wander, begging their bread, \textcolor{benred8}{11}~while eager creditors eye his goods, and strangers divide the fruits of his toil. \textcolor{benred8}{12}~May no friend be left to do him a kindness, none to have pity on his defenceless kin; \textcolor{benred8}{13}~a speedy end to his race, oblivion for his name before a generation passes! \textcolor{benred8}{14}~Still may the sin of his fathers be remembered in the Lord\textquotesingle s sight, his mother\textquotesingle s guilt remain indelible; \textcolor{benred8}{15}~still may the Lord keep it in mind, and wipe out their memory from the earth. \textcolor{benred8}{16}~Did he himself keep mercy in mind, \textcolor{benred8}{17}~when he persecuted the helpless, the destitute, the grief-stricken, and marked them down for death? \textcolor{benred8}{18}~Cursing he loved, upon him let the curse fall; for blessing he cared little, may blessing still pass him by. Let cursing wrap him about, sink like water into his inmost being, soak, like oil, into the marrow of his bones! \textcolor{benred8}{19}~Let it be the garb he wears, cling to him like a girdle that he can never take off.
\textcolor{benred8}{20}~So, in their own coin, may the Lord repay them,\footnote[3]{This is a doubtful interpretation of the Hebrew; it is possible to render \textasciigrave Such are the proceedings of my accusers\textquotesingle .} my accusers that defame me so cruelly. \textcolor{benred8}{21}~But do thou, my Lord and Master, take my part, to defend thy own honour; no mercy is so tender as thine. \textcolor{benred8}{22}~Deliver me in my helpless need; my heart is pierced through with anguish. \textcolor{benred8}{23}~Like a tapering shadow I depart, swept away like a locust on the wing. \textcolor{benred8}{24}~My knees are weak with fasting, my strength pines away unnourished. \textcolor{benred8}{25}~They make a laughing-stock of me, toss their heads in derision as they pass by. \textcolor{benred8}{26}~Help me, O Lord my God; deliver me in thy mercy; \textcolor{benred8}{27}~prove to them that my woes are a visitation from thee, sent by no hand but thine. \textcolor{benred8}{28}~Bless me, thou, and let them curse as they will; disappoint my adversaries, and grant thy servant relief. \textcolor{benred8}{29}~Let these, my accusers, be covered with shame, wrapped in the mantle of their own confusion. \textcolor{benred8}{30}~Loudly will I give the Lord thanks, praise him before multitudes that listen; \textcolor{benred8}{31}~the Lord who has stood at the right hand of the friendless, brought redress to an innocent soul misjudged.
\begin{large}\begin{center}\textsc{Chapter cix}\end{center}\end{large}
(Of David. A psalm.\footnote[1]{This psalm, the meaning of which is in any case obscure, is further confused in the Latin, and in the Septuagint Greek, by the use of the same rendering for two different words in the Hebrew, here translated by \textasciigrave Lord\textquotesingle  and \textasciigrave Master\textquotesingle  respectively. Cf. Mt. 22.44, Mk. 12.36, Lk. 20.42, Ac. 2.34.})
\textcolor{benred8}{1}~To the Master I serve the Lord\textquotesingle s promise was given, Sit here at my right hand while I make thy enemies a footstool under thy feet. \textcolor{benred8}{2}~The Lord will make thy empire spring up like a branch out of Sion; thou art to bear rule in the midst of thy enemies. \textcolor{benred8}{3}~From birth, princely state shall be thine, holy and glorious; thou art my son, born like dew before the day-star rises. \textcolor{benred8}{4}~The Lord has sworn an oath there is no retracting, Thou art a priest for ever in the line of Melchisedech.\footnote[2]{See Heb. 5.6, and elsewhere.} \textcolor{benred8}{5}~At thy right hand, the Lord will beat down kings in the day of his vengeance; \textcolor{benred8}{6}~he will pass sentence on the nations, heap high the bodies, scatter far and wide the heads of the slain. \textcolor{benred8}{7}~Let him but drink of the brook by the wayside, he will lift up his head in victory.\footnote[3]{There is no agreement as to the precise meaning of this verse; some would render, \textasciigrave they will drink\textquotesingle , understanding the first half as an ironic reference to the princes lying slaughtered by the road side.}
\begin{large}\begin{center}\textsc{Chapter cx}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~All my heart goes out to the Lord in praise, before the assembly where the just are gathered. \textcolor{benred8}{2}~Chant we the Lord\textquotesingle s wondrous doings, delight and study of all who love him. \textcolor{benred8}{3}~Ever his deeds are high and glorious, faithful he abides to all eternity. \textcolor{benred8}{4}~Great deeds, that he keeps still in remembrance! \textcolor{benred8}{5}~He, the Lord, is kind and merciful. In abundance he fed the men who feared him, keeping his covenant for ever. \textcolor{benred8}{6}~Lordly the power he shewed his people, \textcolor{benred8}{7}~making the lands of the heathen their possession. No act but shews him just and faithful; of his decrees there is no relenting. \textcolor{benred8}{8}~Perpetual time shall leave them changeless; right and truth are their foundation. \textcolor{benred8}{9}~So he has brought our race deliverance; to all eternity stands his covenant. Unutterable is his name and worshipful; \textcolor{benred8}{10}~vain without his fear is learning. Wise evermore are you who follow it; yours the prize that lasts for ever.
\begin{large}\begin{center}\textsc{Chapter cxi}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~A blessed man is he, who fears the Lord, bearing great love to his commandments. \textcolor{benred8}{2}~Children of his shall win renown in their country; do right, and thy sons shall find a blessing. \textcolor{benred8}{3}~Ease shall dwell in his house, and great prosperity; fame shall ever record his bounty. \textcolor{benred8}{4}~Good men see a light dawn in darkness; his light, who is merciful, kind and faithful. \textcolor{benred8}{5}~It goes well with the man who lends in pity, just and merciful in his dealings. \textcolor{benred8}{6}~Length of days shall leave him still unshaken; men will remember the just for ever. \textcolor{benred8}{7}~No fear shall he have of evil tidings; on the Lord his hope is fixed unchangeably. \textcolor{benred8}{8}~Patient his heart remains and stedfast, quietly he waits for the downfall of his enemies. \textcolor{benred8}{9}~Rich are his alms to the needy; still his bounty abides in memory.\footnote[1]{II Cor. 9.9.} The Lord will lift up his head in triumph; \textcolor{benred8}{10}~ungodly men are ill content to see it. Vainly they gnash their teeth in envy; worldly hopes must fade and perish.
\begin{large}\begin{center}\textsc{Chapter cxii}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Praise the Lord, you that are his servants, praise the name of the Lord together. \textcolor{benred8}{2}~Blessed be the Lord\textquotesingle s name at all times, from this day to all eternity; \textcolor{benred8}{3}~from the sun\textquotesingle s rise to the sun\textquotesingle s setting let the Lord\textquotesingle s name be praised continually. \textcolor{benred8}{4}~The Lord is sovereign king of all the nations; his glory is high above the heavens. \textcolor{benred8}{5}~Who is like the Lord our God, so high above us, \textcolor{benred8}{6}~that stoops to regard both heaven and earth, \textcolor{benred8}{7}~lifting up the poor from the dust he lay in, raising the beggar out of his dung-hill, \textcolor{benred8}{8}~to find him a place among the princes, the princes that rule over his people? \textcolor{benred8}{9}~He gives the barren woman a home to dwell in, a mother rejoicing in her children.
\begin{large}\begin{center}\textsc{Chapter cxiii}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~When Israel came out of Egypt, and the sons of Jacob heard no more a strange language, \textcolor{benred8}{2}~the Lord took Juda for his sanctuary, Israel for his own dominion. \textcolor{benred8}{3}~The seas fled at the sight they witnessed, backward flowed the stream of Jordan; \textcolor{benred8}{4}~up leapt, like rams, the startled mountains, up leapt the hills, like yearling sheep. \textcolor{benred8}{5}~What ailed you, seas, that you fled in terror, Jordan\textquotesingle s stream, what drove thee back? \textcolor{benred8}{6}~Why did you leap up like rams, you mountains, leap up, you hills, like yearling sheep? \textcolor{benred8}{7}~Let earth thrill at its Master\textquotesingle s presence; it is he that comes, the God of Jacob, \textcolor{benred8}{8}~who turned the rock into pools of water, the flint-stone into a springing well.
\textcolor{benred8}{9}~Not to us, Lord, not to us the glory; let thy name alone be honoured; \textcolor{benred8}{10}~thine the merciful, thine the faithful; why must the heathen say, Their God deserts them? \textcolor{benred8}{11}~Our God is a God that dwells in heaven; all that his will designs, he executes. \textcolor{benred8}{12}~The heathen have silver idols and golden, gods which the hands of men have fashioned. \textcolor{benred8}{13}~They have mouths, and yet are silent; eyes they have, and yet are sightless; \textcolor{benred8}{14}~ears they have, and want all hearing; noses, and yet no smell can reach them; \textcolor{benred8}{15}~hands unfeeling, feet unstirring; never a sound their throats may utter. \textcolor{benred8}{16}~Such be the end of all who make them, such the reward of all who trust them. \textcolor{benred8}{17}~It is the Lord that gives hope to the race of Israel, their only help, their only stronghold; \textcolor{benred8}{18}~the Lord that gives hope to the race of Aaron, their only help, their only stronghold; \textcolor{benred8}{19}~the Lord that gives hope to all who fear him, their only help, their only stronghold. \textcolor{benred8}{20}~The Lord keeps us in mind, and grants us blessing, blesses the race of Israel, blesses the race of Aaron; \textcolor{benred8}{21}~all those who fear the Lord, small and great alike, he blesses. \textcolor{benred8}{22}~Still may the Lord grant you increase, you and your children after you; \textcolor{benred8}{23}~the blessing of the Lord be upon you. It is he that made both heaven and earth; \textcolor{benred8}{24}~to the Lord belongs the heaven of heavens, the earth he gives to the children of men. \textcolor{benred8}{25}~From the dead, Lord, thou hast no praises, the men who go down into the place of silence; \textcolor{benred8}{26}~but we bless the Lord, we, the living, from this day to all eternity.
\begin{large}\begin{center}\textsc{Chapter cxiv}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~My heart is aflame, so graciously the Lord listens to my entreaty; \textcolor{benred8}{2}~the Lord, who grants me audience when I invoke his name. \textcolor{benred8}{3}~Death\textquotesingle s noose about me, caught in the snares of the grave, ever I found distress and grief at my side, \textcolor{benred8}{4}~till I called upon the Lord, Save me, Lord, in my peril. \textcolor{benred8}{5}~Merciful the Lord our God is, and just, and full of pity; \textcolor{benred8}{6}~he cares for simple hearts, and to me, when I lay humbled, he brought deliverance. \textcolor{benred8}{7}~Return, my soul, where thy peace lies; the Lord has dealt kindly with thee; \textcolor{benred8}{8}~he has saved my life from peril, banished my tears, kept my feet from falling. \textcolor{benred8}{9}~Mine to walk at ease, enjoying the Lord\textquotesingle s presence, in the land of the living.
\begin{large}\begin{center}\textsc{Chapter cxv}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Praise the Lord, all you Gentiles, let all the nations of the world do him honour. \textcolor{benred8}{2}~Abundant has his mercy been towards us; the Lord remains faithful to his word for ever.
\begin{large}\begin{center}\textsc{Chapter cxvi}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Give thanks to the Lord; the Lord is gracious, his mercy endures for ever. \textcolor{benred8}{2}~Echo the cry, sons of Israel; the Lord is gracious, his mercy endures for ever. \textcolor{benred8}{3}~His mercy endures for ever, echo the cry, sons of Aaron; \textcolor{benred8}{4}~his mercy endures for ever; echo the cry, all you who are the Lord\textquotesingle s worshippers. \textcolor{benred8}{5}~I called on the Lord when trouble beset me, and the Lord listened, and brought me relief. \textcolor{benred8}{6}~With the Lord at my side, I have no fear of the worst man can do; \textcolor{benred8}{7}~with the Lord at my side to aid me, I shall yet see my enemies baffled. \textcolor{benred8}{8}~Better trust the Lord than rely on the help of man; \textcolor{benred8}{9}~better trust the Lord than rely on the word of princes. \textcolor{benred8}{10}~Let all heathendom ring me round, see, in the power of the Lord I crush them! \textcolor{benred8}{11}~They cut me off from every way of escape, but see, in the power of the Lord I crush them! \textcolor{benred8}{12}~They swarm about me like bees, their fury blazes up like fire among thorns, but see, in the power of the Lord, I crush them! \textcolor{benred8}{13}~I reeled under the blow, and had well-nigh fallen, but still the Lord was there to aid me. \textcolor{benred8}{14}~Who but the Lord is my protector, my stronghold; who but the Lord has brought me deliverance?
\textcolor{benred8}{15}~The homes of the just echo, now, with glad cries of victory; the power of the Lord has triumphed. \textcolor{benred8}{16}~The power of the Lord has brought me to great honour, the power of the Lord has triumphed. \textcolor{benred8}{17}~I am reprieved from death, to live on and proclaim what the Lord has done for me. \textcolor{benred8}{18}~The Lord has chastened me, chastened me indeed, but he would not doom me to die. \textcolor{benred8}{19}~Open me the gates where right dwells; let me go in and thank the Lord! \textcolor{benred8}{20}~Here is the gate that leads to the Lord\textquotesingle s presence; here shall just souls find entry. \textcolor{benred8}{21}~Thanks be to thee, Lord, for giving me audience, thanks be to thee, my deliverer. \textcolor{benred8}{22}~The very stone which the builders rejected has become the chief stone at the corner; \textcolor{benred8}{23}~this is the Lord\textquotesingle s doing, and it is marvellous in our eyes.\footnote[1]{vv. 22, 23. Mt. 21.42.} \textcolor{benred8}{24}~This day is a holiday of the Lord\textquotesingle s own choosing; greet this day with rejoicing, greet this day with triumph!
\textcolor{benred8}{25}~Deliverance, Lord, deliverance; Lord, grant us days of prosperity! \textcolor{benred8}{26}~Blessed is he who comes in the name of the Lord! A blessing from the Lord\textquotesingle s house upon your company! \textcolor{benred8}{27}~The Lord is God; his light shines out to welcome us; marshal the procession aright, with a screen of boughs that reaches to the very horns of the altar. \textcolor{benred8}{28}~Thou art my God, mine to thank thee, thou art my God, mine to extol thee. \textcolor{benred8}{29}~Give thanks to the Lord; the Lord is gracious, his mercy endures for ever.
\begin{large}\begin{center}\textsc{Chapter cxvii}\end{center}\end{large}
\textcolor{benred8}{1}~Ah, blessed they, who pass through life\textquotesingle s journey unstained, who follow the law of the Lord! \textcolor{benred8}{2}~Ah, blessed they, who cherish his decrees, make him the whole quest of their hearts! \textcolor{benred8}{3}~Afar from wrong-doing, thy sure paths they tread. \textcolor{benred8}{4}~Above all else it binds us, the charge thou hast given us to keep. \textcolor{benred8}{5}~Ah, how shall my steps be surely guided to keep faith with thy covenant? \textcolor{benred8}{6}~Attentive to all thy commandments, I go my way undismayed. \textcolor{benred8}{7}~A true heart\textquotesingle s worship thou shalt have, thy just awards prompting me. \textcolor{benred8}{8}~All shall be done thy laws demand, so thou wilt not forsake me utterly.
\textcolor{benred8}{9}~Best shall he keep his youth unstained, who is true to thy trust. \textcolor{benred8}{10}~Be thou the whole quest of my heart; never let me turn aside from thy commandments. \textcolor{benred8}{11}~Buried deep in my heart, thy warnings shall keep me clear of sin. \textcolor{benred8}{12}~Blessed art thou, O Lord, teach me to know thy will. \textcolor{benred8}{13}~By these lips let the awards thou makest ever be recorded. \textcolor{benred8}{14}~Blithely as one that has found great possessions, I follow thy decrees. \textcolor{benred8}{15}~Bethinking me still of the charge thou givest, I will mark thy footsteps. \textcolor{benred8}{16}~Be thy covenant ever my delight, thy words kept in memory.
\textcolor{benred8}{17}~Crown thy servant with life, to live faithful to thy commands. \textcolor{benred8}{18}~Clear sight be mine, to contemplate the wonders of thy law. \textcolor{benred8}{19}~Comfort this earthly exile; do not refuse me the knowledge of thy will. \textcolor{benred8}{20}~Crushed lies my spirit, longing ever for thy just awards. \textcolor{benred8}{21}~Chastener of the proud, thy curse lies on all who swerve from thy covenant. \textcolor{benred8}{22}~Clear me of the reproach that shames me, as I was ever attentive to thy claims. \textcolor{benred8}{23}~Closeted together, princes plot against me, thy servant, that thinks only of thy decrees. \textcolor{benred8}{24}~Claims lovingly cherished, decrees that are my counsellors!
\textcolor{benred8}{25}~Deep lies my soul in the dust, restore life to me, as thou hast promised. \textcolor{benred8}{26}~Deign, now, to shew me thy will, thou who hast listened when I opened my heart to thee. \textcolor{benred8}{27}~Direct me in the path thou biddest me follow, and all my musing shall be of thy wonderful deeds. \textcolor{benred8}{28}~Despair wrings tears from me; let thy promises raise me up once more. \textcolor{benred8}{29}~Deliver me from every false thought; make me free of thy covenant. \textcolor{benred8}{30}~Duty\textquotesingle s path my choice, I keep thy bidding ever in remembrance. \textcolor{benred8}{31}~Disappoint me, Lord, never, one that holds fast by thy commandments. \textcolor{benred8}{32}~Do but open my heart wide, and easy lies the path thou hast decreed.
\textcolor{benred8}{33}~Expound, Lord, thy whole bidding to me; faithfully I will keep it. \textcolor{benred8}{34}~Enlighten me, to scan thy law closely, and keep true to it with all my heart. \textcolor{benred8}{35}~Eagerly I long to be guided in the way of thy obedience. \textcolor{benred8}{36}~Ever let my choice be set on thy will, not on covetous thoughts. \textcolor{benred8}{37}~Eyes have I none for vain phantoms; let me find life in following thy ways. \textcolor{benred8}{38}~Establish with me, thy servant, the promise made to thy worshippers. \textcolor{benred8}{39}~Ease me of the reproach my heart dreads, thou, whose awards are gracious. \textcolor{benred8}{40}~Each command of thine I embrace lovingly; do thou in thy faithfulness grant me life.
\textcolor{benred8}{41}~For me too, Lord, thy mercy, for me too the deliverance thou hast promised! \textcolor{benred8}{42}~Fit answer for those who taunt me, that I rely on thy truth. \textcolor{benred8}{43}~Faithful thy promise, let me not boast in vain; in thy covenant lies my hope. \textcolor{benred8}{44}~For ever and for evermore true to thy charge thou shalt find me. \textcolor{benred8}{45}~Freely shall my feet tread, if thy will is all my quest. \textcolor{benred8}{46}~Fearlessly will I talk of thy decrees in the presence of kings, and be never abashed. \textcolor{benred8}{47}~Fain would I have all my comfort in the law I love. \textcolor{benred8}{48}~Flung wide my arms to greet thy law, ever in my thoughts thy bidding.
\textcolor{benred8}{49}~Go not back on the word thou hast pledged to thy servant; there lies all my hope. \textcolor{benred8}{50}~Good news in my affliction, thy promises have brought me life. \textcolor{benred8}{51}~Ground down by the scorn of my oppressors, never from thy law I swerve aside. \textcolor{benred8}{52}~Gracious comfort, Lord, is the memory of thy just dealings in times long past. \textcolor{benred8}{53}~Great ruth have I to see wrong-doers, and how they abandon thy law. \textcolor{benred8}{54}~Gone out into a land of exile, of thy covenant I make my song. \textcolor{benred8}{55}~Gloom of the night finds me still thinking of thy name, Lord, still observant of thy bidding. \textcolor{benred8}{56}~Guerdon I ask no other, but the following of thy will.
\textcolor{benred8}{57}~Heritage, Lord, I claim no other, but to obey thy word. \textcolor{benred8}{58}~Heart-deep my supplication before thee for the mercies thou hast promised. \textcolor{benred8}{59}~Have I not planned out my path, turned aside to follow thy decrees? \textcolor{benred8}{60}~Haste such as mine can brook no delay in carrying out all thy bidding. \textcolor{benred8}{61}~Hemmed in by the snares which sinners laid for me, never was I forgetful of thy law. \textcolor{benred8}{62}~Hearken when I rise at dead of night to praise thee for thy just dealings. \textcolor{benred8}{63}~How well I love the souls that fear thee, and are true to thy trust! \textcolor{benred8}{64}~How thy mercy fills the earth, Lord! Teach me to do thy will.
\textcolor{benred8}{65}~In fulfilment of thy promise, Lord, what kindness thou hast shewn thy servant! \textcolor{benred8}{66}~Inspire, instruct me still; all my hope is in thy covenant. \textcolor{benred8}{67}~Idly I strayed till thou didst chasten me; no more shall thy warnings go unheeded. \textcolor{benred8}{68}~Indeed, indeed thou art gracious; teach me to do thy bidding. \textcolor{benred8}{69}~In vain my oppressors plot against me; thy will is all my quest. \textcolor{benred8}{70}~Inhuman hearts, curdled with scorn! For me, thy law is enough. \textcolor{benred8}{71}~It was in mercy thou didst chasten me, schooling me to thy obedience. \textcolor{benred8}{72}~Is not the law thou hast given dearer to me than rich store of gold and silver?
\textcolor{benred8}{73}~Jealous for the handiwork thou hast made, teach me to understand thy commandments. \textcolor{benred8}{74}~Joy shall be theirs, thy true worshippers, to see the confidence I have in thy word. \textcolor{benred8}{75}~Just are thy awards; I know well, Lord, it was in faithfulness thou didst afflict me. \textcolor{benred8}{76}~Judge me no more; pity and comfort thy servant as thou hast promised. \textcolor{benred8}{77}~Judge me no more; pardon and life for one that loves thy will! \textcolor{benred8}{78}~Just be their fall, who wrong me scornfully; thy law is all my study. \textcolor{benred8}{79}~Joined to my company be every soul that worships thee and heeds thy warnings. \textcolor{benred8}{80}~Jealously let my heart observe thy bidding; let me not hope in vain.
\textcolor{benred8}{81}~Keeping watch for thy aid, my soul languishes, yet I trust in thy word. \textcolor{benred8}{82}~Keeping watch for the fulfilment of thy promise, my eyes languish for comfort still delayed. \textcolor{benred8}{83}~Kitchen-smoke shrivels the wine-skin; so waste I, yet never forget thy will. \textcolor{benred8}{84}~Knowest thou not how short are thy servant\textquotesingle s days? Soon be my wrongs redressed. \textcolor{benred8}{85}~Knaves will be plotting against me still, that are no friends to thy law. \textcolor{benred8}{86}~Knaves they are that wrong me; bring aid, as thy covenant stands unchanging. \textcolor{benred8}{87}~Keep thy bidding I would, though small hope of life they had left me. \textcolor{benred8}{88}~Kind as thou ever wert, preserve me; then utter thy bidding, and I will obey.
\textcolor{benred8}{89}~Lord, the word thou hast spoken stands ever unchanged as heaven. \textcolor{benred8}{90}~Loyal to his promise, age after age, is he who made the enduring earth. \textcolor{benred8}{91}~Long as time lasts, these shall stand, obeying thy decree, Master of all. \textcolor{benred8}{92}~Lest I should sink in my affliction, thou hast given thy covenant to be my comfort. \textcolor{benred8}{93}~Life-giving are thy commands, never by me forgotten. \textcolor{benred8}{94}~Lend me thy aid, for thine I am, and thy bidding is all my quest. \textcolor{benred8}{95}~Let sinners go about to destroy me, I wait on thy will. \textcolor{benred8}{96}~Look where I may, all good things must end; only thy law is wide beyond measure.
\textcolor{benred8}{97}~My delight, Lord, is in thy bidding; ever my thoughts return to it. \textcolor{benred8}{98}~Musing still on thy commandments, I have grown more prudent than my enemies. \textcolor{benred8}{99}~More wisdom have I than all my teachers, so well have I pondered thy decrees. \textcolor{benred8}{100}~More learning have I than my elders, I that hold true to thy charge. \textcolor{benred8}{101}~Mindful of thy warnings, I guide my steps clear of every evil path. \textcolor{benred8}{102}~Meek under thy tuition, thy will I keep ever in view. \textcolor{benred8}{103}~Meat most appetizing are thy promises; never was honey so sweet to my taste. \textcolor{benred8}{104}~Made wise by thy law, I shun every path of evil-doing.
\textcolor{benred8}{105}~No lamp like thy word to guide my feet, to shew light on my path. \textcolor{benred8}{106}~Never will I retract my oath to give thy just commands observance. \textcolor{benred8}{107}~Nothing, Lord, but affliction, never the saving help thou didst promise me? \textcolor{benred8}{108}~Nay, Lord, accept these vows of mine; teach me to do thy bidding. \textcolor{benred8}{109}~Needs must I carry my life in my hands, yet am I ever mindful of thy law. \textcolor{benred8}{110}~Nearly the snares of the wicked caught my feet, yet would I not swerve from thy obedience. \textcolor{benred8}{111}~Now and ever thy covenant is my prize, is my heart\textquotesingle s comfort. \textcolor{benred8}{112}~Now and ever to do thy will perfectly is my heart\textquotesingle s aim.
\textcolor{benred8}{113}~Out upon the men that play traitor to the law I love! \textcolor{benred8}{114}~Other defence, other shield have I none; in thy law I trust. \textcolor{benred8}{115}~Out of my path, lovers of wrong; I will keep my God\textquotesingle s commandments. \textcolor{benred8}{116}~Only let thy promised aid preserve me; do not disappoint me of the hope I cherish. \textcolor{benred8}{117}~Only do thou sustain me in safety, looking ever to thy will. \textcolor{benred8}{118}~Obey thee who will not, shall earn thy disdain; idle is all their scheming. \textcolor{benred8}{119}~Outcasts they are that profane the land with wrong; for me, thy law is enough. \textcolor{benred8}{120}~Overcome is my whole being with the fear of thee; I am adread of thy judgements.
\textcolor{benred8}{121}~Protect the justice of my cause; never leave me at the mercy of my oppressors. \textcolor{benred8}{122}~Pledge thyself still to befriend me; save me from the oppression of my enemies. \textcolor{benred8}{123}~Pining away, I look for thy saving help, the faithful keeping of thy promises. \textcolor{benred8}{124}~Pity thy own servant, and teach him thy decrees. \textcolor{benred8}{125}~Perfect in thy own servant\textquotesingle s heart the knowledge of thy will. \textcolor{benred8}{126}~Put off the hour, Lord, no more; too long thy commandment stands defied. \textcolor{benred8}{127}~Precious beyond gold or jewel I hold thy law. \textcolor{benred8}{128}~Prized be every decree of thine; forsworn be every path of evil-doing.
\textcolor{benred8}{129}~Right wonderful thy decrees are, hard to read, and well my heart heeds them. \textcolor{benred8}{130}~Revelation and light thy words disclose to the simple. \textcolor{benred8}{131}~Rises ever a sigh from my lips as I long after thy covenant. \textcolor{benred8}{132}~Regard and pity me, as thou hast pity for all that love thy name. \textcolor{benred8}{133}~Rule thou my path as thou hast promised; never be wrong-doing my master. \textcolor{benred8}{134}~Rescue me from man\textquotesingle s oppression, to wait henceforth on thy bidding. \textcolor{benred8}{135}~Restore to thy servant the smile of thy living favour, and teach him to know thy will. \textcolor{benred8}{136}~Rivers of tears flow from my eyes, to see thy law forgotten.
\textcolor{benred8}{137}~So just, Lord, thou art, thy awards so truly given! \textcolor{benred8}{138}~Strict justice and utter faithfulness inspire all thy decrees. \textcolor{benred8}{139}~Stung by love\textquotesingle s jealousy, I watch my enemies defy thy bidding. \textcolor{benred8}{140}~Shall not I, thy servant, love thy promises, tested and found true? \textcolor{benred8}{141}~Still despised and disinherited, I do not forget thy charge. \textcolor{benred8}{142}~Stands thy faithfulness eternally, thy law for ever changeless. \textcolor{benred8}{143}~Sorrow and distress have fallen on me; in thy commandments is all my comfort. \textcolor{benred8}{144}~Sentence eternal is thy decree; teach me the wisdom that brings life.
\textcolor{benred8}{145}~Thy audience, Lord, my whole heart claims, a heart true to thy trust. \textcolor{benred8}{146}~To thee I cry, O grant deliverance; I will do all thy bidding. \textcolor{benred8}{147}~Twilight comes, and I awake to plead with thee, hoping ever in thy promises. \textcolor{benred8}{148}~Through the night my eyes keep watch, to ponder thy sayings. \textcolor{benred8}{149}~Thine, Lord, to listen in thy mercy, and grant life according to thy will. \textcolor{benred8}{150}~Treacherous foes draw near, that are strangers to thy covenant. \textcolor{benred8}{151}~Thou, Lord, art close at hand; all thy awards are true. \textcolor{benred8}{152}~Taught long since by thy decrees, I know well thou hast ordained them everlastingly.
\textcolor{benred8}{153}~Unblessed is my lot; look down and rescue me, that still am mindful of thy law. \textcolor{benred8}{154}~Uphold my cause, and deliver me; true to thy promise, grant me life. \textcolor{benred8}{155}~Unknown thy mercy to the sinner that defies thy bidding. \textcolor{benred8}{156}~Unnumbered, Lord, are thy blessings; as thy will is, grant me life. \textcolor{benred8}{157}~Under all the assaults of my oppressors, I keep true to thy charge. \textcolor{benred8}{158}~Unhappy I, that watch thy warnings to the sinner go unheeded! \textcolor{benred8}{159}~Up, Lord, and witness the love I bear thy covenant; in thy mercy bid me live! \textcolor{benred8}{160}~Unchanging truth is thy word\textquotesingle s fountain-head, eternal the force of thy just decrees.
\textcolor{benred8}{161}~Vexed by the causeless malice of princes, my heart still dreads thy warnings. \textcolor{benred8}{162}~Victors rejoice not more over rich spoils, than I in thy promises. \textcolor{benred8}{163}~Villainy I abhor and renounce; thy law is all my love. \textcolor{benred8}{164}~Votive thanks seven times a day I give thee for the just awards thou makest. \textcolor{benred8}{165}~Very great peace is theirs who love thy law; their feet never stumble. \textcolor{benred8}{166}~Valiantly, Lord, I wait on thee for succour, keeping ever true to thy charge. \textcolor{benred8}{167}~Vanquished by great love, my heart is ever obedient to thy will. \textcolor{benred8}{168}~Vigilantly I observe precept and bidding of thine, living always as in thy sight.
\textcolor{benred8}{169}~Wilt thou not admit my cry, Lord, to thy presence, and grant me thy promised gift of wisdom? \textcolor{benred8}{170}~Wilt thou not countenance my plea, redeem thy pledge to deliver me? \textcolor{benred8}{171}~What praise shall burst from my lips, when thou makest known thy will! \textcolor{benred8}{172}~What hymns of thankfulness this tongue shall raise to the author of all just decrees! \textcolor{benred8}{173}~Wouldst thou but lift thy hand to aid me, that take my stand on thy covenant! \textcolor{benred8}{174}~Weary it is, Lord, waiting for deliverance, but thy law is my comfort. \textcolor{benred8}{175}~When will thy just award grant redress, that I may live to praise thee? \textcolor{benred8}{176}~Wayward thou seest me, like a lost sheep; come to look for thy servant, that is mindful still of thy bidding.
\begin{large}\begin{center}\textsc{Chapter cxviii}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~Not unheeded I cry to the Lord in the hour of my distress. \textcolor{benred8}{2}~Lord, have pity and deliver me from the treacherous lips, the perjured tongue. \textcolor{benred8}{3}~Perjurer, he will give thee all thy deserts and more; \textcolor{benred8}{4}~sharp arrows from a warrior\textquotesingle s bow, blazing faggots of broom. \textcolor{benred8}{5}~Unhappy I, that live an exile in Mosoch, or dwell among the tents of Cedar! \textcolor{benred8}{6}~Long banished here among the enemies of peace, \textcolor{benred8}{7}~for peace I plead, and their cry is still for battle.
\begin{large}\begin{center}\textsc{Chapter cxix}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~I lift up my eyes to the hills, to find deliverance; \textcolor{benred8}{2}~from the Lord deliverance comes to me, the Lord who made heaven and earth. \textcolor{benred8}{3}~Never will he who guards thee allow thy foot to stumble; never fall asleep at his post! \textcolor{benred8}{4}~Such a guardian has Israel, one who is never weary, never sleeps; \textcolor{benred8}{5}~it is the Lord that guards thee, the Lord that stands at thy right hand to give thee shelter. \textcolor{benred8}{6}~The sun\textquotesingle s rays by day, the moon\textquotesingle s by night, shall have no power to hurt thee. \textcolor{benred8}{7}~The Lord will guard thee from all evil; the Lord will protect thee in danger; \textcolor{benred8}{8}~the Lord will protect thy journeying and thy home-coming, henceforth and for ever.
\begin{large}\begin{center}\textsc{Chapter cxx}\end{center}\end{large}
(A song of ascents. Of David.)
\textcolor{benred8}{1}~Welcome sound, when I heard them saying, We will go into the Lord\textquotesingle s house! \textcolor{benred8}{2}~Within thy gates, Jerusalem, our feet stand at last; \textcolor{benred8}{3}~Jerusalem, built as a city should be built that is one in fellowship. \textcolor{benred8}{4}~There the tribes meet, the Lord\textquotesingle s own tribes, to give praise, as Israel is ever bound, to the Lord\textquotesingle s name; \textcolor{benred8}{5}~there the thrones are set for judgement, thrones for the house of David. \textcolor{benred8}{6}~Pray for all that brings Jerusalem peace! May all who love thee dwell at ease! \textcolor{benred8}{7}~Let there be peace within thy ramparts, ease in thy strongholds! \textcolor{benred8}{8}~For love of my brethren and my familiar friends, peace is still my prayer for thee; \textcolor{benred8}{9}~remembering the house of the Lord our God, for thy happiness I plead.
\begin{large}\begin{center}\textsc{Chapter cxxi}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~Unto thee I lift up my eyes, unto thee, who dwellest in the heavens. \textcolor{benred8}{2}~See how the eyes of servants are fixed on the hands of their masters, the eyes of a maid on the hand of her mistress! Our eyes, too, are fixed on the Lord our God, waiting for him to shew mercy on us. \textcolor{benred8}{3}~Have mercy on us, Lord, have mercy on us; we have had our fill of man\textquotesingle s derision. \textcolor{benred8}{4}~Our hearts can bear no more to be the scorn of luxury, the derision of the proud.
\begin{large}\begin{center}\textsc{Chapter cxxii}\end{center}\end{large}
(A song of ascents. Of David.)
\textcolor{benred8}{1}~If the Lord had not been on our side, Israel may boast, \textcolor{benred8}{2}~if the Lord had not been on our side when human foes assailed us, \textcolor{benred8}{3}~it seemed as if they must have swallowed us up alive, so fierce their anger threatened us. \textcolor{benred8}{4}~It seemed as if the tide must have sucked us down, the torrent closed above us; \textcolor{benred8}{5}~closed above us the waters that ran so high. \textcolor{benred8}{6}~Blessed be the Lord, who has not let us fall a prey to those ravening mouths! \textcolor{benred8}{7}~Safe, like a bird rescued from the fowler\textquotesingle s snare; the snare is broken and we are safe! \textcolor{benred8}{8}~Such help is ours, the Lord\textquotesingle s help, that made heaven and earth.
\begin{large}\begin{center}\textsc{Chapter cxxiii}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~Those who trust in the Lord are strong as mount Sion itself, that stands \textcolor{benred8}{2}~unmoved for ever. The hills protect Jerusalem; so the Lord protects his people, now and for ever. \textcolor{benred8}{3}~Domain of the just! No longer shall godless men bear rule in it; else the just, too, might soil their hands with guilt. \textcolor{benred8}{4}~Deal kindly, Lord, with the kindly, with the true-hearted. \textcolor{benred8}{5}~Feet that stray into false paths the Lord will punish, as he punishes wrong-doers; but upon Israel there shall be peace.
\begin{large}\begin{center}\textsc{Chapter cxxiv}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~When the Lord gave back Sion her banished sons, we walked like men in a dream; \textcolor{benred8}{2}~in every mouth was laughter, joy was on every tongue. Among the heathen themselves it was said, What favour the Lord has shewn them! \textcolor{benred8}{3}~Favour indeed the Lord has shewn us, and our hearts are rejoiced. \textcolor{benred8}{4}~Deliver us, Lord, from our bondage; our withered hopes, Lord, like some desert water-course renew!\footnote[1]{Some commentators think that the process of deliverance was still incomplete, only a small number of exiles (for example) having returned from captivity; others, that verse 1 expresses, not something which had actually happened, but an imaginary picture.} \textcolor{benred8}{5}~The men who are sowing in tears will reap, one day, with joy. \textcolor{benred8}{6}~Mournful enough they go, but with seed to scatter; ~trust me, they will come back rejoicing, as they carry their sheaves with them.
\begin{large}\begin{center}\textsc{Chapter cxxv}\end{center}\end{large}
(A song of ascents. Of Solomon.)
\textcolor{benred8}{1}~Vain is the builder\textquotesingle s toil, if the house is not of the Lord\textquotesingle s building; vainly the guard keeps watch, if the city has not the Lord for its guardian. \textcolor{benred8}{2}~Vain, that you should be astir before daybreak, and sit on over your tasks late into the night, you whose bread is so hardly won; is it not in the hours of sleep that he blesses the men he loves?\footnote[1]{The verse is not an incitement to idleness; the gist of the psalm lies in what follows. What is the use of a house to a childless man, of a well-built city with a dwindling number of inhabitants; what is the use of toiling day and night, unless you have a family to provide for?} \textcolor{benred8}{3}~Fatherhood itself is the Lord\textquotesingle s gift, the fruitful womb is a reward that comes from him. \textcolor{benred8}{4}~Crown of thy youth, children are like arrows in a warrior\textquotesingle s hand. \textcolor{benred8}{5}~Happy, whose quiver is well filled with these; their cause will not be set aside when they plead against their enemies at the gate.
\begin{large}\begin{center}\textsc{Chapter cxxvi}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~Blessed thou art, if thou dost fear the Lord, and follow his paths! \textcolor{benred8}{2}~Thyself shall eat what thy hands have toiled to win; blessed thou art; all good shall be thine. \textcolor{benred8}{3}~Thy wife shall be fruitful as a vine, in the heart of thy home, the children round thy table sturdy as olive-branches. \textcolor{benred8}{4}~Let a man serve the Lord, such is the blessing that awaits him. \textcolor{benred8}{5}~May the Lord who dwells in Sion bless thee; mayest thou see Jerusalem in prosperity all thy life long. \textcolor{benred8}{6}~Mayest thou live to see thy children\textquotesingle s children, and peace resting upon Israel.
\begin{large}\begin{center}\textsc{Chapter cxxvii}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~Sore have they beset me even from my youth (let this be Israel\textquotesingle s boast); \textcolor{benred8}{2}~sore have they beset me even from my youth, but never once outmatched me. \textcolor{benred8}{3}~I bent my back to the oppressor, and long was the furrow ere the plough turned; \textcolor{benred8}{4}~but the Lord proved faithful, and cut the bonds of tyranny asunder. \textcolor{benred8}{5}~Let them be dismayed and routed, all these enemies of Sion. \textcolor{benred8}{6}~Let them be like the stalks on a house-top, that wither there unharvested; \textcolor{benred8}{7}~never will they be grasped in the reaper\textquotesingle s hand, or fill the binder\textquotesingle s bosom, \textcolor{benred8}{8}~no passer-by will say, The Lord\textquotesingle s blessing on you; we bless you in the name of the Lord.\footnote[1]{Cf. Ruth 2.4.}
\begin{large}\begin{center}\textsc{Chapter cxxviii}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~Out of the depths I cry to thee, O Lord; \textcolor{benred8}{2}~Master, listen to my voice; let but thy ears be attentive to the voice that calls on thee for pardon. \textcolor{benred8}{3}~If thou, Lord, wilt keep record of our iniquities, Master, who has strength to bear it? \textcolor{benred8}{4}~Ah, but with thee there is forgiveness; be thy name ever revered. \textcolor{benred8}{5}~I wait for the Lord, for his word of promise my soul waits; \textcolor{benred8}{6}~patient my soul waits, as ever watchman that looked for the day. Patient as watchman at dawn, \textcolor{benred8}{7}~for the Lord Israel waits, the Lord with whom there is mercy, with whom is abundant power to ransom. \textcolor{benred8}{8}~He it is that will ransom Israel from all his iniquities.
\begin{large}\begin{center}\textsc{Chapter cxxix}\end{center}\end{large}
(A song of ascents. Of David.)
\textcolor{benred8}{1}~Lord, my heart is not lifted up, my eyes not raised from the earth; my mind does not dwell on high things, on marvels that are beyond my reach. \textcolor{benred8}{2}~Bear me witness that I kept my soul ever quiet, ever at peace. The thoughts of a child on its mother\textquotesingle s breast, a child\textquotesingle s thoughts were all my soul knew. \textcolor{benred8}{3}~Let Israel trust in the Lord, henceforth and for ever.
\begin{large}\begin{center}\textsc{Chapter cxxx}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~In David\textquotesingle s reckoning, Lord, let not his patient care be forgotten, \textcolor{benred8}{2}~the oath he swore to the Lord, the vow he made to the great God of Jacob: \textcolor{benred8}{3}~Never will I come beneath the roof of my house, or climb up into the bed that is strewn for me; \textcolor{benred8}{4}~never shall these eyes have sleep, these eye-lids close, \textcolor{benred8}{5}~until I have found the Lord a home, the great God of Jacob a dwelling-place. \textcolor{benred8}{6}~And now, at Ephrata, we have heard tidings of what we looked for, we have found it in the plains of Jaar;\footnote[1]{Literally, \textasciigrave Behold, we have heard of it at Ephrata, and found it in the plains of Jaar\textquotesingle . If this psalm was composed for the dedication of Solomon\textquotesingle s Temple (II Par. 6.41, 42) and commemorates the bringing back of the Ark to Jerusalem (I Par. 13), we should naturally suppose that \textasciigrave it\textquotesingle  means the Ark. It is not clear that the Ark ever rested at Bethlehem-Ephrata; but some think that Ephrata here is another name for Silo (I Kg. 4.3), and that the other half of the verse refers to Cariathiarim, the \textasciigrave city of the woods\textquotesingle  (I Par. 13.5).} \textcolor{benred8}{7}~now to go into his dwelling, pay reverence at his footstool! \textcolor{benred8}{8}~Up, Lord, and take possession of thy resting-place, thou and the ark which is shrine of thy glory! \textcolor{benred8}{9}~Let thy priests go clad in the vesture of innocence, thy faithful people cry aloud with rejoicing.
\textcolor{benred8}{10}~Think of thy servant David, and do not refuse audience to the king thou hast anointed. \textcolor{benred8}{11}~Never will the Lord be false to that inviolable oath he swore to David: I will raise to thy throne heirs of thy own body; \textcolor{benred8}{12}~if thy sons hold fast to my covenant, to the decrees which I make known to them, their sons too shall reign on thy throne for ever. \textcolor{benred8}{13}~The Lord\textquotesingle s choice has fallen upon Sion, this is the dwelling he longed for: \textcolor{benred8}{14}~Here, for ever, is my resting-place, here is my destined home. \textcolor{benred8}{15}~Trust me, I will bless her with abundant store, the poor shall have bread to their heart\textquotesingle s content; \textcolor{benred8}{16}~I will clothe her priests in the vesture of triumph, cries of rejoicing shall echo among her faithful people. \textcolor{benred8}{17}~There the stock of David shall bud, there shall a lamp burn continually for the king I have anointed.\footnote[2]{Literally, \textasciigrave There will I make a horn grow for David, I have trimmed a lamp for my anointed\textquotesingle , both metaphors for the continuance of his royal dynasty (III Kg. 15.4; Lk. 1.69).} \textcolor{benred8}{18}~I will cover his enemies with confusion; on his brow the crown I gave shall shine untarnished.
\begin{large}\begin{center}\textsc{Chapter cxxxi}\end{center}\end{large}
(A song of ascents. Of David.)
\textcolor{benred8}{1}~Gracious the sight, and full of comfort, when brethren dwell united. \textcolor{benred8}{2}~Gracious as balm poured on the head till it flows down on to the beard; balm that flowed down Aaron\textquotesingle s beard, and reached the very skirts of his robe. \textcolor{benred8}{3}~It is as if dew like the dews of Hermon were falling on this hill of Sion;\footnote[1]{Literally, \textasciigrave like the dew of Hermon that falls on the hill of Sion\textquotesingle , which, however, was more than a hundred miles distant. It seems clear, therefore, that the dews of Hermon are only mentioned here as typical of exceptionally heavy dews; unless those authors are right who suspect that the reference is to a mount Sion, differently spelt in Hebrew, which was part of the Hermon range (Deut. 4.48).} here, where the Lord grants benediction and life everlastingly.
\begin{large}\begin{center}\textsc{Chapter cxxxii}\end{center}\end{large}
(A song of ascents.)
\textcolor{benred8}{1}~Come, then, praise the Lord, all you that are the Lord\textquotesingle s servants; you that wait on the Lord\textquotesingle s house at midnight, \textcolor{benred8}{2}~lift up your hands towards the sanctuary and bless the Lord. \textcolor{benred8}{3}~May the Lord who dwells in Sion bless thee, the Lord who made heaven and earth!
\begin{large}\begin{center}\textsc{Chapter cxxxiii}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Praise the Lord\textquotesingle s name; praise the Lord, you that are his servants, \textcolor{benred8}{2}~you who stand in the house of the Lord, in the courts where our God dwells. \textcolor{benred8}{3}~Praise to the Lord, a Lord so gracious, praise to his name, a name so well beloved. \textcolor{benred8}{4}~Has not the Lord made choice of Jacob, claimed Israel for his own?
\textcolor{benred8}{5}~Doubt it never, the Lord is great; he, our Master, is higher than all the gods. \textcolor{benred8}{6}~In heaven and on earth, in the sea and in the deep waters beneath us, the Lord accomplishes his will; \textcolor{benred8}{7}~summoning clouds from the ends of the earth, rain-storm wedding to lightning-flash, bringing winds out of his store-house. \textcolor{benred8}{8}~He it was that smote the first-born of the Egyptians, man and beast alike; \textcolor{benred8}{9}~what wonders and portents, Egypt, thou didst witness, sent to plague Pharao and all his servants! \textcolor{benred8}{10}~He it was that smote nation after nation, and slew the kings in their pride, \textcolor{benred8}{11}~Sehon king of the Amorrhites, and Og the king of Basan, and all the rulers of Chanaan, \textcolor{benred8}{12}~and marked down their lands for a dwelling-place where his own people of Israel should dwell.
\textcolor{benred8}{13}~Lord, thy name abides for ever; age succeeds age, and thou art ever unforgotten. \textcolor{benred8}{14}~The Lord defends his people, takes pity on his servants. \textcolor{benred8}{15}~What are the idols of the heathen but silver and gold, gods which the hands of men have fashioned? \textcolor{benred8}{16}~They have mouths, and yet are silent; eyes they have, and yet are sightless; \textcolor{benred8}{17}~ears they have, and want all hearing, never a breath have they in their mouths. \textcolor{benred8}{18}~Such the end of all who make them, such the reward of all who trust them. \textcolor{benred8}{19}~Bless the Lord, sons of Israel, bless the Lord, sons of Aaron, \textcolor{benred8}{20}~bless the Lord, sons of Levi, bless the Lord, all you that are the Lord\textquotesingle s worshippers. \textcolor{benred8}{21}~Here, in Sion his dwelling-place, here, in Jerusalem, let the Lord\textquotesingle s name be blessed.
\begin{large}\begin{center}\textsc{Chapter cxxxiv}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Give thanks to the Lord for his goodness, his mercy is eternal; \textcolor{benred8}{2}~give thanks to the God of gods, his mercy is eternal; \textcolor{benred8}{3}~give thanks to the Lord of lords, his mercy is eternal. \textcolor{benred8}{4}~Eternal his mercy, who does great deeds as none else can; \textcolor{benred8}{5}~eternal his mercy, whose wisdom made the heavens; \textcolor{benred8}{6}~eternal his mercy, who poised earth upon the floods. \textcolor{benred8}{7}~Eternal his mercy, who made the great luminaries; \textcolor{benred8}{8}~made the sun to rule by day, his mercy is eternal; \textcolor{benred8}{9}~made the moon and the stars to rule by night, his mercy is eternal.
\textcolor{benred8}{10}~Eternal his mercy, who smote the Egyptians by smiting their first-born; \textcolor{benred8}{11}~eternal his mercy, who delivered Israel from their midst, \textcolor{benred8}{12}~with constraining power, with his arm raised on high, his mercy is eternal. \textcolor{benred8}{13}~Eternal the mercy that divided the Red Sea in two, \textcolor{benred8}{14}~eternal the mercy that led Israel through its waters, \textcolor{benred8}{15}~eternal the mercy that drowned in the Red Sea Pharao and Pharao\textquotesingle s men. \textcolor{benred8}{16}~And so he led his people through the wilderness, his mercy is eternal.
\textcolor{benred8}{17}~Eternal the mercy that smote great kings, \textcolor{benred8}{18}~eternal the mercy that slew the kings in their pride, \textcolor{benred8}{19}~Sehon king of the Amorrhites, his mercy is eternal, \textcolor{benred8}{20}~and Og the king of Basan, his mercy is eternal. \textcolor{benred8}{21}~Eternal the mercy that marked down their land to be a dwelling-place; \textcolor{benred8}{22}~a dwelling-place for his servant Israel, his mercy is eternal. \textcolor{benred8}{23}~Eternal the mercy that remembers us in our affliction, \textcolor{benred8}{24}~eternal the mercy that rescues us from our enemies, \textcolor{benred8}{25}~eternal the mercy that gives all living things their food. \textcolor{benred8}{26}~Give thanks to the God of heaven, his mercy is eternal.
\begin{large}\begin{center}\textsc{Chapter cxxxv}\end{center}\end{large}
\lettrine[lines=2]{W}{e} sat down by the streams of Babylon and wept there, remembering Sion. \textcolor{benred8}{2}~Willow-trees grow there, and on these we hung up our harps \textcolor{benred8}{3}~when the men who took us prisoner cried out for a song. We must make sport for our enemies; A stave, there, from the music they sing at Sion! \textcolor{benred8}{4}~What, should we sing the Lord\textquotesingle s song in a strange land? \textcolor{benred8}{5}~Jerusalem, if I forget thee, perish the skill of my right hand! \textcolor{benred8}{6}~Let my tongue stick fast to the roof of my mouth if I cease to remember thee, if I love not Jerusalem dearer than heart\textquotesingle s content! \textcolor{benred8}{7}~Remember, Lord, how the sons of Edom triumphed when Jerusalem fell; O\textquotesingle erthrow it, they cried, o\textquotesingle erthrow it, till the very foundation is left bare. \textcolor{benred8}{8}~Babylon, pitiless queen, blessed be the man who deals out to thee the measure thou hast dealt to us; \textcolor{benred8}{9}~blessed be the man who will catch up thy children, and dash them against the rocks!
\begin{large}\begin{center}\textsc{Chapter cxxxvi}\end{center}\end{large}
(Of David.)
\textcolor{benred8}{1}~My heart\textquotesingle s thanks, Lord, for listening to the prayer I uttered; angels for my witnesses,\footnote[1]{The Hebrew word here translated \textasciigrave angels\textquotesingle  means literally \textasciigrave gods\textquotesingle . Some think it refers, as in Ps. 81, to earthly rulers, cf. verse 4 below.} I will sing of thy praise. \textcolor{benred8}{2}~I bow down in worship towards thy sanctuary, praising thy name for thy mercy and faithfulness; thy own honour and thy pledged word thou hast vindicated for all the world to see. \textcolor{benred8}{3}~To thee I appealed, and thou didst listen to me, didst fill my heart with courage. \textcolor{benred8}{4}~All the kings of the earth, Lord, will praise thee now; were not thy promises made in their hearing? \textcolor{benred8}{5}~Their song shall be of the Lord\textquotesingle s doings, how great is his renown, \textcolor{benred8}{6}~the Lord, who is so high above us, yet looks with favour on the humble, looks on the proud too, but from far off. \textcolor{benred8}{7}~Though affliction surround my path, thou dost preserve me; it is thy power that confronts my enemies\textquotesingle  malice, thy right hand that rescues me. \textcolor{benred8}{8}~My purposes the Lord will yet speed; thy mercy, Lord, endures for ever, and wilt thou abandon us, the creatures of thy own hands?
\begin{large}\begin{center}\textsc{Chapter cxxxvii}\end{center}\end{large}
(To the choir-master. Of David. A psalm.)
\textcolor{benred8}{1}~Lord, I lie open to thy scrutiny; thou knowest me, \textcolor{benred8}{2}~knowest when sit down and when I rise up again, canst read my thoughts from far away. \textcolor{benred8}{3}~Walk I or sleep I, thou canst tell; no movement of mine but thou art watching it. \textcolor{benred8}{4}~Before ever the words are framed on my lips, all my thought is known to thee; \textcolor{benred8}{5}~rearguard and vanguard, thou dost compass me about, thy hand still laid upon me. \textcolor{benred8}{6}~Such wisdom as thine is far beyond my reach, no thought of mine can attain it.
\textcolor{benred8}{7}~Where can I go, then, to take refuge from thy spirit, to hide from thy view? \textcolor{benred8}{8}~If I should climb up to heaven, thou art there; if I sink down to the world beneath, thou art present still. \textcolor{benred8}{9}~If I could wing my way eastwards, or find a dwelling beyond the western sea, \textcolor{benred8}{10}~still would I find thee beckoning to me, thy right hand upholding me. \textcolor{benred8}{11}~Or perhaps I would think to bury myself in darkness; night should surround me, friendlier than day; \textcolor{benred8}{12}~but no, darkness is no hiding-place from thee, with thee the night shines clear as day itself; light and dark are one.
\textcolor{benred8}{13}~Author, thou, of my inmost being, didst thou not form me in my mother\textquotesingle s womb? \textcolor{benred8}{14}~I praise thee for my wondrous fashioning, for all the wonders of thy creation. Of my soul thou hast full knowledge, \textcolor{benred8}{15}~and this mortal frame had no mysteries for thee, who didst contrive it in secret, devise its pattern, there in the dark recesses of the earth. \textcolor{benred8}{16}~All my acts thy eyes have seen, all are set down already in thy record; my days were numbered before ever they came to be.
\textcolor{benred8}{17}~A riddle, O my God, thy dealings with me, so vast their scope! \textcolor{benred8}{18}~As well count the sand, as try to fathom them; and, were that skill mine, thy own being still confronts me. \textcolor{benred8}{19}~O God, wouldst thou but make an end of the wicked! Murderers, keep your distance from me! \textcolor{benred8}{20}~Treacherously they rebel against thee, faithlessly set thee at defiance. \textcolor{benred8}{21}~Lord, do I not hate the men who hate thee, am I not sick at heart over their rebellion? \textcolor{benred8}{22}~Surpassing hatred I bear them, count them my sworn enemies. \textcolor{benred8}{23}~Scrutinize me, O God, as thou wilt, and read my heart; put me to the test, and examine my restless thoughts. \textcolor{benred8}{24}~See if on any false paths my heart is set, and thyself lead me in the ways of old.
\begin{large}\begin{center}\textsc{Chapter cxxxviii}\end{center}\end{large}
(To the choir-master. A psalm. Of David.)
\textcolor{benred8}{2}~Rescue me, Lord, from human malice, save me from the lovers of oppression, \textcolor{benred8}{3}~always plotting treachery in their hearts, always at their quarrelling, \textcolor{benred8}{4}~tongues sharp as the tongues of serpents, lips that hide the poison of adders. \textcolor{benred8}{5}~Preserve me Lord, from the power of sinful men, save me from these lovers of oppression who are plotting to trip my feet. \textcolor{benred8}{6}~What hidden snares they set for me, these tyrants, what nets they spread to catch me, what traps they lay in my path!
\textcolor{benred8}{7}~To the Lord I make my appeal, Thou art my God, listen, Lord, to the voice that pleads with thee. \textcolor{benred8}{8}~My Lord, my Master, my strong deliverer, it is thou that shieldest my head in the day of battle. \textcolor{benred8}{9}~Lord, do not let malice have its way with me, do not prosper its evil designs. \textcolor{benred8}{10}~They carry their heads high as they close in around me; let their conspiracy prove its own undoing; \textcolor{benred8}{11}~let burning coals rain down on them, be they cast into a pit whence they shall rise no more. \textcolor{benred8}{12}~Not long the blasphemer\textquotesingle s time on earth: misfortune will overtake the oppressor unawares. \textcolor{benred8}{13}~Can I doubt that the Lord will avenge the helpless, will grant the poor redress? \textcolor{benred8}{14}~Honest men will yet live to praise thy name; upright hearts to enjoy the smile of thy favour.
\begin{large}\begin{center}\textsc{Chapter cxxxix}\end{center}\end{large}
(A psalm. Of David.)
\textcolor{benred8}{1}~Come quickly, Lord, at my cry for succour; do not let my appeal to thee go unheard. \textcolor{benred8}{2}~Welcome as incense-smoke let my prayer rise up before thee; when I lift up my hands, be it acceptable as the evening sacrifice. \textcolor{benred8}{3}~Lord, set a guard on my mouth, post a sentry before my lips; \textcolor{benred8}{4}~do not turn my heart towards thoughts of evil, and deeds of treachery; never let me take part with the wrong-doers, and share the banquet with them. \textcolor{benred8}{5}~Rather let some just man deal me heavy blows; this shall be his kindness to me; reprove me, and it shall be balm poured over me; such unction never will this head refuse. Their injuries I will still greet with a prayer~\ldots\ 
\textcolor{benred8}{6}~\ldots\  My words have won their hearts, a people that had seen their chieftains hurled down the rock-face,\footnote[1]{It does not seem likely that the word \textasciigrave their\textquotesingle  in this verse refers to the sinners just mentioned; it is more naturally interpreted of the Jewish people generally. This verse, and the verse which follows, are obscure.} \textcolor{benred8}{7}~a people whose bones lie scattered at the grave\textquotesingle s mouth, like seed when the earth is cloven into furrows.
\textcolor{benred8}{8}~To thee these eyes look, my Lord, my Master; in thee I trust; let not my life be forfeit. \textcolor{benred8}{9}~Preserve me from the ambush they have laid for me, from the snares of the wrong-doers. \textcolor{benred8}{10}~Into their own net, sinner upon sinner, may they fall, and I pass on in safety.
\begin{large}\begin{center}\textsc{Chapter cxl}\end{center}\end{large}
(A maskil. Of David, when he was in the cave. A prayer.)
\textcolor{benred8}{2}~Loud is my cry to the Lord, the prayer I utter for the Lord\textquotesingle s mercy, \textcolor{benred8}{3}~as I pour out my complaint before him, tell him of the affliction I endure. \textcolor{benred8}{4}~My heart is ready to faint within me, but thou art watching over my path. They lie in ambush for me, there by the wayside; \textcolor{benred8}{5}~I look to the right of me, and find none to take my part; all hope of escape is cut off from me, none is concerned for my safety. \textcolor{benred8}{6}~To thee, Lord, I cry, claiming thee for my only refuge, all that is left me in this world of living men. \textcolor{benred8}{7}~Listen, then, to my plea; thou seest me all defenceless. Rescue me from persecutors who are too strong for me; restore liberty to a captive soul. \textcolor{benred8}{8}~What thanks, then, will I give to thy name, honest hearts all about me, rejoicing to see thy favour restored!
\begin{large}\begin{center}\textsc{Chapter cxli}\end{center}\end{large}
(A psalm. Of David.)
\textcolor{benred8}{1}~Listen, Lord, to my prayer; give my plea a hearing, as thou art ever faithful; listen, thou who lovest the right. \textcolor{benred8}{2}~Do not call thy servant to account; what man is there living that can stand guiltless in thy presence? \textcolor{benred8}{3}~See how my enemies plot against my life, how they have abased me in the dust, set me down in dark places, like the long-forgotten dead! \textcolor{benred8}{4}~My spirits are crushed within me, my heart is cowed. \textcolor{benred8}{5}~And my mind goes back to past days; I think of all thou didst once, dwell on the proofs thou gavest of thy power. \textcolor{benred8}{6}~To thee I spread out my hands in prayer, for thee my soul thirsts, like a land parched with drought.
\textcolor{benred8}{7}~Hasten, Lord, to answer my prayer; my spirit grows faint. Do not turn thy face away from me, and leave me like one sunk in the abyss. \textcolor{benred8}{8}~Speedily let me win thy mercy, my hope is in thee; to thee I lift up my heart, shew me the path I must follow; \textcolor{benred8}{9}~to thee I fly for refuge, deliver me, Lord, from my enemies. \textcolor{benred8}{10}~Thou art my God, teach me to do thy will; let thy gracious spirit lead me, safe ground under my feet. \textcolor{benred8}{11}~For the honour of thy own name, Lord, grant me life; in thy mercy rescue me from my cruel affliction. \textcolor{benred8}{12}~Have pity on me, and scatter my enemies; thy servant I; make an end of my cruel persecutors.
\begin{large}\begin{center}\textsc{Chapter cxlii}\end{center}\end{large}
(Of David.)
\textcolor{benred8}{1}~Blessed be the Lord, my refuge, who makes these hands strong for battle, these fingers skilled in fight; \textcolor{benred8}{2}~the Lord who pities me and grants me safety, who shelters me and sets me at liberty, who protects me and gives me confidence, bowing down nations to my will. \textcolor{benred8}{3}~Lord, what is Adam\textquotesingle s race, that thou givest heed to it, what is man, that thou carest for him? \textcolor{benred8}{4}~Like the wind he goes, like a shadow his days pass.
\textcolor{benred8}{5}~Bid heaven stoop, Lord, and come down to earth; at thy touch, the mountains will be wreathed in smoke. \textcolor{benred8}{6}~Brandish thy lightnings, to rout my enemies; shoot thy arrows, and throw them into confusion! \textcolor{benred8}{7}~With heavenly aid, from yonder flood deliver me; rescue me from the power of alien foes, \textcolor{benred8}{8}~who make treacherous promises, and lift their hands in perjury. \textcolor{benred8}{9}~Then, O my God, I will sing thee a new song, on a ten-stringed harp I will sound thy praise; \textcolor{benred8}{10}~the God to whom kings must look for victory, the God who has brought his servant David rescue. \textcolor{benred8}{11}~Save me from the cruel sword, deliver me from the power of alien foes, who make treacherous promises, and lift their hands in perjury.
\textcolor{benred8}{12}~So may our sons grow to manhood, tall as the saplings, our daughters shapely as some column at the turn of a building, it may be, the temple itself. \textcolor{benred8}{13}~Our garners full, well stored with every kind of plenty, our sheep bearing a thousand-fold, thronging the pasture in their tens of thousands, \textcolor{benred8}{14}~our oxen straining at the load; no ruined walls, no exile, no lamenting in our streets. \textcolor{benred8}{15}~Happy men call such a people as this; and is not the people happy, that has the Lord for its God?
\begin{large}\begin{center}\textsc{Chapter cxliii}\end{center}\end{large}
(Praises. Of David.)
\textcolor{benred8}{1}~And shall I not extol thee, my God, my king; shall I not bless thy name for ever and for evermore? \textcolor{benred8}{2}~Blessing shall be thine, day after day; for ever and for evermore praised be thy name. \textcolor{benred8}{3}~Can any praise be worthy of the Lord\textquotesingle s majesty, any thought set limits to his greatness? \textcolor{benred8}{4}~Down the ages the story of thy deeds is told, thy power is ever acclaimed; \textcolor{benred8}{5}~each magnifies thy unapproachable glory, makes known thy wonders. \textcolor{benred8}{6}~Fearful are the tales they tell of thy power, proclaiming thy magnificence; \textcolor{benred8}{7}~grateful their memory of all thy goodness, as they boast of thy just dealings. \textcolor{benred8}{8}~How gracious the Lord is, how merciful, how patient, how rich in pity! \textcolor{benred8}{9}~Is he not a loving Lord to his whole creation; does not his mercy reach out to all that he has made?
\textcolor{benred8}{10}~Joining, then, Lord, in thy whole creation\textquotesingle s praise, let thy faithful servants bless thee; \textcolor{benred8}{11}~let them publish the glory of thy kingdom, and discourse of thy power, \textcolor{benred8}{12}~making that power known to the race of men, the glory, the splendour of that kingdom! \textcolor{benred8}{13}~No age shall dawn but shall see thee reigning still; generations pass, and thy rule shall endure. ~O how true the Lord is to all his promises, how gracious in all his dealings! \textcolor{benred8}{14}~Prostrate though men may fall, the Lord will lift them up, will revive their crushed spirits.
\textcolor{benred8}{15}~Quietly, Lord, thy creatures raise their eyes to thee, and thou grantest them, in due time, their nourishment, \textcolor{benred8}{16}~ready to open thy hand, and fill with thy blessing all that lives. \textcolor{benred8}{17}~So faithful the Lord is in all he does, so gracious in all his dealings. \textcolor{benred8}{18}~The Lord draws near to every man that calls upon him, will he but call upon him with a true heart. \textcolor{benred8}{19}~Utter but the wish, you that fear the Lord, and he will grant it, will hear the cry, and bring aid. \textcolor{benred8}{20}~Vigilantly the Lord watches over all that love him, marks down the wicked for destruction. \textcolor{benred8}{21}~While these lips tell of the Lord\textquotesingle s praise, let all that lives bless his holy name, for ever, and for evermore.
\begin{large}\begin{center}\textsc{Chapter cxliv}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{2}~Praise the Lord, my soul; while life lasts, I will praise the Lord; of him, my God, shall my songs be while I am here to sing them. Do not put your trust in princes; \textcolor{benred8}{3}~they are but men, they have no power to save. \textcolor{benred8}{4}~As soon as the breath leaves his body, man goes back to the dust he belongs to; with that, all his designs will come to nothing. \textcolor{benred8}{5}~Happier the man who turns to the God of Jacob for help, puts no confidence but in the Lord his God, \textcolor{benred8}{6}~maker of heaven and earth and sea and all they contain; \textcolor{benred8}{7}~the God who keeps faith for ever, who redresses wrong, and gives food to the hungry. The Lord, who brings release to the prisoner, \textcolor{benred8}{8}~the Lord, who gives sight to the blind, the Lord, who comforts the burdened, the Lord, who befriends the innocent! \textcolor{benred8}{9}~The Lord, who protects the stranger, who defends orphan and widow, who overturns the counsel of the wicked! \textcolor{benred8}{10}~The Lord, reigning for ever, thy God, Sion, reigning from age to age! Alleluia.
\begin{large}\begin{center}\textsc{Chapter cxlv}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Praise the Lord; the Lord is gracious; sing to our God, a God who so claims our love; praise is his right. \textcolor{benred8}{2}~The Lord is rebuilding Jerusalem, is calling the banished sons of Israel home; \textcolor{benred8}{3}~he it is that heals the broken heart, and binds up its wounds. \textcolor{benred8}{4}~Does he not know the number of the stars, and call each by its name? \textcolor{benred8}{5}~How great a Lord is ours, how magnificent his strength, how inscrutable his wisdom! \textcolor{benred8}{6}~The Lord is the defender of the oppressed, and lays the wicked low in the dust. \textcolor{benred8}{7}~Strike up, then, in thanksgiving to the Lord, with the harp\textquotesingle s music praise our God; \textcolor{benred8}{8}~the God who curtains heaven with cloud, and lays up a store of rain for the earth, who clothes the mountain-sides with grass, with corn for man\textquotesingle s need, \textcolor{benred8}{9}~gives food to the cattle, food to the young ravens that cry out to him. \textcolor{benred8}{10}~Not the well-mounted warrior is his choice, not the swift runner wins his favour; \textcolor{benred8}{11}~the Lord\textquotesingle s favour is for those who fear him, and put their trust in his divine mercy.
\begin{large}\begin{center}\textsc{Chapter cxlvi}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Give praise to the Lord in heaven; praise him, all that dwells on high. \textcolor{benred8}{2}~Praise him, all you angels of his, praise him, all his armies. \textcolor{benred8}{3}~Praise him, sun and moon; praise him, every star that shines. \textcolor{benred8}{4}~Praise him, you highest heavens, you waters beyond the heavens. \textcolor{benred8}{5}~Let all these praise the Lord; it was his command that created them. \textcolor{benred8}{6}~He has set them there unageing for ever, given them a law which cannot be altered.
\textcolor{benred8}{7}~Give praise to the Lord on earth, monsters of the sea and all its depths; \textcolor{benred8}{8}~fire and hail, snow and mist, and the storm-wind that executes his decree; \textcolor{benred8}{9}~all you mountains and hills, all you fruit trees and cedars; \textcolor{benred8}{10}~all you wild beasts and cattle, creeping things and birds that fly in air; \textcolor{benred8}{11}~all you kings and peoples of the world, all you that are princes and judges on earth; \textcolor{benred8}{12}~young men and maids, old men and boys together; let them all give praise to the Lord\textquotesingle s name. \textcolor{benred8}{13}~His name is exalted as no other, \textcolor{benred8}{14}~his praise reaches beyond heaven and earth; and now he has given fresh strength to his people. Shall not his faithful servants praise him, the sons of Israel, the people that draw near to him? Alleluia.
\begin{large}\begin{center}\textsc{Chapter cxlvii}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Sing the Lord a new song; here, where the faithful gather, let his praise be heard. \textcolor{benred8}{2}~In him, the maker of Israel, let Israel triumph; for him, the ruler of Sion, let Sion\textquotesingle s children keep holiday; \textcolor{benred8}{3}~let there be dancing in honour of his name, music of tambour and of harp, to praise him. \textcolor{benred8}{4}~Still the Lord shews favour to his people, still he relieves the oppressed, and grants them victory. \textcolor{benred8}{5}~In triumph let thy faithful servants rejoice, rejoice and take their rest. \textcolor{benred8}{6}~Ever on their lips they bear the high praise of God, ever in their hands they carry two-edged swords, \textcolor{benred8}{7}~ready to take vengeance upon the heathen, \textcolor{benred8}{8}~to curb the nations, to chain the kings, and bind princes in fetters of iron. \textcolor{benred8}{9}~Long since their doom is written; boast it is of his true servants that doom to execute. Alleluia.
\begin{large}\begin{center}\textsc{Chapter cxlviii}\end{center}\end{large}
(Alleluia.)
\textcolor{benred8}{1}~Praise God in his sanctuary, praise him on his sovereign throne. \textcolor{benred8}{2}~Praise him for his noble acts, praise him for his surpassing greatness. \textcolor{benred8}{3}~Praise him with the bray of the trumpet, praise him with harp and zither. \textcolor{benred8}{4}~Praise him with the tambour and the dance, praise him with the music of string and of reed. \textcolor{benred8}{5}~Praise him with the clang of the cymbals, the cymbals that ring merrily. \textcolor{benred8}{6}~All creatures that breath have, praise the Lord. Alleluia. \textcolor{benred8}{1}~These proverbs were written by David\textquotesingle s son Solomon, that was king of Israel, \textcolor{benred8}{2}~for the better understanding of true wisdom, and self-command. \textcolor{benred8}{3}~Here is made known the secret of discernment; here men may learn the lesson of insight, the dictates of duty and right and honour. \textcolor{benred8}{4}~Here simplicity is put on its guard; here youth may find instruction and advice both together. \textcolor{benred8}{5}~The wise, too, may be the wiser for hearing them; they will aid even the discerning to guide his course aright; \textcolor{benred8}{6}~he will read both parables and the interpretation of parables, both wise words and the hidden thoughts they signify.


\end{document}






